---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# samplyr

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/samplyr)](https://CRAN.R-project.org/package=samplyr)
[![R-CMD-check](https://gitlab.com/dickoa/samplyr/badges/main/pipeline.svg)](https://gitlab.com/dickoa/samplyr/-/pipelines)
[![Codecov test coverage](https://codecov.io/gl/dickoa/samplyr/branch/main/graph/badge.svg)](https://app.codecov.io/gl/dickoa/samplyr?branch=main)
<!-- badges: end -->

A tidy grammar for survey sampling in R. `samplyr` provides a minimal set of composable verbs for stratified, clustered, multi-stage, and multi-phase sampling designs with PPS methods, sample coordination, and panel rotation.

## Ecosystem Positioning

`samplyr` sits in the middle of a three-layer workflow:

- Use `svyplan` for planning (`n`, precision, power, allocation, budget).
- Use `samplyr` to specify designs, draw samples, and carry design metadata through fieldwork.
- Use `survey`/`srvyr` after data collection for estimation and inference.

These packages are complementary, not competing. Handoffs are explicit (for example, `svyplan` outputs can feed `draw(n = ...)`, and `samplyr` outputs convert via `as_svydesign()`, `as_svrepdesign()`, and `as_survey_design()`).

## Why samplyr?

`samplyr` is built around a simple idea: sampling code should read like its English description.

```{r overview}
library(samplyr)
data(bfa_eas)

# "Stratify by region, proportionally allocate 500 samples, execute"
sampling_design() |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 500) |>
  execute(bfa_eas, seed = 1)
```

Consider a real survey design from Lohr (2022, Example 7.1), based on a 1991 study of bed net use in rural Gambia (D'Alessandro et al., 1994):

> Malaria morbidity can be reduced by using bed nets impregnated with insecticide, but this is only effective if the bed nets are in widespread use. In 1991, a nationwide survey was designed to estimate the prevalence of bed net use in rural areas of the Gambia (D'Alessandro et al., 1994).
>
> The sampling frame consisted of all rural villages of fewer than 3,000 people. The villages were **stratified by three geographic regions** (eastern, central, and western) and by **whether the village had a public health clinic (PHC)** or not. In each region **five districts were chosen with probability proportional to the district population**. In each district **four villages were chosen, again with probability proportional to census population**: two PHC villages and two non-PHC villages. Finally, **six compounds were chosen** more or less randomly from each village.

In `samplyr`, this three-stage stratified cluster design translates directly into code:

```{r}
design <- sampling_design(title = "Gambia bed nets") |>
  add_stage() |>
    stratify_by(region) |>
    cluster_by(district) |>
    draw(n = 5, method = "pps_brewer", mos = population) |>
  add_stage() |>
    stratify_by(phc) |>
    cluster_by(village) |>
    draw(n = 2, method = "pps_brewer", mos = population) |>
  add_stage() |>
    draw(n = 6)
design
```

The `samplyr` code mirrors the verbal description verb for verb.

*Lohr, S. L. (2022). Sampling: Design and Analysis (3rd ed.). CRC Press.*

## Installation

```{r, eval = FALSE}
# Install sondage first (sampling algorithms backend)
pak::pkg_install("gitlab::dickoa/sondage")

# Install svyplan (sample size, precision, power, and stratification)
pak::pkg_install("gitlab::dickoa/svyplan")

# Install samplyr
pak::pkg_install("gitlab::dickoa/samplyr")
```

## The Grammar

`samplyr` uses 5 verbs and 1 modifier:

| Function            | Purpose                               |
|---------------------|---------------------------------------|
| `sampling_design()` | Create a new sampling design          |
| `stratify_by()`     | Define stratification and allocation  |
| `cluster_by()`      | Define cluster/PSU variable           |
| `draw()`            | Specify sample size and method        |
| `execute()`         | Run the design on a frame             |
| `add_stage()`       | Delimit stages in multi-stage designs |

### Frame-Independent Design

`stratify_by()` and `cluster_by()` take bare column names. The design is stored as a specification and resolved only when a frame is available
(`validate_frame()`, `execute()`, `as_svydesign()`), so design specification stays separate from execution.

```{r frame-independent}
design <- sampling_design() |>
  stratify_by(region, alloc = "proportional") |>
  cluster_by(ea_id) |>
  draw(n = 300)

sample <- execute(design, bfa_eas, seed = 2)
sample
```

## Quick Start

```{r quick-start}
library(samplyr)
data(bfa_eas)

# Simple random sample
srs_smpl <- sampling_design() |>
  draw(n = 100) |>
  execute(bfa_eas, seed = 321)

srs_smpl

# Stratified proportional allocation
strata_smpl <- sampling_design() |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 300) |>
  execute(bfa_eas, seed = 12)

strata_smpl

# PPS cluster sampling
cluster_smpl <- sampling_design() |>
  cluster_by(ea_id) |>
  draw(n = 50, method = "pps_brewer", mos = households) |>
  execute(bfa_eas, seed = 123)

cluster_smpl
```

## Multi-Stage Sampling

Use `add_stage()` to define multi-stage designs. This example selects districts with PPS, then samples EAs within each:

```{r multi-stage}
library(dplyr, warn.conflicts = FALSE)
data(zwe_eas)

# Add district-level measure of size
zwe_frame <- zwe_eas |>
  mutate(district_hh = sum(households), .by = district)

# Two-stage design: 10 districts, 5 EAs per district
sample <- sampling_design() |>
  add_stage(label = "Districts") |>
    cluster_by(district) |>
    draw(n = 10, method = "pps_brewer", mos = district_hh) |>
  add_stage(label = "EAs") |>
    draw(n = 5) |>
  execute(zwe_frame, seed = 12345)

sample
```

### Operational Sampling

Execute stages separately when fieldwork happens between stages:

```{r operational}
design <- sampling_design() |>
  add_stage(label = "EA") |>
    stratify_by(urban_rural) |>
    cluster_by(ea_id) |>
    draw(n = 10, method = "pps_brewer", mos = households) |>
  add_stage(label = "HH") |>
    draw(n = 5)

# Execute stage 1 only
selected_eas <- execute(design, zwe_eas, stages = 1, seed = 1)
selected_eas

# ... fieldwork ...
# list all households in selected EAs
selected_eas_list <- selected_eas |>
  slice(rep(seq_len(n()), households)) |>
  mutate(hh_id = row_number())

# Execute stage 2
final_sample <- design |>
  execute(selected_eas_list, seed = 2)
final_sample
```

## Selection Methods

### Equal Probability

| Method       | Sample Size | Description                                          |
|--------------|-------------|------------------------------------------------------|
| `srswor`     | Fixed       | Simple random sampling without replacement (default) |
| `srswr`      | Fixed       | Simple random sampling with replacement              |
| `systematic` | Fixed       | Systematic sampling                                  |
| `bernoulli`  | Random      | Bernoulli sampling                                   |

### Probability Proportional to Size (PPS)

| Method            | Sample Size | Description                   |
|-------------------|-------------|-------------------------------|
| `pps_brewer`      | Fixed       | Brewer's method (recommended) |
| `pps_systematic`  | Fixed       | PPS systematic                |
| `pps_cps`         | Fixed       | Conditional Poisson sampling  |
| `pps_poisson`     | Random      | PPS Poisson (PRN) |
| `pps_sps`         | Fixed       | Sequential Poisson sampling (PRN) |
| `pps_pareto`      | Fixed       | Pareto piPS sampling (PRN)        |
| `pps_multinomial` | Fixed       | PPS with replacement          |
| `pps_chromy`      | Fixed       | PPS with minimum replacement  |

### Balanced Sampling

| Method     | Sample Size | Description                                     |
|------------|-------------|-------------------------------------------------|
| `balanced` | Fixed       | Cube method (Deville & Tille 2004), uses `aux`  |

## Allocation Methods

When stratifying, control how the total sample is distributed:

| Method         | Description                                             |
|----------------|---------------------------------------------------------|
| (none)         | `n` applies per stratum                                 |
| `equal`        | Same sample size in each stratum                        |
| `proportional` | Proportional to stratum size                            |
| `neyman`       | Minimize variance (requires `variance`)                 |
| `optimal`      | Minimize cost-variance (requires `variance` and `cost`) |
| `power`        | Compromise allocation (requires `cv` and `importance`)  |

### Sample Size Bounds

Use `min_n` and `max_n` in `draw()` to constrain stratum sample sizes when using allocation methods:

```{r bounds}
data(bfa_eas_variance)

# Ensure at least 2 per stratum (minimum for variance estimation)
sampling_design() |>
  stratify_by(region, alloc = "neyman", variance = bfa_eas_variance) |>
  draw(n = 300, min_n = 2) |>
  execute(bfa_eas, seed = 321)
```

### Custom Allocation

For custom stratum-specific sizes or rates, pass a data frame to `n` or `frac` in `draw()`:

```{r custom-alloc, eval = FALSE}
# Custom allocation with data frame
sizes_df <- data.frame(
  region = c("North", "South", "East", "West"),
  n = c(100, 200, 150, 100)
)

sample <- sampling_design() |>
  stratify_by(region) |>
  draw(n = sizes_df) |>
  execute(frame, seed = 42)
```

```{r neyman}
# Neyman allocation
data(bfa_eas_variance)

sample <- sampling_design() |>
  stratify_by(region, alloc = "neyman", variance = bfa_eas_variance) |>
  draw(n = 300) |>
  execute(bfa_eas, seed = 4321)

sample
```

## Sample Coordination (PRN)

Permanent random numbers enable coordinated sampling across survey waves. Assign a stable uniform random number to each frame unit, then pass it via `prn`:

```{r prn, eval = FALSE}
frame$prn <- runif(nrow(frame))

# Wave 1: sequential Poisson sampling with PRN
wave1 <- sampling_design() |>
  draw(n = 500, method = "pps_sps", mos = size, prn = prn) |>
  execute(frame, seed = 10)

# Wave 2: same PRN -> high overlap (positive coordination)
wave2 <- sampling_design() |>
  draw(n = 500, method = "pps_sps", mos = size, prn = prn) |>
  execute(frame, seed = 20)
```

PRN is supported for `bernoulli`, `pps_poisson`, `pps_sps`, and `pps_pareto`.

## Balanced Sampling

The cube method (Deville & Tille 2004) produces samples whose Horvitz-Thompson estimates of auxiliary totals match the population totals, improving precision:

```{r balanced}
# Balanced sample using population and area as auxiliary variables
balanced_smpl <- sampling_design() |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 300, method = "balanced", mos = households,
       aux = c(population, area_km2)) |>
  execute(bfa_eas, seed = 24)

balanced_smpl
```

The `aux` parameter accepts bare column names. When combined with `mos`, inclusion probabilities are proportional to size while the sample remains balanced on the auxiliary variables. Stratified designs use the stratified cube algorithm (Chauvet 2009) in a single call. Balanced sampling is supported for up to 2 stages.

## Custom Sampling Methods

Any unequal probability sampling method can be plugged into `samplyr` by registering it with `sondage::register_method()`. The registered method is then available in `draw()` using the `pps_<name>` convention.

For example, to use the elimination procedure of Tille (1996) from the sampling package:

```{r custom-method}
# Wrap sampling::UPtille to return selected indices
tille_fn <- function(pik, n = NULL, prn = NULL, ...) {
  which(as.logical(sampling::UPtille(pik)))
}

# Joint inclusion probabilities, restricted to the sampled units
tille_joint_fn <- function(pik, sample_idx = NULL, ...) {
  pi2 <- sampling::UPtillepi2(pik)
  if (!is.null(sample_idx))
    pi2 <- pi2[sample_idx, sample_idx, drop = FALSE]
  pi2
}

# Register with both sampling and joint probability functions
sondage::register_method(
  "tille", type = "wor",
  sample_fn = tille_fn,
  joint_fn = tille_joint_fn
)

# Use it like any built-in method
sample_tille <- sampling_design() |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 300, method = "pps_tille", mos = households) |>
  execute(bfa_eas, seed = 1)

sample_tille

# Clean up
sondage::unregister_method("tille")
```

Custom methods flow through the full pipeline: stratification, multi-stage designs, certainty selection, survey export, and joint probabilities. Providing a `joint_fn` enables exact variance estimation via `joint_expectation()` and `survey::ppsmat()`. See `?sondage::register_method` for the full contract.

## Survey Export

Convert samples to `survey` or `srvyr` objects for analysis:

```{r survey-export, eval = FALSE}
svy <- as_svydesign(sample)
survey::svymean(~y, svy)

# Exact or high-entropy joint inclusion probabilities
jip <- joint_expectation(sample, frame, stage = 1)
svy_exact <- as_svydesign(sample, pps = survey::ppsmat(jip[[1]]))
```

By default `as_svydesign()` uses Brewer variance approximation. For tighter variance estimates, `joint_expectation()` computes pairwise joint inclusion probabilities from the original frame. The result is
exact for CPS, systematic, and Poisson whereas high-entropy approximation O(N^2) is by  Brewer, SPS, Pareto and Balanced sampling using the Cube algorithm (exact recursive formulas exist for some of these methods but are usually O(N^3) and impractical for large frames). See `?joint_expectation` for the full method-by-method breakdown.

## Panel Partitioning

Partition the sample into rotation groups for panel surveys:

```{r panels}
panel_sample <- sampling_design() |>
  stratify_by(region) |>
  draw(n = 200) |>
  execute(bfa_eas, seed = 1, panels = 4)

table(panel_sample$.panel)
```

Panels are assigned by systematic interleaving within strata. For multi-stage designs, panels are assigned at the PSU level and propagated to all units. Per-panel analysis is also possible by multiplying the weights by the number of panels since the weights reflect the full-sample inclusion probability.

## Replicated Sampling

Draw multiple independent samples from the same frame and design with `reps`. This is useful for simulation studies, variance estimation via repeated sampling, or quality control:

```{r reps}
rep_sample <- sampling_design() |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 300) |>
  execute(bfa_eas, seed = 42, reps = 5)

table(rep_sample$.replicate)
```

Each replicate is an independent draw. Replicate `r` uses seed `seed + r - 1`. To analyse a single replicate, filter first:

```{r reps-filter, eval = FALSE}
rep1 <- rep_sample |> filter(.replicate == 1)
as_svydesign(rep1)
```

## Two-Phase Sampling

Pipe a `tbl_sample` into `execute()` for multi-phase designs:

```{r two-phase}
# Phase 1: large screening sample
phase1 <- sampling_design() |>
  draw(n = 500) |>
  execute(bfa_eas, seed = 1)

# Phase 2: subsample from phase 1
phase2 <- sampling_design() |>
  draw(n = 50) |>
  execute(phase1, seed = 2)

phase2
```

Weights compound automatically across phases.

## Diagnostics

```{r diagnostics}
summary(strata_smpl)
```

## Validation

For statistical validation on synthetic populations with known truths, see `vignette("validation")`. It combines deterministic invariants (weights, FPC, certainty, stage compounding) with Monte Carlo checks of bias, standard-error calibration, and coverage.

## Included Datasets

Synthetic sampling frames for learning and testing:

| Dataset           | Description                                       | Rows    |
|-------------------|---------------------------------------------------|---------|
| `bfa_eas`         | Household survey EA frame (Burkina Faso)          | 44,570  |
| `zwe_eas`         | Two-stage cluster survey EA frame (Zimbabwe)      | 107,250 |
| `ken_enterprises` | Enterprise survey frame (Kenya)                   | 6,823   |

Plus auxiliary data: `bfa_eas_variance`, `bfa_eas_cost`

## Comparison with SAS and SPSS

### SAS PROC SURVEYSELECT

```sas
proc surveyselect data=frame method=pps n=50 seed=12345;
  strata region;
  cluster school;
  size enrollment;
run;
```

```{r sas-compare, eval = FALSE}
sampling_design() |>
  stratify_by(region) |>
  cluster_by(school) |>
  draw(n = 50, method = "pps_brewer", mos = enrollment) |>
  execute(frame, seed = 1)
```

### SAS Allocation with Bounds

```sas
proc surveyselect data=frame method=srs n=500 seed=42;
  strata region / alloc=neyman var=variance_data allocmin=2 allocmax=100;
run;
```

```{r sas-bounds, eval = FALSE}
sampling_design() |>
  stratify_by(region, alloc = "neyman", variance = variance_data) |>
  draw(n = 500, min_n = 2, max_n = 100) |>
  execute(frame, seed = 2)
```

### SAS Rounding Control

```sas
proc surveyselect data=frame method=sys samprate=0.02 seed=2 round=nearest;
  strata State;
run;
```

```{r sas-rounding, eval = FALSE}
sampling_design() |>
  stratify_by(State) |>
  draw(frac = 0.02, method = "systematic", round = "nearest") |>
  execute(frame, seed = 3)
```

### SPSS CSPLAN

```spss
CSPLAN SAMPLE
  /PLAN FILE='myplan.csplan'
  /DESIGN STRATA=region CLUSTER=school
  /METHOD TYPE=PPS_WOR
  /SIZE VALUE=50
  /MOS VARIABLE=enrollment.
```

```{r spss-compare, eval = FALSE}
sampling_design() |>
  stratify_by(region) |>
  cluster_by(school) |>
  draw(n = 50, method = "pps_brewer", mos = enrollment) |>
  execute(frame, seed = 4)
```
