---
title: "Design semantics and assumptions"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Design semantics and assumptions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
has_suggests <- requireNamespace("survey", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_suggests
)
```

```{r setup, message = FALSE, echo = FALSE}
library(samplyr)
library(dplyr)
```

This vignette documents what each `samplyr` operation assumes, what it computes, and what is preserved or lost at `survey` export.

Use this as a topic reference while working through the tutorials. The main
reading path is `vignette("introduction")`, then
`vignette("survey-analysis")`.

## Components, scope and transitions

A design is a sequence of stages. Each stage declares optional strata and
allocation, an optional sampling-unit key, and one selection specification.
`draw()` completes the stage, which then takes no further strata, clusters or
selection, so a design is extended with `add_stage()`. `execute()` interprets
it against frames.
Without allocation, scalar `n` applies per stratum. With allocation it is a
total over those strata. At later stages, both meanings apply within each
selected parent pool. For random-size methods, the target is an expectation.

| Operation | Input | Result and statistical meaning |
|---|---|---|
| `sampling_design()` and specification verbs | Column names and selection rules | An inspectable specification, with no sample drawn |
| `frame_summary(design, frame)` | Specification and registers | Resolved pool targets and chances, without random draws |
| `execute(design, frame)` | Specification and population register(s) | A realized sample with selection metadata and weights |
| `execute(partial_sample, listing)` | An unfinished design's sample and its next register | Continuation with completed-stage probabilities retained |
| `execute(new_design, previous_sample)` | A new design and an earlier realized sample | New phase with conditional selection factors compounded with prior weights |
| `execute(master, wave = t)` | A master sample with a panel schedule | Activation by conditional selection under the frozen panel quotas |
| `as_svydesign(sample)` or `as_svrepdesign(sample)` | A supported sample and collected observations | An analysis representation, subject to the inference restrictions |

The composition laws are that previews agree with execution's resolved
chances, continuation under the same frame order and uninterrupted RNG stream
agrees with one-call execution, and serialization preserves the declared
specification. Weight interpretation must survive every transformation, and
an export must preserve the required stage information or refuse the design.
Agreement of numerical targets does not turn approximate probabilities into
exact ones. Phase-wise conditional factors also need not be unconditional
marginal inclusion probabilities.

## Assumptions at a glance

For a compact selection table, see `?selection-methods`, and for how each method's variance is estimated, `?variance-estimation`. A complete three-stage fieldwork example is in
`vignette("introduction")`.

The following table summarizes what `samplyr` assumes and what the user must verify. Each row is expanded in the sections below.

| Assumption                                 | Applies to                 | samplyr behavior                                                                                     | User must verify                                                                                      |
|:-------------------------------------------|:---------------------------|:-----------------------------------------------------------------------------------------------------|:------------------------------------------------------------------------------------------------------|
| Strata are exhaustive and disjoint         | `stratify_by()`            | Partitions the frame by unique combinations of strata variables                                      | Strata variables define a valid partition                                                             |
| Independent sampling across strata         | `stratify_by() |> draw()` | Each stratum sampled with a separate call to `sondage`, except stratified `cube`                      | Correct by construction when strata are disjoint. Stratified `cube` balances across strata in one call |
| Cluster variables constant within clusters | `cluster_by()`             | Validated at execution: MOS and strata variables may not vary within a cluster                       | Cluster ID, read within its strata and earlier-stage units, identifies the group                      |
| Conditional selection laws | `add_stage()` | Stage \(k\) sampling depends only on the units selected at stage \(k - 1\) and its declared inputs | Later frame construction and selection depend on prior randomization only through selected ancestors and recorded design variables |
| FPC records method-specific stage information | `execute()` | Stores a population or cluster count for WOR and `Inf` for WR or PMR. PPS WOR is converted to \(\pi_i\) at survey export | Frame size at execution is treated as the population, one of several definitions when a population changes between waves |
| Weights are inverse selection chances, unadjusted | All methods | Uses inclusion probabilities for WOR and expected hit counts for WR or PMR | Downstream adjustment is the analyst's responsibility |
| PRN values are iid \(\text{U}(0, 1)\)      | `draw(prn = ...)`          | Validated: strictly in \((0, 1)\), no NAs, numeric                                                   | Independence and uniformity of the PRN column                                                         |

## Notation

Throughout this vignette:

- \(U = \{1, \ldots, N\}\): finite population (the frame).
- \(S \subseteq U\): selected population units for a WOR design.
- \(K_i\): number of realized hits for unit \(i\) in a WR or PMR design.
- \(q_i\): stage chance, equal to \(\pi_i = \Pr(i \in S)\) for WOR and
  \(E(K_i)\) for WR or PMR.
- \(\pi_{kl} = \Pr(k \in S \text{ and } l \in S)\): joint (second-order)
  inclusion probability for WOR.
- \(E(K_k K_l)\): joint expected hits for WR or PMR.
- \(w_i = 1 / q_i\): design weight per selected unit or realized occurrence.
- \(N_h\): number of the stage's sampling units in stratum \(h\)
  (\(h = 1, \ldots, H\)), so PSUs at a clustered stage.
- \(n_h\): sample size drawn from stratum \(h\).
- \(K\): number of sampling stages.
- \(\pi_i^{(k \mid S^{(k-1)})}\): conditional inclusion probability of unit
  \(i\) at stage \(k\), given the set \(S^{(k-1)}\) selected at stage \(k - 1\).

`execute()` produces **design weights only**. Calibration, trimming, and nonresponse adjustment are out of scope and belong in the estimation layer (`survey`, `srvyr`, or similar).

One declared transformation departs from that, and says so in the object it returns: `share_weights()` produces **estimation weights** for a population other than the one that was sampled. Its semantics, and the contract that keeps the two kinds of weight apart, are the subject of a section below.

## What each verb means

### `sampling_design()`: frame-independent plan

A `sampling_design` object is a specification, not a computation. Column names referenced in `stratify_by()`, `cluster_by()`, and `draw()` are stored as strings (deferred resolution). No data is touched until `execute()` binds the design to a frame. This separation means the same design can be applied to different frames or re-executed with different seeds.

```{r design-inert, message = FALSE}
library(samplyr)
library(dplyr)

design <- sampling_design(title = "SRS WOR") |>
  stratify_by(region) |>
  draw(n = 50, method = "srswor")

# No data has been sampled yet
design
```

### `stratify_by()`: partition into independent sub-populations

`stratify_by()` partitions the frame into \(H\) strata defined by the Cartesian product of its variables:

$$
U = U_1 \cup U_2 \cup \cdots \cup U_H, \qquad U_h \cap U_{h'} = \emptyset
\; \text{for } h \neq h'.
$$

Sampling within each stratum is independent. This holds by construction because `samplyr` invokes `sondage` algorithms separately for each stratum. Stratified `cube` is the exception: its global flight phase balances across strata, so its strata are not selected independently.

When an allocation method is specified, the total sample size \(n\) is distributed across strata:

| Allocation       | Formula                                 | Requires           |
|:-----------------|:----------------------------------------|:-------------------|
| `"equal"`        | \(n_h = n / H\)                         |                    |
| `"proportional"` | \(n_h \propto N_h\), or \(X_h\) if given  | `importance`, optional |
| `"neyman"`       | \(n_h \propto N_h \, S_h\)              | `variance`         |
| `"optimal"`      | \(n_h \propto N_h \, S_h / \sqrt{C_h}\) | `variance`, `cost` |
| `"power"`        | \(n_h \propto \text{CV}_h \cdot X_h^q\) | `cv`, `importance` |

Each formula gives a *factor* \(w_h\). The allocation is the solution of

\[
n_h = \operatorname{clamp}(\lambda w_h, \; \ell_h, \; u_h), \qquad
\ell_h = \min(\texttt{min\_n}, N_h), \qquad
u_h = \min(\texttt{max\_n}, N_h)
\]

with \(\lambda\) chosen so that \(\sum_h n_h = n\). The population size \(N_h\) is an upper bound whether or not `max_n` was supplied, so a fixed-size without-replacement design satisfies \(0 \le n_h \le N_h\) and \(\sum_h n_h = n\) for every feasible request.

Saturated strata are fixed at their bound and removed from the problem, then \(\lambda\) is recomputed over the strata still free. Redistribution therefore follows the same factors \(w_h\) the method is defined by, which is what keeps a Neyman allocation Neyman after a stratum saturates: redistributing in proportion to unused capacity instead would silently change the criterion. When every free stratum has \(w_h = 0\) the criterion cannot distinguish them and the remainder is split equally in stratum order.

One consequence is worth stating plainly: a named rule can be departed from when a stratum is too small to hold its share. With populations \((10, 490, 500)\) and \(n = 300\), `"equal"` yields 10/145/145, not 100/100/100. `execute()` reports this once per run.

The constrained real-valued targets are then rounded with largest-remainder (Hare-Niemeyer) rounding that preserves \(\sum n_h = n\) and stays inside every bound.

With-replacement (`srswr`, `pps_multinomial`) and minimum-replacement (`pps_chromy`) designs can select a unit more than once, so \(N_h\) is not a bound on the number of draws and only `min_n` and `max_n` apply. Per-stratum sizes supplied directly, rather than through `alloc`, are instructions and are never redistributed.

### `cluster_by()`: define the sampling unit

`cluster_by()` declares the variable(s) that identify the primary sampling unit (PSU) at the current stage. Operationally, `samplyr`:

1. Groups the frame by the cluster variable(s).
2. Draws a sample of **clusters** (not individual units).
3. Returns all population units within selected clusters.

Any variable used in PPS size measures (`mos`) or in stratification must be constant within each cluster. `samplyr` validates this at execution time and raises an error if cluster-level variables vary within groups.

`cluster_by()` is purely structural. It does not affect inclusion probabilities directly. They are determined by `draw()` applied to the cluster-level aggregates.

Cluster ids may be numbered within the level above them. A unit is identified by its id together with the units earlier stages selected and the strata of its own stage, so town 1 of county A and town 1 of county B are two towns. `cluster_by(nest = FALSE)` instead declares the ids unique across strata and turns that into a check, which refuses a cluster coded into two strata instead of splitting it. `?cluster_by` gives the rules, including what a separate register for a later stage must carry.

### `draw()`: specify the selection mechanism

`draw()` attaches a selection method and sample size (or fraction) to the current stage. The inclusion probabilities and weights depend on the method
class:

**Equal-probability without replacement** (`srswor`, `systematic`, `bernoulli`):

$$
\pi_i = \frac{n}{N} \quad (\text{or } \pi_i = f \text{ for bernoulli}),
\qquad w_i = \frac{1}{\pi_i}.
$$

Fixed-size for `srswor` and `systematic` and random-size for `bernoulli`. Systematic sampling induces implicit stratification by row order. Use `control` to specify the sort variable.

**PPS methods** (`pps_brewer`, `pps_systematic`, `pps_cps`,
`pps_sampford`, `pps_poisson`, `pps_sps`, `pps_pareto`):

$$
\pi_i = n \, \frac{\text{mos}_i}{\sum_j \text{mos}_j}
$$

computed by `sondage::inclusion_prob()`, which iteratively caps units with \(\pi_i \geq 1\) and redistributes. These "certainty" units (\(\pi_i = 1\))
receive stage weight 1 and contribute zero first-stage variance. In multi-stage designs, final `.weight` still compounds all stages, so certainty at one stage does not imply overall weight 1. For Poisson sampling, \(\pi_i = f \cdot N \cdot \text{mos}_i / \sum \text{mos}\), capped at 1 (random-size design).

Two of the seven attain that \(\pi_i\) only approximately. `pps_sps` and
`pps_pareto` are order sampling designs, which fix the sample size at the cost
of the marginals: the quantity above is the **target** inclusion probability
handed to the algorithm, and the realized inclusion probability is close to it
but not equal, approaching it as \(n\) and \(N\) grow [@tille2020, sec. 5.11;
@rosen1997a]. Both sit in the `"approximate"` probabilities tier. Everywhere
below that reads \(\pi_i\) off `.weight` inherits that distinction. The other
five methods satisfy it exactly.

**With-replacement and PMR** (`srswr`, `pps_multinomial`, `pps_chromy`):

The quantity of interest is the expected number of hits \(E(n_i) = n p_i = n \, \text{mos}_i / \sum \text{mos}\). If unit \(i\) is selected \(k\) times, `samplyr` creates \(k\) rows with a `.draw` column indexing each occurrence. The weight on every output row is \(w_i = 1 / E(n_i) = 1 / (n p_i)\). For WR the draws are independent and this is the per-hit Hansen-Hurwitz estimator. No finite population correction applies (FPC = \(\infty\)).

Chromy's sequential PPS (`pps_chromy`) is classified as *probability minimum replacement* (PMR): each unit receives \(\lfloor E(n_i) \rfloor\) or \(\lceil E(n_i) \rceil\) hits. Its occurrences are not independent WR draws. At survey export, samplyr uses the Hansen-Hurwitz approximation documented below. When all \(E(n_i) < 1\), the selection itself reduces to WOR.

**Balanced and spatially balanced sampling** (`cube`, `lpm2`, `scps`):

The cube method [@devilletille2004] selects a fixed-size sample that satisfies (or nearly satisfies) the balancing equations $\sum_{i \in S} a_{ij} / \pi_i = \sum_{i \in U} a_{ij}$ for auxiliary variables $a_j$ specified via `aux`. Without `mos`, $\pi_i = n/N$ and probabilities are equal. With `mos`, $\pi_i$ is computed by `sondage::inclusion_prob()`. A `bound()` marker adds hard adjacent-integer category-count constraints.

When stratified, the stratified cube algorithm [@chauvet2009] runs flight phases per stratum, then a global flight phase, then per-stratum landing phases.

LPM2 and SCPS instead use the coordinates supplied through `spread`. They do not accept cube auxiliaries or count bounds. At most 2 stages may use a balanced-family method. Certainty selection is not supported, and of the two only SCPS accepts a PRN.

**PRN-compatible methods** (`bernoulli`, `pps_poisson`, `pps_sps`, `pps_pareto`, `scps`):

When `prn` is specified, the permanent random numbers are passed through to `sondage`. Selection becomes deterministic given the PRN values, so the
`seed` argument to `execute()` has no effect on the selection outcome. For `scps` it is also fixed by the order in which the pool's units are visited.

### `add_stage()`: delimit stages

`add_stage()` is syntactic: it closes the current stage and opens a new one. It carries no statistical content of its own. Stages are numbered
\(1, 2, \ldots, K\) in definition order.

### `execute()`: materialize the sample

`execute()` binds the design to one or more frames and runs the selection:

1. For each stage \(k\), the frame is restricted to units belonging to clusters selected at stage \(k - 1\) (or the full frame for \(k = 1\)).
2. Strata and cluster groupings are resolved against the (subsetted) frame.
3. The selection algorithm from `draw()` is invoked via `sondage`.
4. The `.weight`, `.fpc`, `.draw`, and `.certainty` columns are computed where they apply and suffixed with the stage number (`?sample-columns`).

**One frame or one per stage.** The number of frames schedules the stages. `execute(design, hierarchy)` runs every stage against one table, and `execute(design, schools, classes, students)` gives each stage its own register, mapped by position. Any other count is refused. The frames may equally be held as a value with `execute(design, list(schools, classes, students))`. A name on a list element is a diagnostic label for that frame.

A register is supplied whole. samplyr restricts it to the units its parent stage selected and carries onto it the variables earlier stages introduced, so a class register needs neither pre-filtering to the sampled schools nor a copy of the school-level stratification column. A register that omits such a carried variable is fine. One whose own copy disagrees with the parent is an error.

**Partial execution.** `execute(design, psu_frame, stages = 1)` runs only the first stage and returns a `tbl_sample`. That object stores both the unchanged design and the realized PSU selection. Use it as the first argument of the later call and supply the new listing as the frame: `execute(psu_sample, listing_frame)`. Execution resumes at the next unexecuted stage, carries forward the PSU weights and FPC, and remains one multi-stage design.

The `stages` argument may select any initial contiguous batch beginning at stage 1 and on a partial `tbl_sample`, it may select the next contiguous batch. Omitting it executes all remaining stages, with one exception. If more than one stage remains and a single frame is supplied, that frame could be the next stage's register or a hierarchy covering the rest. The two draw different samples, so `execute()` asks for `stages` rather than guessing.

**The three spellings are one implementation.** A three-stage design executed against one hierarchy, against three registers in a single call, or stage by stage as a continuation runs the same stage transition each time. All three draw the same sample, provided the continuation runs under a single RNG stream and names `stages` on every intermediate call:

```r
one_call     <- execute(design, schools, classes, students, seed = 7)
hierarchical <- execute(design, hierarchy, seed = 7)

continued <- withr::with_seed(7, {
  s1 <- execute(design, schools, stages = 1)
  s2 <- execute(s1, classes, stages = 2)
  execute(s2, students, stages = 3)
})
```

Note where the seed goes. One seed wraps the whole chain, and the individual calls take none. Seeding each call separately is a different and equally valid RNG boundary, but it is not this equivalence and will select different units. The equivalence is a tested invariant rather than an incidental property. The continued sample is therefore the one the single call would have drawn.

**Two-phase sampling.** A new phase starts with a new design: `execute(design2, phase1_sample)`. Here the phase-1 `tbl_sample` is the frame argument, not the execution object. samplyr records a previous-phase link and compounds the phase-1 and phase-2 weights multiplicatively. The conditional phase-2 weight is computed at survey export time. Thus `execute(psu_sample, listing_frame)` means “continue this design's stages,” whereas `execute(new_design, prior_sample)` means “sample a new phase.”

## Weight compounding across stages

**Single-stage.** Let $q_i$ be the stage selection quantity. For WOR,
$q_i = \pi_i$. For WR or PMR, $q_i = E(K_i)$, the expected hit count.
The design weight is:

$$
w_i = \frac{1}{q_i}.
$$

**Multi-stage.** Let \(q_i^{(k)}\) denote the quantity resolved within the
clusters selected at prior stages. The compound weight is:

$$
w_i = \prod_{k=1}^{K} \frac{1}{q_i^{(k)}} = \prod_{k=1}^{K} w_i^{(k)}.
$$

For an all-WOR exact design this is the inverse product of conditional
inclusion probabilities. With a WR or PMR stage, each output row is a realized
occurrence and that stage contributes an inverse expected-hit factor.
`samplyr` joins each stage's weight to its parent by the shared cluster
variables and multiplies. The per-stage weights remain in `.weight_1`,
`.weight_2`, and so on. The `.weight` column always equals their product.

```{r weight-compound}
two_stage <- sampling_design() |>
  add_stage("Districts") |>
    stratify_by(province) |>
    cluster_by(district) |>
    draw(n = 2, method = "srswor") |>
  add_stage("EAs") |>
    draw(n = 5, method = "srswor") |>
  execute(zwe_eas, seed = 123456)

# Compound weight equals the product of per-stage weights
all.equal(two_stage$.weight,
          two_stage$.weight_1 * two_stage$.weight_2)
```

The warning is the design meeting the frame. Bulawayo is a single-district province, so a take of two districts selects the one that exists. `execute()` reports every pool it could not fill, once per stage, because a capped pool makes the design non-self-weighting. The weights above still compound exactly and Bulawayo's first-stage weight is simply 1.

**Two-phase.** When a new phase-2 design is executed with the phase-1 `tbl_sample` as its frame, the phase-1 weight is stored internally and multiplied into the final `.weight`. The conditional phase-2 weight \(w_{\text{cond}} = w_{\text{overall}} / w_{\text{phase 1}}\) is computed during survey export for use with `survey::twophase()`.

## Finite population corrections

`samplyr` stores the FPC as a population count in `.fpc_k`. The interpretation depends on the method, and the conversion for survey export happens automatically in `as_svydesign()`:

| Method class                  | `.fpc_k` stores        | Passed to survey as | survey interpretation                 |
|:------------------------------|:-----------------------|:--------------------|:--------------------------------------|
| EP-WOR (srswor, systematic)   | \(N_h\) (stage units)  | \(N_h\) directly    | Sampling fraction \(f_h = n_h / N_h\) |
| PPS-WOR (pps_brewer, ...)     | \(N_h\) (stage units)  | \(1 / w_k = \pi_i\) | Inclusion probability                 |
| WR / PMR (srswr, pps_chromy, ...) | \(\infty\) (synthetic) | \(\infty\)          | No FPC (Hansen-Hurwitz variance)     |

The PPS conversion merits explanation. `survey::svydesign()` interprets FPC values less than 1 as inclusion probabilities and values \(\geq 1\) as
population sizes. `samplyr` stores \(N_h\) for every without-replacement method, then at export time, `as_svydesign()` creates a per-row column `.fpc_pi_k = 1 / .weight_k = \pi_i` for PPS-WOR stages. The transformation is lossless: the per-stage weight already encodes \(\pi_i\), so nothing is approximated in converting it. What is approximate for `pps_sps` and `pps_pareto` is the quantity being transformed, since their \(\pi_i\) is a target rather than the realized inclusion probability.

For WR and PMR methods, a synthetic column filled with `Inf` is created. The survey package interprets this as "no finite population correction," which gives the Hansen-Hurwitz variance estimator.

The table describes one represented stage, or several without a PPS-WOR stage. `survey::svydesign()` needs every FPC term on the same scale, so a multi-stage design with a PPS-WOR stage passes fractions instead: \(1 / w_k\) for every without-replacement stage and 0 for WR and PMR stages. `?as_svydesign` gives both encodings.

**The population the correction refers to.** `samplyr` treats the frame as executed as the population, so \(N_h\) is counted from the frame the stage drew from. That is one convention among several. Where the sampling fraction is not negligible and the population itself changes between occasions, the finite population correction matters more and admits several competing definitions, particularly for measures of change across coordinated or overlapping samples [@tam1984; @laniel1988; @nordberg2000; @berger2004change]. `samplyr` does not choose among them for you: it records the frame size it saw, and a different convention has to be imposed downstream.

## Independence across stages

`samplyr` assumes that the stage-\(k\) selection mechanism depends on stage \(k - 1\) only through the **set** of selected clusters, not through
the specific randomization that produced it. Formally:

$$
\Pr(S^{(k)} \mid S^{(k-1)}) \text{ depends on } S^{(k-1)} \text{ as a set.}
$$

This is the standard assumption in multi-stage sampling theory
[@cochran1977, ch. 11; @sarndal1992, ch. 4.3].

**Consequence for variance estimation.** Under this assumption, the recursive
multi-stage variance formula applies [@sarndal1992, ch. 4.3]. For methods with
a supported linearization estimator, `as_svydesign()` exports every executed
stage to `survey::svydesign()`: one `ids` term, one `fpc` term, and (when the
stage is stratified) one `strata` term per stage. A first-stage census, for
example, correctly attributes all variance to the later stages.

A stage that samples with replacement is handled differently: later stages
nested in it contribute nothing beyond the between-draw variability, which the
Hansen-Hurwitz estimator already captures.

Bounded cube, LPM2, and SCPS do not currently have supported linearization
estimators. Their variance needs `as_svrepdesign(type = "subbootstrap")` or
`"mrbbootstrap"`, which `?variance-estimation` describes.

## Weights for a population other than the one sampled

Two transformations of an executed sample produce a weight for units that
were never in the frame. Neither changes the draw.

`share_weights()` is the generalized weight share method [@lavallee2007]. A
sample is selected from population \(U^A\), and estimates are wanted for a
second population \(U^B\) linked to it: children through their parents,
establishments through their enterprises, persons through their dwellings.
Every target unit in a reached cluster \(i\) receives

$$
w_i = \sum_j \frac{I(j \in S^A)}{\pi_j} \frac{L_{ji}}{L_i},
$$

the source weights carried across the links and divided by the
**population** number of links \(L_i\) to that cluster. \(\pi_j\) is needed
only for selected source units, which `.weight` already holds. \(L_i\) sums
over the whole of \(U^A\), so counting a supplied link table asserts that the
table is a complete population register. That assertion has no default:
either name a column holding the population multiplicity, or state it with
`complete_links()`. `?share_weights` gives the target-cluster choices and
weighted links.

`stack_frames()` collects samples selected independently from two or more
registers of the same population, each incomplete. The multiplicity estimator
for overlapping frames is the weight share method with the standardized link
matrix, since the number of links to a unit is the number of frames
containing it [@lohr2021, section 3.2; @mecatti2007]. A stack records which
frames each sampled unit belongs to and leaves the compositing factor to the
export, because the appropriate factor depends on the estimand. A stack does
not assert that the frames together cover the target population.

### The weight contract

A shared weight is an estimation weight and is not generally the reciprocal
of a target unit's inclusion probability. The design the object carries
describes selection of the **source** population. The result records which
kind of weight it holds, and every statistical consumer states what it does
about that record rather than treating the sample as ordinary.

| Consumer | On a shared weight |
|---|---|
| `design_effect()`, `effective_n()` | Supported. They read `.weight` and nothing else |
| `as_svrepdesign()` | Supported. It replicates the source design and shares inside every replicate |
| The data-manipulation methods | Preserve the record |
| `as_svydesign()` | Supported. It exports one row per source-target contribution |
| `joint_expectation()`, `varcomp()` | Refuse |
| A further `execute()`, and the panel verbs | Refuse |

Each refusal has its own condition class and all of them are also catchable as
`samplyr_error_weight_contract`. Neither `as_tbl_sample()` nor an ordinary
dplyr reconstruction can launder an estimation weight back into a design-weight
sample.

A target cluster with no link to the source population can never be surveyed,
and the estimator understates totals by exactly its share. `share_weights()`
records that finding without warning, and the warning fires where the estimate
is formed. For a `stack_frames()` collection it is evaluated over the union of
the frames and names only the clusters no frame reaches.

## What is lost at survey export

`as_svydesign()` translates a `tbl_sample` into a `survey::svydesign()` object. The translation is faithful for standard variance estimators, but some `samplyr` metadata is not representable in the `survey` framework.

### 1. Element sampling before later stages

An unclustered element-sampling stage followed by further stages is not nested cluster sampling: the later selections operate on the realized element sample, not within larger units, so there is no hierarchy for the multi-stage variance recursion (or for replicate-weight resampling) to use. This is phase sampling. `execute()` refuses this shape (`samplyr_error_stage_parent_id`), and the supported path is a two-phase sample: execute the element stage under its first-phase design, execute a new second-phase design with that sample as its frame, and export with `as_svydesign()`, which uses `survey::twophase()`. A *final* unclustered stage is fully supported: it is exported with a synthesized row-identity `ids` term and its own FPC.

### 2. Variance by selection method

How each selection method is linearized, what a joint matrix from
`joint_expectation()` changes, how with-replacement and Chromy stages are
represented, and how certainty units form a take-all stratum are in
`?variance-estimation`, with the measured error of the approximations that
were measured.

### 3. Panel assignments

The `.panel` column is not passed to survey. Panels are randomized fixed-quota partitions for rotation or workload distribution. The full-sample weights in `.weight` are correct for the combined sample. Panel assignment is not an additional probability-sampling phase, so a single panel does not automatically have inclusion probability \(\pi_i / k\). Conditional on the frozen quotas, its probability is the block's realized quota over the block size. Multiplying one panel's weights by \(k\) is therefore not generally valid for population inference.

### 4. Design metadata

`samplyr` stores the full `sampling_design` object, stage labels, seed, and execution history as attributes of the `tbl_sample`. These are accessible via `get_design()` and `get_stages_executed()` but are not round-tripped through the survey export.

## Panel partitioning semantics

There are two distinct operations here and they have different statistical
status. **Partitioning** (`panels = k` with no schedule) splits one execution
into \(k\) rotation groups by randomized fixed quota inside ordered blocks.
It changes no inclusion probability and no `.weight`, so the combined sample
is analyzed with the stored weights. **Wave activation** (a scheduled `panels`
plus `execute(wave = t)`) selects a subset of those groups, and that *is* a
second sampling phase. Its activation probability is the block's frozen quota
for the active panels over the block size, and the weight is multiplied by the
inverse: the block size over that quota. For a block of eight units with four
activated, the factor is 2. It is exact and generally not the nominal
\(k / r\) for \(r\) active panels of \(k\), and it is why the block sizes
and quotas are recorded with the sample.

Neither operation replenishes the sample. A program that draws refreshment
cohorts against later frame vintages is a registry of separate executions, not
a partition of one. See `rotation_program()`. `?panel-assignment` gives the assignment rules,
the choice of assignment stage and what it costs, certainty units, schedules
and small pools.

## PRN and sample coordination semantics

`samplyr` passes permanent random numbers through to `sondage` without transformation. PRN-compatible methods (`bernoulli`, `pps_poisson`, `pps_sps`, `pps_pareto`, `scps`) produce a deterministic selection given the PRN values, and for `scps` the visiting order. When PRN is supplied, the `seed` argument to `execute()` does not affect selection.

Positive and negative coordination across survey waves is a user-level
workflow, not a package feature. `samplyr` passes PRNs to the selection method,
the user manages their persistence and update. The negative-coordination
update is \(u_{\text{new}} = (u - \pi) \bmod 1\) [@ohlsson1995].

## References
