---
title: "Design Semantics and Assumptions"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Design Semantics and Assumptions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, echo = FALSE}
library(samplyr)
library(dplyr)
```

This vignette documents what each `samplyr` operation assumes, what it computes, and what is preserved or lost at `survey` export. The contracts stated here are asserted; `vignette("validation")` checks them against synthetic populations with known truths, and reports what a run at that replication count can and cannot establish.

## Assumptions at a glance

The following table summarizes what `samplyr` assumes and what the user must verify. Each row is expanded in the sections below.

| Assumption                                 | Applies to                 | samplyr behavior                                                                                     | User must verify                                                                                      |
|:-------------------------------------------|:---------------------------|:-----------------------------------------------------------------------------------------------------|:------------------------------------------------------------------------------------------------------|
| Strata are exhaustive and disjoint         | `stratify_by()`            | Partitions the frame by unique combinations of strata variables                                      | Strata variables define a valid partition                                                             |
| Independent sampling across strata         | `stratify_by() |> draw()` | Each stratum sampled with a separate call to `sondage`                                                 | Correct by construction when strata are disjoint                                                      |
| Cluster variables constant within clusters | `cluster_by()`             | Validated at execution: MOS and strata variables may not vary within a cluster                       | Cluster ID uniquely identifies the group                                                              |
| Conditional independence across stages     | `add_stage()`              | Stage \(k\) sampling depends only on the set of units selected at stage \(k - 1\)                    | Standard multi-stage assumption; holds unless second-stage frame depends on first-stage randomization |
| FPC stored as population count             | `execute()`                | `.fpc_k` contains \(N_h\) (or cluster count); converted to \(\pi_i\) at survey export for PPS stages | Frame size at execution is treated as the population, which is one of several definitions in a population that changes between waves |
| Weights are \(1/\pi_i\), unadjusted        | All methods                | No calibration, trimming, or non-response adjustment                                                 | Downstream adjustment is the analyst's responsibility                                                 |
| PRN values are iid \(\text{U}(0, 1)\)      | `draw(prn = ...)`          | Validated: strictly in \((0, 1)\), no NAs, numeric                                                   | Independence and uniformity of the PRN column                                                         |

## Notation

Throughout this vignette:

- \(U = \{1, \ldots, N\}\): finite population (the frame).
- \(S \subseteq U\): selected sample.
- \(\pi_i = \Pr(i \in S)\): first-order inclusion probability.
- \(\pi_{kl} = \Pr(k \in S \text{ and } l \in S)\): joint (second-order)
  inclusion probability.
- \(w_i = 1 / \pi_i\): design weight (sampling weight before any adjustment).
- \(N_h\): population size of stratum \(h\) (\(h = 1, \ldots, H\)).
- \(n_h\): sample size drawn from stratum \(h\).
- \(K\): number of sampling stages.
- \(\pi_i^{(k \mid S^{(k-1)})}\): conditional inclusion probability of unit
  \(i\) at stage \(k\), given the set \(S^{(k-1)}\) selected at stage \(k - 1\).

`samplyr` produces **design weights only**. Calibration, trimming, and non-response adjustment are out of scope and belong in the estimation layer (`survey`, `srvyr`, or similar).

## What each verb means

### `sampling_design()`: frame-independent plan

A `sampling_design` object is a specification, not a computation. Column names referenced in `stratify_by()`, `cluster_by()`, and `draw()` are stored as strings (deferred resolution). No data is touched until `execute()` binds the design to a frame. This separation means the same design can be applied to different frames or re-executed with different seeds.

```{r design-inert}
design <- sampling_design(title = "SRS WOR") |>
  stratify_by(region) |>
  draw(n = 50, method = "srswor")

# No data has been sampled yet
design
```

### `stratify_by()`: partition into independent sub-populations

`stratify_by()` partitions the frame into \(H\) strata defined by the Cartesian product of its variables:

$$
U = U_1 \cup U_2 \cup \cdots \cup U_H, \qquad U_h \cap U_{h'} = \emptyset
\; \text{for } h \neq h'.
$$

Sampling within each stratum is independent. This holds by construction because `samplyr` invokes `sondage` algorithms separately for each stratum.

When an allocation method is specified, the total sample size \(n\) is distributed across strata:

| Allocation       | Formula                                 | Requires           |
|:-----------------|:----------------------------------------|:-------------------|
| `"equal"`        | \(n_h = n / H\)                         |                    |
| `"proportional"` | \(n_h \propto N_h\)                     |                    |
| `"neyman"`       | \(n_h \propto N_h \, S_h\)              | `variance`         |
| `"optimal"`      | \(n_h \propto N_h \, S_h / \sqrt{C_h}\) | `variance`, `cost` |
| `"power"`        | \(n_h \propto \text{CV}_h \cdot X_h^q\) | `cv`, `importance` |

Each formula gives a *factor* \(w_h\); the allocation is the solution of

\[
n_h = \operatorname{clamp}(\lambda w_h, \; \ell_h, \; u_h), \qquad
\ell_h = \min(\texttt{min\_n}, N_h), \qquad
u_h = \min(\texttt{max\_n}, N_h)
\]

with \(\lambda\) chosen so that \(\sum_h n_h = n\). The population size \(N_h\) is an upper bound whether or not `max_n` was supplied, so a fixed-size without-replacement design satisfies \(0 \le n_h \le N_h\) and \(\sum_h n_h = n\) for every feasible request.

Saturated strata are fixed at their bound and removed from the problem, then \(\lambda\) is recomputed over the strata still free. Redistribution therefore follows the same factors \(w_h\) the method is defined by, which is what keeps a Neyman allocation Neyman after a stratum saturates: redistributing in proportion to unused capacity instead would silently change the criterion. When every free stratum has \(w_h = 0\) the criterion cannot distinguish them and the remainder is split equally in stratum order.

One consequence is worth stating plainly: a named rule can be departed from when a stratum is too small to hold its share. With populations \((10, 490, 500)\) and \(n = 300\), `"equal"` yields 10/145/145, not 100/100/100. `execute()` reports this once per run.

The constrained real-valued targets are then rounded with largest-remainder (Hare--Niemeyer) rounding that preserves \(\sum n_h = n\) and stays inside every bound.

With-replacement and Poisson-multinomial designs draw units repeatedly, so \(N_h\) is not a bound on the number of draws and only `min_n` and `max_n` apply. Per-stratum sizes supplied directly, rather than through `alloc`, are instructions and are never redistributed.

### `cluster_by()`: define the sampling unit

`cluster_by()` declares the variable(s) that identify the primary sampling unit (PSU) at the current stage. Operationally, `samplyr`:

1. Groups the frame by the cluster variable(s).
2. Draws a sample of **clusters** (not individual units).
3. Returns all population units within selected clusters.

Any variable used in PPS size measures (`mos`) or in stratification must be constant within each cluster. `samplyr` validates this at execution time and raises an error if cluster-level variables vary within groups.

`cluster_by()` is purely structural. It does not affect inclusion probabilities directly. They are determined by `draw()` applied to the cluster-level aggregates.

### `draw()`: specify the selection mechanism

`draw()` attaches a selection method and sample size (or fraction) to the current stage. The inclusion probabilities and weights depend on the method
class:

**Equal-probability without replacement** (`srswor`, `systematic`, `bernoulli`):

$$
\pi_i = \frac{n}{N} \quad (\text{or } \pi_i = f \text{ for bernoulli}),
\qquad w_i = \frac{1}{\pi_i}.
$$

Fixed-size for `srswor` and `systematic` and random-size for `bernoulli`. Systematic sampling induces implicit stratification by row order. Use `control` to specify the sort variable.

**PPS without replacement** (`pps_brewer`, `pps_systematic`, `pps_cps`,
`pps_sampford`, `pps_poisson`, `pps_sps`, `pps_pareto`):

$$
\pi_i = n \, \frac{\text{mos}_i}{\sum_j \text{mos}_j}
$$

computed by `sondage::inclusion_prob()`, which iteratively caps units with \(\pi_i \geq 1\) and redistributes. These "certainty" units (\(\pi_i = 1\))
receive stage weight 1 and contribute zero first-stage variance. In multi-stage designs, final `.weight` still compounds all stages, so certainty at one stage does not imply overall weight 1. For Poisson sampling, \(\pi_i = f \cdot N \cdot \text{mos}_i / \sum \text{mos}\), capped at 1 (random-size design).

Two of the seven attain that \(\pi_i\) only approximately. `pps_sps` and
`pps_pareto` are order sampling designs, which fix the sample size at the cost
of the marginals: the quantity above is the **target** inclusion probability
handed to the algorithm, and the realized inclusion probability is close to it
but not equal, approaching it as \(n\) and \(N\) grow [@tille2020, sec. 5.11;
@rosen1997a]. Both sit in the `"approximate"` probabilities tier. Everywhere
below that reads \(\pi_i\) off `.weight` inherits that distinction; the other
five methods satisfy it exactly.

**With-replacement and PMR** (`srswr`, `pps_multinomial`, `pps_chromy`):

The quantity of interest is the expected number of hits \(E(n_i) = n \, \text{mos}_i / \sum \text{mos}\). Each draw is an independent selection; if unit \(i\) is selected \(k\) times, `samplyr` creates \(k\) rows with a `.draw` column indexing each hit. The weight per draw is \(w_i = 1 / p_i\), where \(p_i\) is the single-draw selection probability. No finite population correction applies (FPC = \(\infty\)).

Chromy's sequential PPS (`pps_chromy`) is classified as *probability minimum replacement* (PMR): each unit receives \(\lfloor E(n_i) \rfloor\) or \(\lceil E(n_i) \rceil\) hits. When all \(E(n_i) < 1\), this reduces to WOR.

**Balanced and spatially balanced sampling** (`cube`, `lpm2`, `scps`):

The cube method [@devilletille2004] selects a fixed-size sample that satisfies (or nearly satisfies) the balancing equations $\sum_{i \in S} a_{ij} / \pi_i = \sum_{i \in U} a_{ij}$ for auxiliary variables $a_j$ specified via `aux`. Without `mos`, $\pi_i = n/N$ (equal probability); with `mos`, $\pi_i$ is computed by `sondage::inclusion_prob()`. A `bound()` marker adds hard adjacent-integer category-count constraints.

When stratified, the stratified cube algorithm [@chauvet2009] runs flight phases per stratum, then a global flight phase, then per-stratum landing phases.

LPM2 and SCPS instead use the coordinates supplied through `spread`; they do not accept cube auxiliaries or count bounds. At most 2 stages may use a balanced-family method. PRN and certainty selection are not supported.

**PRN-compatible methods** (`bernoulli`, `pps_poisson`, `pps_sps`, `pps_pareto`):

When `prn` is specified, the permanent random numbers are passed through to `sondage`. Selection becomes deterministic given the PRN values, so the
`seed` argument to `execute()` has no effect on the selection outcome.

### `add_stage()`: delimit stages

`add_stage()` is syntactic: it closes the current stage and opens a new one. It carries no statistical content of its own. Stages are numbered
\(1, 2, \ldots, K\) in definition order.

### `execute()`: materialize the sample

`execute()` binds the design to one or more frames and runs the selection:

1. For each stage \(k\), the frame is restricted to units belonging to clusters selected at stage \(k - 1\) (or the full frame for \(k = 1\)).
2. Strata and cluster groupings are resolved against the (subsetted) frame.
3. The selection algorithm from `draw()` is invoked via `sondage`.
4. The `.weight`, `.fpc`, `.draw`, and `.certainty` columns are computed and suffixed with the stage number.

**One frame or one per stage.** The number of frames schedules the stages. `execute(design, hierarchy)` runs every stage against one table, and `execute(design, schools, classes, students)` gives each stage its own register, mapped by position. Any other count is refused. The frames may equally be held as a value, `execute(design, list(schools, classes, students))`, which is the same call; a name on a list element is a diagnostic label for that frame.

A register is supplied whole. samplyr restricts it to the units its parent stage selected and carries onto it the variables earlier stages introduced, so a class register needs neither pre-filtering to the sampled schools nor a copy of the school-level stratification column. A register that omits such a carried variable is fine; one whose own copy of it disagrees with the parent is an error.

**Partial execution.** `execute(design, psu_frame, stages = 1)` runs only the first stage and returns a `tbl_sample`. That object stores both the unchanged design and the realized PSU selection. Use it as the first argument of the later call and supply the new listing as the frame: `execute(psu_sample, listing_frame)`. Execution resumes at the next unexecuted stage, carries forward the PSU weights and FPC, and remains one multi-stage design.

The `stages` argument may select any initial contiguous batch beginning at stage 1; on a partial `tbl_sample`, it may select the next contiguous batch. Omitting it executes all remaining stages, with one exception. If more than one stage remains and a single frame is supplied, that frame could be the next stage's register or a hierarchy covering the rest. The two draw different samples, so `execute()` asks for `stages` rather than guessing.

**The three spellings are one implementation.** A three-stage design executed against one hierarchy, against three registers in a single call, or stage by stage as a continuation runs the same stage transition each time. All three draw the same sample, provided the continuation runs under a single RNG stream and names `stages` on every intermediate call:

```r
one_call     <- execute(design, schools, classes, students, seed = 7)
hierarchical <- execute(design, hierarchy, seed = 7)

continued <- withr::with_seed(7, {
  s1 <- execute(design, schools, stages = 1)
  s2 <- execute(s1, classes, stages = 2)
  execute(s2, students, stages = 3)
})
```

Note where the seed goes. One seed wraps the whole chain; the individual calls take none. Seeding each call separately is a different and equally valid RNG boundary, but it is not this equivalence and will select different units. The equivalence is a tested invariant rather than an incidental property, which is what makes a continuation safe to interleave with fieldwork: the sample is the one the single call would have drawn.

**Two-phase sampling.** A new phase starts with a new design: `execute(design2, phase1_sample)`. Here the phase-1 `tbl_sample` is the frame argument, not the execution object. samplyr records a previous-phase link and compounds the phase-1 and phase-2 weights multiplicatively. The conditional phase-2 weight is computed at survey export time. Thus `execute(psu_sample, listing_frame)` means “continue this design's stages,” whereas `execute(new_design, prior_sample)` means “sample a new phase.”

## Weight compounding across stages

**Single-stage.** The design weight is:

$$
w_i = \frac{1}{\pi_i}.
$$

**Multi-stage.** Let \(\pi_i^{(k)}\) denote the conditional inclusion probability of unit \(i\) at stage \(k\), given the set of clusters selected at all prior stages. The compound weight is:

$$
w_i = \prod_{k=1}^{K} \frac{1}{\pi_i^{(k)}} = \prod_{k=1}^{K} w_i^{(k)}.
$$

`samplyr` computes this by joining each stage's weight to the previous stage's weight on the shared cluster variables and multiplying. The per-stage weights are preserved as `.weight_1`, `.weight_2`, etc. The `.weight` column always equals their product.

```{r weight-compound}
two_stage <- sampling_design() |>
  add_stage("Districts") |>
    stratify_by(province) |>
    cluster_by(district) |>
    draw(n = 2, method = "srswor") |>
  add_stage("EAs") |>
    draw(n = 5, method = "srswor") |>
  execute(zwe_eas, seed = 42)

# Compound weight equals the product of per-stage weights
all.equal(two_stage$.weight,
          two_stage$.weight_1 * two_stage$.weight_2)
```

The warning is the design meeting the frame: Bulawayo is a single-district province, so a take of two districts selects the one that exists. `execute()` reports every pool it could not fill, once per stage, because a capped pool makes the design non-self-weighting. The weights above still compound exactly; Bulawayo's first-stage weight is simply 1.

**Two-phase.** When a new phase-2 design is executed with the phase-1 `tbl_sample` as its frame, the phase-1 weight is stored internally and multiplied into the final `.weight`. The conditional phase-2 weight \(w_{\text{cond}} = w_{\text{overall}} / w_{\text{phase 1}}\) is computed during survey export for use with `survey::twophase()`.

## Finite population corrections

`samplyr` stores the FPC as a population count in `.fpc_k`. The interpretation depends on the method, and the conversion for survey export happens automatically in `as_svydesign()`:

| Method class                  | `.fpc_k` stores        | Passed to survey as | survey interpretation                 |
|:------------------------------|:-----------------------|:--------------------|:--------------------------------------|
| EP-WOR (srswor, systematic)   | \(N_h\) (stratum pop.) | \(N_h\) directly    | Sampling fraction \(f_h = n_h / N_h\) |
| PPS-WOR (pps_brewer, ...)     | \(N_h\) (stratum pop.) | \(1 / w_k = \pi_i\) | Inclusion probability                 |
| WR / PMR (srswr, chromy, ...) | \(\infty\) (synthetic) | \(\infty\)          | No FPC (Hansen--Hurwitz variance)     |

The PPS conversion merits explanation. `survey::svydesign()` interprets FPC values less than 1 as inclusion probabilities and values \(\geq 1\) as
population sizes. `samplyr` stores \(N_h\) uniformly for all methods, then at export time, `as_svydesign()` creates a per-row column `.fpc_pi_k = 1 / .weight_k = \pi_i` for PPS-WOR stages. The transformation is lossless: the per-stage weight already encodes \(\pi_i\), so nothing is approximated in converting it. What is approximate for `pps_sps` and `pps_pareto` is the quantity being transformed, since their \(\pi_i\) is a target rather than the realized inclusion probability.

For WR and PMR methods, a synthetic column filled with `Inf` is created. The survey package interprets this as "no finite population correction," which gives the Hansen--Hurwitz variance estimator.

**The population the correction refers to.** `samplyr` treats the frame as executed as the population, so \(N_h\) is counted from the frame the stage drew from. That is one convention among several. Where the sampling fraction is not negligible and the population itself changes between occasions, the finite population correction matters more and admits several competing definitions, particularly for measures of change across coordinated or overlapping samples [@tam1984; @laniel1988; @nordberg2000; @berger2004change]. `samplyr` does not choose among them for you: it records the frame size it saw, and a different convention has to be imposed downstream.

## Independence across stages

`samplyr` assumes that the stage-\(k\) selection mechanism depends on stage \(k - 1\) only through the **set** of selected clusters, not through
the specific randomization that produced it. Formally:

$$
\Pr(S^{(k)} \mid S^{(k-1)}) \text{ depends on } S^{(k-1)} \text{ as a set.}
$$

This is the standard assumption in multi-stage sampling theory
[@cochran1977, ch. 11; @sarndal1992, ch. 4.3].

**Consequence for variance estimation.** Under this assumption, the recursive
multi-stage variance formula applies [@sarndal1992, ch. 4.3]. For methods with
a supported linearization estimator, `as_svydesign()` exports every executed
stage to `survey::svydesign()`: one `ids` term, one `fpc` term, and (when the
stage is stratified) one `strata` term per stage. A first-stage census, for
example, correctly attributes all variance to the later stages.

A stage that samples with replacement is handled differently: later stages
nested in it contribute nothing beyond the between-draw variability, which the
Hansen--Hurwitz estimator already captures.

Bounded cube, LPM2, and SCPS do not currently have supported linearization
estimators; they use the generic bootstrap approximation described below.

## What is lost at survey export

`as_svydesign()` translates a `tbl_sample` into a `survey::svydesign()` object. The translation is faithful for standard variance estimators, but some `samplyr` metadata is not representable in the `survey` framework.

### 1. Element sampling before later stages

An unclustered element-sampling stage followed by further stages is not nested cluster sampling: the later selections operate on the realized element sample, not within larger units, so there is no hierarchy for the multi-stage variance recursion (or for replicate-weight resampling) to use. This is phase sampling. `as_svydesign()` raises an error for this shape and the supported path is a two-phase sample: execute the element stage under its first-phase design, execute a new second-phase design with that sample as its frame, and export with `as_svydesign()`, which uses `survey::twophase()`. A *final* unclustered stage is fully supported: it is exported with a synthesized row-identity `ids` term and its own FPC.

### 2. Joint inclusion probabilities

By default, `as_svydesign()` uses Brewer's variance approximation [@berger2004]
for fixed-size PPS WOR methods, including Sampford. Here Brewer names the
variance estimator, not the selection algorithm. The approximation derives
\(\pi_{kl}\) from marginal \(\pi_i\) without computing the full
\(N \times N\) matrix.

`joint_expectation()` computes second-order quantities directly instead. What it
returns depends on the method:

- **Exact** for CPS, Sampford, systematic, and Poisson.
- **The high-entropy approximation** [@hajek1964; @brewerdonadio2003], which is
  \(O(N^2)\), for generalized Brewer, SPS, Pareto, and unconstrained cube.
- **Unavailable** for bounded cube, LPM2, and SCPS. Their supported analysis
  route is `as_svrepdesign(type = "subbootstrap")`, a generic PPS bootstrap
  approximation that does not recreate the constraints or the spatial
  algorithm.

Exact recursive formulas exist for generalized Brewer [@brewer2002, ch. 9] but
are \(O(N^3)\) and impractical for large frames.

Units with \(\pi_i = 1\) (certainty selections) are handled internally. The
joint matrix is computed on the stochastic part and reassembled with
\(\pi_{kl} = 1\) for certainty-certainty pairs and \(\pi_{kl} = \pi_l\) for
certainty-stochastic pairs.

To use the joint matrix for single-stage variance estimation, pass it via
`pps = survey::ppsmat(joint_matrix)`. See `?joint_expectation` for the full
method-by-method breakdown.

Rows and columns follow first appearance in the sample. A WR stage contains
one row per distinct selected population unit rather than one row per draw.
At a later stage below a WR parent, each parent draw occurrence defines its
own conditional block; cross-occurrence entries are products of marginal
conditional chances. Pair the matrix with stage-specific unit identities, not
blindly with every descendant sample row.

### 3. With-replacement row structure

For WR methods (`srswr`, `pps_multinomial`), each draw produces one row with `.draw_k` indexing the hit. At export, `.draw_k` becomes the sampling-unit identifier in the `ids` formula, and the FPC is set to \(\infty\). The survey package then uses the Hansen--Hurwitz variance estimator, which is the standard approach for WR designs.

### 4. Chromy / PMR semantics

`pps_chromy` is treated identically to WR at export (no FPC, no `pps` argument). The exact Sen--Yates--Grundy formula for PMR requires \(E(n_i) E(n_j) - E(n_i n_j)\) as pairwise weights [@chromy2009, eq. 5], but `survey::ppsmat()` reads \(\pi_i\) from the diagonal, which for PMR contains \(E(n_i^2)\), not \(E(n_i)\). The Hansen--Hurwitz approximation is conservative and more stable [@chauvet2019].

### 5. Panel assignments

The `.panel` column is not passed to survey. Panels are a sample-management concept: randomized fixed-quota partitions for rotation or workload distribution. The full-sample weights in `.weight` are correct for the combined sample. Panel assignment is not an additional probability-sampling phase, so a single panel does not automatically have inclusion probability \(\pi_i / k\); conditional on the frozen quotas it is the block's realized quota over the block size. Multiplying one panel's weights by \(k\) is therefore not generally valid for population inference.

### 6. Certainty flag

`samplyr` stores `.certainty_k` as a logical column. At export, certainty units (\(\pi_i = 1\)) are placed in a synthetic stratum (`.cert_stratum = "certainty"`), added to the strata formula alongside user-defined strata. The certainty stratum contributes zero variance (it is a census) and does not inflate degrees of freedom [@cochran1977, ch. 11; @sarndal1992, ch. 3.5].

### 7. Design metadata

`samplyr` stores the full `sampling_design` object, stage labels, seed, and execution history as attributes of the `tbl_sample`. These are accessible via `get_design()` and `get_stages_executed()` but are not round-tripped through the survey export.

## Panel partitioning semantics

There are two distinct operations here and they have different statistical
status. **Partitioning** (`panels = k` with no schedule) splits one execution
into \(k\) groups for rotation or workload; it does not change any inclusion
probability and does not change `.weight`. **Wave activation** (a scheduled
`panels` plus `execute(wave = t)`) selects a subset of those groups, and that
*is* a second sampling phase with its own inclusion probability and its own
weight factor. The rules below are the partitioning rules; activation is
covered under "Wave activation" after them.

Neither operation replenishes the sample. A programme that draws refreshment
cohorts against later frame vintages is a registry of separate executions, not
a partition of one; see `rotation_program()` and
`vignette("rotating-panels")`.

`execute(..., panels = k)` partitions the sample into \(k\) non-overlapping rotation groups by randomized fixed quota inside frozen ordered blocks:

- **Assignment pools**: each selection stratum of the assignment stage is a pool, cut into consecutive blocks of \(k \lceil 2 / r_{\min} \rceil\) assignment units, where \(r_{\min}\) is the fewest panels any declared wave activates. Without a schedule \(r_{\min} = 1\) and the block is \(2k\); a schedule keeping two of four panels live blocks at 4 rather than 8. The rule is that the take of \(r_{\min}\) panels must leave at least two units per block, which is the smallest take carrying a within-block variance estimate. Every block carries a fixed quota per panel and its labels are permuted within the block, so each unit carries each panel with probability \(1/k\) and panel sizes within a pool differ by at most one.
- **Clustered designs**: panels are assigned at the assignment stage's cluster or occurrence level, then propagated to every unit below. By default that stage is the first, so a PSU carries one panel and all its members inherit it. Under a with-replacement stage the assignment unit is the realized draw, so one population cluster drawn twice may carry two panels.
- **Which stage assigns**: `panel_stage` moves the assignment to another stage, and every later stage still inherits its ancestor's panel. Assigning at stage 1 rotates whole primary units; assigning lower down rotates units inside parents that stay in the survey, which is the address-panel design.
- **What moving it costs**: a pool becomes the stage's own strata inside each realized parent and never crosses a parent, so pools are smaller and `small_pool` binds more often. Selection certainty then counts only at the assignment stage, in both directions: a certainty primary unit does not make its households permanent, and a certainty selection below the assignment stage does not keep its household in every wave.
- **Control sorting**: the `control` argument to `draw()` orders the units before they are cut into blocks, so every panel inherits the same spread over that order.
- **Certainty units**: labelled from their own pools. A certainty unit is in the sample at every occasion and consumes no rotating quota.

The block sizes and the realized block-by-panel quotas are recorded with the sample and written by `write_design()`, because activating a subset of the panels is a simple random subsample without replacement of the block quota, not of \(1/k\).

Moving the assignment below stage 1 is not a default and should not be treated as one. Holding the parent fixed while its members rotate can bias cross-sectional estimates over time, and a unit that cannot move between parents is balanced for net change but not for gross change. `vignette("rotating-panels")` works through the case.

Weights are **not** adjusted by partitioning. Panels structure the output but do not affect the selection mechanism or inclusion probabilities. Analyze the combined sample with the stored weights.

### Wave activation

Activating a subset of the panels is the second operation, and it is a probability subsample rather than a filter. `execute(master, wave = t)` returns a two-phase sample: the master is phase 1, the activation is phase 2, the blocks are the phase-2 strata and their frozen quotas the phase-2 population counts. Each weight is multiplied by the inverse of its unit's activation probability, which is its block's frozen quota for the active panels over the block size. That factor is exact and is generally *not* the nominal \(k / r\) implied by counting groups, because a block that took the remainder carries a different quota. Certainty units are activated with probability one and their weights are untouched.

This is why the block sizes and realized block-by-panel quotas are recorded with the sample rather than recomputed: without them the subsample would have to be assumed to be \(1/k\), which it is not.

## PRN and sample coordination semantics

`samplyr` passes permanent random numbers through to `sondage` without transformation. PRN-compatible methods (`bernoulli`, `pps_poisson`, `pps_sps`, `pps_pareto`) produce a deterministic selection given the PRN values. When PRN is supplied, the `seed` argument to `execute()` does not affect selection.

Positive and negative coordination across survey waves is a user-level workflow, not a package feature. `samplyr` provides the infrastructure (PRN passthrough, panel partitioning, tidy output) and the user implements the update rule. See `vignette("sampling-coordination")` for the recommended workflow, including the negative-coordination formula \(u_{\text{new}} = (u - \pi) \bmod 1\) [@tille2020, sec. 8.2.4].

## References
