---
title: "Get Started with samplyr"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Get Started with samplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, echo = FALSE}
library(samplyr)
library(dplyr)
```

## The basic workflow

`samplyr` separates a sampling design from the frame on which it runs. You
describe the design first, validate or inspect it, and then execute it against
one or more frames.

The smallest complete example is a simple random sample without replacement.

```{r}
library(samplyr)
library(dplyr)

data(bfa_eas)

design <- sampling_design(title = "Burkina Faso EA sample") |>
  draw(n = 100)

sample <- execute(design, bfa_eas, seed = 24)
sample |>
  as.data.frame() |>
  dplyr::select(ea_id, region, households, .weight) |>
  head(4)
```

The frame has one row per enumeration area. The result keeps its frame columns
and adds the design information needed for analysis.

`.weight` is the overall design weight. In a one-stage sample it equals
`.weight_1`. A multistage sample also carries one weight column per stage and
multiplies them into `.weight`.

## The design grammar

Six functions cover the ordinary workflow.

| Function | Role |
|:--|:--|
| `sampling_design()` | Start a reusable design |
| `stratify_by()` | Divide a stage into sampling strata |
| `cluster_by()` | Name the units selected as clusters |
| `draw()` | Choose the size and selection method |
| `add_stage()` | Start another sampling stage |
| `execute()` | Run the design against its frame or frames |

Column names are stored in the design and resolved when a frame is supplied.
This lets the same design be validated and reused against compatible frame
vintages.

```{r}
stratified_design <- sampling_design(title = "Stratified EA sample") |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 260)

validate_frame(stratified_design, bfa_eas)
```

`validate_frame()` returns invisibly when the required columns and values are
usable. It reports problems before selection begins.

## Stratified sampling

Stratification samples independently within groups. Here `n = 260` is a total
allocated across regions in proportion to their frame sizes.

```{r}
stratified_sample <- execute(stratified_design, bfa_eas, seed = 25)

stratified_sample |>
  count(region, name = "n_sampled")
```

If `stratify_by()` has no allocation rule, a scalar `n` is interpreted per
stratum. You can also pass a named vector or a `svyplan::n_alloc()` result to
`draw()` when each stratum needs its own take.

Use equal allocation when comparable precision within each stratum is the
goal. Use proportional allocation for a self-weighting element design when
the other design features permit it. Neyman and cost-weighted allocations
need prior variability or cost information and optimize the criterion supplied
for that planning variable. They are not universal optima for every outcome
and domain.

Sample-size bounds can protect small reporting domains.

```{r}
bounded_design <- sampling_design() |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 260, min_n = 12, max_n = 30)

bounded_sample <- execute(bounded_design, bfa_eas, seed = 26)
bounded_sample |>
  count(region, name = "n_sampled")
```

For without-replacement sampling, the population of a stratum remains a hard
upper bound. If a requested allocation cannot be reached, `execute()` reports
the affected pools and uses the attainable take.

## Cluster and PPS sampling

`cluster_by()` changes the sampling unit. The next example selects enumeration
areas with probability related to their household counts.

```{r}
pps_design <- sampling_design(title = "PPS EA sample") |>
  cluster_by(ea_id) |>
  draw(n = 80, method = "pps_brewer", mos = households)

pps_sample <- execute(pps_design, bfa_eas, seed = 27)
nrow(pps_sample)
```

`mos` is the measure of size. A large measure gives a unit a larger selection
probability. If a resolved probability reaches one, that unit is selected with
certainty and contributes no variance at that stage. Certainty at one stage
does not make the final multistage weight equal to one.

`pps_brewer` is one fixed-size PPS method. Method choice also depends on the
second-order information required for variance estimation, whether a random
sample size is acceptable, and whether permanent random numbers are needed.
The comparison is in `?selection-methods`.

## Continue inside selected units

`add_stage()` starts another selection within each selected parent. For
example, adding `add_stage() |> draw(n = 12)` after the EA selection declares
a household draw within each selected EA. Execution needs household rows,
with EA identifiers linking them to their parents.

The next tutorial, `vignette("three-stage-sampling")`, works through this
process in full: commune and EA registers, household listing, continuation,
and survey export. It also explains the scope of `n` at every stage.

## Fractions, replacement, and ordering

`draw()` also accepts `frac`. Fixed-size methods convert the fraction to an
integer take, while Bernoulli sampling keeps a random realized size.

```{r}
fraction_sample <- sampling_design() |>
  draw(frac = 0.01) |>
  execute(bfa_eas, seed = 29)

nrow(fraction_sample)
```

With-replacement methods may select the same population unit several times.
Their samples carry `.draw_1`, `.draw_2`, and later-stage equivalents so every
draw occurrence remains distinct.

Systematic and sequential methods depend on frame order. Use `control` when an
explicit ordering is part of the design. Their variance properties depend on
that ordering, so an SRS variance approximation for systematic selection need
not be conservative.

## Inspecting the result

A `tbl_sample` is a tibble with a design and an execution receipt attached.
Ordinary one-to-one joins and new analysis columns preserve it. Removing or
duplicating rows changes the realized sample and prevents survey export.

```{r}
summary(stratified_sample)
```

`frame_summary()` reports what the execution resolved in each selection pool.

```{r}
frame_summary(stratified_sample, detail = "pool") |>
  select(stage, region, N, n_target, n_realized) |>
  slice_head(n = 6)
```

For domain estimates, convert the complete sample first and subset the survey
design afterwards. See `vignette("survey-analysis")` for that workflow.

## Choose your next step

The main learning path is:

1. **Get started**: the design grammar and a first sample, covered here.
2. **Select a three-stage sample**: `vignette("three-stage-sampling")` follows
   fieldwork from area registers to households and observations.
3. **Analyze the sample**: `vignette("survey-analysis")` covers outcome joins,
   domain estimates and variance choices, with optional two-phase and
   overlapping-frame examples.

Use the other articles when their task arises:

| Task | Article |
|:--|:--|
| Choose sizes and allocations before selection | `vignette("survey-planning")` |
| Coordinate new draws with permanent random numbers | `vignette("sampling-coordination")` |
| Schedule panels and refreshment cohorts | `vignette("rotating-panels")` |
| Save or replay a design | `vignette("serialization")` |

For a method or inference lookup, use `?selection-methods`. Consult
`vignette("design-semantics")` for the underlying contracts and
`vignette("validation")` for simulation diagnostics.

## A compact checklist

1. Make sure one row represents the unit selected at the current stage.
2. Declare strata, clusters, and selection methods in their sampling order.
3. Validate the design against the intended frame vintage.
4. Use a recorded seed when the selection must be reproducible.
5. Keep the returned design columns unchanged.
6. Archive the executed sample and its frame provenance.

Continue with `vignette("three-stage-sampling")`, or go directly to
`vignette("survey-analysis")` if you already have a selected sample.

## References
