---
title: "Get started with samplyr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with samplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, echo = FALSE}
library(samplyr)
library(dplyr)
```

## The basic workflow

`samplyr` separates a sampling design from the frame on which it runs. You
describe the design first, validate or inspect it, and then execute it against
one or more frames.

The smallest complete example is a simple random sample without replacement.

```{r}
library(samplyr)
library(dplyr)

data(bfa_eas)

design <- sampling_design(title = "Burkina Faso EA sample") |>
  draw(n = 100)

sample <- execute(design, bfa_eas, seed = 24)
sample |>
  as.data.frame() |>
  dplyr::select(ea_id, region, households, .weight) |>
  head(4)
```

The frame has one row per enumeration area. The result keeps its frame columns
and adds the design information needed for analysis.

`.weight` is the overall design weight. In a one-stage sample it equals
`.weight_1`. A multi-stage sample also carries one weight column per stage and
multiplies them into `.weight`.

## The design grammar

Six functions cover the ordinary workflow.

| Function | Role |
|:--|:--|
| `sampling_design()` | Start a reusable design |
| `stratify_by()` | Divide a stage into sampling strata |
| `cluster_by()` | Name the units selected as clusters |
| `draw()` | Choose the size and selection method |
| `add_stage()` | Start another sampling stage |
| `execute()` | Run the design against its frame or frames |

Column names are stored in the design and resolved when a frame is supplied.
This lets the same design be validated and reused against compatible frame
vintages.

```{r}
stratified_design <- sampling_design(title = "Stratified EA sample") |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 260)

validate_frame(stratified_design, bfa_eas)
```

`validate_frame()` returns invisibly when the required columns and values are
usable. It reports problems before selection begins.

## Stratified sampling

Stratification samples independently within groups. Here `n = 260` is a total
allocated across regions in proportion to their frame sizes.

```{r}
stratified_sample <- execute(stratified_design, bfa_eas, seed = 25)

stratified_sample |>
  count(region, name = "n_sampled")
```

If `stratify_by()` has no allocation rule, a scalar `n` is interpreted per
stratum. You can also pass a named vector or a `svyplan::n_alloc()` result to
`draw()` when each stratum needs its own take.

Equal allocation gives comparable precision within each stratum only when
the strata vary about equally, so it suits strata that are reporting domains
of similar variability. Proportional allocation divides the total by the
number of units each stratum holds at that stage, which makes an element
sample self-weighting. At a clustered stage it counts clusters, not the
households inside them. Given `importance`, such as each stratum's household
total, it allocates clusters in proportion to that size instead. Neyman and cost-weighted allocations
need prior variability or cost information and optimize the criterion supplied
for that planning variable. They are not universal optima for every outcome
and domain.

Sample-size bounds can protect small reporting domains.

```{r}
bounded_design <- sampling_design() |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 260, min_n = 12, max_n = 30)

bounded_sample <- execute(bounded_design, bfa_eas, seed = 26)
bounded_sample |>
  count(region, name = "n_sampled")
```

For without-replacement sampling, the population of a stratum remains a hard
upper bound. If a requested allocation cannot be reached, `execute()` reports
the affected pools and uses the attainable take.

## Cluster and PPS sampling

`cluster_by()` changes the sampling unit. The next example selects enumeration
areas with probability related to their household counts.

```{r}
pps_design <- sampling_design(title = "PPS EA sample") |>
  cluster_by(ea_id) |>
  draw(n = 80, method = "pps_brewer", mos = households)

pps_sample <- execute(pps_design, bfa_eas, seed = 27)
nrow(pps_sample)
```

`mos` is the measure of size. A large measure gives a unit a larger selection
probability. If a resolved probability reaches one, that unit is selected with
certainty and contributes no variance at that stage. Certainty at one stage
does not make the final multi-stage weight equal to one.

`pps_brewer` is one fixed-size PPS method. Method choice also depends on the
second-order information required for variance estimation, whether a random
sample size is acceptable, and whether permanent random numbers are needed.
The comparison is in `?selection-methods`.

`cluster_by()` also names each unit's identifier, and a frame with one row per
unit can use it too. With the same seed, the design below selects the same EAs
with the same weights as `draw(n = 200)` alone, and it records `ea_id` as the
identity of each selected unit.

```{r unit-identity}
screening <- sampling_design(title = "Screening sample") |>
  cluster_by(ea_id) |>
  draw(n = 200)

screened <- execute(screening, bfa_eas, seed = 25)
nrow(screened)
```

Declare it whenever a second phase may later subsample this sample. The two
phases are linked through an identifier both designs declare. An ID column
the design does not name does not link them, so the survey export of the
second phase would be refused. The examples in `?execute` select a second
phase, and `?as_svydesign` gives its export.

## A three-stage household sample

`add_stage()` starts another selection within each selected parent. This
section selects communes within regions, enumeration areas (EAs) within those
communes, and households within those EAs. Selection pauses while the
household listing is prepared, then continues under the same design.

`bfa_eas` is a derived Burkina Faso frame. Its modeled EAs, household counts
and historical administrative labels are teaching inputs, not a current
official household register. The household listing and interview outcomes
below are synthetic.

### Give each register its own sampling unit and size

```{r registers}
ea_register <- bfa_eas |>
  mutate(
    region,
    commune_id = paste(region, province, commune, sep = "/"),
    ea_id,
    ea_households = households,
    .keep = "none"
  )

commune_register <- ea_register |>
  summarise(commune_households = sum(ea_households),
            .by = c(region, commune_id))

head(commune_register, 3)
```

The measure of size belongs to the unit being selected, so each level has its
own register. Their order in `execute()` maps them to stages, and names on a
list are labels, not a join specification.

### Declare all three stages

```{r design}
household_design <- sampling_design("Three-stage household survey") |>
  add_stage("Communes") |>
    stratify_by(region) |>
    cluster_by(commune_id) |>
    draw(n = 4, method = "pps_brewer", mos = commune_households) |>
  add_stage("EAs") |>
    cluster_by(ea_id) |>
    draw(n = 3, method = "pps_brewer", mos = ea_households) |>
  add_stage("Households") |>
    draw(n = 12)

household_design
```

The takes are four communes **per region**, three EAs **per selected
commune**, and twelve households **per selected EA**. The first stage has no
allocation rule, so `n = 4` is per stratum. A take that differs by EA is a
table keyed by `ea_id`, described under `n` in `?draw`.

### Select areas, then continue from the listing

```{r select-areas}
selected_eas <- execute(
  household_design,
  list(commune_register, ea_register),
  stages = 1:2,
  seed = 91
)

selected_eas |>
  as.data.frame() |>
  count(region, name = "selected_eas")
```

The EA register covers all communes, and execution restricts it to the
selected parents. In fieldwork, a listing team then enumerates households in
the selected EAs. Here an artificial listing gives each selected EA as many
households as its modeled count.

```{r listing}
household_listing <- selected_eas |>
  as.data.frame() |>
  reframe(hh_id = seq_len(ea_households), .by = c(commune_id, ea_id))
```

A real listing rarely finds exactly the modeled count. That does not bias the
weights, because the household stage uses the take over the listed count.
Passing the listing to `execute()` with the partial sample continues the
design.

```{r select-households}
issued <- execute(selected_eas, household_listing, seed = 92)

issued |>
  as.data.frame() |>
  select(commune_id, ea_id, hh_id, .weight_1, .weight_2, .weight_3, .weight) |>
  head(4)
```

The warning names the EAs listed with fewer households than the take. Each is
taken whole, which keeps its weight exact but makes the sample
non-self-weighting. The final weight is the reciprocal of the product of the
three conditional selection probabilities. Continuation keeps the earlier
factors, whereas starting a new design on `selected_eas` would declare a new
sampling phase.

### Join observations and hand off to analysis

Keep every issued row and join the collected outcomes to it. The artificial
consumption values stand for an outcome collected in the field.

```{r analysis, eval = requireNamespace("survey", quietly = TRUE)}
set.seed(93)
interviews <- issued |>
  as.data.frame() |>
  mutate(ea_id, hh_id,
         consumption = round(rlnorm(n(), log(100000), 0.5)),
         .keep = "none")

observed <- issued |>
  left_join(interviews, by = c("ea_id", "hh_id"), relationship = "one-to-one")

survey::svymean(~consumption, as_svydesign(observed))
```

This export uses Brewer's variance approximation for the PPS stages.
`?variance-estimation` gives the alternatives, including RWYB replicates, and
`vignette("survey-analysis")` continues with outcome joins, domains and
nonresponse.

## Fractions, replacement, and ordering

`draw()` also accepts `frac`. Fixed-size methods convert the fraction to an
integer take, while Bernoulli sampling keeps a random realized size.

```{r}
fraction_sample <- sampling_design() |>
  draw(frac = 0.01) |>
  execute(bfa_eas, seed = 29)

nrow(fraction_sample)
```

With-replacement methods may select the same population unit several times.
Their samples carry `.draw_1`, `.draw_2`, and later-stage equivalents so every
draw occurrence remains distinct.

Systematic and sequential methods depend on frame order. Use `control` when an
explicit ordering is part of the design. Their variance properties depend on
that ordering, so an SRS variance approximation for systematic selection need
not be conservative.

## Coordinate draws with permanent random numbers

A permanent random number (PRN) is a uniform value stored with each unit in
the register. Methods that accept one select deterministically from it, so a
later draw that reuses the PRNs selects the same units, and a draw on shifted
PRNs avoids them. `bernoulli`, `pps_poisson`, `pps_sps`, `pps_pareto` and
`scps` accept a `prn` column.

```{r prn}
set.seed(30)
register <- bfa_eas |>
  mutate(prn = runif(n()))

coordinated <- sampling_design() |>
  draw(n = 100, method = "pps_sps", mos = households, prn = prn)

first <- execute(coordinated, register, seed = 1)
second <- execute(coordinated, register, seed = 2)
setequal(first$ea_id, second$ea_id)
```

The seeds differ, but the samples do not. `?draw` gives the methods that
accept `prn` and what fixes an `scps` selection. The book chapter
[Coordinate samples with permanent random numbers](https://www.ahmadoudicko.com/sampling-design-r-book/chapters/coordination.html)
covers negative coordination, register updates and response burden.

## Inspecting the result

A `tbl_sample` is a tibble with a design and an execution receipt attached.
Ordinary one-to-one joins and new analysis columns preserve it. Removing or
duplicating rows changes the realized sample and prevents survey export.

```{r}
summary(stratified_sample)
```

`frame_summary()` reports what the execution resolved in each selection pool.

```{r}
frame_summary(stratified_sample, detail = "pool") |>
  select(stage, region, N, n_target, n_realized) |>
  slice_head(n = 6)
```

For domain estimates, convert the complete sample first and subset the survey
design afterwards. See `vignette("survey-analysis")` for that workflow.

## Choose your next step

This article covered the design grammar, a first sample, and a three-stage
household sample followed from area registers to households and observations.
The next step on the main path is to analyze the sample:
`vignette("survey-analysis")` covers outcome joins, domain estimates and
variance choices.

Use the other articles when their task arises:

| Task | Where |
|:--|:--|
| Choose sizes and allocations before selection | `vignette("svyplan", package = "svyplan")`, then [Plan a sample with svyplan](https://www.ahmadoudicko.com/sampling-design-r-book/chapters/plan-to-draw.html) |
| Coordinate new draws with permanent random numbers | `?draw`, argument `prn`, and [Coordinate samples with permanent random numbers](https://www.ahmadoudicko.com/sampling-design-r-book/chapters/coordination.html) |
| Schedule panels and refreshment cohorts | `?panel-assignment`, `?rotation_program`, and [Plan and run a rotating panel](https://www.ahmadoudicko.com/sampling-design-r-book/chapters/rotating-panels.html) |
| Save or replay a design | `vignette("serialization")` |

For a method lookup, use `?selection-methods`, and for how each method's
variance is estimated, `?variance-estimation`. Consult
`vignette("design-semantics")` for the underlying contracts and
[Validation on synthetic populations](https://www.ahmadoudicko.com/sampling-design-r-book/chapters/validation.html)
for simulation diagnostics. The book
[Survey Sampling Design with R](https://www.ahmadoudicko.com/sampling-design-r-book/) collects these
chapters with the `svyplan` planning guides.

## A compact checklist

1. Make sure one row represents the unit selected at the current stage.
2. Declare strata, clusters, and selection methods in their sampling order.
3. Validate the design against the intended frame vintage.
4. Use a recorded seed when the selection must be reproducible.
5. Keep the returned design columns unchanged.
6. Archive the executed sample and its frame provenance.

Continue with `vignette("survey-analysis")` to analyze the selected sample.

