---
title: "Get started with samplyr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with samplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, echo = FALSE}
library(samplyr)
library(dplyr)
```

## The basic workflow

`samplyr` separates a sampling design from the frame on which it runs. You
describe the design first, validate or inspect it, and then execute it against
one or more frames.

The smallest complete example is a simple random sample without replacement.

```{r}
library(samplyr)
library(dplyr)

data(bfa_eas)

design <- sampling_design(title = "Burkina Faso EA sample") |>
  draw(n = 100)

sample <- execute(design, bfa_eas, seed = 24)
sample |>
  as.data.frame() |>
  dplyr::select(ea_id, region, households, .weight) |>
  head(4)
```

The frame has one row per enumeration area. The result keeps its frame columns
and adds the design information needed for analysis.

`.weight` is the overall design weight. In a one-stage sample it equals
`.weight_1`. A multi-stage sample also carries one weight column per stage and
multiplies them into `.weight`.

## The design grammar

Six functions cover the ordinary workflow.

| Function | Role |
|:--|:--|
| `sampling_design()` | Start a reusable design |
| `stratify_by()` | Divide a stage into sampling strata |
| `cluster_by()` | Name the units selected as clusters |
| `draw()` | Choose the size and selection method |
| `add_stage()` | Start another sampling stage |
| `execute()` | Run the design against its frame or frames |

Column names are stored in the design and resolved when a frame is supplied.
This lets the same design be validated and reused against compatible frame
vintages.

```{r}
stratified_design <- sampling_design(title = "Stratified EA sample") |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 260)

validate_frame(stratified_design, bfa_eas)
```

`validate_frame()` returns invisibly when the required columns and values are
usable. It reports problems before selection begins.

## Stratified sampling

Stratification samples independently within groups. Here `n = 260` is a total
allocated across regions in proportion to their frame sizes.

```{r}
stratified_sample <- execute(stratified_design, bfa_eas, seed = 25)

stratified_sample |>
  count(region, name = "n_sampled")
```

If `stratify_by()` has no allocation rule, a scalar `n` is interpreted per
stratum. You can also pass a named vector or a `svyplan::n_alloc()` result to
`draw()` when each stratum needs its own take.

Equal allocation gives comparable precision within each stratum only when
the strata vary about equally, so it suits strata that are reporting domains
of similar variability. Proportional allocation divides the total by the
number of units each stratum holds at that stage, which makes an element
sample self-weighting. At a clustered stage it counts clusters, not the
households inside them. Given `importance`, such as each stratum's household
total, it allocates clusters in proportion to that size instead. Neyman and cost-weighted allocations
need prior variability or cost information and optimize the criterion supplied
for that planning variable. They are not universal optima for every outcome
and domain.

Sample-size bounds can protect small reporting domains.

```{r}
bounded_design <- sampling_design() |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 260, min_n = 12, max_n = 30)

bounded_sample <- execute(bounded_design, bfa_eas, seed = 26)
bounded_sample |>
  count(region, name = "n_sampled")
```

For without-replacement sampling, the population of a stratum remains a hard
upper bound. If a requested allocation cannot be reached, `execute()` reports
the affected pools and uses the attainable take.

## Cluster and PPS sampling

`cluster_by()` changes the sampling unit. The next example selects enumeration
areas with probability related to their household counts.

```{r}
pps_design <- sampling_design(title = "PPS EA sample") |>
  cluster_by(ea_id) |>
  draw(n = 80, method = "pps_brewer", mos = households)

pps_sample <- execute(pps_design, bfa_eas, seed = 27)
nrow(pps_sample)
```

`mos` is the measure of size. A large measure gives a unit a larger selection
probability. If a resolved probability reaches one, that unit is selected with
certainty and contributes no variance at that stage. Certainty at one stage
does not make the final multi-stage weight equal to one.

`pps_brewer` is one fixed-size PPS method. Method choice also depends on the
second-order information required for variance estimation, whether a random
sample size is acceptable, and whether permanent random numbers are needed.
The comparison is in `?selection-methods`.

`cluster_by()` also names each unit's identifier, and a frame with one row per
unit can use it too. With the same seed, the design below selects the same EAs
with the same weights as `draw(n = 200)` alone, and it records `ea_id` as the
identity of each selected unit.

```{r unit-identity}
screening <- sampling_design(title = "Screening sample") |>
  cluster_by(ea_id) |>
  draw(n = 200)

screened <- execute(screening, bfa_eas, seed = 25)
nrow(screened)
```

Declare it whenever a second phase may later subsample this sample. The two
phases are linked through an identifier both designs declare. An ID column
the design does not name does not link them, so the survey export of the
second phase would be refused. `vignette("survey-analysis")` works through a
screening sample followed by a detailed interview.

## A three-stage household sample

`add_stage()` starts another selection within each selected parent. This
section selects communes within regions, enumeration areas (EAs) within those
communes, and households within those EAs. The first two stages use PPS
without replacement and the last uses simple random sampling without
replacement. Selection pauses while the household listing is prepared, then
continues under the same design.

`bfa_eas` is a derived Burkina Faso frame. Its modeled EAs, household counts
and historical administrative labels are teaching inputs, not a current
official household register. The household listing and interview outcomes
below are synthetic.

### Give each register its own sampling unit and size

```{r registers}
ea_register <- bfa_eas |>
  mutate(
    region,
    commune_id = paste(region, province, commune, sep = "/"),
    ea_id,
    ea_households = households,
    .keep = "none"
  )

commune_register <- ea_register |>
  summarise(commune_households = sum(ea_households),
            .by = c(region, commune_id))

head(commune_register, 3)
```

| Register            | One row represents           | Key                                   | Measure of size                            |
|---------------------|------------------------------|---------------------------------------|--------------------------------------------|
| `commune_register`  | A commune                    | `commune_id`, qualified by geography  | Sum of EA household counts in that commune |
| `ea_register`       | An EA                        | `ea_id`, with its parent `commune_id` | That EA's modeled household count          |
| `household_listing` | A household in a selected EA | `ea_id` and local `hh_id`             | None, equal-probability sampling           |

The measure of size belongs to the unit being selected. In one shared
hierarchy frame, a commune's total would have to repeat on each of its EA
rows. Separate registers make the two levels explicit. Their order in
`execute()` maps them to stages, and names on a list are labels, not a join
specification.

### Declare all three stages

```{r design}
household_design <- sampling_design("Three-stage household survey") |>
  add_stage("Communes") |>
    stratify_by(region) |>
    cluster_by(commune_id) |>
    draw(n = 4, method = "pps_brewer", mos = commune_households) |>
  add_stage("EAs") |>
    cluster_by(ea_id) |>
    draw(n = 3, method = "pps_brewer", mos = ea_households) |>
  add_stage("Households") |>
    draw(n = 12)

household_design
```

The takes are four communes **per region**, three EAs **per selected
commune**, and twelve households **per selected EA**. The first stage has no
allocation rule, so `n = 4` is per stratum. A pool smaller than its take is
a census, which execution reports. Region stratification does not carry over
as a stratum specification to later stages.

### Select areas, then pause for listing

```{r select-areas}
selected_eas <- execute(
  household_design,
  list(commune_register, ea_register),
  stages = 1:2,
  seed = 91
)

selected_eas |>
  as.data.frame() |>
  count(region, name = "selected_eas")
```

The EA register covers all communes, and execution restricts it to the
selected parents. The result holds the two completed stages and their weights,
and it is the starting point for continuation.

In fieldwork, a listing team enumerates households in the selected EAs. Here
an artificial listing gives each selected EA as many households as its modeled
count. It creates identifiers, not household records.

```{r listing}
household_listing <- selected_eas |>
  as.data.frame() |>
  reframe(hh_id = seq_len(ea_households), .by = c(commune_id, ea_id))

stopifnot(!anyDuplicated(household_listing[c("ea_id", "hh_id")]))
head(household_listing, 3)
```

The listing is a plain data frame because it enumerates candidates for the
next stage. The parent EA tells apart household IDs that repeat across EAs.

A real listing rarely finds exactly the number of households the EA was
selected on. That does not bias the weights. The EA stage used the measure of
size in its register, and the household stage uses the take over the listed
count, so each factor is the probability of the selection actually made. What
the difference costs is self-weighting. The final weight varies with the ratio
of listed to modeled households, and an EA listed with fewer households than
the take is taken whole, which execution reports.

When a selected EA is too large to list, it is divided into segments, the
segments are counted, and one is selected with probability proportional to its
count. That is one more stage between EAs and households,
`add_stage("Segments") |> cluster_by(segment_id) |> draw(n = 1, method = "pps_brewer", mos = segment_households)`,
with a segment register as its frame. Its probability enters the weight like
any other stage's, and the household listing then covers the selected segment
only.

### Continue from the listing

```{r select-households}
issued <- execute(selected_eas, household_listing, seed = 92)

issued |>
  as.data.frame() |>
  select(commune_id, ea_id, hh_id, .weight_1, .weight_2, .weight_3, .weight) |>
  head(4)

stopifnot(isTRUE(all.equal(
  issued$.weight,
  issued$.weight_1 * issued$.weight_2 * issued$.weight_3
)))
frame_summary(issued)
```

The final weight is the reciprocal of the product of the three conditional
selection probabilities. Continuation keeps the earlier factors, whereas
starting a new design on `selected_eas` would declare a new sampling phase.
The two seeds specify separate reproducible randomization steps, so matching a
one-call execution exactly would need the same uninterrupted RNG stream.

A person stage continues the same way from a household roster, for example
`add_stage("Person") |> draw(n = 1, on_empty = "silent")` for one eligible
person per household. A selected household with no eligible member has no
roster rows. By default that refuses the stage, because a missing roster is
more often an error than an empty household. `on_empty = "silent"` or
`"warn"` accepts such households: they stay in the design, contribute zero to
every total, and are counted by `summary()`. An EA in which no household has
an eligible member is left with no row at all. `as_svydesign()` then warns
(`samplyr_warning_export_empty_parent`), because survey computes that EA's
term within its commune without it, and RWYB refuses the sample. A commune
left with no row is refused by every export route.

### Set a household take per EA before selection

The household stage above takes twelve households in every EA. A take that
differs by EA and is fixed before any selection is a table keyed by `ea_id`.
Stratifying the household stage by that identifier applies each row within
its own EA. The table can cover every EA in the register, although only the
selected ones will be listed.

```{r per-ea-take}
takes <- ea_register |>
  mutate(
    ea_id,
    # Never more than the EA's modeled count.
    n = pmin(if_else(ea_households < 150, 8, 12), ea_households),
    .keep = "none"
  )

per_ea_design <- sampling_design("Three-stage survey, take per EA") |>
  add_stage("Communes") |>
    stratify_by(region) |>
    cluster_by(commune_id) |>
    draw(n = 4, method = "pps_brewer", mos = commune_households) |>
  add_stage("EAs") |>
    cluster_by(ea_id) |>
    draw(n = 3, method = "pps_brewer", mos = ea_households) |>
  add_stage("Households") |>
    stratify_by(ea_id) |>
    draw(n = takes)

areas <- execute(
  per_ea_design,
  list(commune_register, ea_register),
  stages = 1:2,
  seed = 91
)
area_listing <- areas |>
  as.data.frame() |>
  reframe(hh_id = seq_len(ea_households), .by = c(commune_id, ea_id))

per_ea <- execute(areas, area_listing, seed = 92)
per_ea |>
  as.data.frame() |>
  count(ea_id, name = "households") |>
  left_join(takes, by = "ea_id") |>
  head(4)
```

Each selected EA takes its own row, and the rows for EAs that were not
selected are simply unused. A mistyped `ea_id` that matches no selected EA is
therefore not caught at this point. Check the table against the register
before selection, for example with `setdiff(takes$ea_id, ea_register$ea_id)`.

### Join observations and hand off to analysis

Every issued household responds in this example. The artificial consumption
values stand for an outcome collected in the field.

```{r observations}
set.seed(93)
interviews <- issued |>
  as.data.frame() |>
  mutate(ea_id, hh_id,
         consumption = round(rlnorm(n(), log(100000), 0.5)),
         .keep = "none")

observed <- issued |>
  left_join(interviews, by = c("ea_id", "hh_id"), relationship = "one-to-one")
stopifnot(nrow(observed) == nrow(issued))
```

With real nonresponse, keep the issued rows and join response indicators and
measurements to them. Sampling weights describe selection, and dropping
nonrespondents is not a response adjustment. The analyst chooses and
justifies that adjustment downstream.

```{r analysis, eval = requireNamespace("survey", quietly = TRUE)}
linearized <- as_svydesign(observed)
survey::svymean(~consumption, linearized)
```

This export uses Brewer's variance approximation for the PPS stages. Exact
first-order probabilities do not make that variance approximation exact.
svrep's RWYB replicates are an alternative, also approximate for PPS without
replacement.

```{r replication, eval = requireNamespace("svrep", quietly = TRUE)}
set.seed(94)
replicated <- as_svrepdesign(observed, type = "rwyb", replicates = 500)
survey::svymean(~consumption, replicated)
```

The warning describes the PPS approximation. More replicates reduce
simulation error, not the error of the variance approximation.
`vignette("survey-analysis")` continues with outcome joins, domains and other
analysis designs.

## Fractions, replacement, and ordering

`draw()` also accepts `frac`. Fixed-size methods convert the fraction to an
integer take, while Bernoulli sampling keeps a random realized size.

```{r}
fraction_sample <- sampling_design() |>
  draw(frac = 0.01) |>
  execute(bfa_eas, seed = 29)

nrow(fraction_sample)
```

With-replacement methods may select the same population unit several times.
Their samples carry `.draw_1`, `.draw_2`, and later-stage equivalents so every
draw occurrence remains distinct.

Systematic and sequential methods depend on frame order. Use `control` when an
explicit ordering is part of the design. Their variance properties depend on
that ordering, so an SRS variance approximation for systematic selection need
not be conservative.

## Inspecting the result

A `tbl_sample` is a tibble with a design and an execution receipt attached.
Ordinary one-to-one joins and new analysis columns preserve it. Removing or
duplicating rows changes the realized sample and prevents survey export.

```{r}
summary(stratified_sample)
```

`frame_summary()` reports what the execution resolved in each selection pool.

```{r}
frame_summary(stratified_sample, detail = "pool") |>
  select(stage, region, N, n_target, n_realized) |>
  slice_head(n = 6)
```

For domain estimates, convert the complete sample first and subset the survey
design afterwards. See `vignette("survey-analysis")` for that workflow.

## Choose your next step

This article covered the design grammar, a first sample, and a three-stage
household sample followed from area registers to households and
observations. The next step on the main path is to analyze the sample:
`vignette("survey-analysis")` covers outcome joins, domain estimates and
variance choices, with optional two-phase and overlapping-frame examples.

Use the other articles when their task arises:

| Task | Article |
|:--|:--|
| Choose sizes and allocations before selection | `vignette("survey-planning")` |
| Coordinate new draws with permanent random numbers | `vignette("sampling-coordination")` |
| Schedule panels and refreshment cohorts | `vignette("rotating-panels")` |
| Save or replay a design | `vignette("serialization")` |

For a method lookup, use `?selection-methods`, and for how each method's
variance is estimated, `?variance-estimation`. Consult
`vignette("design-semantics")` for the underlying contracts and
`vignette("validation")` for simulation diagnostics.

## A compact checklist

1. Make sure one row represents the unit selected at the current stage.
2. Declare strata, clusters, and selection methods in their sampling order.
3. Validate the design against the intended frame vintage.
4. Use a recorded seed when the selection must be reproducible.
5. Keep the returned design columns unchanged.
6. Archive the executed sample and its frame provenance.

Continue with `vignette("survey-analysis")` to analyze the selected sample.

