---
title: "Introduction to samplyr"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Introduction to samplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, echo = FALSE}
library(samplyr)
library(dplyr)
```

## Overview

`samplyr` provides a tidy grammar for specifying and executing survey sampling designs. The package is built around a minimal set of composable verbs that handle stratified, clustered, and multi-stage sampling.

For the statistical semantics and assumptions underlying each operation, see `vignette("design-semantics")`.

The core idea is that sampling code should read like its English description:

```{r, eval = FALSE}
library(samplyr)
library(dplyr)

# "Stratify by region, proportionally allocate 500 samples, execute"
sampling_design(title = "My sampling design") |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 500) |>
  execute(frame, seed = 1)
```

To see how this works in practice, consider a real survey design from @lohr2022 (Example 7.1), based on a 1991 study of bed net use in rural Gambia (D'Alessandro et al., 1994):

> Malaria morbidity can be reduced by using bed nets impregnated with insecticide, but this is only effective if the bed nets are in widespread use. In 1991, a nationwide survey was designed to estimate the prevalence of bed net use in rural areas of the Gambia (D'Alessandro et al., 1994).
>
> The sampling frame consisted of all rural villages of fewer than 3,000 people. The villages were **stratified by three geographic regions** (eastern, central, and western) and by **whether the village had a public health clinic (PHC)** or not. In each region **five districts were chosen with probability proportional to the district population**. In each district **four villages were chosen, again with probability proportional to census population**: two PHC villages and two non-PHC villages. Finally, **six compounds were chosen** more or less randomly from each village.

In `samplyr`, this three-stage stratified cluster design translates directly into code:


```{r}
design <- sampling_design(title = "Gambia bed nets") |>
  add_stage() |>
    stratify_by(region) |>
    cluster_by(district) |>
    draw(n = 5, method = "pps_brewer", mos = population) |>
  add_stage() |>
    stratify_by(phc) |>
    cluster_by(village) |>
    draw(n = 2, method = "pps_brewer", mos = population) |>
  add_stage() |>
    draw(n = 6)
design
```

And you can then use `execute` to get your sample.

```{r, eval = FALSE}
sampling_design(title = "Gambia bed net") |>
  add_stage() |>
    stratify_by(region) |>
    cluster_by(district) |>
    draw(n = 5, method = "pps_brewer", mos = population) |>
  add_stage() |>
    stratify_by(phc) |>
    cluster_by(village) |>
    draw(n = 2, method = "pps_brewer", mos = population) |>
  add_stage() |>
    draw(n = 6) |>
  execute(frame, seed = 1991)

# or
execute(design, frame, seed = 1991)
```

The `samplyr` code mirrors the verbal description verb for verb.

## The Grammar

`samplyr` uses 5 verbs and 1 modifier:

| Function | Purpose |
|----------|---------|
| `sampling_design()` | Create a new sampling design |
| `stratify_by()` | Define stratification variables and allocation |
| `cluster_by()` | Define cluster/PSU variable |
| `draw()` | Specify sample size, method |
| `execute()` | Run the design on a frame |
| `add_stage()` | Delimit stages in multi-stage designs |

## Deferred Column Resolution

`stratify_by()` and `cluster_by()` accept bare column names and store
them in the design object. Column names are validated only when
a frame is available (`execute()`, `as_svydesign()`), which
keeps design specification separate from execution.

```{r, eval = FALSE}
design <- sampling_design(title = "Stratified cluster sampling") |>
  stratify_by(region, alloc = "proportional") |>
  cluster_by(ea_id) |>
  draw(n = 200)

sample <- execute(design, bfa_eas, seed = 101)
```

## Example Data

We'll use the `bfa_eas` dataset throughout this vignette. It is an enumeration area frame for Burkina Faso, with 44,570 EAs across 13 regions.

```{r}
data(bfa_eas)

bfa_eas |>
  glimpse()
```

---

## Simple Random Sampling

The most basic design selects n units at random from the frame.

```{r}
design <- sampling_design(title = "Simple Random Sampling") |>
  draw(n = 100)
design
```

The `sampling_design` object stores design metadata and can be reused.

```{r}
sample <- execute(design, bfa_eas, seed = 24)
nrow(sample)
```


The result includes the original columns plus sampling metadata: `.weight` (sampling weight), `.weight_1`, `.weight_2`, etc. (per-stage weights). For with-replacement methods, `.draw_1`, `.draw_2`, etc. number each independent draw.

```{r}
sample |>
  as_tibble() |>
  select(ea_id, region, urban_rural, .weight, .weight_1)
```

### Selection Methods

The `method` argument controls how units are selected. By default, `samplyr` uses simple random sampling without replacement (`srswor`).

Systematic sampling selects units at fixed intervals, which can improve precision when the frame is ordered.

```{r}
sample_sys <- sampling_design(title = "Systematic") |>
  draw(n = 100, method = "systematic") |>
  execute(bfa_eas, seed = 2021)
sample_sys
```

Sampling with replacement allows the same unit to be selected multiple times, useful for bootstrap procedures.

```{r}
sample_wr <- sampling_design(title = "SRS WR") |>
  draw(n = 100, method = "srswr") |>
  execute(bfa_eas, seed = 123)
sample_wr
```

### Using Sampling Fractions

Instead of a fixed sample size, you can specify a sampling fraction with `frac`.

```{r}
sample <- sampling_design() |>
  draw(frac = 0.05) |>
  execute(bfa_eas, seed = 1789)
nrow(sample)
```

Bernoulli sampling selects each unit independently with the specified probability. This gives a random sample size, which can simplify field protocols.

```{r}
sample <- sampling_design() |>
  draw(frac = 0.05, method = "bernoulli") |>
  execute(bfa_eas, seed = 1960)
nrow(sample)
```

### Rounding Behavior

When using `frac`, the computed sample size (N × frac) must be rounded to an integer. By default, `samplyr` rounds up (ceiling), which matches SAS SURVEYSELECT's default behavior. Use the `round` argument to control this:

- `"up"` (default): round up, ensures adequate sample size
- `"down"`: round down, minimum sample size
- `"nearest"`: Standard mathematical rounding

```{r}
# Frame of 105 units, frac = 0.1 -> 10.5
frame_105 <- head(bfa_eas, 105)

# Default: rounds up to 11
nrow(sampling_design() |> draw(frac = 0.1) |> execute(frame_105, seed = 3))

# Round down to 10
nrow(sampling_design() |> draw(frac = 0.1, round = "down") |> execute(frame_105, seed = 3))

# Round to nearest (10)
nrow(sampling_design() |> draw(frac = 0.1, round = "nearest") |> execute(frame_105, seed = 3))
```

---

## Stratified Sampling

Stratification partitions the frame into non-overlapping groups (strata) and samples from each. This ensures representation from all subgroups and often improves precision.

### Per-Stratum Sample Size

Without an allocation method, `n` specifies the sample size within each stratum.

```{r}
sample <- sampling_design() |>
  stratify_by(region) |>
  draw(n = 20) |>
  execute(bfa_eas, seed = 123)

sample |>
  count(region)
```

### Allocation Methods

With an allocation method, `n` becomes the total sample size to distribute across strata. Allocation methods are defined for total sample-size allocation, so `draw(frac = ...)` is not supported when `alloc` is set.

**Proportional allocation** distributes the sample proportionally to stratum sizes, ensuring the sample mirrors the population structure.

```{r}
sample <- sampling_design(title = "Proportional Allocation") |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 300) |>
  execute(bfa_eas, seed = 42)

sample |>
  count(region) |>
  mutate(pct = n / sum(n))
```

**Equal allocation** assigns the same sample size to each stratum, regardless of population size. This maximizes precision for comparisons between strata.

```{r}
sample <- sampling_design(title = "Equal Allocation") |>
  stratify_by(region, alloc = "equal") |>
  draw(n = 160) |>
  execute(bfa_eas, seed = 2026)

sample |>
  count(region)
```

**Neyman allocation** minimizes the variance of the overall estimate by allocating more sample to strata with higher variability. It requires prior information about stratum variances.
Use explicit stratification variable names when passing `variance`.

```{r}
data(bfa_eas_variance)

bfa_eas_variance

sample <- sampling_design(title = "Neyman Allocation") |>
  stratify_by(region, alloc = "neyman", variance = bfa_eas_variance) |>
  draw(n = 300) |>
  execute(bfa_eas, seed = 12345)

sample |>
  count(region)
```

**Optimal allocation** extends Neyman allocation by also accounting for differential costs across strata. It minimizes variance for a fixed budget (or cost for fixed precision).
Use explicit stratification variable names when passing `variance`/`cost`.

```{r}
data(bfa_eas_cost)

bfa_eas_cost

sample <- sampling_design(title = "Optimal Allocation") |>
  stratify_by(region, alloc = "optimal",
              variance = bfa_eas_variance,
              cost = bfa_eas_cost) |>
  draw(n = 300) |>
  execute(bfa_eas, seed = 9876)

sample |>
  count(region)
```

**Power allocation** [@bankier1988] uses \(n_h \propto C_h X_h^q\), where `cv` supplies \(C_h\), `importance` supplies \(X_h\), and `power` is \(q \in [0, 1]\).

```{r}
cv_df <- data.frame(
  region = levels(bfa_eas$region),
  cv = c(0.40, 0.35, 0.12, 0.20, 0.30, 0.18,
         0.15, 0.38, 0.22, 0.32, 0.17, 0.45, 0.25)
)

importance_df <- data.frame(
  region = levels(bfa_eas$region),
  importance = c(60, 40, 120, 70, 80, 65,
                 50, 55, 90, 75, 45, 35, 30)
)

sample <- sampling_design(title = "Power Allocation") |>
  stratify_by(
    region,
    alloc = "power",
    cv = cv_df,
    importance = importance_df,
    power = 0.5
  ) |>
  draw(n = 300) |>
  execute(bfa_eas, seed = 777)

sample |>
  count(region)
```

### Sample Size Bounds

When using allocation methods, you can constrain stratum sample sizes with `min_n` and `max_n` in `draw()`.

**Minimum sample size** ensures every stratum gets at least the specified number of units. This is essential for variance estimation (requires n ≥ 2 per stratum) or when you need reliable subgroup estimates.

```{r}
# Neyman allocation with minimum 5 per stratum
sample <- sampling_design() |>
  stratify_by(region, alloc = "neyman", variance = bfa_eas_variance) |>
  draw(n = 300, min_n = 20) |>
  execute(bfa_eas, seed = 1)

sample |>
  count(region)
```

**Maximum sample size** caps large strata to prevent them from dominating the sample. This is useful when operational constraints limit how many units you can handle in any one stratum.

```{r}
# Proportional allocation but cap at 60 per stratum
sample <- sampling_design() |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 300, max_n = 55) |>
  execute(bfa_eas, seed = 40)

sample |>
  count(region)
```

**Both bounds together** create a feasible range for each stratum:

```{r}
sample <- sampling_design() |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 300, min_n = 20, max_n = 55) |>
  execute(bfa_eas, seed = 2003)

sample |>
  count(region)
```

When a stratum's population is smaller than `min_n`, the entire stratum is selected (capped at population size).

**Custom allocation** lets you specify exact sample sizes or rates per stratum by passing a data frame to `draw()`.

```{r}
sizes_df <- data.frame(
  region = levels(bfa_eas$region),
  n = c(20, 12, 25, 18, 22, 16, 14, 15, 20, 18, 12, 10, 8)
)

sample <- sampling_design() |>
  stratify_by(region) |>
  draw(n = sizes_df) |>
  execute(bfa_eas, seed = 101)

sample |>
  count(region)
```

### Multiple Stratification Variables

You can stratify by multiple variables to create crossed strata. Here we stratify by both region and urban/rural status.

```{r}
sample <- sampling_design() |>
  stratify_by(region, urban_rural, alloc = "proportional") |>
  draw(n = 300) |>
  execute(bfa_eas, seed = 42)

sample |>
  count(region, urban_rural) |>
  head(10)
```

---

## Cluster Sampling

Cluster sampling selects groups (clusters) rather than individual units. This is practical when a complete list of individuals isn't available or when travel costs make dispersed sampling expensive.

Use `cluster_by()` to specify the cluster identifier. All units within selected clusters are included.

```{r}
sample <- sampling_design() |>
  cluster_by(ea_id) |>
  draw(n = 50) |>
  execute(bfa_eas, seed = 120)

sample |>
  summarise(
    n_clusters = n_distinct(ea_id),
    n_units = n()
  )
```

---

## PPS Sampling

Probability Proportional to Size (PPS) sampling selects units with probability proportional to a size measure. This is standard for cluster sampling where clusters vary in size.

Use the `mos` argument to specify the measure of size variable. Here we select EAs with probability proportional to their household count.

```{r}
sample_pps <- sampling_design() |>
  cluster_by(ea_id) |>
  draw(n = 50, method = "pps_brewer", mos = households) |>
  execute(bfa_eas, seed = 365)

sample_pps |>
  summarise(
    mean_hh = mean(households),
    median_hh = median(households)
  )
```

Compared to simple random sampling, PPS tends to select larger clusters:

```{r}
sample_srs <- sampling_design() |>
  cluster_by(ea_id) |>
  draw(n = 50) |>
  execute(bfa_eas, seed = 42)

bind_rows(
  sample_pps |> summarise(method = "PPS", mean_hh = mean(households)),
  sample_srs |> summarise(method = "SRS", mean_hh = mean(households))
)
```

### PPS Methods

`samplyr` supports several PPS methods:

| Method            | Description                                                           |
|-------------------|-----------------------------------------------------------------------|
| `pps_brewer`      | Brewer's method - fast with good properties (recommended)             |
| `pps_systematic`  | PPS systematic - simple but can have periodicity issues               |
| `pps_cps`         | Conditional Poisson sampling (maximum entropy)                        |
| `pps_sampford`    | Sampford fixed-size PPS sampling                                      |
| `pps_poisson`     | PPS Poisson - random sample size (accepts `n` or `frac`; supports `prn`) |
| `pps_sps`         | Sequential Poisson sampling (supports `prn`)                          |
| `pps_pareto`      | Pareto sampling (supports `prn`)                                      |
| `pps_multinomial` | PPS with replacement                                                  |
| `pps_chromy`      | Chromy's sequential PPS / minimum replacement                         |

### Certainty Selection

In PPS sampling, very large units can have expected inclusion probabilities above 1. Certainty selection handles this by selecting those units with probability 1 and sampling the rest normally.

Use `certainty_size` for an absolute threshold or `certainty_prop` for a proportional one.

```{r}
sample_cert <- sampling_design(title = "Certainty selection") |>
  draw(n = 50, method = "pps_brewer", mos = households,
       certainty_size = 800) |>
  execute(bfa_eas, seed = 123)

# How many EAs with more than 800 households
filter(bfa_eas, households > 800)

# Check which units were selected with certainty
count(sample_cert, .certainty_1)
```

With `certainty_prop`, units whose MOS proportion exceeds the threshold are taken with certainty. The process is iterative: after removing certainty units, proportions are recomputed.

```{r}
sample_certp <- sampling_design() |>
  stratify_by(region) |>
  draw(n = 50, method = "pps_systematic", mos = households,
       certainty_prop = 0.05) |>
  execute(bfa_eas, seed = 12345)

# How many EAs with more than 5% of the total hh by region
bfa_eas |>
  mutate(mosprop = households / sum(households),
         .by = region) |>
  filter(mosprop >= 0.05)

filter(sample_certp, .certainty_1)
```

### Permanent Random Numbers (PRN)

Some PPS methods support permanent random numbers for sample coordination across survey waves. Assign a stable U(0,1) value to each frame unit, then pass it via `prn`:

```{r}
set.seed(1)
bfa_eas$prn <- runif(nrow(bfa_eas))

sample_prn <- sampling_design(title = "Samples coordination with PRN") |>
  cluster_by(ea_id) |>
  draw(n = 50, method = "pps_sps", mos = households, prn = prn) |>
  execute(bfa_eas, seed = 1)
sample_prn
```

Because the PRN values are stable, re-executing the same design on the same frame produces highly overlapping samples (positive coordination). PRN is supported for `bernoulli`, `pps_poisson`, `pps_sps`, and `pps_pareto`. You can learn more about coordination with permanent random numbers in the following vignette `vignette("sampling-coordination")`.

---

## Control Sorting and Serpentine Ordering

Control sorting orders the frame before selection, providing implicit stratification. This is effective with systematic and sequential methods, where it ensures the sample spreads evenly across the sorted variables.

```{r}
# Nested sorting: standard ascending order
sample_sort <- sampling_design() |>
  draw(n = 100, method = "systematic",
       control = c(region, province)) |>
  execute(bfa_eas, seed = 98765)
sample_sort
```

`serp()` implements serpentine (snake) sorting, which alternates direction at each level of the hierarchy. This minimizes jumps at boundaries, matching SAS PROC SURVEYSELECT's `SORT=SERP`.

```{r}
# Serpentine sorting
sample_serp <- sampling_design() |>
  draw(n = 100, method = "systematic",
       control = serp(region, province)) |>
  execute(bfa_eas, seed = 1)
sample_serp
```

Control sorting can be combined with explicit stratification. The sorting is then applied within each stratum.

```{r}
sample_serp2 <- sampling_design() |>
  stratify_by(urban_rural) |>
  draw(n = 100, method = "systematic",
       control = serp(region, province)) |>
  execute(bfa_eas, seed = 2)
sample_serp2
```

---

## Balanced Sampling

Balanced sampling selects a sample whose Horvitz-Thompson estimates of auxiliary totals match (or nearly match) the population totals. This improves precision for any variable correlated with the auxiliary variables. `samplyr` implements the cube method (Deville & Tillé 2004) via `method = "cube"`. The older `method = "balanced"` spelling remains a compatibility alias, but designs are stored and printed with the canonical `"cube"` name.

Specify auxiliary variables with `aux`. Without `mos`, inclusion probabilities are equal; with `mos`, they are proportional to size:

```{r}
# Equal-probability balanced sample
balanced_eq <- sampling_design() |>
  draw(n = 100, method = "cube",
       aux = c(population, households)) |>
  execute(bfa_eas, seed = 42)
balanced_eq
```

The balancing property means the HT estimate of auxiliary totals is close to the true total:

```{r}
pop_total <- sum(bfa_eas$population)
ht_total <- sum(balanced_eq$population * balanced_eq$.weight)
c(population = pop_total, ht_estimate = ht_total)
```

Balanced sampling combines naturally with stratification. When stratified, `samplyr` calls the stratified cube algorithm (Chauvet 2009) in a single pass, preserving the allocation computed by `stratify_by()`:

```{r}
# Stratified PPS balanced sample
balanced_strat <- sampling_design() |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 300, method = "cube", mos = households,
       aux = c(population, area_km2)) |>
  execute(bfa_eas, seed = 1960)
balanced_strat
```

When used with `cluster_by()`, auxiliary values are automatically aggregated (summed) to the cluster level before selection.

### Controlled category counts

Place `bound()` around a categorical variable inside `aux` to require its realized sample count to be either the floor or ceiling of the expected count. Bare auxiliary variables and count bounds can be used together:

```{r, eval = FALSE}
controlled <- sampling_design() |>
  draw(
    n = 300,
    method = "cube",
    mos = households,
    aux = c(population, bound(region), bound(urban_rural))
  ) |>
  execute(bfa_eas, seed = 1960)
```

`bound(region)` controls each observed region margin separately. Use separate `bound()` calls to control several margins. If the cube landing step cannot satisfy every requested bound, execution fails rather than silently relaxing a constraint.

### Spatially balanced sampling

LPM2 and SCPS spread the sample through a coordinate space. Select the algorithm explicitly and pass finite numeric coordinates through `spread`:

```{r, eval = FALSE}
spatial <- sampling_design() |>
  draw(
    n = 300,
    method = "lpm2",
    mos = households,
    spread = c(longitude, latitude)
  ) |>
  execute(bfa_eas, seed = 1960)
```

Use `method = "scps"` for spatially correlated Poisson sampling. Spatial methods cannot be combined with cube auxiliaries or `bound()` constraints. Linearized survey export and joint inclusion probabilities are unavailable for bounded cube, LPM2, and SCPS designs. `as_svrepdesign(type = "subbootstrap")` provides a generic PPS bootstrap approximation: it is not a native design-specific variance estimator and does not recreate the bounds or spatial algorithm inside each replicate.

Balanced-family sampling supports up to 2 stages. At stage 1, element-level auxiliary variables are aggregated to PSU totals; at stage 2, the method runs on the element-level frame within each selected PSU.

---

## Custom Sampling Methods

Methods registered with `sondage::register_method()` use a family prefix in `draw()`. Registered `type = "wor"` and `type = "wr"` methods use `pps_<name>` and require `mos`. Registered `type = "balanced"` methods use `balanced_<name>` and may omit `mos` for equal probabilities. A balanced method can declare `supports_aux = TRUE` to receive `aux`, or `supports_spread = TRUE` to require and receive `spread` coordinates.

For example, the elimination procedure of @tille1996 is available in the sampling package. We can register it as a custom method along with its exact joint inclusion probabilities:

```{r custom-method}
# Wrap sampling::UPtille to return selected indices
tille_fn <- function(pik, n = NULL, prn = NULL, ...) {
  which(as.logical(sampling::UPtille(pik)))
}

# Joint inclusion probabilities, restricted to the sampled units
tille_joint_fn <- function(pik, sample_idx = NULL, ...) {
  pi2 <- sampling::UPtillepi2(pik)
  if (!is.null(sample_idx))
    pi2 <- pi2[sample_idx, sample_idx, drop = FALSE]
  pi2
}

sondage::register_method(
  "tille", type = "wor",
  sample_fn = tille_fn,
  joint_fn = tille_joint_fn,
  probabilities = "exact"
)
```

Once registered, the method works like any built-in PPS method:

```{r custom-method-execute}
set.seed(1)
frame <- bfa_eas[sample(seq_len(nrow(bfa_eas)), 500), ]

sample_tille <- sampling_design() |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 50, method = "pps_tille", mos = households) |>
  execute(frame, seed = 1)

sample_tille
```

Because we provided a `joint_fn`, exact variance estimation is available via `joint_expectation()`:

```{r custom-method-joint, eval = requireNamespace("survey", quietly = TRUE)}
jip <- joint_expectation(sample_tille, frame)
svy <- as_svydesign(sample_tille, pps = survey::ppsmat(jip$stage_1))
survey::svymean(~population, svy)
```

```{r custom-method-cleanup, echo = FALSE}
sondage::unregister_method("tille")
```

Custom methods flow through the full pipeline: stratification, certainty selection, multi-stage designs, survey export, and summary diagnostics. See `?sondage::register_method` for the full contract including WR methods and PRN support.

samplyr weights every sample by `1 / pik`, the chance vector it resolves and hands to the registered method. Those weights, and the chances the frame digest records, are only valid when the method honors `pik` as its first-order target. Registered methods state their tier with `probabilities` (sondage >= 0.8.8): `"exact"` when the design's true first-order inclusion probabilities equal `pik` (UPtille above), `"approximate"` when they are honored to a documented approximation (as Pareto sampling is), and `"unknown"` (the default) when `pik` is a selection weight only. `draw()` refuses `"unknown"` methods rather than produce biased weights, so declaring the tier is part of registering a method for samplyr. The classic trap the default protects against: `sample(prob = pik)` yields expected hits exactly equal to `pik` with replacement, but its without-replacement inclusion probabilities differ from `pik`, so a WOR wrapper's tier really is unknown.

---

## Multi-Stage Sampling

Multi-stage designs sample in stages: first select primary sampling units (PSUs), then sample within them. This is the standard approach for large-scale surveys when a complete frame of ultimate units isn't available upfront.

Use `add_stage()` to delimit each stage of the design.

### Two-Stage Design

For this example, we'll use `zwe_eas` to demonstrate a two-stage design: first select districts as PSUs, then sample EAs within selected districts.

```{r}
data(zwe_eas)

# First, let's see the structure: EAs nested within districts
zwe_eas |>
  count(province, district) |>
  head(10)
```

We need to create a measure of size for districts. Here we'll use total households per district.

```{r}
# Add district-level households as MOS
zwe_frame <- zwe_eas |>
  mutate(district_hh = sum(households),
         .by = district)
```

Now we can run a two-stage design: select 10 districts with PPS, then sample 5 EAs within each.

```{r}
twostage_design <- sampling_design(title = "Two-stage cluster sampling") |>
  add_stage(label = "Districts") |>
    cluster_by(district) |>
    draw(n = 10, method = "pps_brewer", mos = district_hh) |>
  add_stage(label = "EAs") |>
  draw(n = 5)
twostage_design
```

Let's apply it to the frame.

```{r}
sample_eas <- execute(twostage_design,
                      zwe_frame, seed = 3)

sample_eas |>
  summarise(n_districts = n_distinct(district),
            n_eas = n())
```

### Stratified Multi-Stage Design

Combining stratification with multi-stage sampling is standard for national surveys. Here we stratify by province (which is constant within districts), then select districts and EAs within each stratum.

```{r}
sample_eas2 <- sampling_design(title = "Stratified two-stage cluster sampling") |>
  add_stage(label = "Districts") |>
    stratify_by(province) |>
    cluster_by(district) |>
    draw(n = 2, method = "pps_brewer", mos = district_hh) |>
  add_stage(label = "EAs") |>
    draw(n = 3) |>
  execute(zwe_frame, seed = 4)

sample_eas2 |>
  count(province, district)
```

---

## Operational Multi-Stage Sampling

In practice, multi-stage surveys often execute stages at different times. After selecting PSUs, field teams may need to conduct listing or other operations before the second stage can proceed.

The `stages` argument lets you execute only specific stages.

```{r}
design <- sampling_design() |>
  add_stage(label = "EA") |>
    stratify_by(urban_rural) |>
    cluster_by(ea_id) |>
    draw(n = 10, method = "pps_brewer", mos = households) |>
  add_stage(label = "HH") |>
    draw(n = 5)

# Execute stage 1 only
selected_eas <- execute(design, zwe_eas, stages = 1, seed = 1)
selected_eas
```

Once the EAs are selected, households are listed in each of them.

```{r}
selected_eas_list <- selected_eas |>
  slice(rep(seq_len(n()), households)) |>
  mutate(hh_id = row_number())
head(selected_eas_list)
```

You can also use `tidyr::uncount()` for the listing step.

```{r}
selected_eas_list <- selected_eas |>
  tidyr::uncount(weights = households, .id = "hh_id", .remove = FALSE)
```

After fieldwork or preparation is complete, execute the remaining stages
by piping the stage-1 result into `execute()` with the listing frame.
This continuation pattern preserves the full design metadata needed for
survey export.

```{r}
final_sample <- selected_eas |>
  execute(selected_eas_list, seed = 2)
final_sample
```

> **Note:** All dplyr verbs (`mutate()`, `filter()`, `*_join()`, etc.) preserve the
> `tbl_sample` class automatically. Operations that change the row set
> (removing, adding, or duplicating rows) or that drop, rename, or
> overwrite internal design columns such as `.weight` or `.fpc_1` keep
> the class but mark the sample as
> modified: it still works as data, but `as_svydesign()` and other
> design-based computations refuse it, because the stored design no
> longer describes the rows. For subpopulation estimates, export first
> and subset the survey design (see the survey analysis vignette).
> Extracting one complete replicate with `filter(.replicate == r)` stays
> fully supported. If you manipulate a sample with base R functions
> like `merge()` or `cbind()`, both the class and its attributes are lost. Prefer the
> dplyr equivalents (`left_join()`, `bind_cols()`). See `?as_tbl_sample` for the full
> list.

### A Self-Weighting Design

Many household surveys want every household in a domain to have the same
inclusion probability `f`. Equal probabilities give equal weights, which
simplifies analysis and keeps weight variation from inflating the
variance. The classic recipe selects EAs with PPS on a household count,
then samples households within each selected EA at a rate that
compensates for the first stage. The product of the two stages is
constant by construction: `pi1 * frac = f`.

```{r}
domain <- zwe_eas |> filter(province == "Bulawayo")

m <- 30                                # EAs to select
b <- 5                                 # target household take per EA
f <- b * m / sum(domain$households)    # overall household rate

pi1 <- sondage::inclusion_prob(domain$households, m)
rates <- tibble(ea_id = domain$ea_id, frac = f / pi1)
```

`sondage::inclusion_prob()` computes the same first-stage probabilities
the EA stage will use, including certainty truncation when a large EA
reaches `pik = 1` (a certainty EA simply gets `frac = f`). The rate table
covers every EA in the domain because the design is written before stage
1 runs; the household stage only consults the EAs that were selected.

Choosing `f = b * m / M` makes the expected take exactly `b` households
in every EA. Set `round = "nearest"` so the realized take stays at `b`
instead of being rounded up:

```{r}
design_sw <- sampling_design("Self-weighting household survey") |>
  add_stage(label = "EA") |>
    cluster_by(ea_id) |>
    draw(n = m, method = "pps_systematic", mos = households) |>
  add_stage(label = "HH") |>
    stratify_by(ea_id) |>
    draw(frac = rates, round = "nearest")
```

`stratify_by(ea_id)` at the household stage makes each selected EA its
own stratum, which is what lets the sampling rate vary by EA. Execution
follows the same listing pattern as above:

```{r}
eas_sw <- execute(design_sw, domain, stages = 1, seed = 101)
listing_sw <- eas_sw |>
  tidyr::uncount(households, .id = "hh_no", .remove = FALSE)

hh_sw <- eas_sw |> execute(listing_sw, seed = 102)

range(hh_sw$.weight)
1 / f
```

Every household carries the same weight `1/f`.

Why a compensating rate instead of a fixed take of `b` households
(`draw(n = b)`)? With a fixed take the design is self-weighting only when
the listing finds exactly the household count used as the measure of
size. Fresh listings differ from census counts. The compensating rate
keeps the weights centered on `1/f` whatever the listing finds; only the
rounding of small takes moves them. A fixed take passes listing changes
straight into the weights:

```{r}
set.seed(7)
fresh_counts <- eas_sw |>
  mutate(listed = pmax(1L, as.integer(round(households * runif(n(), 0.85, 1.35)))))
listing_fresh <- fresh_counts |>
  tidyr::uncount(listed, .id = "hh_no", .remove = FALSE)

hh_fresh <- eas_sw |> execute(listing_fresh, seed = 103)

range(hh_fresh$.weight) * f    # stays near 1
```

Feasibility requires `f <= pi1` in every EA, so that `frac <= 1`. Very
small EAs can violate this. The standard fixes are to merge small EAs
with a neighbor before sampling, or to cap their rate with
`frac = pmin(1, f / pi1)` and accept a weight above `1/f` in those EAs.

---

## Replicated Sampling

The `reps` parameter in `execute()` draws multiple independent samples from the same frame under the same design. The output is a single stacked `tbl_sample` with a `.replicate` column identifying each replicate. This is useful for simulation studies, repeated-sampling variance estimation, or quality control.

```{r}
rep_sample <- sampling_design() |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 300) |>
  execute(bfa_eas, seed = 42, reps = 5)

rep_sample |>
  count(.replicate)
```

Each replicate is a fully independent draw. Replicate `r` uses seed `seed + r - 1`, so the first replicate matches a standalone execution with the same seed:

```{r}
standalone <- sampling_design() |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 300) |>
  execute(bfa_eas, seed = 42)

rep1 <- rep_sample |> filter(.replicate == 1)
identical(sort(rep1$ea_id), sort(standalone$ea_id))
```

Replicated sampling also works with multi-stage designs. When continuing from a replicated partial execution, each replicate is continued independently:

```{r}
design <- sampling_design() |>
  add_stage(label = "Districts") |>
    cluster_by(district) |>
    draw(n = 10, method = "pps_brewer", mos = district_hh) |>
  add_stage(label = "EAs") |>
    draw(n = 5)

stage1 <- execute(design, zwe_frame, stages = 1, seed = 1, reps = 3)
final <- execute(stage1, zwe_frame, seed = 100)

final |>
  summarise(n_districts = n_distinct(district), n_eas = n(),
            .by = .replicate)
```

Survey export functions require a single replicate. Filter first:

```{r, eval = FALSE}
one_rep <- rep_sample |> filter(.replicate == 1)
svy <- as_svydesign(one_rep)
```

Note: `reps` cannot be combined with `panels` or with stages that use permanent random numbers (PRN). PRN produces identical samples across replicates, so coordination across waves should use an explicit loop with different PRN vectors instead.

---

## Working with Results

### Accessing the Design

The sampling design is attached to the result and can be retrieved with `get_design()`.

```{r}
sample <- sampling_design() |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 200) |>
  execute(bfa_eas, seed = 2024)

design <- get_design(sample)
design
```

and accessing element of the design

```{r}
design$stages[[1]]$strata$vars
```

### Weight Diagnostics

The `summary()` method reports weight diagnostics including Kish's effective sample size and design effect due to unequal weighting. For the full design
effect (including clustering and stratification), use `survey::svymean()` with `deff = TRUE` after converting with `as_svydesign()`.

```{r}
summary(sample)
```

### The Frame Digest

Every execution records a summary frame digest by default: a compact record of
the selection pools and chances the design resolved, with no original unit
identifiers. The `n/N` line in the printed header and the per-stage realization
lines in `summary()` come from it, and `frame_summary()` returns it as tibbles.
The original frame is not needed:

```{r}
ea_sample <- sampling_design(title = "EA sample") |>
  stratify_by(region) |>
  cluster_by(ea_id) |>
  draw(n = 25, method = "pps_brewer", mos = households) |>
  execute(bfa_eas, seed = 42)

frame_summary(ea_sample)
```

The digest mode is an explicit performance and downstream-capability choice:

- `frame_digest = "summary"` is the default and keeps compact pool and chance
  summaries.
- `frame_digest = "none"` skips digest and selection-trace construction for
  the lowest execution overhead.
- `frame_digest = "full"` retains exact per-unit chances for
  unequal-probability element stages. Use it when `joint_expectation()` must
  work later without the original frame.

The companion package samplens, in development, draws the digest as a visual sampling card.

---

## Method Reference

### Selection Methods

| Method            | Sample Size | Replacement                                                    | Notes                                        |
|-------------------|-------------|----------------------------------------------------------------|----------------------------------------------|
| `srswor`          | Fixed       | No                                                             | Default, general purpose                     |
| `srswr`           | Fixed       | Yes                                                            | Bootstrap, rare populations                  |
| `systematic`      | Fixed       | No                                                             | Ordered frames                               |
| `bernoulli`       | Random      | No                                                             | Simple field protocols                       |
| `pps_brewer`      | Fixed       | No                                                             | Recommended PPS method                       |
| `pps_systematic`  | Fixed       | No                                                             | Simple PPS                                   |
| `pps_cps`         | Fixed       | No                                                             | Maximum entropy, exact joint inclusion probs |
| `pps_sampford`    | Fixed       | No                                                             | Sampford PPS, exact joint inclusion probs    |
| `pps_poisson`     | Random      | No                                                             | PPS with random size                         |
| `pps_sps`         | Fixed       | No                                                             | Sequential Poisson, supports PRN             |
| `pps_pareto`      | Fixed       | No                                                             | Pareto sampling, supports PRN                |
| `pps_multinomial` | Fixed       | Yes                                                            | PPS with replacement                         |
| `pps_chromy`      | Fixed       | PMR.^[Probability Minimum Replacement based on expected hits.] | Chromy's sequential PPS                      |
| `cube`            | Fixed       | No                                                             | Cube method, balances on `aux` variables     |
| `lpm2`            | Fixed       | No                                                             | Local pivotal, requires `spread`              |
| `scps`            | Fixed       | No                                                             | Spatially correlated Poisson, requires `spread` |

### Allocation Methods

| Method         | Description                           | Requirements                             |
|----------------|---------------------------------------|------------------------------------------|
| (none)         | n per stratum                         |                                          |
| `equal`        | Same n per stratum                    |                                          |
| `proportional` | n ∝ stratum size                      |                                          |
| `neyman`       | Minimize variance                     | `variance` data frame                    |
| `optimal`      | Minimize variance/cost                | `variance` + `cost` data frames          |
| `power`        | Compromise: \(n_h \propto C_h X_h^q\) | `cv` + `importance` (+ optional `power`) |

For custom stratum-specific sizes or rates, pass a data frame directly to `n` or `frac` in `draw()`.
When `alloc` is specified, use `n` (not `frac`).

### Sample Size Bounds

| Argument | Description                                                        |
|----------|--------------------------------------------------------------------|
| `min_n`  | Minimum sample size per stratum (for variance estimation, set ≥ 2) |
| `max_n`  | Maximum sample size per stratum (to cap dominant strata)           |

These only apply when using allocation methods (`equal`, `proportional`, `neyman`, `optimal`, `power`).

### Rounding Methods

| Value       | Description                              |
|-------------|------------------------------------------|
| `"up"`      | Round up (ceiling), default, matches SAS |
| `"down"`    | Round down (floor)                       |
| `"nearest"` | Standard mathematical rounding           |

These apply when using `frac` to specify sampling rates. When computed sample size < 1, it is always rounded up to 1.

---

## Best Practices

1. **Always set a seed** for reproducibility
2. **Use meaningful stage labels** for documentation
3. **Validate designs** with `validate_frame()` before execution
4. **Check weight distributions** with `summary()` on the sample

```{r}
design <- sampling_design(title = "National Health Survey 2024") |>
  add_stage(label = "Enumeration Areas") |>
    stratify_by(region, urban_rural) |>
    cluster_by(ea_id) |>
    draw(n = 5, method = "pps_brewer", mos = households) |>
  add_stage(label = "Households") |>
    draw(n = 20)

print(design)
```

### Frame Validation

Use `validate_frame()` to check that a frame has all the variables a design requires before executing.

```{r, error = TRUE}
# This passes
validate_frame(design, bfa_eas)

# This fails with a clear message
bad_frame <- data.frame(id = 1:100, value = rnorm(100))
validate_frame(design, bad_frame)
```

## Reference
