---
title: "Rotating Panels and Longitudinal Designs"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Rotating Panels and Longitudinal Designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, echo = FALSE}
library(samplyr)
library(sondage)
library(svyplan)
library(dplyr)
```

## What this vignette covers

A longitudinal survey visits the same units more than once. @smith2009 identify
five broad types of longitudinal design, and four of them need nothing beyond
the verbs in `vignette("introduction")`. The fifth, the rotating panel, is what
`execute(panels = )` and `execute(wave = )` are for.

This vignette maps the five designs onto `samplyr`, shows what the panel
machinery builds and what it does not, and states the limits the code cannot
show you. Sample coordination with permanent random numbers is a separate
mechanism that answers a separate question; see
`vignette("sampling-coordination")` for it. The two compose, and one note below
says how.

## Setup

The examples use `ken_enterprises`, a synthetic business register, with a
permanent random number attached so the master can be drawn by `pps_sps`.

```{r}
library(samplyr)
library(sondage)
library(svyplan)
library(dplyr)

data(ken_enterprises)

n_wave <- 200

set.seed(1)
frame <- ken_enterprises |>
  mutate(u = runif(n()))
```

## Vocabulary

Longitudinal work overloads words that already mean something specific in a
sampling design. In `samplyr`:

| Term            | Meaning                                                |
|:----------------|:-------------------------------------------------------|
| stage           | selection within units selected at an earlier stage     |
| phase           | probability subsampling from an earlier sample          |
| frame           | a source covering a target population                   |
| wave            | a programme or calendar occasion                        |
| panel           | a randomized rotation, reserve, or workload group       |
| cohort          | a sample's entry batch or refreshment vintage           |
| unit identity   | the substantive entity followed across occasions        |

Four consequences follow. A wave is not a stage: `add_stage()` selects within
what an earlier stage selected, whereas a wave revisits units already
selected. A panel group is not a refreshment cohort: `panels` partitions one
draw, a cohort is a separate draw. Repeated observation of every eligible unit
adds no selection probability, so interviewing a fixed panel five times does
not change `.weight`. And response, attrition and tracing failure are not
selection operations; they belong to the estimation layer.

The printed token is `.panel`. There is deliberately no `.wave` column:
`samplyr` realizes designs against frames and does not model time.

## Defining the longitudinal population

Before choosing a design, decide which population the estimates are about
[@smith2009, sec. 2.3.1]. There are three answers and they lead to different
designs:

- **Static**: the population as of wave 1. Deaths occur, births cannot enter.
  Population information is fixed for all waves, so weighting is
  straightforward, and the natural design is a fixed panel. The risk is that
  readers take the results for the third option.
- **Intersection**: units present at every wave. Both births and deaths are
  excluded, auxiliary totals are hard to construct, and the definition changes
  whenever a wave is added.
- **Union**: units present at any wave. This suits a genuinely dynamic
  population, requires a design that introduces births, and is what most
  household panels use. Adding a wave changes the population and may require
  revising earlier estimates.

This is a declaration, not a computation, so it is not an argument to any
verb. It is what determines whether you need births at all, and therefore
which design below applies.

## The five designs

| Design                  | What it is                                                                          | In `samplyr`                                                                                                                                                                                                     |
|:------------------------|:------------------------------------------------------------------------------------|:-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| Fixed panel             | The same units on every occasion; no additions after wave 1                          | One `execute()`. Waves 2 to *T* are repeated measurement on the same `tbl_sample`; the base selection weight does not change because another interview happened.                                                    |
| Fixed panel plus births | A fixed panel with a sample of recent entrants added at each occasion                | The original `execute()`, plus a separate `execute()` against an entrant register for each refreshment cohort. The components stay separate objects until their target population and estimator are defined.        |
| Repeated panel          | A series of panels, each representing an equivalent population defined at a later time | Reuse the same frame-independent `sampling_design` in a fresh `execute(design, frame_j, seed = seed_j)`. Not `replay_design()`, which reproduces one recorded execution and refuses a changed frame.                |
| Rotating panel          | A fixed proportion of units replaced at each occasion                                | `execute(panels = k)` for the start-up partition; the replenishment is ordinary repeated execution. See below, because this is the one that is easy to overclaim.                                                    |
| Split panel             | A panel plus a fresh cross-sectional sample at each occasion [@kish1987, pp. 181--183] | Two independent executions kept as separate objects. In the *P* = 1 variation a subset of units is included on every occasion, so longitudinal estimates are possible for that stratum; those units already carry `.certainty_k`.                                             |

A rotating panel is a repeated panel whose overlap pattern is fixed and
usually balanced, which is what allows the live groups to be combined for a
cross-sectional estimate [@kaltoncitro1993]. Labour force surveys in many
countries use one [@steel1997].

## What `panels` builds, and what it does not

A real rotating panel replenishes itself: at each entry time it draws a fresh
group from the population as it stands then, so the live groups carry
different population vintages [@lynn2012, sec. 3]. That is exactly what makes
them combinable.

`execute(panels = k)` partitions **one** execution against **one** frame.
Every group comes from the same frame vintage, and no entrant can appear in
any of them. Over a long horizon that cannot represent the cross-sectional
population at later waves.

Where the partition is exactly right is start-up. The "immediate start"
option draws the full sample in year 1 and subdivides it into groups
designated to remain for one, two, three, ... years, with a fresh group added
each year thereafter; @kaltoncitro1993 prefer it to a gradual start because it
delivers the full sample size from year 1 rather than filling up over the
rotation cycle. So the coherent shape is:

```text
year 1     execute(design, frame, panels = k)     start-up partition
year 2+    execute(design, frame_vintage_t)       one fresh group per period
```

`panels` builds the start-up partition and the workload rotation over a frozen
master. Steady-state replenishment is ordinary repeated execution against
successive frame vintages, and `rotation_program()` is what links the resulting
executions into one programme; see below.

`design_overlap()` describes such a programme from its schedule alone,
before any frame exists, and draws it. Here each group is live for two
consecutive occasions, which is the pattern used throughout this vignette.

```{r rotation-chart, fig.width = 7, fig.height = 3.6, fig.alt = "Rotation chart for a design keeping each group in sample for two consecutive periods, started immediately. Seven samples enter on successive periods, two are live in every period, and the steady state is reached at period 1."}
plot(design_overlap("2"), type = "schedule", start = "immediate")
```

The chart is the whole programme, not one execution. Two groups are live in
every period and the steady state is reached immediately, which is the property
@kaltoncitro1993 prefer it for. `execute(panels = k)` builds the first column of
that picture and nothing to its right; every later column is a fresh execution
against a later vintage, which is what `rotation_program()` registers.

Combining the live groups is then not a matter of pooling weights. A unit that
entered the population between periods 1 and 2 could not have been drawn into
group 1, so the combined sample under-represents joiners unless their
probabilities are corrected. For an initial sample and a later one
[@lynn2012, sec. 8.1]:

```text
stayers   P ~= n1/N1 + n2/N2      small-fraction approximation
leavers   P = n1/N1
joiners   P = n2/N2
w = 1/P
```

This is still an inverse inclusion probability, so it stays inside the
single-frame estimator. What it needs is data `samplyr` cannot invent: for
each unit, whether it was eligible for each earlier draw. That comes from
survey questions establishing population membership at the earlier date, and
it is easy to forget to ask them. Supply it as a column on the frame; two
execution receipts cannot reveal it.

One practical note: `panels` composes with PRN coordination. The conflict
check in `execute()` is against `reps`, not `prn`, so a PRN-rotated draw can
also be partitioned into workload groups. See
`vignette("sampling-coordination")` for the coordination side.

## Scheduling the groups

Declaring *when* each group is live turns the partition into a rotation you
can materialize. Pass `panels` a data frame with `panel` and `wave` columns
and an optional logical `active` column instead of a count. Here four groups
rotate two at a time, so each group is live for two consecutive occasions.

```{r}
rotation <- data.frame(
  panel  = rep(1:4, times = 4),
  wave   = rep(1:4, each = 4),
  active = c(TRUE,  TRUE,  FALSE, FALSE,
             FALSE, TRUE,  TRUE,  FALSE,
             FALSE, FALSE, TRUE,  TRUE,
             TRUE,  FALSE, FALSE, TRUE)
)

master <- sampling_design() |>
  draw(n = n_wave, method = "pps_sps", mos = revenue_millions, prn = u) |>
  execute(frame, seed = 7, panels = rotation)

table(master$.panel)
```

The schedule is read at the draw, not only later: the fewest panels any wave
activates sets the assignment block size, so a schedule that keeps two of
four groups live blocks more finely than one that keeps a single group live.

`execute(master, wave = t)` then realizes one occasion.

```{r}
wave_2 <- execute(master, wave = 2)

c(master = nrow(master), wave_2 = nrow(wave_2))
table(rotating = !wave_2$.certainty_1)
```

Two of four groups is a little under half the master, plus every certainty
unit: `pps_sps` on a skewed register resolves the largest enterprises to
$\pi_k = 1$, and a permanent unit is in the sample at every occasion.

Activation is a probability subsample, not a filter, so the wave carries
weights that estimate the same population total the master does.

```{r}
c(master = sum(master$.weight), wave_2 = sum(wave_2$.weight))
```

They are close rather than equal, which is the point: the activation is a
second sampling phase and its estimator is unbiased, not exact.

The factor applied to each weight is the inverse of the unit's activation
probability, which is its block's frozen quota for the active panels over the
block size. That is exact, and it is generally not the nominal `k / r`
implied by counting groups: a block that took the remainder carries a
different quota, and its units carry a different factor. Certainty units are
activated with probability one and their weights are untouched.

A wave is a two-phase sample and exports as one. It retains the master as its
first phase, and `as_svydesign()` passes the activation to `twophase()` as the
second: the blocks become the phase-2 strata and their frozen sizes the
phase-2 population counts, which is what makes the subsampling variance exact
rather than assumed. `method` is the only choice the export takes, `"full"` by
default with `"simple"` and `"approx"` available where their assumptions hold,
and none of them is ever selected as a silent fallback for another.

```{r}
equal_master <- sampling_design() |>
  draw(n = n_wave) |>
  execute(frame, seed = 7, panels = rotation)

equal_wave <- execute(equal_master, wave = 2)

class(as_svydesign(equal_wave))
```

The master drawn above is a different case. `twophase()` accepts no `pps`
specification at its first phase, so a master with unequal inclusion
probabilities has no exact linearization there, and `as_svydesign(wave_2)`
refuses with that reason rather than returning the with-replacement
approximation unannounced. The activation weights stay exact and estimate
totals correctly; it is the variance that has no exact form. The master's own
export is unaffected, and so is any wave of an equal-probability master,
clustered or multistage.

## Which unit rotates

By default the assignment belongs to the first stage, so a clustered design
rotates whole primary units and every unit below inherits its ancestor's
group. `panel_stage` moves it to another stage. That is what expresses the
address-panel design [@smith2009, sec. 2.3.4]: the selected areas stay in the
survey while the addresses inside them rotate.

```{r}
addresses <- data.frame(
  ea      = rep(sprintf("EA%02d", 1:40), each = 12),
  address = rep(1:12, times = 40)
)

area_panel <- sampling_design() |>
  add_stage() |> cluster_by(ea) |> draw(n = 12) |>
  add_stage() |> cluster_by(address) |> draw(n = 8) |>
  execute(addresses, seed = 11, panels = rotation, panel_stage = 2)

# Panels per selected area: every area carries all four.
table(tapply(area_panel$.panel, area_panel$ea, function(x) length(unique(x))))
```

Every area is in every wave, and half its addresses turn over between
consecutive occasions.

```{r}
area_wave <- execute(area_panel, wave = 2)

c(areas_in_master = length(unique(area_panel$ea)),
  areas_in_wave   = length(unique(area_wave$ea)),
  addresses_in_wave = nrow(area_wave))
```

Two things change with the assignment stage, and both matter operationally.

**Pools shrink, so `small_pool` binds far more often.** A pool becomes the
assignment stage's strata *inside* each realized parent, so pools are as small
as the within-parent take rather than as large as a stratum. A pool holding
`panels - r_min` assignment units **or fewer** is refused, rather than silently
leaving areas out of some wave. Here `r_min` is the fewest panels any declared
wave activates, the same quantity that sets the block size above: with four
panels and two live per wave, the refusal is at two addresses per area or fewer.

**Selection certainty counts only at the assignment stage, in both
directions.** A certainty area does not make its addresses permanent, and a
certainty selection *below* the address does not keep that address in every
wave. A person selected with probability one inside a rotating address leaves
the survey with the address, which is a property of the design rather than a
defect to work around.

This is not a better default, and it should not be adopted as one. Holding the
parent fixed while its members rotate can bias cross-sectional estimates over
time [@smith2009, sec. 2.3.4]. A sampling unit that cannot move between
parents is balanced for cross-sectional estimates and for net change, but not
for gross change. And the frame has an operational consequence that is easy to
get wrong: a vacant address found at one wave stays in the sample for later
waves, so the frame must not be filtered to occupied units between waves.

## How much two occasions share

How far two waves overlap is fixed at the master's draw, so it can be asked
before either has been materialized.

```{r}
overlap_12 <- joint_expectation(master, waves = c(1, 2))
overlap_12
```

One row per block, carrying counts and probabilities side by side. `units` is
the block size; `take_1`, `take_2` and `take_both` are the frozen quotas, the
number of the block's units live at the first wave, at the second, and at both.
`prob_1` and `prob_2` are those quotas as chances for a single unit,
`joint_same` the chance it is live at both, and `joint_distinct` the chance that
two different units of the block are live at one wave each. `pool`, `stratum`,
`class` and `activation` identify where the block came from and whether it
rotates. Take the same wave twice, `waves = c(2, 2)`, for the joint expectation
within a single wave.

These follow from the frozen quotas rather than from counting groups. In a
block of `m` assignment units where the panels live at the two waves take
$a_t$ and $a_s$ units and those live at both take $a_\cap$:

$$P(i \in W_t,\, j \in W_s) = \frac{a_t a_s - a_\cap}{m(m-1)}, \qquad i \neq j$$

with $a_\cap / m$ for a single unit at both waves. Units in different blocks
are independent. A block that took the remainder carries different quotas and
therefore different values, which is why these are read from the record.

Aggregating `take_both` against `take_1` gives the overlap proportion the
rotation literature quotes. `design_overlap()` computes the same proportion
from the schedule alone, so the two can be compared.

### What the pattern implies

The literature figures are properties of the pattern, not of any master. The US
CPS rotates each group four months in, eight out and four in again.

```{r}
cps <- design_overlap("4-8-4")
cps[c(1, 12)]

design_overlap("15")[c(1, 12)]
```

Three quarters of the sample is retained one month later and half of it one
year later, against 93% and 20% for an in-for-15 rotation. The lag-12 return is
the point of the design, and the profile shows it better than the two numbers
do.

```{r overlap-profile, fig.width = 7, fig.height = 3.6, fig.alt = "Overlap by lag for the 4-8-4 rotation. Overlap falls from 0.75 at lag 1 to zero at lags 4 through 8, then returns to a second peak of 0.50 at lag 12 before decaying again."}
plot(cps, type = "overlap")
```

The notation carries the distinction that makes those figures reproducible.
`"4-8-4"` names spell lengths, in then out then in; `"1-1-0-0-1-1"` names one
flag per occasion. A string of all ones can be read either way, and it is
refused rather than guessed.

```{r, error = TRUE}
design_overlap("1-1-1-1")
```

That is why a 4-8-4 has to be declared with a finite life rather than cycled: a
group that returns to its first spell keeps seven of eight groups month to
month instead of six.

### What this master delivers

The design in this vignette keeps each group live for two consecutive
occasions, so its schedule is `"2"`. Its nominal overlap and what the master
actually delivers are not the same number.

```{r}
nominal  <- design_overlap("2")[[1]]
realized <- sum(overlap_12$take_both) / sum(overlap_12$take_1)

c(nominal = nominal, realized = round(realized, 4))
```

The gap is the permanent block. Most rotating blocks hold four assignment units
and put two in each wave and one in both, and the few remainder blocks carry
their own quotas, but the rotating blocks together still come to exactly one
half. The block holding the certainty units is live in full at every wave and
contributes one, which is what lifts the total.

```{r}
overlap_12 |>
  summarise(blocks = n(), units = sum(units),
            take_1 = sum(take_1), take_both = sum(take_both),
            .by = activation)
```

A schedule cannot know this, because certainty is a property of the draw rather
than of the pattern. `design_overlap()` is the planning quantity, available
before there is a frame; `joint_expectation()` is what this realized master
records. Quote the first when sizing a design and the second when describing
one that exists.

**Say which overlap you mean.** Three qualifiers are needed before the number
is interpretable. The figures below are illustrative rather than computed from
any design shown here: they stand for a single asymmetric master under which
0.75, 0.53 and 0.33 are all correct answers, differing only in which question
is being asked.

- **reference set**: the units the design selected, or the units that
  responded. `samplyr` knows the first exactly. The second needs response
  records, which it does not have.
- **direction**: which wave is the denominator. Where the waves differ in
  size, an asymmetric schedule gives 0.50 forward and 0.33 reversed.
- **measure**: units counted, or weighted. A `pps_sps` master with permanent
  certainty units reports 0.75 by count and 0.53 by weight, because those
  units are live at every wave while carrying weight one.

Written out, the count-based directional quantity is

$$g_{\text{design}}(t \to s) = \frac{\#(\text{selected at both } t \text{ and } s)}{\#(\text{selected at } t)} = \frac{\sum_b a_{\cap b}}{\sum_b a_{tb}}$$

This is the $g$ that appears in the rotation literature [@bergerpriam2016].
It is **not** the $k$ of the resampling shortcut in @cantydavison1999, which
is the responding overlap: they warn explicitly that the realized proportion
should be used there rather than the theoretical one, and the difference is
nonresponse.

Under `samplyr`'s frozen-activation contract the scheduled and the selected
overlap coincide exactly. A rotating unit is live at both waves precisely when
its panel lies in both active sets, and a permanent certainty unit is live at
every wave whatever its panel, so in both cases the realized count is
determined rather than random. That is a property of the contract rather than
of rotation in general: an operation that drew again at wave time, or
subsampled for attrition, would separate them.

These are conditional on the units the master selected and on the frozen
quotas. They are not the unconditional joint inclusion probabilities of the
whole two-phase design, which also carry the master's own pairwise term.

One limit remains. A wave cannot be replayed, because it is derived from a
recorded execution rather than being one; save and replay the master instead.

Finally, the schedule states which groups are live when. It does not
replenish the sample: every group still comes from the frame vintage the
master was drawn from, which is the scoping point above. Replenishment is the
next section.

## Replenishing across vintages

A rotating panel that replenishes is not one execution. It is a start-up master
against the first frame vintage, then one refreshment cohort against each later
one. `rotation_program()` is the registry that links those executions and says
which components are live at each occasion.

```{r}
startup <- sampling_design() |>
  draw(n = n_wave) |>
  execute(frame, seed = 7, panels = rotation)

# A refreshment cohort drawn from an entrant register: units that joined the
# population after the master's vintage, so no unit can be on both frames.
set.seed(99)
entrants <- ken_enterprises |>
  slice_sample(n = 900) |>
  mutate(enterprise_id = paste0("NEW_", sprintf("%04d", seq_len(900))))

intake_2 <- sampling_design() |>
  draw(n = 60) |>
  execute(entrants, seed = 8)

program <- rotation_program(
  cohorts = list(startup = startup, intake_2 = intake_2),
  entry_wave = c(startup = 1, intake_2 = 2),
  schedule = rbind(
    transform(rotation, cohort = "startup"),
    data.frame(cohort = "intake_2", panel = 1, wave = 1:4,
               active = c(FALSE, TRUE, TRUE, TRUE))
  )
)
program
```

`entry_wave` is declared rather than inferred from the schedule. A cohort can
be drawn early and held in reserve, so its first active wave is not necessarily
its entry, and entry is what anchors which population vintage the cohort
represents.

`execute(program, wave = t)` materializes one occasion across every live cohort.

```{r}
wave_2_program <- execute(program, wave = 2)
wave_2_program
```

The result is a collection with one member per live cohort, each keeping its own
receipt and its own exact activation factor. **It is not row-bound and its
weights are not combined**, and that is a statement about what the registry
knows rather than caution. Whether a unit could have been drawn into more than
one cohort depends on facts no receipt carries: whether it was in the population
at each vintage, what chance it would have had in a draw it was not selected
into, and whether the draws were independent. Two executions with identical
inclusion probabilities can be independent or identical, and nothing stored
tells them apart.

The case where the analyst does know is the one above. An **entrant register**
holds only units that joined since the previous vintage, so the cohorts are
disjoint by construction, every unit belongs to exactly one, and the component
weights are already right for the combined live set. A refresher drawn from the
*whole* population at a later vintage is the other case: a unit present at both
times had two chances of selection, and combining then needs a union
probability, which is a multiple-frame problem rather than a scheduling one and
is out of scope for `samplyr`.

A program is a registry, so `write_design()` does not accept one. Archive each
cohort's executed sample and the schedule that links them, and rebuild the
program on read; see `vignette("serialization")`.

## Estimating change across waves

Two waves exported separately give two variances and no covariance, and the
covariance is most of the answer. `stack_waves()` performs the structural half
of the handoff: it verifies that the waves come from one executed master and
returns a plain long table for an inference layer.

```{r}
tab <- stack_waves(
  execute(equal_master, wave = 1),
  execute(equal_master, wave = 2)
)

dim(tab)
names(tab)[1:4]
```

Two of those names carry the contract. `design_weight` is the exact design
weight including the activation factor, and is **not** a final weight: nothing
here is adjusted for nonresponse or calibrated, so a routine asking for
`w_final` is asking for something the analysis must supply. `master_id`
matches units across waves of one master and means nothing outside it; in
particular it is not a primary sampling unit, and a clustered design's own
cluster variable is carried through for that.

Verification is the reason to prefer this to `rbind()`. Two executions of one
design produce the same `.sample_id` values, so waves of *different* masters
stack silently even when the two frames share no unit at all, and the result
matches units that were never the same unit.

### Choosing a variance-of-change route

Write $V_\Delta = V_t + V_s - 2\rho\sqrt{V_t V_s}$. The two marginal variances
come from `samplyr`'s per-wave export; the correlation is the part that has to
be estimated, and it is where the routes differ.

| Route | Needs | Behaviour |
|---|---|---|
| Berger--Priam form B [@bergerpriam2016] | Union sample, wave indicators, design strata when stratified, design weights | The recommended route. Calibrated from 7% to 93% overlap in the fixtures measured here |
| Berger--Priam form A | The same | Anti-conservative at high overlap |
| `vardpoor::vardchanges()` | A calibrated `w_final`, stable PSU | Reproduces form A exactly, and inherits its behaviour |
| Independence | Nothing | Conservative while the omitted covariance is positive, which under a rotation it is; badly inefficient at high overlap |
| Matched-sample correlation [@kish1965] | The common sample | Not recommended at any overlap |

The last row is the one worth naming, because it looks right and is not. Taking
the correlation from the matched sample and combining it with the full-sample
variances omits the overlap factor $g$: it measures how correlated a *unit* is
with itself over time, where what is wanted is how correlated the two
*estimators* are.

Form B is the recommended route, and it is one call to `lm()` once the union
sample is assembled. Use the unstratified form below only for an unstratified
master; a stratified one needs the stratified specification of @bergerpriam2016
[§4.1], wave-specific stratum indicators and their interactions in place of the
bare wave indicators, and it additionally requires the common-sample fraction to
be the same in every stratum, since a single $g$ is applied.

```{r}
w1 <- execute(equal_master, wave = 1)
w2 <- execute(equal_master, wave = 2)
w1$measure <- w1$revenue_millions
w2$measure <- w2$revenue_millions * 1.03

tab <- stack_waves(w1, w2)

# Marginal variances: samplyr's per-wave export.
v1 <- survey::SE(survey::svytotal(~measure, as_svydesign(w1)))^2
v2 <- survey::SE(survey::svytotal(~measure, as_svydesign(w2)))^2

# The union, with each wave's weighted measure zero-filled outside it.
ids <- union(tab$master_id[tab$wave == 1], tab$master_id[tab$wave == 2])
at <- function(w) match(ids, tab$master_id[tab$wave == w])
y <- function(w) {
  i <- at(w)
  ifelse(is.na(i), 0,
         (tab$measure * tab$design_weight)[tab$wave == w][i])
}
common <- !is.na(at(1)) & !is.na(at(2))
g <- sum(common) / sum(!is.na(at(1)))

# Form B: the common-sample residual correlation, times g. The model is the
# paper's own -- wave indicators, their interaction, and no intercept.
z1 <- as.integer(!is.na(at(1)))
z2 <- as.integer(!is.na(at(2)))
fit <- stats::lm(cbind(ifelse(common, y(1), 0), ifelse(common, y(2), 0)) ~
                   -1 + z1 * z2)
V <- stats::estVar(fit)
rho <- V[1, 2] / sqrt(V[1, 1] * V[2, 2]) * g

c(overlap = g, correlation = rho,
  se_change = sqrt(v1 + v2 - 2 * rho * sqrt(v1 * v2)),
  se_if_independent = sqrt(v1 + v2))
```

The two standard errors differ by the covariance, which is the quantity two
separate exports cannot supply.

Compute the correlation rather than predicting it from the overlap. The measure
here is perfectly correlated across waves and the estimator correlation comes
out at the overlap itself, but that is a coincidence of this example: the two
waves also share the master's own selection variation, so the correlation can
sit above the overlap times the within-unit correlation.

### Where these routes stop being calibrated

All of them are approximate, and their consistency arguments assume a negligible
sampling fraction and a high-entropy master. Four limits are worth knowing
before the numbers are used, each measured on one population and one rotation
and therefore an empirical landmark rather than a threshold:

- **The wave fraction, not the master fraction, is what counts.** A rotation
  keeping half its panels live puts half the master in each wave. Form B was
  calibrated through a 5% wave fraction, showed measurable degradation at 10%,
  and was clearly anti-conservative at 25%. The mechanism is that the
  correlation construction carries no finite population correction while the
  correlation it estimates does.
- **A systematic master is outside the validated guarantee.** Its marginal
  variances are exported through the equal-probability path and are approximate
  for that design, so a reported change variance combines two errors that can
  point in opposite directions and cancel. A plausible-looking number is not
  evidence that either component is right; see `?as_svydesign`.
- **Sensitivity runs the wrong way.** A correlation slightly too high understates
  the variance by a multiplier of $\rho/(\rho - 1)$ at equal marginal variances,
  which is about $-13$ at $\rho = 0.93$. The multiplier is a property of the
  correlation and the margins, not of the rotation pattern, so compute it for
  the design in hand.
- **Nothing here is validated for multiple cohorts, nonresponse, calibration, or
  unequal-probability masters.**

The measurements behind these landmarks, with their fixtures, are recorded in
`dev/t1-2b-variance-of-change-plan.md` in the package sources.

## Attaching later waves to the sample

A `tbl_sample` records the selection. Joining later-wave responses one to one,
in wide form, preserves that: one row per selected unit, design columns
untouched. Reshaping to one row per unit-wave does not. Duplicated unit-wave
rows are repeated measurements of one selected unit, not independently sampled
observations, and exporting them with `as_svydesign()` would count each unit
once per interview. Export the wide form, or reshape after export.

## Limits to state before fieldwork

**Transient stratifiers.** `sampling_design()` stores stratification variables
as names and resolves them at execution, so re-executing a design against a
later frame vintage re-classifies units. For a repeated panel that is correct,
since each panel represents the population as of its own time. For a fixed
panel it is wrong: a unit must keep its wave-1 stratum. Unequal sampling
fractions across a classification that units drift out of can improve
precision initially and reduce it after several waves [@smith2009, sec.
2.4.4].

**The sampling unit and the followed unit may differ, deliberately.** The
standard device for avoiding movers is a sampling unit that cannot move: a
labour force survey that samples addresses balances movers out against movers
in. Such a design is balanced for cross-sectional estimates and for *net*
change, but not for *gross* change [@smith2009, sec. 2.3.5]. `cluster_by()`
expresses it and needs nothing new. The caveat matters because a
dwelling-based rotating panel quietly fails the analysis most people want a
panel for. It also constrains the frame: an address found vacant at one wave
stays in the sample for later waves, so the frame must not be filtered to
occupied units between waves.

**Entrant coverage is a property of the frame, not of the draw.** `samplyr`
can draw from a supplied entrant register. It cannot establish that the
register covers the entrant flow, or that it covers it promptly enough for
short-lived entrants to be represented at all.

**Sampling that depends on the existing sample is out of scope.** Adding
household members reached through an existing sample member gives selection
probabilities that cannot be known from the design, and model-based approaches
with untestable assumptions are required [@lynn2012, sec. 8.1]. Attrition,
measurement error and imputation are likewise estimation-layer problems:
`samplyr` produces design weights only.

## Where this fits

- `vignette("sampling-coordination")` covers permanent random numbers, the
  separate mechanism for controlling overlap between samples. It composes with
  the panel machinery here but answers a different question.
- `vignette("design-semantics")` states the panel partitioning contract: block
  sizes, which stage assigns, and why weights are unchanged without a schedule.
- `vignette("survey-analysis")` covers exporting a wave to `survey` and `srvyr`.
- `vignette("serialization")` covers archiving a master, and what a
  `rotation_program()` cannot store.
- `vignette("survey-planning")` sizes a repeated survey before it runs:
  `design_overlap()` for what the rotation shares, `prec_change()` for the
  precision that buys, and `n_panel()` for recruitment under attrition. It is
  the before-fieldwork counterpart to the change estimation above.

## References
