---
title: "Plan and Run a Rotating Panel"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Plan and Run a Rotating Panel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, echo = FALSE}
library(samplyr)
library(svyplan)
library(dplyr)
```

## Start with the fieldwork question

A longitudinal survey observes some units more than once. A rotating panel
replaces part of the sample at regular intervals, so it can support both
cross-sectional estimates and estimates of change.

This article follows one small design from planning to selection. Each cohort
is interviewed twice. At a steady state, half of one occasion's issued sample
also appears at the next occasion. The example is deliberately simple and
avoids permanent random numbers, unequal-probability sampling, and a choice of
longitudinal variance estimator until the basic workflow is clear.

Two packages have different jobs here. `svyplan` converts precision, response,
retention, and overlap assumptions into fieldwork numbers. `samplyr` selects
the start-up and refreshment samples and records which panels are active.

## Choose the kind of repeated survey

These common designs should not be treated as interchangeable
[@smith2009; @lynn2012].

| Design | What happens | Basic implementation |
|:--|:--|:--|
| Fixed panel | One sample is interviewed repeatedly | Run `execute()` once and join later responses to the selected units |
| Fixed panel with entrants | The original panel is kept and new population entrants are added | Draw the original sample and each entrant cohort separately |
| Repeated cross-section | A fresh population sample is drawn on every occasion | Reuse the design with a new frame and seed |
| Rotating panel | Several cohorts are live and one is replaced regularly | Use a start-up partition, then draw one refreshment cohort per occasion |
| Split panel | A continuing panel and a fresh cross-section run together | Keep the two samples and their estimators distinct |

Before choosing among them, state the target population. A fixed wave-1
population cannot represent later births. A cross-sectional target at each
occasion needs refreshed frame coverage. `samplyr` cannot infer that choice
from the data.

## Plan a two-occasion rotation

`design_rotation()` describes the time a cohort spends in and out of sample.
The string `"2"` means that a cohort is interviewed on two consecutive
occasions and then leaves, and `design_overlap()` reports the overlap it
produces.

```{r rotation-chart, fig.width = 7, fig.height = 3.6, fig.alt = "Rotation chart for cohorts interviewed on two consecutive occasions.", message = FALSE}
library(samplyr)
library(svyplan)
library(dplyr)

rotation <- design_rotation("2")
overlap_plan <- design_overlap(rotation)
overlap_plan[1]
plot(overlap_plan, type = "schedule", start = "immediate")
```

This is issued-sample overlap. Below full response, the overlap among
respondents may differ because response can persist across occasions.

Next, define the responding sample needed for a proportion and allow for
response at recruitment and retention into the second interview.

```{r}
precision_target <- n_prop(p = 0.5, moe = 0.08)

panel_plan <- n_panel(
  precision_target,
  retention = 0.90,
  resp_rate = 0.80,
  design = "rotating",
  start = "immediate"
)

panel_plan

fieldwork_plan <- design_schedule(
  panel_plan,
  rotation,
  horizon = 3,
  horizon_policy = "continuing",
  refreshment = "entrant_register",
  rounding = "ceiling"
)
```

The two numbers needed for selection are explicit in the result.

```{r}
intake_n <- ceiling(panel_plan$n_entrants)

selection_sizes <- c(
  startup = panel_plan$n_cohorts * intake_n,
  intake_each_occasion = intake_n
)
selection_sizes
```

`startup` is the number issued across the two cohorts present at the first
occasion. `intake_each_occasion` is the fresh cohort added later. They are not
two names for the same sample size.

`n_panel()` uses the response and retention rates supplied by the planner. It
does not estimate attrition, adjust weights for nonresponse, or guarantee that
every realized occasion reaches the target. Its optional `assurance` argument
can add a probability-of-reaching-target calculation when that is needed.

## Draw and partition the start-up sample

The immediate start contains two panels recruited together at occasion 1, and
both are then at their first interview. One panel is recruited into a
two-interview life and is active on occasions 1 and 2. The other is recruited
into a one-interview life and is active only on occasion 1. They differ in life
length, not in how far they have already progressed through the retention
chain.

```{r}
data(ken_enterprises)

base_design <- sampling_design() |>
  draw(n = selection_sizes[["startup"]])

startup <- execute(
  base_design,
  ken_enterprises,
  seed = 7,
  panels = fieldwork_plan
)

count(startup, .panel)
```

Passing the plan here lets samplyr extract the startup panels and freeze their
full activity schedule when the sample is drawn. The partition does not draw
future entrants or create later frame vintages. It only organizes this one
executed sample.

Materialize an occasion from the recorded schedule with `execute(wave = )`.

```{r}
startup_wave_1 <- execute(startup, wave = 1)
startup_wave_2 <- execute(startup, wave = 2)

c(
  wave_1 = nrow(startup_wave_1),
  wave_2 = nrow(startup_wave_2)
)
```

Wave activation is a probability subsample, not a row filter. The resulting
object carries an activation factor in its weight and can be exported as a
two-phase survey design.

```{r}
class(as_svydesign(startup_wave_2))
```

Certainty units and permanent small pools do not follow the overlap declared by
this automatic route, so it refuses them. They can still use an explicit
data-frame panel schedule when their operational and inference treatment has
been decided. Unequal-probability and multistage designs also need care in
variance estimation. `as_svydesign()` refuses unsupported combinations rather
than silently replacing their variance estimator.

## Add a refreshment cohort

At occasion 2, the older start-up cohort has left and a fresh cohort enters.
For a dynamic population, that cohort should be drawn from the frame vintage
and eligibility definition specified by the survey design.

The synthetic example below creates a separate entrant register. Its distinct
identifiers make the cohort frames disjoint by construction.

```{r}
set.seed(99)

entrants <- ken_enterprises |>
  slice_sample(n = 500, replace = TRUE) |>
  mutate(
    enterprise_id = paste0("NEW_", sprintf("%04d", row_number()))
  )

intake_design <- sampling_design() |>
  draw(n = selection_sizes[["intake_each_occasion"]])

intake_2 <- execute(intake_design, entrants, seed = 8)
```

`rotation_program()` records how separately executed cohorts fit into one
fieldwork programme.

```{r}
program <- rotation_program(
  cohorts = list(startup = startup, intake_2 = intake_2),
  schedule = fieldwork_plan,
  through = 2
)

occasion_2 <- execute(program, wave = 2)
occasion_2
```

The result is a collection with one component per live cohort. It is not
automatically row-bound and its weights are not combined. The correct combined
estimator depends on who was eligible for each cohort. Disjoint entrant frames
are simpler. Refreshment draws from the whole population can give a unit more
than one chance of selection and require a union-probability or multiple-frame
estimator.

For a longer run, increase `horizon` when creating `fieldwork_plan`, repeat the
intake step once per occasion, and register the cohorts through the latest
fielded occasion. The planned value `panel_plan$n_entrants` is the steady-state
intake for each draw.

## Check overlap in an executed master

`svyplan::design_overlap()` reports overlap from the planned rotation.
`samplyr::joint_expectation()` reports what an executed master records,
including block-size remainders. For manually scheduled designs, it also
handles permanent certainty units.

```{r}
executed_overlap <- joint_expectation(startup, waves = c(1, 2))

c(
  planned = overlap_plan[1],
  executed = sum(executed_overlap$take_both) / sum(executed_overlap$take_1))
```

Be explicit about the denominator when wave sizes differ. Also say whether
the overlap counts issued units or respondents. `samplyr` knows the first, but
response data are needed for the second.

## Attach responses without changing the design

A `tbl_sample` records selection. Later responses can be joined one-to-one by
a stable unit identifier while the sample remains one row per selected unit.
Duplicating a selected row once per interview and exporting that long table as
a new survey design would count the unit as though it had been sampled more
than once.

For two waves derived from the same master, `stack_waves()` checks their common
origin and returns a long table for a longitudinal analysis package.

The establishment frame and the observations below are synthetic. Each
occasion supplies a new measurement for every issued establishment. Retain
the frame's baseline revenue under another name, and join the occasion's
interview by `enterprise_id`. The join preserves the issued rows.

```{r wave-observations}
set.seed(101)
interviews_1 <- startup_wave_1 |>
  as.data.frame() |>
  transmute(enterprise_id,
            revenue_millions = revenue_millions * exp(rnorm(n(), 0, 0.1)))
interviews_2 <- startup_wave_2 |>
  as.data.frame() |>
  transmute(enterprise_id,
            revenue_millions = revenue_millions * exp(rnorm(n(), 0.05, 0.1)))

observed_wave_1 <- startup_wave_1 |>
  rename(revenue_baseline = revenue_millions) |>
  left_join(interviews_1, by = "enterprise_id", relationship = "one-to-one")
observed_wave_2 <- startup_wave_2 |>
  rename(revenue_baseline = revenue_millions) |>
  left_join(interviews_2, by = "enterprise_id", relationship = "one-to-one")

survey::svytotal(~revenue_millions, as_svydesign(observed_wave_1))
survey::svytotal(~revenue_millions, as_svydesign(observed_wave_2))

wave_table <- stack_waves(observed_wave_1, observed_wave_2)
wave_table |>
  select(wave, master_id, enterprise_id, revenue_millions, design_weight) |>
  head(4)
```

Both exports use the current occasion's revenue, including where the retained
master carries a same-named baseline column. A missing current measurement
would remain missing. It would not fall back to the baseline. The master and
the other occasion remain unchanged.

These are separate estimates for the start-up cohort's frame. The second
does not include the refreshment cohort or estimate a growing population's
total automatically. This example assumes complete response on both occasions.

`design_weight` is the design weight, including activation. It is not a final
nonresponse-adjusted or calibrated weight. `master_id` links the same selected
unit across these waves. It is not a cluster identifier.

Two separate cross-sectional variance estimates do not contain the covariance
needed for a variance of change. The appropriate longitudinal estimator
depends on the design, response process, calibration, and available joint
inclusion information. `stack_waves()` validates the handoff but deliberately
does not choose that estimator.

## Where PRNs fit

Permanent random numbers control overlap between separate draws. Panel
schedules control how long an already selected cohort remains active. A survey
may use both, but neither substitutes for the other. Start with this article
when the fieldwork rule is “keep each cohort for two occasions.” Use
`vignette("sampling-coordination")` when the rule is “make the next draw share
more or fewer units with the current draw.”

## Limits to decide before fieldwork

- State whether estimates refer to the wave-1 population, units present at
  every occasion, or the population at each occasion.
- Keep wave-1 strata fixed for a fixed panel. Re-executing a design against a
  new frame instead reclassifies units using that frame.
- Decide whether the sampled unit is also the followed unit. For example,
  rotating addresses while areas remain fixed has different longitudinal
  properties from rotating whole areas.
- Verify entrant coverage and identifier continuity in the frame. A sampling
  function cannot establish either one.
- Plan nonresponse adjustment, calibration, attrition treatment, and the
  variance of change in the analysis layer.

## The short workflow

1. Use `design_rotation()` to describe the cohort life.
2. Use `n_panel()` to obtain `n_in_sample` and `n_entrants`.
3. Draw the start-up sample and partition it with `panels`.
4. Materialize scheduled occasions with `execute(wave = )`.
5. Draw each refreshment cohort against its intended frame vintage.
6. Register the cohorts with `rotation_program()`.
7. Export components with their own designs and use a longitudinal inference
   method suited to the survey.

## References
