---
title: "Sampling Coordination with Permanent Random Numbers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sampling Coordination with Permanent Random Numbers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, echo = FALSE}
library(samplyr)
library(sondage)
library(dplyr)
```

## Why Coordinate Samples?

Repeated surveys such as business panels, labour force surveys, agricultural censuses draw fresh samples at regular intervals from the same population.
Similarly, multiple survey programmes often sample from the same register simultaneously. Two natural questions arise:

- **Positive coordination**: How can we maximise overlap between samples to reduce respondent burden and improve change estimates?
- **Negative coordination**: How can we minimise overlap to spread the burden and maximise cumulative coverage?

These questions apply both *temporally* (across waves of the same survey)
and *cross-sectionally* (across different surveys on the same frame also
known as **collocated sampling**).

The **permanent random number** (PRN) technique, introduced by
Brewer et al. (1972) and developed by Ohlsson (1995) and Cotton & Hesse
(1992), answers both questions with a single mechanism: attach a stable
$U(0,1)$ value to each frame unit and update it between waves.

samplyr supports PRN-based coordination for four selection methods:
`bernoulli`, `pps_poisson`, `pps_sps`, and `pps_pareto`.

## Setup

We use `nigeria_business`, a synthetic business register with 11,617
enterprises.

```{r}
library(samplyr)
library(sondage)
library(dplyr)

data(nigeria_business)
glimpse(nigeria_business)
```

Each enterprise has an `annual_turnover` that will serve as our measure of
size for PPS sampling. We target $n = 200$ enterprises per wave.

```{r}
n_wave <- 200
```

### Assigning permanent random numbers

Every frame unit gets a $U(0,1)$ value drawn once and stored permanently
in the register. These values are reused or updated across waves.

```{r}
set.seed(2026)

frame <- nigeria_business |>
  mutate(u = runif(n()))
```

### Computing inclusion probabilities

Coordination requires knowing each unit's first-order inclusion probability
$\pi_k$. For PPS without replacement, `sondage::inclusion_prob()` computes
these from the measure of size and sample size. This must be done on the
**full frame**, not just the sample.

```{r}
frame <- frame |>
  mutate(pik = inclusion_prob(annual_turnover, n_wave))

summary(frame$pik)
```

## Positive Coordination: Maximum Overlap

To maximise overlap between waves, reuse the same PRN values. With
sequential Poisson sampling (SPS), two waves drawn from the same frame
with identical PRNs produce identical samples.

```{r}
wave1 <- sampling_design() |>
  draw(n = n_wave, method = "pps_sps", mos = annual_turnover, prn = u) |>
  execute(frame, seed = 1)

wave2 <- sampling_design() |>
  draw(n = n_wave, method = "pps_sps", mos = annual_turnover, prn = u) |>
  execute(frame, seed = 1)

overlap_positive <- length(intersect(wave1$enterprise_id, wave2$enterprise_id))
overlap_positive
```

Perfect overlap. This is useful when the goal is to track the same units
over time, for example, measuring quarterly changes in turnover within
a fixed panel.

## Negative Coordination: Minimum Overlap

To minimise overlap, update the PRN values between waves using Tillé's
(2006, §8.2) formula:

$$u_k^{(t+1)} = \bigl(u_k^{(t)} - \pi_k\bigr) \bmod 1$$

Selected units (those with small $u_k$) get pushed to the high end of
$(0,1)$, making them unlikely to be selected again. Non-selected units
shift downward, increasing their chance of selection.

```{r}
# Wave 1
wave1 <- sampling_design() |>
  draw(n = n_wave, method = "pps_sps", mos = annual_turnover, prn = u) |>
  execute(frame, seed = 1)

# Update PRNs on the full frame
frame <- frame |>
  mutate(u = (u - pik) %% 1)

# Wave 2
wave2 <- sampling_design() |>
  draw(n = n_wave, method = "pps_sps", mos = annual_turnover, prn = u) |>
  execute(frame, seed = 1)

overlap_negative <- length(intersect(wave1$enterprise_id, wave2$enterprise_id))
overlap_negative
```

### Why isn't the overlap zero?

Not all overlap can be eliminated. Units with large $\pi_k$ (big enterprises)
are selected in almost every wave because their inclusion probability is close
to or equal to 1. The theoretical minimum overlap known as the **Anticipated Lower
Bound** (ALB) is:

$$\text{ALB} = \sum_{k=1}^{N} \max(0,\; 2\pi_k - 1)$$

Only units with $\pi_k > 0.5$ contribute to forced overlap.

```{r}
alb <- sum(pmax(0, 2 * frame$pik - 1))
expected_independent <- sum(frame$pik^2)

cat("ALB (theoretical minimum):", round(alb), "\n")
cat("Observed overlap:         ", overlap_negative, "\n")
cat("Expected if independent:  ", round(expected_independent), "\n")
cat("AUB (theoretical maximum):", n_wave, "\n")
```

The negative coordination brings the overlap close to the ALB, well below
what independent sampling would produce.

## Comparison: Three Strategies

```{r}
# Independent sampling: fresh PRNs for wave 2
frame_ind <- nigeria_business |>
  mutate(
    u1 = withr::with_seed(2026, runif(n())),
    u2 = runif(n())
  )

wave1_ind <- sampling_design() |>
  draw(n = n_wave, method = "pps_sps", mos = annual_turnover, prn = u1) |>
  execute(frame_ind, seed = 1)

wave2_ind <- sampling_design() |>
  draw(n = n_wave, method = "pps_sps", mos = annual_turnover, prn = u2) |>
  execute(frame_ind, seed = 1)

overlap_ind <- length(intersect(wave1_ind$enterprise_id, wave2_ind$enterprise_id))

tibble(
  strategy = c("Positive (same PRN)", "Independent (fresh PRN)", "Negative (Tillé update)"),
  overlap = c(overlap_positive, overlap_ind, overlap_negative),
  bound = c(n_wave, round(expected_independent), round(alb))
) |>
  mutate(pct = paste0(round(100 * overlap / n_wave), "%"))
```

## Multi-Wave Rotation

In practice, surveys run for many waves. The PRN update is applied after
each wave, rotating units through the sample.

```{r}
set.seed(2026)

frame <- nigeria_business |>
  mutate(
    u = runif(n()),
    pik = inclusion_prob(annual_turnover, n_wave)
  )

n_waves <- 5
waves <- vector("list", n_waves)

for (t in seq_len(n_waves)) {
  waves[[t]] <- sampling_design() |>
    draw(n = n_wave, method = "pps_sps", mos = annual_turnover, prn = u) |>
    execute(frame, seed = 1)

  # Update PRNs for next wave
  frame <- frame |>
    mutate(u = (u - pik) %% 1)
}
```

### Pairwise overlap matrix

```{r}
ids <- lapply(waves, `[[`, "enterprise_id")

overlap_mat <- outer(ids, ids, Vectorize(\(a, b) length(intersect(a, b))))
dimnames(overlap_mat) <- rep(list(paste("Wave", seq_len(n_waves))), 2)

overlap_mat
```

Consecutive waves have low overlap (near the ALB). Non-consecutive waves show
varying overlap as the rotation cycle progresses.

### Cumulative coverage

After $T$ waves, how many distinct enterprises have been surveyed?

```{r}
cumulative <- integer(n_waves)
seen <- character(0)

for (t in seq_len(n_waves)) {
  seen <- union(seen, waves[[t]]$enterprise_id)
  cumulative[t] <- length(seen)
}

tibble(
  wave = seq_len(n_waves),
  new_units = c(n_wave, diff(cumulative)),
  cumulative = cumulative,
  pct_frame = paste0(round(100 * cumulative / nrow(nigeria_business), 1), "%")
)
```

Negative coordination maximises the number of distinct units covered,
which is valuable for building comprehensive registers or spreading
respondent burden.

## Handling Births and Deaths

Real business registers change between waves. New enterprises (**births**)
enter the frame; closed enterprises (**deaths**) leave it.

- **Deaths**: Simply remove them from the frame. Their PRN values are
  discarded. No special treatment needed.
- **Births**: Assign fresh $U(0,1)$ values. These new units enter the
  coordination cycle at a random position.

```{r}
# Simulate: remove 200 enterprises (deaths), add 500 new ones (births)
set.seed(42)

n_births <- 500

# Deaths: remove a random subset
deaths <- sample(frame$enterprise_id, 200)
frame_t2 <- frame |>
  filter(!enterprise_id %in% deaths)

# Births: new enterprises with fresh PRNs
# Resample turnover from the existing frame to get a realistic size distribution
births <- tibble(
  enterprise_id = paste0("NEW_", sprintf("%05d", seq_len(n_births))),
  zone = sample(levels(nigeria_business$zone), n_births, replace = TRUE) |>
    factor(levels = levels(nigeria_business$zone)),
  state = sample(levels(nigeria_business$state), n_births, replace = TRUE) |>
    factor(levels = levels(nigeria_business$state)),
  sector = sample(levels(nigeria_business$sector), n_births, replace = TRUE) |>
    factor(levels = levels(nigeria_business$sector)),
  size_class = sample(levels(nigeria_business$size_class), n_births, replace = TRUE) |>
    factor(levels = levels(nigeria_business$size_class)),
  employees = sample(nigeria_business$employees, n_births, replace = TRUE),
  annual_turnover = sample(nigeria_business$annual_turnover, n_births, replace = TRUE),
  u = runif(n_births)  # fresh PRNs
)

frame_t2 <- bind_rows(frame_t2, births)

# Recompute inclusion probabilities on the new frame
frame_t2 <- frame_t2 |>
  mutate(pik = inclusion_prob(annual_turnover, n_wave))

# Draw wave from updated frame
wave_t2 <- sampling_design() |>
  draw(n = n_wave, method = "pps_sps", mos = annual_turnover, prn = u) |>
  execute(frame_t2, seed = 1)

# How many births were selected?
n_births_selected <- sum(wave_t2$enterprise_id %in% births$enterprise_id)
cat("Births selected:", n_births_selected, "out of", n_births, "\n")
```

The coordination mechanism automatically handles frame changes. Surviving
units retain their updated PRN values; births start fresh.

## Using Pareto Sampling

The `pps_pareto` method also supports PRN-based coordination. In theory,
Pareto sampling has maximum entropy among fixed-size PPS designs that
use PRNs, making it particularly well-suited for coordination.

```{r}
set.seed(2026)

frame <- nigeria_business |>
  mutate(
    u = runif(n()),
    pik = inclusion_prob(annual_turnover, n_wave)
  )

wave1_par <- sampling_design() |>
  draw(n = n_wave, method = "pps_pareto", mos = annual_turnover, prn = u) |>
  execute(frame, seed = 1)

frame <- frame |> mutate(u = (u - pik) %% 1)

wave2_par <- sampling_design() |>
  draw(n = n_wave, method = "pps_pareto", mos = annual_turnover, prn = u) |>
  execute(frame, seed = 1)

overlap_par <- length(intersect(wave1_par$enterprise_id, wave2_par$enterprise_id))

cat("Overlap (Pareto, negative):", overlap_par, "\n")
cat("ALB:", round(sum(pmax(0, 2 * frame$pik - 1))), "\n")
```

## Stratified Coordination

When the design is stratified, inclusion probabilities must be computed
**within each stratum**. The PRN update is then applied globally the per-stratum $\pi_k$ values ensure correct rotation within each group.

```{r}
set.seed(2026)

frame <- nigeria_business |>
  mutate(u = runif(n()))

# Compute within-stratum pik
frame <- frame |>
  mutate(
    pik = inclusion_prob(annual_turnover, 50),
    .by = size_class
  )

cat("Max pik:", max(frame$pik), "\n")
cat("ALB:", round(sum(pmax(0, 2 * frame$pik - 1))), "\n")

wave1_str <- sampling_design() |>
  stratify_by(size_class) |>
  draw(n = 50, method = "pps_sps", mos = annual_turnover, prn = u) |>
  execute(frame, seed = 1)

frame <- frame |> mutate(u = (u - pik) %% 1)

wave2_str <- sampling_design() |>
  stratify_by(size_class) |>
  draw(n = 50, method = "pps_sps", mos = annual_turnover, prn = u) |>
  execute(frame, seed = 1)

overlap_str <- length(intersect(wave1_str$enterprise_id, wave2_str$enterprise_id))

cat("Overlap (stratified, negative):", overlap_str, "out of", nrow(wave1_str), "\n")
```

With smaller within-stratum $\pi_k$ values (all below 0.5), the ALB is zero
and perfect negative coordination is achievable.

## Collocated Sampling: Coordinating Across Surveys

The examples above coordinate samples from the **same survey** across time
(waves). A related problem is coordinating **different surveys** drawn from
the same frame at the same time point. This is called *collocated sampling*
(Ernst 1986; Ohlsson 1995) and is common in business survey programmes
where a single register serves multiple survey operations.

The mechanism is identical: all surveys share the same PRN column.

- **Positive coordination** (maximise overlap): both surveys use the same PRN.
  Useful when surveys share common modules or when joint analysis is
  planned.
- **Negative coordination** (minimise overlap): one survey uses shifted PRNs
  ($u' = (u + 0.5) \bmod 1$). Useful when the goal is to spread respondent
  burden across the register.

### Example: Turnover survey and employment survey

Suppose a statistical agency runs two surveys on the same business register:
a turnover survey ($n = 200$, PPS on `annual_turnover`) and an employment
survey ($n = 150$, PPS on `employees`). The surveys have different sample
sizes, different measures of size, and different stratification but they
share the same frame and the same PRN column.

```{r}
set.seed(2026)

frame <- nigeria_business |>
  mutate(u = runif(n()))

# Survey A: turnover survey (PPS on annual_turnover, stratified by sector)
survey_a <- sampling_design() |>
  draw(n = 200, method = "pps_sps", mos = annual_turnover, prn = u) |>
  execute(frame, seed = 1)

# Survey B: employment survey (PPS on employees) — positive coordination
survey_b <- sampling_design() |>
  draw(n = 150, method = "pps_sps", mos = employees, prn = u) |>
  execute(frame, seed = 1)

overlap_pos <- length(intersect(survey_a$enterprise_id, survey_b$enterprise_id))
```

For negative coordination, shift the PRN by 0.5:

```{r}
# Survey B with negative coordination
frame_neg <- frame |>
  mutate(u_neg = (u + 0.5) %% 1)

survey_b_neg <- sampling_design() |>
  draw(n = 150, method = "pps_sps", mos = employees, prn = u_neg) |>
  execute(frame_neg, seed = 1)

overlap_neg <- length(intersect(survey_a$enterprise_id, survey_b_neg$enterprise_id))
```

Compare with independent sampling (fresh PRNs):

```{r}
frame_ind <- frame |>
  mutate(u_fresh = runif(n()))

survey_b_ind <- sampling_design() |>
  draw(n = 150, method = "pps_sps", mos = employees, prn = u_fresh) |>
  execute(frame_ind, seed = 1)

overlap_ind <- length(intersect(survey_a$enterprise_id, survey_b_ind$enterprise_id))
```

### Overlap bounds for different surveys

When two surveys have different inclusion probabilities $\pi_k^A$ and $\pi_k^B$,
the bounds generalise naturally:

```{r}
pik_a <- inclusion_prob(frame$annual_turnover, 200)
pik_b <- inclusion_prob(frame$employees, 150)

alb_ab <- sum(pmax(0, pik_a + pik_b - 1))
aub_ab <- sum(pmin(pik_a, pik_b))
exp_ind <- sum(pik_a * pik_b)

tibble(
  strategy = c("Positive (same PRN)", "Independent (fresh PRN)", "Negative (shifted PRN)"),
  overlap = c(overlap_pos, overlap_ind, overlap_neg),
  bound = c(round(aub_ab), round(exp_ind), round(alb_ab))
)
```

Because `annual_turnover` and `employees` are correlated (large firms have
both high turnover and many employees), the overlap under positive coordination
is substantial. Negative coordination via PRN shift effectively spreads the
burden.

### Scaling to more than two surveys

For $K$ surveys, shift each survey's PRN by $j/K$ for $j = 0, \ldots, K-1$:

```{r, eval = FALSE}
# Negative coordination across 4 surveys
for (j in seq_len(4)) {
  frame_j <- frame |>
    mutate(u_j = (u + (j - 1) / 4) %% 1)

  samples[[j]] <- sampling_design() |>
    draw(n = n_j, method = "pps_sps", mos = mos_j, prn = u_j) |>
    execute(frame_j, seed = 1)
}
```

This simple shift strategy works well when the surveys have similar
$\pi_k$. For more complex programmes, NSOs use dedicated
coordination systems such as the Rivière method (Rivière 1998),
the Netherlands EDS (De Ree 1983; Van Huis et al. 1994), or the
Swiss method (Graf & Qualité 2014). These methods track cumulative
burden per unit and permute PRNs within microstrata to achieve
near-optimal coordination across the full survey programme. See
Tillé (2020, §8.2.7--8.2.9) for a detailed treatment.

## Summary

The PRN coordination workflow in samplyr is:

1. **Assign PRNs** once: `mutate(u = runif(n()))`
2. **Compute $\pi_k$** on the full frame: `mutate(pik = inclusion_prob(mos, n))`
3. **Draw the sample**: `sampling_design() |> draw(method = "pps_sps", prn = u) |> execute(frame)`
4. **Update or shift PRNs**:
   - Temporal negative coordination (waves): `mutate(u = (u - pik) %% 1)`
   - Temporal positive coordination: keep `u` unchanged
   - Cross-survey negative coordination: shift by `(u + 0.5) %% 1`
   - Cross-survey positive coordination: reuse same `u`

| Goal | PRN update | Expected overlap |
|------|-----------|-----------------|
| Maximum overlap | None (reuse) | $\sum \min(\pi_k^A, \pi_k^B)$ |
| Independent | Fresh $U(0,1)$ | $\sum \pi_k^A \cdot \pi_k^B$ |
| Minimum overlap | Shift or Tillé update | $\sum \max(0, \pi_k^A + \pi_k^B - 1)$ |

## References

- Brewer, K. R. W., Early, L. J. & Joyce, S. F. (1972). Selecting several
  samples from a single population. *Australian Journal of Statistics*, 14(3),
  231--239.
- Cotton, F. & Hesse, C. (1992). Tirages coordonnés d'échantillons.
  *Actes des Journées de Méthodologie Statistique*, INSEE.
- De Ree, S. J. M. (1983). The EDS sampling system for business surveys at
  the Netherlands Central Bureau of Statistics. *Proceedings of the Section
  on Survey Research Methods*, ASA, 197--202.
- Ernst, L. R. (1986). Variance of the estimated mean for several imputation
  procedures. *Proceedings of the Section on Survey Research Methods*, ASA.
- Graf, E. & Qualité, L. (2014). The Swiss coordination system.
  *Statistics in Transition New Series*, 15(3), 371--396.
- Ohlsson, E. (1995). Coordination of samples using permanent random
  numbers. In *Business Survey Methods* (eds Cox et al.), Wiley.
- Rivière, P. (1998). Méthode de micro-stratification pour les tirages
  coordonnés d'échantillons. *Actes des Journées de Méthodologie
  Statistique*, INSEE.
- Tillé, Y. (2006). *Sampling Algorithms*. Springer. Chapter 8.
- Tillé, Y. (2020). *Sampling and Estimation from Finite Populations*. Wiley.
  Section 8.2.
- Van Huis, L. T., De Ree, S. J. M. & Grotendorst, A. (1994). A business
  survey redesign at Statistics Netherlands. *International Statistical
  Review*, 62(3), 357--376.
