---
title: "Sampling Coordination with Permanent Random Numbers"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Sampling Coordination with Permanent Random Numbers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, echo = FALSE}
library(samplyr)
library(sondage)
library(dplyr)
```

## Why Coordinate Samples?

Repeated surveys (business panels, labor force surveys, agricultural censuses) draw fresh samples at regular intervals from the same population.
Similarly, multiple survey programs often sample from the same register simultaneously. Two natural questions arise:

- **Positive coordination**: How can we maximize overlap between samples to reduce respondent burden and improve change estimates?
- **Negative coordination**: How can we minimize overlap to spread the burden and maximize cumulative coverage?

These questions apply both *temporally* (across waves of the same survey) and *cross-sectionally* (across different surveys on the same frame, known as **collocated sampling**).

The **permanent random number** (PRN) technique, introduced by @brewer1972 and developed by @ohlsson1995 and @cotton1992, answers both questions with a single mechanism: attach a stable $U(0,1)$ value to each frame unit and update it between waves.

`samplyr` supports PRN-based coordination for four selection methods: `bernoulli`, `pps_poisson`, `pps_sps`, and `pps_pareto`.

`pps_poisson` keeps $\pi_i$ dependent only on unit $i$ and the frame total, which is what makes coordination stable across waves. The cost is that a unit large enough to reach $\pi_i > 1$ has its chance clipped to 1, and the expected sample size falls below the target. On a business register this is the ordinary case, since revenue is heavily skewed. `execute()` reports it with a warning of class `samplyr_warning_poisson_shortfall`; handle the dominant units explicitly with `certainty_size` or `certainty_prop`. Note that this re-resolves the remaining units over a smaller MOS total, so their $\pi_i$ rise and now depend on which units were declared certain. Accepting the shortfall is what keeps $\pi_i$ a function of unit $i$ and the frame total alone. See `?draw`.

## Setup

We use `ken_enterprises`, a synthetic business register with 17,004 enterprises.

```{r}
library(samplyr)
library(sondage)
library(dplyr)

data(ken_enterprises)
glimpse(ken_enterprises)
```

Each enterprise has a `revenue_millions` column that serves as our measure of size for PPS sampling. We target $n = 200$ enterprises per wave.

```{r}
n_wave <- 200
```

### Assigning permanent random numbers

Every frame unit gets a $U(0,1)$ value drawn once and stored permanently in the register. These values are reused or updated across waves.

```{r}
set.seed(1)

frame <- ken_enterprises |>
  mutate(u = runif(n()))
```

### Computing inclusion probabilities

Coordination requires knowing each unit's first-order inclusion probability $\pi_k$. For PPS without replacement, `sondage::inclusion_prob()` computes
these from the measure of size and sample size. This must be done on the **full frame**, not just the sample.

```{r}
frame <- frame |>
  mutate(pik = inclusion_prob(revenue_millions, n_wave))

round(quantile(frame$pik), 4)
```

## Positive Coordination: Maximum Overlap

To maximize overlap between waves, reuse the same PRN values. With sequential Poisson sampling (SPS), two waves drawn from the same frame with identical PRNs produce identical samples.

```{r}
wave1 <- sampling_design() |>
  draw(n = n_wave, method = "pps_sps", mos = revenue_millions, prn = u) |>
  execute(frame, seed = 42)

wave2 <- sampling_design() |>
  draw(n = n_wave, method = "pps_sps", mos = revenue_millions, prn = u) |>
  execute(frame, seed = 123)

overlap_positive <- length(intersect(wave1$enterprise_id,
                                     wave2$enterprise_id))
overlap_positive
```

Perfect overlap. This is useful when the goal is to track the same units over time, for example measuring quarterly changes in revenue within a fixed panel.

Overlap is harder to keep when the *design* changes rather than the frame. Reusing PRNs preserves overlap only while strata and inclusion probabilities stay put; a redesign moves both, and maximizing the retained sample under new strata and new probabilities is a separate problem with its own literature [@kishscott1971; @hidiroglou1991].

## Negative Coordination: Minimum Overlap

To minimize overlap, update the PRN values between waves using @tille2006 [§8.2] formula:

$$u_k^{(t+1)} = \bigl(u_k^{(t)} - \pi_k\bigr) \bmod 1$$

Selected units (those with small $u_k$) get pushed to the high end of $(0,1)$, making them unlikely to be selected again. Non-selected units shift downward, increasing their chance of selection.

```{r}
# Wave 1
wave1 <- sampling_design() |>
  draw(n = n_wave, method = "pps_sps", mos = revenue_millions, prn = u) |>
  execute(frame, seed = 1)

# Update PRNs on the full frame
frame <- frame |>
  mutate(u = (u - pik) %% 1)

# Wave 2
wave2 <- sampling_design() |>
  draw(n = n_wave, method = "pps_sps", mos = revenue_millions, prn = u) |>
  execute(frame, seed = 1)

overlap_negative <- length(intersect(wave1$enterprise_id, wave2$enterprise_id))
overlap_negative
```

### Why isn't the overlap zero?

Not all overlap can be eliminated. Units with large $\pi_k$ (big enterprises)
are selected in almost every wave because their inclusion probability is close
to or equal to 1. The **Anticipated Lower Bound** (ALB) for expected overlap is:

$$\text{ALB} = \sum_{k=1}^{N} \max(0,\; 2\pi_k - 1)$$

Only units with $\pi_k > 0.5$ contribute to forced overlap.

```{r}
alb <- sum(pmax(0, 2 * frame$pik - 1))
expected_independent <- sum(frame$pik^2)

tibble(x = c("ALB (expected lower bound)",
             "Observed overlap",
             "Expected if independent",
             "AUB (expected upper bound)"),
       y = c(round(alb), overlap_negative,
             round(expected_independent), n_wave))
```

The negative coordination brings the realized overlap near the anticipated
lower-bound benchmark, well below what independent sampling would produce. The
ALB applies to expected overlap across repeated coordinated draws; one realized
overlap can fall on either side of its rounded value.

## Comparison: Three Strategies

```{r}
# Independent sampling: fresh PRNs for wave 2
set.seed(1)
frame_ind <- ken_enterprises |>
  mutate(u1 = runif(n()),
         u2 = runif(n()))

wave1_ind <- sampling_design() |>
  draw(n = n_wave, method = "pps_sps", mos = revenue_millions, prn = u1) |>
  execute(frame_ind, seed = 1)

wave2_ind <- sampling_design() |>
  draw(n = n_wave, method = "pps_sps", mos = revenue_millions, prn = u2) |>
  execute(frame_ind, seed = 1)

overlap_ind <- length(intersect(wave1_ind$enterprise_id,
                                wave2_ind$enterprise_id))

tibble(strategy = c("Positive (same PRN)", "Independent (fresh PRN)", "Negative (Tille update)"),
       overlap = c(overlap_positive, overlap_ind, overlap_negative),
       bound = c(n_wave, round(expected_independent), round(alb))) |>
  mutate(pct = paste0(round(100 * overlap / n_wave), "%"))
```

## Longitudinal Designs

Coordination is a mechanism; what you point it at is a design. @smith2009
identify five broad types of longitudinal survey design. Four of them need
nothing beyond the verbs used so far. This section maps them onto `samplyr`
and states the limits that the code cannot show you.

### Vocabulary

Longitudinal work overloads words that already mean something specific in a
sampling design. In `samplyr`:

| Term            | Meaning                                                |
|:----------------|:-------------------------------------------------------|
| stage           | selection within units selected at an earlier stage     |
| phase           | probability subsampling from an earlier sample          |
| frame           | a source covering a target population                   |
| wave            | a programme or calendar occasion                        |
| panel           | a randomized rotation, reserve, or workload group       |
| cohort          | a sample's entry batch or refreshment vintage           |
| unit identity   | the substantive entity followed across occasions        |

Four consequences follow. A wave is not a stage: `add_stage()` selects within
what an earlier stage selected, whereas a wave revisits units already
selected. A panel group is not a refreshment cohort: `panels` partitions one
draw, a cohort is a separate draw. Repeated observation of every eligible unit
adds no selection probability, so interviewing a fixed panel five times does
not change `.weight`. And response, attrition and tracing failure are not
selection operations; they belong to the estimation layer.

The printed token is `.panel`. There is deliberately no `.wave` column:
`samplyr` realizes designs against frames and does not model time.

### Defining the longitudinal population

Before choosing a design, decide which population the estimates are about
[@smith2009, sec. 2.3.1]. There are three answers and they lead to different
designs:

- **Static**: the population as of wave 1. Deaths occur, births cannot enter.
  Population information is fixed for all waves, so weighting is
  straightforward, and the natural design is a fixed panel. The risk is that
  readers take the results for the third option.
- **Intersection**: units present at every wave. Both births and deaths are
  excluded, auxiliary totals are hard to construct, and the definition changes
  whenever a wave is added.
- **Union**: units present at any wave. This suits a genuinely dynamic
  population, requires a design that introduces births, and is what most
  household panels use. Adding a wave changes the population and may require
  revising earlier estimates.

This is a declaration, not a computation, so it is not an argument to any
verb. It is what determines whether you need births at all, and therefore
which design below applies.

### The five designs

| Design                  | What it is                                                                          | In `samplyr`                                                                                                                                                                                                     |
|:------------------------|:------------------------------------------------------------------------------------|:-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| Fixed panel             | The same units on every occasion; no additions after wave 1                          | One `execute()`. Waves 2 to *T* are repeated measurement on the same `tbl_sample`; the base selection weight does not change because another interview happened.                                                    |
| Fixed panel plus births | A fixed panel with a sample of recent entrants added at each occasion                | The original `execute()`, plus a separate `execute()` against an entrant register for each refreshment cohort. The components stay separate objects until their target population and estimator are defined.        |
| Repeated panel          | A series of panels, each representing an equivalent population defined at a later time | Reuse the same frame-independent `sampling_design` in a fresh `execute(design, frame_j, seed = seed_j)`. Not `replay_design()`, which reproduces one recorded execution and refuses a changed frame.                |
| Rotating panel          | A fixed proportion of units replaced at each occasion                                | `execute(panels = k)` for the start-up partition; the replenishment is ordinary repeated execution. See below, because this is the one that is easy to overclaim.                                                    |
| Split panel             | A panel plus a fresh cross-sectional sample at each occasion [@kish1987, pp. 181--183] | Two independent executions kept as separate objects. In the *P* = 1 variation a subset of units is included on every occasion, so longitudinal estimates are possible for that stratum; those units already carry `.certainty_k`.                                             |

A rotating panel is a repeated panel whose overlap pattern is fixed and
usually balanced, which is what allows the live groups to be combined for a
cross-sectional estimate [@kaltoncitro1993]. Labour force surveys in many
countries use one [@steel1997].

### What `panels` builds, and what it does not

A real rotating panel replenishes itself: at each entry time it draws a fresh
group from the population as it stands then, so the live groups carry
different population vintages [@lynn2012, sec. 3]. That is exactly what makes
them combinable.

`execute(panels = k)` partitions **one** execution against **one** frame.
Every group comes from the same frame vintage, and no entrant can appear in
any of them. Over a long horizon that cannot represent the cross-sectional
population at later waves.

Where the partition is exactly right is start-up. The "immediate start"
option draws the full sample in year 1 and subdivides it into groups
designated to remain for one, two, three, ... years, with a fresh group added
each year thereafter; it is preferred to a gradual start because it delivers
the full sample size from year 1. So the coherent shape is:

```text
year 1     execute(design, frame, panels = k)     start-up partition
year 2+    execute(design, frame_vintage_t)       one fresh group per period
```

`panels` builds the start-up partition and the workload rotation over a frozen
master. Steady-state replenishment is ordinary repeated execution against
successive frame vintages, which is what the rest of this vignette does.

Combining the live groups is then not a matter of pooling weights. A unit that
entered the population between periods 1 and 2 could not have been drawn into
group 1, so the combined sample under-represents joiners unless their
probabilities are corrected. For an initial sample and a later one
[@lynn2012, sec. 8.1]:

```text
stayers   P ~= n1/N1 + n2/N2      small-fraction approximation
leavers   P = n1/N1
joiners   P = n2/N2
w = 1/P
```

This is still an inverse inclusion probability, so it stays inside the
single-frame estimator. What it needs is data `samplyr` cannot invent: for
each unit, whether it was eligible for each earlier draw. That comes from
survey questions establishing population membership at the earlier date, and
it is easy to forget to ask them. Supply it as a column on the frame; two
execution receipts cannot reveal it.

One practical note: `panels` composes with PRN coordination. The conflict
check in `execute()` is against `reps`, not `prn`, so a PRN-rotated draw can
also be partitioned into workload groups.

### Scheduling the groups

Declaring *when* each group is live turns the partition into a rotation you
can materialize. Pass `panels` a data frame with `panel` and `wave` columns
and an optional logical `active` column instead of a count. Here four groups
rotate two at a time, so each group is live for two consecutive occasions:

```{r}
rotation <- data.frame(
  panel  = rep(1:4, times = 4),
  wave   = rep(1:4, each = 4),
  active = c(TRUE,  TRUE,  FALSE, FALSE,
             FALSE, TRUE,  TRUE,  FALSE,
             FALSE, FALSE, TRUE,  TRUE,
             TRUE,  FALSE, FALSE, TRUE)
)

master <- sampling_design() |>
  draw(n = n_wave, method = "pps_sps", mos = revenue_millions, prn = u) |>
  execute(frame, seed = 7, panels = rotation)

table(master$.panel)
```

The schedule is read at the draw, not only later: the fewest panels any wave
activates sets the assignment block size, so a schedule that keeps two of
four groups live blocks more finely than one that keeps a single group live.

`execute(master, wave = t)` then realizes one occasion:

```{r}
wave_2 <- execute(master, wave = 2)

c(master = nrow(master), wave_2 = nrow(wave_2))
table(rotating = !wave_2$.certainty_1)
```

Two of four groups is a little under half the master, plus every certainty
unit: `pps_sps` on a skewed register resolves the largest enterprises to
$\pi_k = 1$, and a permanent unit is in the sample at every occasion.

Activation is a probability subsample, not a filter, so the wave carries
weights that estimate the same population total the master does:

```{r}
c(master = sum(master$.weight), wave_2 = sum(wave_2$.weight))
```

They are close rather than equal, which is the point: the activation is a
second sampling phase and its estimator is unbiased, not exact.

The factor applied to each weight is the inverse of the unit's activation
probability, which is its block's frozen quota for the active panels over the
block size. That is exact, and it is generally not the nominal `k / r`
implied by counting groups: a block that took the remainder carries a
different quota, and its units carry a different factor. Certainty units are
activated with probability one and their weights are untouched.

A wave is a two-phase sample and exports as one. It retains the master as its
first phase, and `as_svydesign()` passes the activation to `twophase()` as the
second: the blocks become the phase-2 strata and their frozen sizes the
phase-2 population counts, which is what makes the subsampling variance exact
rather than assumed. `method` is the only choice the export takes, `"full"` by
default with `"simple"` and `"approx"` available where their assumptions hold,
and none of them is ever selected as a silent fallback for another.

```{r}
equal_master <- sampling_design() |>
  draw(n = n_wave) |>
  execute(frame, seed = 7, panels = rotation)

equal_wave <- execute(equal_master, wave = 2)

class(as_svydesign(equal_wave))
```

The master drawn above is a different case. `twophase()` accepts no `pps`
specification at its first phase, so a master with unequal inclusion
probabilities has no exact linearization there, and `as_svydesign(wave_2)`
refuses with that reason rather than returning the with-replacement
approximation unannounced. The activation weights stay exact and estimate
totals correctly; it is the variance that has no exact form. The master's own
export is unaffected, and so is any wave of an equal-probability master,
clustered or multistage.

### Which unit rotates

By default the assignment belongs to the first stage, so a clustered design
rotates whole primary units and every unit below inherits its ancestor's
group. `panel_stage` moves it to another stage. That is what expresses the
address-panel design of chapter 2, section 2.3.4: the selected areas stay in
the survey while the addresses inside them rotate.

```{r}
addresses <- data.frame(
  ea      = rep(sprintf("EA%02d", 1:40), each = 12),
  address = rep(1:12, times = 40)
)

area_panel <- sampling_design() |>
  add_stage() |> cluster_by(ea) |> draw(n = 12) |>
  add_stage() |> cluster_by(address) |> draw(n = 8) |>
  execute(addresses, seed = 11, panels = rotation, panel_stage = 2)

# Panels per selected area: every area carries all four.
table(tapply(area_panel$.panel, area_panel$ea, function(x) length(unique(x))))
```

Every area is in every wave, and half its addresses turn over between
consecutive occasions:

```{r}
area_wave <- execute(area_panel, wave = 2)

c(areas_in_master = length(unique(area_panel$ea)),
  areas_in_wave   = length(unique(area_wave$ea)),
  addresses_in_wave = nrow(area_wave))
```

Two things change with the assignment stage, and both matter operationally. A
pool becomes the assignment stage's strata *inside* each realized parent, so
pools are as small as the within-parent take rather than as large as a
stratum; `small_pool` therefore binds far more often, and a design with fewer
than `panels - r` addresses per area is refused rather than silently leaving
areas out of some wave. And selection certainty counts only at the assignment
stage, in both directions: a certainty area does not make its addresses
permanent, and a certainty selection *below* the address does not keep that
address in every wave. A person selected with probability one inside a
rotating address leaves the survey with the address, which is a property of
the design rather than a defect to work around.

This is not a better default, and it should not be adopted as one. Holding the
parent fixed while its members rotate can bias cross-sectional estimates over
time [@smith2009, sec. 2.3.4]. A sampling unit that cannot move between
parents is balanced for cross-sectional estimates and for net change, but not
for gross change. And the frame has an operational consequence that is easy to
get wrong: a vacant address found at one wave stays in the sample for later
waves, so the frame must not be filtered to occupied units between waves.

### How much two occasions share

How far two waves overlap is fixed at the master's draw, so it can be asked
before either has been materialized:

```{r}
joint_expectation(master, waves = c(1, 2))
```

One row per block. `prob_1` and `prob_2` are the chances a unit of that block
is live at each wave, `joint_same` the chance it is live at both, and
`joint_distinct` the chance that two different units of the block are live at
one wave each. Take the same wave twice, `waves = c(2, 2)`, for the joint
expectation within a single wave.

These follow from the frozen quotas rather than from counting groups. In a
block of `m` assignment units where the panels live at the two waves take
$a_t$ and $a_s$ units and those live at both take $a_\cap$:

$$P(i \in W_t,\, j \in W_s) = \frac{a_t a_s - a_\cap}{m(m-1)}, \qquad i \neq j$$

with $a_\cap / m$ for a single unit at both waves. Units in different blocks
are independent. A block that took the remainder carries different quotas and
therefore different values, which is why these are read from the record.

Aggregating `take_both` against `take_1` gives the overlap proportion the
rotation literature quotes: the US CPS 4-8-4 pattern gives 75% from one month
to the next and 50% a year apart, and an in-for-15 rotation 93% and 20%.
Reproducing those figures needs the two spells of the 4-8-4 declared with a
finite life rather than cycled, because a group that returns to its first
spell keeps 7 of 8 groups month to month instead of 6.

**Say which overlap you mean.** Three qualifiers are needed before the number
is interpretable, and the same design honestly reports 0.75, 0.53 or 0.33
depending on which was intended:

- **reference set**: the units the design selected, or the units that
  responded. `samplyr` knows the first exactly. The second needs response
  records, which it does not have.
- **direction**: which wave is the denominator. Where the waves differ in
  size, an asymmetric schedule gives 0.50 forward and 0.33 reversed.
- **measure**: units counted, or weighted. A `pps_sps` master with permanent
  certainty units reports 0.75 by count and 0.53 by weight, because those
  units are live at every wave while carrying weight one.

Written out, the count-based directional quantity is

$$g_{\text{design}}(t \to s) = \frac{\#(\text{selected at both } t \text{ and } s)}{\#(\text{selected at } t)} = \frac{\sum_b a_{\cap b}}{\sum_b a_{tb}}$$

This is the $g$ that appears in the rotation literature [@bergerpriam2016].
It is **not** the $k$ of the resampling shortcut in @cantydavison1999, which
is the responding overlap: they warn explicitly that the realized proportion
should be used there rather than the theoretical one, and the difference is
nonresponse.

Under `samplyr`'s frozen-activation contract the scheduled and the selected
overlap coincide exactly. A rotating unit is live at both waves precisely when
its panel lies in both active sets, and a permanent certainty unit is live at
every wave whatever its panel, so in both cases the realized count is
determined rather than random. That is a property of the contract rather than
of rotation in general: an operation that drew again at wave time, or
subsampled for attrition, would separate them.

These are conditional on the units the master selected and on the frozen
quotas. They are not the unconditional joint inclusion probabilities of the
whole two-phase design, which also carry the master's own pairwise term.

One limit remains. A wave cannot be replayed, because it is derived from a
recorded execution rather than being one; save and replay the master instead.

Finally, the schedule states which groups are live when. It does not
replenish the sample: every group still comes from the frame vintage the
master was drawn from, which is the scoping point above.

### Estimating change across waves

Two waves exported separately give two variances and no covariance, and the
covariance is most of the answer. `stack_waves()` performs the structural half
of the handoff: it verifies that the waves come from one executed master and
returns a plain long table for an inference layer.

```{r}
tab <- stack_waves(
  execute(equal_master, wave = 1),
  execute(equal_master, wave = 2)
)

dim(tab)
names(tab)[1:4]
```

Two of those names carry the contract. `design_weight` is the exact design
weight including the activation factor, and is **not** a final weight: nothing
here is adjusted for nonresponse or calibrated, so a routine asking for
`w_final` is asking for something the analysis must supply. `master_id`
matches units across waves of one master and means nothing outside it; in
particular it is not a primary sampling unit, and a clustered design's own
cluster variable is carried through for that.

Verification is the reason to prefer this to `rbind()`. Two executions of one
design produce the same `.sample_id` values, so waves of *different* masters
stack silently even when the two frames share no unit at all, and the result
matches units that were never the same unit.

#### Two ways to get the variance of change wrong

Write $V_\Delta = V_t + V_s - 2\rho\sqrt{V_t V_s}$. Both mistakes are
common and they fail in opposite directions.

Treating the waves as independent sets $\rho = 0$. That is conservative, and
at high overlap it is severely so: on a 93% rotation it returns **thirteen
times** the correct variance, so intervals are more than three times too wide
and real changes go undetected.

Inserting a correlation that is slightly too high is the dangerous one,
because the error is amplified. To first order

$$\text{RB}(\hat V_\Delta) \approx \frac{-2\rho\sqrt{V_t V_s}}{V_\Delta}\,\text{RB}(\hat\rho)$$

which is $\rho/(\rho-1)$ when the two waves have equal variances. At
$\rho = 0.93$ that multiplier is $-13.3$: a correlation **1% too high**
understates the variance by 13%. The multiplier is a property of the
correlation and the margins, not of the overlap, so it should be computed for
the design in hand rather than read off a rotation pattern.

The classical shortcut of taking the correlation from the matched sample and
combining it with the full-sample variances [@kish1965] fails this way at
every overlap, because it omits the factor $g$: it estimates how correlated a
*unit* is with itself over time, where what is wanted is how correlated the
two *estimators* are. It returned 14% of the correct variance in our
simulations, with 95% intervals covering 50% of the time.

#### Which route to use

| Route | Needs | Behaviour |
|---|---|---|
| Berger--Priam form B [@bergerpriam2016] | Union sample, wave indicators, design strata when stratified, design weights | Calibrated from 7% to 93% overlap in the fixtures measured here: equal-probability masters, small wave fractions, randomized selection |
| Berger--Priam form A | The same | Anti-conservative at high overlap: 78% of the variance, 89.5% coverage |
| `vardpoor::vardchanges()` | A calibrated `w_final`, stable PSU | Reproduces form A exactly, and inherits its behaviour |
| Independence | Nothing | Conservative; badly inefficient at high overlap |
| Matched-sample correlation | The common sample | Not recommended at any overlap |

All of these are approximate. Their consistency arguments assume a negligible
sampling fraction and a high-entropy master, and the two assumptions do not
behave alike here.

**The sampling fraction matters, and the number is smaller than it sounds.**
What counts is the fraction of the population in a *wave*, not in the master,
and a rotation that keeps half its panels live puts half the master in each
wave. Form B returns the right variance with nominal coverage at wave
fractions of 2.5% and 5%. By 10% it returns 88% of the variance, by 15% its
coverage is falling, and at 25% it returns 72% at 90% coverage. Form A is
worse throughout.

The mechanism is that the correlation construction carries no finite
population correction while the correlation it estimates does: the true value
falls as the fraction grows, the estimate stays put, and the resulting
positive bias is amplified.

In this high-overlap, equal-probability fixture, performance was calibrated
through a 5% wave fraction, showed measurable degradation at 10%, and was
clearly anti-conservative at 25%. **These are empirical landmarks, not
universal cut-offs**: one population and one rotation cannot establish a
threshold for a design you have not measured.

**A systematic master is outside the validated guarantee**, and its failures
are worth understanding because of how they present.

Two things combine in any reported variance: the correlation, and the marginal
variances. `samplyr` exports a systematic master through the
equal-probability path, so those margins are the simple-random ones and are
approximate for that design. The two errors can point in opposite directions
and hide each other.

They do. On a randomly ordered frame the correlation component is already
displaced, 1.13 of the truth against 1.02 for simple random selection, while
the reported variances look alike. Order the frame by a trend and the
correlation component is 12% **anti**-conservative while the complete
procedure is 29% **conservative**, the marginal approximation having
overcompensated in the other direction. Give the outcome a period equal to the
sampling interval and the correlation is under-estimated by 0.11 with the two
waves almost perfectly correlated, yet the reported variance is 1.04 of the
truth at 96% coverage.

That last number looks healthy and is healthy only by cancellation, and the
cancellation is enormous. Measured on the same fixture without any wave
structure at all, a single systematic sample of a periodic frame has a true
variance about **1,700 times** the one `as_svydesign()` reports, and its 95%
intervals cover **10%** of the time. The change estimate looks fine because
that error and the correlation error run opposite ways and very nearly cancel.

So a reported number that looks right is not evidence that either component
is, and the marginal variances of a systematic master deserve their own
attention: see `?as_svydesign`.

None of these routes has been validated here for multiple cohorts,
nonresponse, calibration, or unequal-probability masters.

#### A worked route

Form B estimates the correlation from the common sample and multiplies by the
overlap, which is the factor the classical shortcut omits. The marginal
variances come from `samplyr`; only the correlation needs fitting, and that is
one call to `lm()`.

**This is the unstratified form**, which is what the master below uses. A
stratified master needs the stratified specification of @bergerpriam2016
[§4.1]: wave-specific stratum indicators and their interactions in place of
the bare wave indicators. Form B additionally requires the common-sample
fraction to be **the same** in every stratum, since it applies a single factor
`g`, and unequal quotas can break that.

```{r}
w1 <- execute(equal_master, wave = 1)
w2 <- execute(equal_master, wave = 2)
w1$measure <- w1$revenue_millions
w2$measure <- w2$revenue_millions * 1.03

tab <- stack_waves(w1, w2)

# Marginal variances: samplyr's per-wave export.
v1 <- survey::SE(survey::svytotal(~measure, as_svydesign(w1)))^2
v2 <- survey::SE(survey::svytotal(~measure, as_svydesign(w2)))^2

# The union, with each wave's weighted measure zero-filled outside it.
ids <- union(tab$master_id[tab$wave == 1], tab$master_id[tab$wave == 2])
at <- function(w) match(ids, tab$master_id[tab$wave == w])
y <- function(w) {
  i <- at(w)
  ifelse(is.na(i), 0,
         (tab$measure * tab$design_weight)[tab$wave == w][i])
}
common <- !is.na(at(1)) & !is.na(at(2))
g <- sum(common) / sum(!is.na(at(1)))

# Form B: the common-sample residual correlation, times g. The model is the
# paper's own -- wave indicators, their interaction, and no intercept.
z1 <- as.integer(!is.na(at(1)))
z2 <- as.integer(!is.na(at(2)))
fit <- stats::lm(cbind(ifelse(common, y(1), 0), ifelse(common, y(2), 0)) ~
                   -1 + z1 * z2)
V <- stats::estVar(fit)
rho <- V[1, 2] / sqrt(V[1, 1] * V[2, 2]) * g

c(overlap = g, correlation = rho,
  se_change = sqrt(v1 + v2 - 2 * rho * sqrt(v1 * v2)),
  se_if_independent = sqrt(v1 + v2))
```

The two standard errors differ by the covariance, which is the quantity two
separate exports cannot supply.

The measure here is perfectly correlated across the two waves, and the
estimator correlation comes out at the overlap itself. Do not read that as a
rule. The overlap is the main driver, which is why a rotation keeping half its
sample will not usually produce an estimator correlation much above a half
however stable the measure is, and why reaching the amplified region takes a
**high-overlap** design rather than merely a persistent one. But the two waves
also share the master's own selection variation, so the correlation can sit
**above** the overlap times the within-unit correlation: with 7% overlap and a
near-perfectly correlated measure it was measured at 0.15, not 0.07. Compute
it rather than predicting it.

### Attaching later waves to the sample

A `tbl_sample` records the selection. Joining later-wave responses one to one,
in wide form, preserves that: one row per selected unit, design columns
untouched. Reshaping to one row per unit-wave does not. Duplicated unit-wave
rows are repeated measurements of one selected unit, not independently sampled
observations, and exporting them with `as_svydesign()` would count each unit
once per interview. Export the wide form, or reshape after export.

### Limits to state before fieldwork

**Transient stratifiers.** `sampling_design()` stores stratification variables
as names and resolves them at execution, so re-executing a design against a
later frame vintage re-classifies units. For a repeated panel that is correct,
since each panel represents the population as of its own time. For a fixed
panel it is wrong: a unit must keep its wave-1 stratum. Unequal sampling
fractions across a classification that units drift out of can improve
precision initially and reduce it after several waves [@smith2009, sec.
2.4.4].

**The sampling unit and the followed unit may differ, deliberately.** The
standard device for avoiding movers is a sampling unit that cannot move: a
labour force survey that samples addresses balances movers out against movers
in. Such a design is balanced for cross-sectional estimates and for *net*
change, but not for *gross* change [@smith2009, sec. 2.3.5]. `cluster_by()`
expresses it and needs nothing new. The caveat matters because a
dwelling-based rotating panel quietly fails the analysis most people want a
panel for. It also constrains the frame: an address found vacant at one wave
stays in the sample for later waves, so the frame must not be filtered to
occupied units between waves.

**Entrant coverage is a property of the frame, not of the draw.** `samplyr`
can draw from a supplied entrant register. It cannot establish that the
register covers the entrant flow, or that it covers it promptly enough for
short-lived entrants to be represented at all.

**Sampling that depends on the existing sample is out of scope.** Adding
household members reached through an existing sample member gives selection
probabilities that cannot be known from the design, and model-based approaches
with untestable assumptions are required [@lynn2012, sec. 8.1]. Attrition,
measurement error and imputation are likewise estimation-layer problems:
`samplyr` produces design weights only.

## Multi-Wave Rotation

In practice, surveys run for many waves. The PRN update is applied after each wave, rotating units through the sample.

```{r}
set.seed(2026)

frame <- ken_enterprises |>
  mutate(u = runif(n()),
         pik = inclusion_prob(revenue_millions, n_wave))

n_waves <- 5
waves <- vector("list", n_waves)

for (t in seq_len(n_waves)) {
  waves[[t]] <- sampling_design() |>
    draw(n = n_wave, method = "pps_sps", mos = revenue_millions, prn = u) |>
    execute(frame, seed = 1)

  # Update PRNs for next wave
  frame <- frame |>
    mutate(u = (u - pik) %% 1)
}
```

### Pairwise overlap matrix

```{r}
ids <- lapply(waves, `[[`, "enterprise_id")

overlap_mat <- outer(ids, ids, Vectorize(\(a, b) length(intersect(a, b))))
dimnames(overlap_mat) <- rep(list(paste("Wave", seq_len(n_waves))), 2)

overlap_mat
```

Consecutive waves have low overlap around the anticipated lower-bound
benchmark. Non-consecutive waves show varying overlap as the rotation cycle
progresses.

### Cumulative coverage

After $T$ waves, how many distinct enterprises have been surveyed?

```{r}
cumulative <- integer(n_waves)
seen <- character(0)

for (t in seq_len(n_waves)) {
  seen <- union(seen, waves[[t]]$enterprise_id)
  cumulative[t] <- length(seen)
}

tibble(wave = seq_len(n_waves),
       new_units = c(n_wave, diff(cumulative)),
       cumulative = cumulative,
       pct_frame = paste0(round(100 * cumulative / nrow(ken_enterprises), 1), "%"))
```

Negative coordination aims to maximize the number of distinct units covered,
which is valuable for building comprehensive registers or spreading respondent
burden.

## Handling Births and Deaths

Real business registers change between waves. New enterprises (**births**) enter the frame; closed enterprises (**deaths**) leave it.

- **Deaths**: Simply remove them from the frame. Their PRN values are
  discarded. No special treatment needed.
- **Births**: Assign fresh $U(0,1)$ values. These new units enter the
  coordination cycle at a random position.

```{r}
# Simulate: remove 200 enterprises (deaths), add 500 new ones (births)
set.seed(123)
n_births <- 500

# Deaths: remove a random subset
deaths <- sample(frame$enterprise_id, 200)
frame_t2 <- frame |>
  filter(!enterprise_id %in% deaths)

# Births: new enterprises with fresh PRNs
# Sample complete donor records so county-region pairs and business attributes
# remain internally coherent.
births <- ken_enterprises |>
  slice_sample(n = n_births, replace = TRUE) |>
  mutate(
    enterprise_id = paste0("NEW_", sprintf("%05d", seq_len(n_births))),
    u = runif(n_births)
  )

frame_t2 <- bind_rows(frame_t2, births)

# Recompute inclusion probabilities on the new frame
frame_t2 <- frame_t2 |>
  mutate(pik = inclusion_prob(revenue_millions, n_wave))

# Draw wave from updated frame
wave_t2 <- sampling_design() |>
  draw(n = n_wave, method = "pps_sps", mos = revenue_millions, prn = u) |>
  execute(frame_t2, seed = 1)

# How many births were selected?
n_births_selected <- sum(wave_t2$enterprise_id %in% births$enterprise_id)
n_births_selected
n_births_selected / n_births
```

The PRN mechanism carries coordination across this revision: surviving units keep their updated values and births enter with fresh ones. That is the whole of what it does. It does not establish that the revised frame covers the entrant flow, that an identifier still refers to the same entity it did last wave, or that PRN values survived whatever process rebuilt the register. Those are properties of frame construction, and a coordinated draw against a re-keyed or incomplete frame is coordinated with respect to the wrong population. `validate_frame()` checks that the columns the design needs are present and usable; continuity of the register is yours to verify.

## Using Pareto Sampling

The `pps_pareto` method also supports PRN-based coordination. In theory, Pareto sampling has maximum entropy among fixed-size PPS designs that use PRNs, making it particularly well-suited for coordination.

```{r}
set.seed(12345)

frame <- ken_enterprises |>
  mutate(u = runif(n()),
         pik = inclusion_prob(revenue_millions, n_wave))

wave1_par <- sampling_design() |>
  draw(n = n_wave, method = "pps_pareto", mos = revenue_millions, prn = u) |>
  execute(frame, seed = 1)

frame <- frame |> mutate(u = (u - pik) %% 1)

wave2_par <- sampling_design() |>
  draw(n = n_wave, method = "pps_pareto", mos = revenue_millions, prn = u) |>
  execute(frame, seed = 1)

overlap_par <- length(intersect(wave1_par$enterprise_id,
                                wave2_par$enterprise_id))

overlap_par
alb <- round(sum(pmax(0, 2 * frame$pik - 1)))
alb
```

## Stratified Coordination

When the design is stratified, inclusion probabilities must be computed **within each stratum**. The PRN update is then applied globally. The per-stratum $\pi_k$ values ensure correct rotation within each group.

```{r}
set.seed(1980)

frame <- ken_enterprises |>
  mutate(u = runif(n()))

# Compute within-stratum pik
frame <- frame |>
  mutate(pik = inclusion_prob(revenue_millions, 50),
         .by = size_class)

max_pik <- max(frame$pik)
max_pik
alb <- round(sum(pmax(0, 2 * frame$pik - 1)))
alb

wave1_str <- sampling_design() |>
  stratify_by(size_class) |>
  draw(n = 50, method = "pps_pareto", mos = revenue_millions, prn = u) |>
  execute(frame, seed = 1)

frame <- frame |> mutate(u = (u - pik) %% 1)

wave2_str <- sampling_design() |>
  stratify_by(size_class) |>
  draw(n = 50, method = "pps_pareto", mos = revenue_millions, prn = u) |>
  execute(frame, seed = 1)

overlap_str <- length(intersect(wave1_str$enterprise_id,
                                wave2_str$enterprise_id))
overlap_str
```

With smaller within-stratum $\pi_k$ values (all below 0.5), the ALB is zero:
no overlap is forced by certainty or near-certainty units. This is an expected
lower bound, however, not a guarantee that one randomized pair of samples will
have zero realized overlap.

## Collocated Sampling: Coordinating Across Surveys

The examples above coordinate samples from the **same survey** across time (waves). A related problem is coordinating **different surveys** drawn from the same frame at the same time point. This is called *collocated sampling* [@ernst1986; @ohlsson1995] and is common in business survey programs where a single register serves multiple survey operations.

The mechanism is identical: all surveys share the same PRN column.

- **Positive coordination** (maximize overlap): both surveys use the same PRN. Useful when surveys share common modules or when joint analysis is planned.
- **Negative coordination** (minimize overlap): one survey uses shifted PRNs ($u' = (u + 0.5) \bmod 1$). Useful when the goal is to spread respondent burden across the register.

### Example: Turnover survey and employment survey

Suppose a statistical agency runs two surveys on the same business register:
a revenue survey ($n = 200$, PPS on `revenue_millions`) and an employment survey ($n = 150$, PPS on `employees`). The surveys have different sample sizes, different measures of size, and different stratification, but they share the same frame and the same PRN column.

```{r}
set.seed(123)

frame <- ken_enterprises |>
  mutate(u = runif(n()))

# Survey A: revenue survey (PPS on revenue_millions)
survey_a <- sampling_design() |>
  draw(n = 200, method = "pps_sps", mos = revenue_millions, prn = u) |>
  execute(frame, seed = 1)

# Survey B: employment survey (PPS on employees), positive coordination
survey_b <- sampling_design() |>
  draw(n = 150, method = "pps_sps", mos = employees, prn = u) |>
  execute(frame, seed = 1)

overlap_pos <- length(intersect(survey_a$enterprise_id,
                                survey_b$enterprise_id))
overlap_pos
```

For negative coordination, shift the PRN by 0.5:

```{r}
# Survey B with negative coordination
frame_neg <- frame |>
  mutate(u_neg = (u + 0.5) %% 1)

survey_b_neg <- sampling_design() |>
  draw(n = 150, method = "pps_sps", mos = employees, prn = u_neg) |>
  execute(frame_neg, seed = 1)

overlap_neg <- length(intersect(survey_a$enterprise_id,
                                survey_b_neg$enterprise_id))
overlap_neg
```

Compare with independent sampling (fresh PRNs):

```{r}
frame_ind <- frame |>
  mutate(u_fresh = runif(n()))

survey_b_ind <- sampling_design() |>
  draw(n = 150, method = "pps_sps", mos = employees, prn = u_fresh) |>
  execute(frame_ind, seed = 1)

overlap_ind <- length(intersect(survey_a$enterprise_id,
                                survey_b_ind$enterprise_id))
overlap_ind
```

### Overlap bounds for different surveys

When two surveys have different inclusion probabilities $\pi_k^A$ and $\pi_k^B$, the bounds generalize naturally:

```{r}
pik_a <- inclusion_prob(frame$revenue_millions, 200)
pik_b <- inclusion_prob(frame$employees, 150)

alb_ab <- sum(pmax(0, pik_a + pik_b - 1))
aub_ab <- sum(pmin(pik_a, pik_b))
exp_ind <- sum(pik_a * pik_b)

tibble(strategy = c("Positive (same PRN)", "Independent (fresh PRN)", "Negative (shifted PRN)"),
       overlap = c(overlap_pos, overlap_ind, overlap_neg),
       bound = c(round(aub_ab), round(exp_ind), round(alb_ab)))
```

Because `revenue_millions` and `employees` are correlated (large firms have both high revenue and many employees), the overlap under positive coordination
is substantial. Negative coordination via PRN shift effectively spreads the burden.

### Scaling to more than two surveys

For $K$ surveys, shift each survey's PRN by $j/K$ for $j = 0, \ldots, K-1$:

```{r, eval = FALSE}
# Negative coordination across 4 surveys
n_surveys <- 4
for (j in seq_len(n_surveys)) {
  frame_j <- frame |>
    mutate(u_j = (u + (j - 1) / n_surveys) %% 1)

  samples[[j]] <- sampling_design() |>
    draw(n = n_j, method = "pps_sps", mos = mos_j, prn = u_j) |>
    execute(frame_j, seed = 1)
}
```

This simple shift strategy works well when the surveys have similar $\pi_k$. For more complex programs (varying burden weights, priority orderings, or dozens of surveys), NSOs use dedicated coordination systems such as the Rivière method [@riviere2001], the Netherlands EDS [@deree1983; @vanhuis1994], or the Swiss method [@graf2014]. These methods track cumulative burden per unit and permute PRNs within microstrata to achieve near-optimal coordination across the full survey program. See @tille2020 [§8.2.7--8.2.9] for a detailed treatment.

## Summary

The PRN coordination workflow in `samplyr` is:

1. **Assign PRNs** once: `mutate(u = runif(n()))`
2. **Compute $\pi_k$** on the full frame: `mutate(pik = inclusion_prob(mos, n))`
3. **Draw the sample**: `sampling_design() |> draw(method = "pps_sps", prn = u) |> execute(frame)`
4. **Update or shift PRNs**:
   - Temporal negative coordination (waves): `mutate(u = (u - pik) %% 1)`
   - Temporal positive coordination: keep `u` unchanged
   - Cross-survey negative coordination: shift by `(u + 0.5) %% 1`
   - Cross-survey positive coordination: reuse same `u`

| Goal            | PRN update            | Expected overlap                      |
|-----------------|-----------------------|---------------------------------------|
| Maximum overlap | None (reuse)          | $\sum \min(\pi_k^A, \pi_k^B)$         |
| Independent     | Fresh $U(0,1)$        | $\sum \pi_k^A \cdot \pi_k^B$          |
| Minimum overlap | Shift or Tillé update | $\sum \max(0, \pi_k^A + \pi_k^B - 1)$ |

## Reference
