---
title: "Sampling Coordination with Permanent Random Numbers"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Sampling Coordination with Permanent Random Numbers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, echo = FALSE}
library(samplyr)
library(sondage)
library(dplyr)
```

## Why Coordinate Samples?

Repeated surveys (business panels, labor force surveys, agricultural censuses) draw fresh samples at regular intervals from the same population.
Similarly, multiple survey programs often sample from the same register simultaneously. Two natural questions arise:

- **Positive coordination**: How can we maximize overlap between samples to reduce respondent burden and improve change estimates?
- **Negative coordination**: How can we minimize overlap to spread the burden and maximize cumulative coverage?

These questions apply both *temporally* (across waves of the same survey) and *cross-sectionally* (across different surveys on the same frame, known as **collocated sampling**).

The **permanent random number** (PRN) technique, introduced by @brewer1972 and developed by @ohlsson1995 and @cotton1992, answers both questions with a single mechanism: attach a stable $U(0,1)$ value to each frame unit and update it between waves.

`samplyr` supports PRN-based coordination for four selection methods: `bernoulli`, `pps_poisson`, `pps_sps`, and `pps_pareto`.

Coordination controls *which units* successive samples share. It says nothing
about how a survey that revisits its units is designed, scheduled or analysed
over time; that is `vignette("rotating-panels")`, which covers the five
longitudinal designs, `execute(panels = )` and `execute(wave = )`,
`rotation_program()`, and estimating change across waves. The two mechanisms
compose and are otherwise independent.

## Setup

We use `ken_enterprises`, a synthetic business register with 17,004 enterprises.

```{r}
library(samplyr)
library(sondage)
library(dplyr)

data(ken_enterprises)
glimpse(ken_enterprises)
```

Each enterprise has a `revenue_millions` column that serves as our measure of size for PPS sampling. We target 200 enterprises per wave.

```{r}
n_wave <- 200
```

### Assigning permanent random numbers

Every frame unit gets a $U(0,1)$ value drawn once and stored permanently in the register. These values are reused or updated across waves.

```{r}
set.seed(1)

frame <- ken_enterprises |>
  mutate(u = runif(n()))
```

## Positive Coordination: Maximum Overlap

Begin with the simplest design that accepts a PRN. Under `bernoulli` a unit is
selected exactly when its permanent random number falls below the sampling
fraction, and nothing else enters the rule. Two waves that reuse the same `u` at
the same fraction therefore select the same units, whichever seed each is given.

```{r}
f <- n_wave / nrow(frame)

bern_a <- sampling_design() |>
  draw(frac = f, method = "bernoulli", prn = u) |>
  execute(frame, seed = 42)

bern_b <- sampling_design() |>
  draw(frac = f, method = "bernoulli", prn = u) |>
  execute(frame, seed = 123)

c(wave1 = nrow(bern_a), wave2 = nrow(bern_b),
  overlap = length(intersect(bern_a$enterprise_id, bern_b$enterprise_id)))
```

Perfect overlap. This is what a survey wants when the goal is to track the same
units over time, for example measuring quarterly changes in revenue within a
fixed panel. Note that the waves are not exactly 200 units: `bernoulli` fixes
each unit's chance of selection and lets the realized size vary. That is the
trade the fixed-size methods reverse, and we return to it below.

## Negative Coordination: Minimum Overlap

To minimize overlap, subtract each unit's inclusion probability from its PRN
between waves and wrap the result back into $(0,1)$ [@tille2020, sec. 8.2.4]:

$$u_k^{(t+1)} = \bigl(u_k^{(t)} - \pi_k\bigr) \bmod 1$$

Under `bernoulli` this rule is exactly true, and the reason is worth seeing
once. A unit is selected when $u_k < \pi_k$ and not otherwise, so the selected
units are precisely those occupying the interval $[0, \pi_k)$. Subtracting
$\pi_k$ moves that interval to the top of $(0,1)$ and slides everyone else down
by the same amount. A unit selected at wave $t$ therefore cannot be selected at
wave $t+1$ unless its two probabilities together exceed 1. Tillé derives the
update this way and concludes that the resulting negative coordination is
optimal [@tille2020, sec. 8.2.4].

Here every unit shares the same probability `f`, far below 0.5, so the two waves
come out disjoint.

```{r}
bern1 <- sampling_design() |>
  draw(frac = f, method = "bernoulli", prn = u) |>
  execute(frame, seed = 1)

frame_bern <- frame |> mutate(u = (u - f) %% 1)

bern2 <- sampling_design() |>
  draw(frac = f, method = "bernoulli", prn = u) |>
  execute(frame_bern, seed = 1)

c(wave1 = nrow(bern1), wave2 = nrow(bern2),
  overlap = length(intersect(bern1$enterprise_id, bern2$enterprise_id)))
```

That is the whole mechanism. Everything below keeps this update and changes the
selection method underneath it.

## Unequal Probabilities and a Fixed Sample Size

A business register is skewed, and a survey of one usually wants large
enterprises selected more often, plus a sample size it can staff. Each of those
two wishes costs something the Bernoulli story had for free.

Selecting with probability proportional to size gives every unit its own
first-order inclusion probability $\pi_k$. `sondage::inclusion_prob()` computes
these from the measure of size and the sample size, on the **full frame** rather
than on a sample.

```{r}
frame <- frame |>
  mutate(pik = inclusion_prob(revenue_millions, n_wave))

round(quantile(frame$pik), 4)
```

What `pik` means depends on the method it is handed to. Under `bernoulli` and
`pps_poisson` a unit is selected exactly when $u_k < \pi_k$, as above, so
$\pi_k$ is its inclusion probability. `pps_sps` and `pps_pareto` are order
sampling designs [@rosen1997a; @ohlsson1998]: they fix the sample size, and they
pay for it in the marginals. There $\pi_k$ is the **target** inclusion
probability supplied to the algorithm, and Tillé states the consequence plainly
-- "the sample size is fixed but the inclusion probabilities are not exactly
satisfied" [@tille2020, sec. 5.11]. The realized probability approaches the
target as $n$ and $N$ grow, and `samplyr` records the distinction by placing both
methods in the `"approximate"` probabilities tier.

Every bound and benchmark computed from `pik` below inherits that. For
`bernoulli` and `pps_poisson` they are exact; for `pps_sps` and `pps_pareto` they
are approximations whose error is the gap between target and realized
probability.

The other route keeps the probabilities exact and gives up the fixed size.
`pps_poisson` keeps $\pi_i$ dependent only on unit $i$ and the frame total, which
is what makes coordination stable across waves.

The cost is that a unit large enough to reach $\pi_i > 1$ has its chance clipped
to 1, and the expected sample size falls below the target. On a business
register this is the ordinary case, since revenue is heavily skewed.
`execute()` reports it with a warning of class
`samplyr_warning_poisson_shortfall`.

Handle the dominant units explicitly with `certainty_size` or `certainty_prop`
if the shortfall is unacceptable, but note the trade. Declaring units certain
re-resolves the remaining ones over a smaller MOS total, so their $\pi_i$ rise
and now depend on which units were declared certain. Accepting the shortfall
instead is what keeps $\pi_i$ a function of unit $i$ and the frame total alone.
See `?draw`.

The examples that follow use `pps_sps`, sequential Poisson sampling, which is the
fixed-size choice.

### Reusing PRNs under a fixed-size method

Positive coordination carries over unchanged: two waves drawn from the same frame
with identical PRNs produce identical samples.

```{r}
wave1 <- sampling_design() |>
  draw(n = n_wave, method = "pps_sps", mos = revenue_millions, prn = u) |>
  execute(frame, seed = 42)

wave2 <- sampling_design() |>
  draw(n = n_wave, method = "pps_sps", mos = revenue_millions, prn = u) |>
  execute(frame, seed = 123)

overlap_positive <- length(intersect(wave1$enterprise_id,
                                     wave2$enterprise_id))
overlap_positive
```

Overlap is harder to keep when the *design* changes rather than the frame. Reusing PRNs preserves overlap only while strata and inclusion probabilities stay put; a redesign moves both, and maximizing the retained sample under new strata and new probabilities is a separate problem with its own literature [@kishscott1971; @hidiroglou1991]. @tille2020 [§8.2.5] gives a textbook treatment of the Kish and Scott substitution method.

### The same update under a fixed-size method

`pps_sps` orders units by $u_k / \pi_k$ and takes the first $n$, so the shift no
longer reproduces the Bernoulli argument: the key depends on the random number
and the target together, and a large unit with a small key can survive a shift
that would have pushed a Bernoulli unit clear. The update remains the standard
construction for order sampling [@ohlsson1995], and it works well, but its
guarantee is Bernoulli's and not `pps_sps`'s.

```{r}
# Wave 1
wave1 <- sampling_design() |>
  draw(n = n_wave, method = "pps_sps", mos = revenue_millions, prn = u) |>
  execute(frame, seed = 1)

# Update PRNs on the full frame
frame <- frame |>
  mutate(u = (u - pik) %% 1)

# Wave 2
wave2 <- sampling_design() |>
  draw(n = n_wave, method = "pps_sps", mos = revenue_millions, prn = u) |>
  execute(frame, seed = 1)

overlap_negative <- length(intersect(wave1$enterprise_id, wave2$enterprise_id))
overlap_negative
```

## Why isn't the overlap zero?

Not all overlap can be eliminated. Units with large $\pi_k$ (big enterprises)
are selected in almost every wave because their inclusion probability is close
to or equal to 1. The **Absolute Lower Bound** (ALB) on the *expected* overlap
is

$$\text{ALB} = \sum_{k=1}^{N} \max(0,\; 2\pi_k - 1)$$

and the Absolute Upper Bound is $\text{AUB} = \sum_k \min(\pi_k^{(1)},
\pi_k^{(2)})$, here $n$ [@tille2020, sec. 8.2.3]. Only units with $\pi_k > 0.5$
contribute to forced overlap. Tillé attaches a caveat that explains the output
below: except in some very special cases, like simple random sampling, ALB and
AUB cannot be reached.

```{r}
alb <- sum(pmax(0, 2 * frame$pik - 1))
expected_independent <- sum(frame$pik^2)

tibble(quantity  = c("ALB, absolute lower bound",
                     "Expected overlap if independent",
                     "AUB, absolute upper bound"),
       benchmark = c(round(alb), round(expected_independent), n_wave))
```

The realized overlap of `r overlap_negative` is far below what independent
sampling would produce, and it is also a little below the ALB. That is not a
contradiction, and the three reasons are worth separating. The benchmarks bound
an *expectation* over repeated coordinated draws, so a single realized overlap
can fall on either side. The bounds are not targets the method aims at and are
generally not attainable in either direction. And because `pps_sps` is order
sampling, the $\pi_k$ these benchmarks are built from are targets rather than
realized probabilities, so the benchmarks themselves are approximate.

## Comparison: Three Strategies

```{r}
# Independent sampling: fresh PRNs for wave 2
set.seed(1)
frame_ind <- ken_enterprises |>
  mutate(u1 = runif(n()),
         u2 = runif(n()))

wave1_ind <- sampling_design() |>
  draw(n = n_wave, method = "pps_sps", mos = revenue_millions, prn = u1) |>
  execute(frame_ind, seed = 1)

wave2_ind <- sampling_design() |>
  draw(n = n_wave, method = "pps_sps", mos = revenue_millions, prn = u2) |>
  execute(frame_ind, seed = 1)

overlap_ind <- length(intersect(wave1_ind$enterprise_id,
                                wave2_ind$enterprise_id))

tibble(strategy = c("Positive (same PRN)",
                    "Independent (fresh PRN)",
                    "Negative (Tille update)"),
       realized_overlap = c(overlap_positive, overlap_ind, overlap_negative),
       benchmark = c(n_wave, round(expected_independent), round(alb)),
       benchmark_is = c("AUB", "expectation", "ALB")) |>
  mutate(pct_of_wave = paste0(round(100 * realized_overlap / n_wave), "%"))
```

The `benchmark` column is a property of the $\pi_k$ alone, not something the
method was aiming at: the two bounds hold for *any* coordination scheme on these
probabilities, and only the independent row's benchmark is an expectation the
strategy is supposed to reproduce. Positive coordination attains the AUB here
because reusing the PRNs makes the two samples identical, which is the special
case where a bound is reached.

## Multi-Wave Rotation

In practice, surveys run for many waves. The PRN update is applied after each wave, rotating units through the sample.

```{r}
set.seed(2026)

frame <- ken_enterprises |>
  mutate(u = runif(n()),
         pik = inclusion_prob(revenue_millions, n_wave))

n_waves <- 5
waves <- vector("list", n_waves)

for (t in seq_len(n_waves)) {
  waves[[t]] <- sampling_design() |>
    draw(n = n_wave, method = "pps_sps", mos = revenue_millions, prn = u) |>
    execute(frame, seed = 1)

  # Update PRNs for next wave
  frame <- frame |>
    mutate(u = (u - pik) %% 1)
}
```

### Pairwise overlap matrix

```{r}
ids <- lapply(waves, `[[`, "enterprise_id")

overlap_mat <- outer(ids, ids, Vectorize(\(a, b) length(intersect(a, b))))
dimnames(overlap_mat) <- rep(list(paste("Wave", seq_len(n_waves))), 2)

overlap_mat
```

Consecutive waves have low overlap, near the absolute lower bound. Non-consecutive
waves show varying overlap as the rotation cycle progresses.

### Cumulative coverage

After five waves, how many distinct enterprises have been surveyed?

```{r}
cumulative <- integer(n_waves)
seen <- character(0)

for (t in seq_len(n_waves)) {
  seen <- union(seen, waves[[t]]$enterprise_id)
  cumulative[t] <- length(seen)
}

tibble(wave = seq_len(n_waves),
       new_units = c(n_wave, diff(cumulative)),
       cumulative = cumulative,
       pct_frame = paste0(round(100 * cumulative / nrow(ken_enterprises), 1), "%"))
```

Negative coordination aims to maximize the number of distinct units covered,
which is valuable for building comprehensive registers or spreading respondent
burden.

## Handling Births and Deaths

Real business registers change between waves. New enterprises (**births**) enter the frame; closed enterprises (**deaths**) leave it.

- **Deaths**: Remove them from the frame. Their PRN values are discarded, and
  the next cross-sectional draw needs nothing further. This holds because the
  target here is the population at each wave. When the target is longitudinal
  the same event is attrition, which is an estimation-layer problem rather than
  a frame operation; see `vignette("rotating-panels")`.
- **Births**: Assign fresh $U(0,1)$ values. These new units enter the
  coordination cycle at a random position.

```{r}
# Simulate: remove 200 enterprises (deaths), add 500 new ones (births)
set.seed(123)
n_births <- 500

# Deaths: remove a random subset
deaths <- sample(frame$enterprise_id, 200)
frame_t2 <- frame |>
  filter(!enterprise_id %in% deaths)

# Births: new enterprises with fresh PRNs
# Sample complete donor records so county-region pairs and business attributes
# remain internally coherent.
births <- ken_enterprises |>
  slice_sample(n = n_births, replace = TRUE) |>
  mutate(
    enterprise_id = paste0("NEW_", sprintf("%05d", seq_len(n_births))),
    u = runif(n_births)
  )

frame_t2 <- bind_rows(frame_t2, births)

# Recompute inclusion probabilities on the new frame
frame_t2 <- frame_t2 |>
  mutate(pik = inclusion_prob(revenue_millions, n_wave))

# Draw wave from updated frame
wave_t2 <- sampling_design() |>
  draw(n = n_wave, method = "pps_sps", mos = revenue_millions, prn = u) |>
  execute(frame_t2, seed = 1)

# How many births were selected?
n_births_selected <- sum(wave_t2$enterprise_id %in% births$enterprise_id)
n_births_selected
n_births_selected / n_births
```

The PRN mechanism carries coordination across this revision: surviving units keep their updated values and births enter with fresh ones. That is the whole of what it does. It does not establish that the revised frame covers the entrant flow, that an identifier still refers to the same entity it did last wave, or that PRN values survived whatever process rebuilt the register. Those are properties of frame construction, and a coordinated draw against a re-keyed or incomplete frame is coordinated with respect to the wrong population. `validate_frame()` checks that the columns the design needs are present and usable; continuity of the register is yours to verify.

## Using Pareto Sampling

The `pps_pareto` method also supports PRN-based coordination. Two properties
recommend it here. It reaches a fixed sample size, and within the class of order
designs with fixed shape parameters it minimises the variance estimator
[@rosen1997b; @tille2020, sec. 5.11]. That is a minimum-variance result inside
order sampling, not a maximum-entropy result across fixed-size PPS designs: the
maximum-entropy fixed-size design is conditional Poisson, which `samplyr` ships
as `pps_cps` and which does not take a PRN. Pareto's own targets remain
approximate, as with `pps_sps`.

```{r}
set.seed(12345)

frame <- ken_enterprises |>
  mutate(u = runif(n()),
         pik = inclusion_prob(revenue_millions, n_wave))

wave1_par <- sampling_design() |>
  draw(n = n_wave, method = "pps_pareto", mos = revenue_millions, prn = u) |>
  execute(frame, seed = 1)

frame <- frame |> mutate(u = (u - pik) %% 1)

wave2_par <- sampling_design() |>
  draw(n = n_wave, method = "pps_pareto", mos = revenue_millions, prn = u) |>
  execute(frame, seed = 1)

overlap_par <- length(intersect(wave1_par$enterprise_id,
                                wave2_par$enterprise_id))

overlap_par
alb <- round(sum(pmax(0, 2 * frame$pik - 1)))
alb
```

## Stratified Coordination

When the design is stratified, inclusion probabilities must be computed **within each stratum**. The PRN update is then applied globally. The per-stratum $\pi_k$ values ensure correct rotation within each group.

```{r}
set.seed(1980)

frame <- ken_enterprises |>
  mutate(u = runif(n()))

# Compute within-stratum pik
frame <- frame |>
  mutate(pik = inclusion_prob(revenue_millions, 50),
         .by = size_class)

max_pik <- max(frame$pik)
max_pik
alb <- round(sum(pmax(0, 2 * frame$pik - 1)))
alb

wave1_str <- sampling_design() |>
  stratify_by(size_class) |>
  draw(n = 50, method = "pps_pareto", mos = revenue_millions, prn = u) |>
  execute(frame, seed = 1)

frame <- frame |> mutate(u = (u - pik) %% 1)

wave2_str <- sampling_design() |>
  stratify_by(size_class) |>
  draw(n = 50, method = "pps_pareto", mos = revenue_millions, prn = u) |>
  execute(frame, seed = 1)

overlap_str <- length(intersect(wave1_str$enterprise_id,
                                wave2_str$enterprise_id))
overlap_str
```

With smaller within-stratum $\pi_k$ values (all below 0.5), the ALB is zero:
no overlap is forced by certainty or near-certainty units. It is a lower bound
on the *expected* overlap, not a guarantee that one randomized pair of samples
will have zero realized overlap, and it is computed from `pps_pareto`'s target
probabilities rather than its realized ones.

## Collocated Sampling: Coordinating Across Surveys

The examples above coordinate samples from the **same survey** across time (waves). A related problem is coordinating **different surveys** drawn from the same frame at the same time point. This is called *collocated sampling* [@ernst1986; @ohlsson1995] and is common in business survey programs where a single register serves multiple survey operations.

The mechanism is identical: all surveys share the same PRN column.

- **Positive coordination** (maximize overlap): both surveys use the same PRN. Useful when surveys share common modules or when joint analysis is planned.
- **Negative coordination** (minimize overlap): one survey uses shifted PRNs, `(u + 0.5) %% 1`. Useful when the goal is to spread respondent burden across the register.

The half-turn shift is a construction, not a rule the design guarantees. Under
`bernoulli` it is exact for equal probabilities: the two surveys select from
$[0, \pi^A)$ and from an interval starting at 0.5, which are disjoint whenever
$\pi^A + \pi^B \le 1$. Under order sampling with different measures of size it
is a heuristic that works well when the two surveys are of similar size and
their $\pi_k$ are similar, and degrades as they diverge.

### Example: Turnover survey and employment survey

Suppose a statistical agency runs two surveys on the same business register:
a revenue survey of 200 units, PPS on `revenue_millions`, and an employment survey of 150 units, PPS on `employees`. The surveys have different sample sizes, different measures of size, and different stratification, but they share the same frame and the same PRN column.

```{r}
set.seed(123)

frame <- ken_enterprises |>
  mutate(u = runif(n()))

# Survey A: revenue survey (PPS on revenue_millions)
survey_a <- sampling_design() |>
  draw(n = 200, method = "pps_sps", mos = revenue_millions, prn = u) |>
  execute(frame, seed = 1)

# Survey B: employment survey (PPS on employees), positive coordination
survey_b <- sampling_design() |>
  draw(n = 150, method = "pps_sps", mos = employees, prn = u) |>
  execute(frame, seed = 1)

overlap_pos <- length(intersect(survey_a$enterprise_id,
                                survey_b$enterprise_id))
overlap_pos
```

For negative coordination, shift the PRN by 0.5.

```{r}
# Survey B with negative coordination
frame_neg <- frame |>
  mutate(u_neg = (u + 0.5) %% 1)

survey_b_neg <- sampling_design() |>
  draw(n = 150, method = "pps_sps", mos = employees, prn = u_neg) |>
  execute(frame_neg, seed = 1)

overlap_neg <- length(intersect(survey_a$enterprise_id,
                                survey_b_neg$enterprise_id))
overlap_neg
```

Compare with independent sampling, drawn on fresh PRNs.

```{r}
frame_ind <- frame |>
  mutate(u_fresh = runif(n()))

survey_b_ind <- sampling_design() |>
  draw(n = 150, method = "pps_sps", mos = employees, prn = u_fresh) |>
  execute(frame_ind, seed = 1)

overlap_ind <- length(intersect(survey_a$enterprise_id,
                                survey_b_ind$enterprise_id))
overlap_ind
```

### Overlap bounds for different surveys

When two surveys have different inclusion probabilities $\pi_k^A$ and $\pi_k^B$, the bounds generalize naturally.

```{r}
pik_a <- inclusion_prob(frame$revenue_millions, 200)
pik_b <- inclusion_prob(frame$employees, 150)

alb_ab <- sum(pmax(0, pik_a + pik_b - 1))
aub_ab <- sum(pmin(pik_a, pik_b))
exp_ind <- sum(pik_a * pik_b)

tibble(strategy = c("Positive (same PRN)",
                    "Independent (fresh PRN)",
                    "Negative (shifted PRN)"),
       realized_overlap = c(overlap_pos, overlap_ind, overlap_neg),
       benchmark = c(round(aub_ab), round(exp_ind), round(alb_ab)),
       benchmark_is = c("AUB", "expectation", "ALB"))
```

Because `revenue_millions` and `employees` are correlated (large firms have both high revenue and many employees), the overlap under positive coordination
is substantial. Negative coordination via PRN shift effectively spreads the burden.

### Scaling to more than two surveys

Across a programme of `K` surveys, survey `j` shifts its PRN by `j/K`.

```{r, eval = FALSE}
# Negative coordination across 4 surveys
n_surveys <- 4
for (j in seq_len(n_surveys)) {
  frame_j <- frame |>
    mutate(u_j = (u + (j - 1) / n_surveys) %% 1)

  samples[[j]] <- sampling_design() |>
    draw(n = n_j, method = "pps_sps", mos = mos_j, prn = u_j) |>
    execute(frame_j, seed = 1)
}
```

Like the half-turn, the `j/K` shift is a construction whose behaviour depends on
the surveys resembling each other; it carries no optimality claim across a
programme of unequal surveys. For more complex programs (varying burden weights, priority orderings, or dozens of surveys), NSOs use dedicated coordination systems such as the Rivière method [@riviere2001], the Netherlands EDS [@deree1983; @vanhuis1994], or the Swiss method [@graf2014]. These methods track cumulative burden per unit and permute PRNs within microstrata, which lets them coordinate across a whole survey programme rather than a pair of surveys at a time. See @tille2020 [§8.2.7--8.2.9] for a detailed treatment.

## Summary

The PRN coordination workflow in `samplyr` is:

1. **Assign PRNs** once: `mutate(u = runif(n()))`
2. **Compute the inclusion probabilities** on the full frame: `mutate(pik = inclusion_prob(mos, n))`
3. **Draw the sample**: `sampling_design() |> draw(method = "pps_sps", prn = u) |> execute(frame)`
4. **Update or shift PRNs**:
   - Temporal negative coordination (waves): `mutate(u = (u - pik) %% 1)`
   - Temporal positive coordination: keep `u` unchanged
   - Cross-survey negative coordination: shift by `(u + 0.5) %% 1`
   - Cross-survey positive coordination: reuse same `u`

| Goal            | PRN update            | Benchmark for the expected overlap    | What it is  |
|-----------------|-----------------------|---------------------------------------|-------------|
| Maximum overlap | None (reuse)          | $\sum \min(\pi_k^A, \pi_k^B)$         | AUB         |
| Independent     | Fresh $U(0,1)$        | $\sum \pi_k^A \cdot \pi_k^B$          | expectation |
| Minimum overlap | Shift or Tillé update | $\sum \max(0, \pi_k^A + \pi_k^B - 1)$ | ALB         |

The two bounds hold for any coordination scheme on these probabilities and are
generally not attainable; only the middle row is an expectation its strategy
reproduces. Under `pps_sps` and `pps_pareto` all three are computed from target
rather than realized inclusion probabilities.

## References
