---
title: "Coordinate Samples with Permanent Random Numbers"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Coordinate Samples with Permanent Random Numbers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, echo = FALSE}
library(samplyr)
library(sondage)
library(dplyr)
```

## What coordination does

When several samples use the same register, we can deliberately control how
much they overlap. Positive coordination keeps more of the same units. This can
help with estimates of change, but it concentrates repeated response burden on
those units. Negative coordination does the opposite. It spreads response
burden and increases the number of distinct units covered.

Permanent random numbers, or PRNs, provide a practical way to coordinate
samples. Each frame unit receives one random number between zero and one. The
number stays with the unit and is reused or updated for later selections. This
approach has a long history in business surveys [@brewer1972; @ohlsson1995].

This article starts with Bernoulli sampling because its coordination rule is
exact and easy to see. It then shows the fixed-size PPS workflow commonly used
for business surveys. `samplyr` accepts PRNs with `bernoulli`, `pps_poisson`,
`pps_sps`, and `pps_pareto`.

Coordination is not the same as a longitudinal design. A PRN controls overlap
between draws. A longitudinal plan also decides how long units remain in the
sample, how many units enter at each wave, and how change will be estimated.
See `vignette("rotating-panels")` for that workflow.

## Prepare the register

We use the synthetic `ken_enterprises` register and target about 200
enterprises per wave.

```{r, message = FALSE}
library(samplyr)
library(sondage)
library(dplyr)

data(ken_enterprises)

set.seed(1)
frame <- ken_enterprises |>
  mutate(prn = runif(n()))

n_wave <- 200
```

In production, store `prn` in the register rather than recreating it in every
sampling script. A stable identifier and a stable PRN are both needed for
coordination to survive register updates.

## The simplest positive coordination

Under Bernoulli sampling, a unit is selected when its PRN is below the sampling
fraction. Reusing both the PRN and the fraction therefore selects the same
units. The realized sample size can vary around the target.

```{r}
fraction <- n_wave / nrow(frame)

design_bernoulli <- sampling_design() |>
  draw(frac = fraction, method = "bernoulli", prn = prn)

wave_1 <- execute(design_bernoulli, frame, seed = 10)
wave_2 <- execute(design_bernoulli, frame, seed = 20)

tibble(
  size_wave_1 = nrow(wave_1),
  size_wave_2 = nrow(wave_2),
  overlap = length(intersect(wave_1$enterprise_id, wave_2$enterprise_id))
)
```

The seeds differ, but the samples do not. Once a supported method receives a
PRN, selection is determined by the PRN values and the design.

## The simplest negative coordination

For a second Bernoulli wave, subtract the inclusion probability from each PRN
and wrap the result back into the unit interval [@ohlsson1995].

```{r}
frame_wave_2 <- frame |>
  mutate(prn = (prn - fraction) %% 1)

wave_2_negative <- execute(design_bernoulli, frame_wave_2, seed = 20)

tibble(
  size_wave_1 = nrow(wave_1),
  size_wave_2 = nrow(wave_2_negative),
  overlap = length(intersect(
    wave_1$enterprise_id,
    wave_2_negative$enterprise_id
  ))
)
```

Here the two samples are disjoint. That conclusion is exact for this equal-
probability Bernoulli example because twice the sampling fraction is below
one. More generally, high-probability units may have to appear in both samples.

The important operational rule is simple: update PRNs on the full frame, not
only on the selected rows. Otherwise the next draw no longer follows the
intended sampling design.

## A fixed-size PPS sample

Business surveys often need a fixed sample size and want large enterprises to
be selected more often. Sequential Poisson sampling (`pps_sps`) provides that
combination and supports PRNs [@ohlsson1998].

First compute each unit's target inclusion probability on the full frame.

```{r}
frame_pps <- frame |>
  mutate(pik = inclusion_prob(revenue_millions, n_wave))

summary(frame_pps$pik)
```

Then declare the design once and execute it.

```{r}
design_pps <- sampling_design() |>
  draw(
    n = n_wave,
    method = "pps_sps",
    mos = revenue_millions,
    prn = prn
  )

pps_wave_1 <- execute(design_pps, frame_pps, seed = 1)

frame_pps_2 <- frame_pps |>
  mutate(prn = (prn - pik) %% 1)

pps_wave_2 <- execute(design_pps, frame_pps_2, seed = 1)

tibble(
  size_wave_1 = nrow(pps_wave_1),
  size_wave_2 = nrow(pps_wave_2),
  overlap = length(intersect(
    pps_wave_1$enterprise_id,
    pps_wave_2$enterprise_id
  ))
)
```

The update encourages negative coordination, but the exact Bernoulli argument
does not carry over to `pps_sps`. Sequential Poisson sampling orders all units
and fixes the sample size. Its supplied `pik` values are targets rather than
exact first-order inclusion probabilities. The approximation improves for
large populations and samples [@rosen1997a; @ohlsson1998]. The same distinction
applies to `pps_pareto`.

If exact first-order probabilities matter more than a fixed realized size, use
`pps_poisson`. Its sample size is random. Very large measures of size can also
produce certainty units or reduce the expected sample size after probabilities
are capped at one. `execute()` reports that case. See `?draw` for the certainty
options.

## Repeat the update across waves

For a stable frame, a short loop is enough. Each sample uses the current PRN,
then every frame unit receives its next PRN.

```{r}
set.seed(2026)

frame_many <- ken_enterprises |>
  mutate(
    prn = runif(n()),
    pik = inclusion_prob(revenue_millions, n_wave)
  )

n_waves <- 4
waves <- vector("list", n_waves)

for (wave in seq_len(n_waves)) {
  waves[[wave]] <- execute(design_pps, frame_many, seed = 1)
  frame_many <- frame_many |>
    mutate(prn = (prn - pik) %% 1)
}

ids <- lapply(waves, `[[`, "enterprise_id")
overlap <- outer(
  ids,
  ids,
  Vectorize(\(x, y) length(intersect(x, y)))
)
dimnames(overlap) <- rep(list(paste("Wave", seq_len(n_waves))), 2)
overlap
```

The matrix is descriptive. There is no general promise that adjacent fixed-
size PPS samples will be disjoint or that a later pair will have a particular
overlap.

## Stratified samples

Compute `pik` within the same strata used by the design. The PRN update itself
still applies to every row.

```{r}
set.seed(1980)

frame_stratified <- ken_enterprises |>
  mutate(prn = runif(n())) |>
  mutate(
    pik = inclusion_prob(revenue_millions, 50),
    .by = size_class
  )

design_stratified <- sampling_design() |>
  stratify_by(size_class) |>
  draw(
    n = 50,
    method = "pps_sps",
    mos = revenue_millions,
    prn = prn
  )

stratified_wave_1 <- execute(design_stratified, frame_stratified, seed = 1)

frame_stratified_2 <- frame_stratified |>
  mutate(prn = (prn - pik) %% 1)

stratified_wave_2 <- execute(
  design_stratified,
  frame_stratified_2,
  seed = 1
)

length(intersect(
  stratified_wave_1$enterprise_id,
  stratified_wave_2$enterprise_id
))
```

If allocations, strata, or measures of size change, recompute the probabilities
before updating the PRNs. Reusing a PRN still coordinates the draws, but it does
not by itself maximize retention under a redesigned sample [@kishscott1971].

## Births, deaths, and register maintenance

For a cross-sectional survey, surviving units keep their PRNs, births receive
fresh independent PRNs, and units no longer in the target population leave the
frame. Recompute inclusion probabilities after the register changes.

These steps do not solve longitudinal attrition. If the target follows a fixed
cohort, a closed enterprise or a nonrespondent needs an analysis rule as well
as a frame update. PRNs also cannot detect identifier reuse, duplicate units,
or incomplete coverage. Those are register-quality checks.

## Coordination across surveys

Different surveys using the same register can share a PRN column. Reusing the
same PRN tends to increase overlap. A half-turn shift is a simple way to seek
less overlap.

```{r, eval = FALSE}
frame_for_second_survey <- frame |>
  mutate(prn_second = (prn + 0.5) %% 1)
```

For equal-probability Bernoulli samples, separated intervals give an exact
result when their sampling fractions fit without overlap. For fixed-size PPS
samples with different sizes, strata, or size measures, the half-turn is only
a heuristic. Large survey programmes generally need a dedicated coordination
system rather than repeated ad hoc shifts [@ohlsson1995; @riviere2001].

## A practical checklist

1. Add one stable PRN column to the full register.
2. Compute inclusion probabilities on the full selection pool, within strata
   when the design is stratified.
3. Pass the PRN column to a supported method.
4. Keep the PRN for positive coordination, or update it for negative
   coordination.
5. Save the updated full frame before the next wave.
6. Recompute probabilities whenever the frame or design changes.

Start with the Bernoulli example if you are learning the mechanism. Move to
`pps_sps` only when you need both unequal selection probabilities and an exact
realized sample size. Use `vignette("rotating-panels")` when the requirement is
a controlled entry-and-exit schedule rather than simply more or less overlap.

## References
