---
title: "Saving and Sharing Designs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Saving and Sharing Designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, echo = FALSE}
library(samplyr)
library(dplyr)
```

## Overview

A `samplyr` design is a frame-independent specification. It describes *how* to sample, not *what* to sample from. That makes a design a natural artifact to save, version, reuse, and share alongside the survey protocol, in a git repository, or with a collaborator who will execute it on the confidential frame you cannot send.

Three functions cover this workflow:

| Function         | Purpose                                                        |
|:-----------------|:---------------------------------------------------------------|
| `write_design()` | Save a design (or an executed sample's design) to a JSON file  |
| `read_design()`  | Restore a design from a file or JSON string                    |
| `design_json()`  | Render the same format as an in-memory string (databases, APIs) |

The file format is versioned, human-readable JSON and is diffable in version
control, so the sampling plan can be inspected without evaluating R code. It
is currently an experimental, samplyr-native persistence format, not the
finalized cross-tool interchange standard. Its present guarantee is
behavioral: a restored design executes identically to the original in a
compatible samplyr environment.

```{r, eval = FALSE}
execute(read_design("plan.json"), frame, seed = 1)
# is identical to
execute(design, frame, seed = 1)
```

## Saving a design

Any design can be written, from a simple SRS to a multi-stage PPS plan with control sorting:

```{r}
design <- sampling_design(title = "BFA EA Survey") |>
  stratify_by(region, alloc = "proportional") |>
  draw(
    n = 300,
    method = "systematic",
    control = c(serp(province, commune))
  )

path <- tempfile(fileext = ".json")
write_design(design, path)
```

The file is plain JSON:

```{r, comment = ""}
cat(readLines(path), sep = "\n")
```

Everything the design holds is written out in full, including per-stratum allocations, auxiliary data frames for Neyman or optimal allocation, and certainty thresholds. Nothing is lost between sessions:

```{r}
design_named <- sampling_design() |>
  stratify_by(urban_rural) |>
  draw(n = c(Urban = 120, Rural = 180))

cat(design_json(design_named, pretty = TRUE))
```

Scalars are written as JSON numbers, named vectors as objects, and data-frame specifications as arrays of rows. The three shapes are unambiguous, so files can be written or edited by other tools without special type annotations.

## Declarative design and samplyr metadata

The JSON separates declarative design content from implementation metadata.
`design`, `frame`, and `execution` avoid executable R expressions, while
`tools.samplyr` retains exact method names, R classes, frame hashes, and replay
environment details. A samplyr file records both an internal semantic method
descriptor and its exact native spelling:

```json
"method": {
  "id": "generalized_brewer_probability_proportional_to_size",
  "family": "probability_proportional_to_size",
  "algorithm": "generalized_brewer",
  "replacement": "without_replacement",
  "sample_size": "fixed",
  "probabilities": "unequal",
  "standards": [{
    "vocabulary": "DDI SamplingProcedure",
    "version": "1.1.4",
    "code": "Probability"
  }]
}
```

```json
"tools": {
  "samplyr": {
    "design": {
      "stages": [{"method": {"name": "pps_brewer"}}]
    }
  }
}
```

The internal vocabulary distinguishes the built-in algorithms and links them
to the broader [DDI Sampling Procedure controlled vocabulary](https://rdf-vocabulary.ddialliance.org/ddi-cv/SamplingProcedure/1.1.4/SamplingProcedure.html).
It is useful for inspecting samplyr files, but it is not presented as the
finalized external method vocabulary. `read_design()` can restore a recognized
internal method without a `tools.samplyr` block. Custom methods use the
identifier `tool_specific`; the same method must be registered through
`sondage::register_method()` before replay. The file records a
fingerprint of the registered implementation (formals and body of
`sample_fn` and `joint_fn`), and replay refuses a function that
differs from the recorded one even when its registry metadata matches.

The complete, machine-readable dictionary ships as `schema/sampling-methods-v1.json` inside the installed package. Its identifier and version are declared in each document's `schema.method_vocabulary` block.

Unknown tool namespaces are retained when a design is read and written again,
but preservation alone is not a claim of cross-tool design equivalence.

## Reading a design back

`read_design()` restores a `sampling_design` that behaves exactly like the original:

```{r}
restored <- read_design(path)
restored
```

```{r}
s_original <- execute(design, bfa_eas, seed = 123)
s_restored <- execute(restored, bfa_eas, seed = 123)

identical(s_original$ea_id, s_restored$ea_id)
identical(s_original$.weight, s_restored$.weight)
```

`read_design()` also accepts a JSON string directly, which pairs with `design_json()` when designs are stored in a database or sent over an API rather than as files:

```{r}
json <- design_json(design)
identical(
  execute(read_design(json), bfa_eas, seed = 7)$ea_id,
  execute(design, bfa_eas, seed = 7)$ea_id
)
```

## Frame requirements and fingerprints

We made the design files frame-independent, so the frame data is never written. Two derived blocks describe the frame instead.

The **requirements** block is always recorded. It lists the columns each stage needs, tagged by role (stratification, clustering, measure of size, PRN, auxiliary, and control sorting), so that a candidate frame can be checked before execution:

```{r}
info <- attr(restored, "frame_info")

bind_rows(info$required_variables)
```

This complements `validate_frame()`, which performs the full check against an actual frame:

```{r}
validate_frame(restored, bfa_eas)
```

The **fingerprint** block is written when you pass the frame to `write_design()`. Its common part records dimensions and portable column types such as `integer`, `categorical`, and `date`. R-specific details (the expression used for the frame, native R classes, and the `rlang::hash()` content hash) are isolated in `tools.samplyr.frame`. Together they are enough for samplyr to verify the exact frame without shipping any data, while another tool can still use the structural metadata:

```{r}
write_design(design, path, frame = bfa_eas)

fingerprint <- attr(read_design(path), "frame_info")$fingerprint
fingerprint$name
fingerprint$nrow
fingerprint$hash
```

The hash covers the column names, the column values, and the row order. It does not depend on whether the frame is a tibble or a plain data frame, or on the order of its columns. Frames that differ only in those respects are the same population.

To verify a frame months later, pass it to `validate_frame()` together with the restored design. The check is silent when the frame matches the stored fingerprint. If the frame has been revised since the design was written, it explains what changed and still runs the variable checks, because a design remains executable on any compatible frame:

```{r}
validate_frame(read_design(path), bfa_eas[-(1:10), ])
```

Set `fingerprint = "warn"` to escalate the message to a warning, or `"ignore"` to skip the comparison entirely.

## Execution receipts

Passing an executed `tbl_sample` to `write_design()` saves the design together
with an execution receipt: every `execute()` argument that affects the result
(the seed, executed stages, and `panels` or `reps` when used), plus the number
of selected units and execution timestamp. `tools.samplyr.execution` records
the execution-time R and package versions and the RNG configuration.

```{r}
sample <- execute(design, bfa_eas, seed = 2026, panels = 4)

write_design(sample, path, frame = bfa_eas)

receipt <- attr(read_design(path), "execution")
receipt$seed
unlist(receipt$stages_executed)
receipt$panels
receipt$n_selected
```

Together with the frame fingerprint, the receipt makes the sample exactly reproducible. `replay_design()` passes the recorded arguments back to `execute()`, so nothing has to be reconstructed by hand:

```{r}
replayed <- replay_design(read_design(path), bfa_eas)
```

The contract is full equality of the materialized sample when the same frame,
compatible package implementations, and any required registered custom
methods are available. `replay_design()` restores the recorded RNG
configuration for execution and then restores the caller's RNG state. The
replayed `tbl_sample` has the same rows in the same order, the same weights and
design columns, and the same `.panel` and `.replicate` assignments; only the
execution timestamp differs.

```{r}
identical(replayed$ea_id, sample$ea_id)
identical(replayed$.weight, sample$.weight)
identical(replayed$.panel, sample$.panel)
```

`replay_design()` first compares the frame against the stored fingerprint and
errors by default if it differs. Use `fingerprint = "warn"`, `"inform"`, or
`"ignore"` only when you intentionally accept that the result will not be the
recorded sample. It also warns when the recorded R or package versions differ
and checks the replayed row count against the receipt.

Two situations fall outside the single-call contract, and `write_design()` warns about both. A sample built by several `execute()` calls (a stage continuation or a multi-phase pipeline) is flagged as `chained` in its receipt: the receipt records only the final call, so `replay_design()` refuses it. Save and replay each phase or stage batch separately. A sample whose rows or design columns were modified after execution is flagged as `modified`: its receipt describes the original execution, not the modified object.

The sampled rows themselves are not stored, because a design file is a *recipe*, not a dataset. Save the selected units with the usual data tools (`write.csv()`, arrow, ...) and keep the design file next to them as the audit trail.

## Drift detection with the execution digest

The receipt also carries a frame digest: the pool sizes and selection chances that execution resolved, with no unit identifiers. `validate_frame()` uses it to answer a question a fingerprint cannot: does a candidate frame still imply the recorded sample?

A fingerprint only says that bytes changed. The digest comparison says what the change means for the design. Rescaling a size measure changes every byte of the column but not one inclusion probability, and the message says so:

```{r}
pps_sample <- sampling_design(title = "EA PPS") |>
  stratify_by(region) |>
  cluster_by(ea_id) |>
  draw(n = 25, method = "pps_brewer", mos = households) |>
  execute(bfa_eas, seed = 7)

rescaled <- bfa_eas |>
  mutate(households = households * 10)

validate_frame(pps_sample, rescaled)
```

Shifting sizes for some units does change the chances the design would resolve today, and the message points at the stage:

```{r}
shifted <- bfa_eas |>
  mutate(households = ifelse(
    province == "Soum", households + 500, households
  ))

validate_frame(pps_sample, shifted)
```

Structural drift is reported the same way: removing or adding units names each affected pool with its recorded and current sizes. All of this is informational. A drifted frame remains executable; it just no longer reproduces the recorded sample, so replaying or extending the sample on it would give a different result.

## Control expressions and safety

`draw(control = ...)` accepts R expressions, but a design file represents their meaning as plain data. Each control term has a type and an explicit array of variables:

```json
"control": [
  {
    "type": "serpentine",
    "variables": ["province", "commune"]
  }
]
```

The three types are `ascending` for a bare column name, `descending` for
`desc()`, and `serpentine` for `serp()`. This declarative grammar can be
inspected without parsing R source code. `write_design()` refuses expressions
outside this grammar rather than writing a file that would not round-trip:

```{r, error = TRUE}
odd_design <- sampling_design() |>
  draw(n = 50, control = c(population / households))

design_json(odd_design)
```

The reader validates the type and variables as data and constructs the
restricted expression directly; it does not parse them as R code. An unknown
type is rejected, even in a file received from a third party.

```{r, error = TRUE}
payload <- jsonlite::fromJSON(design_json(design), simplifyVector = FALSE)
payload$design$stages[[1]]$draw$control[[1]]$type <- "unknown"
tampered <- jsonlite::toJSON(payload, auto_unbox = TRUE, null = "null")

read_design(tampered)
```

## What round-trips, and what does not

The round-trip contract is **execution equality**. A restored design draws the same sample as the original under the same seed, while the objects themselves may differ in inconsequential ways:

| Aspect                          | Behavior after a round-trip                                         |
|:--------------------------------|:---------------------------------------------------------------------|
| Stages, strata, clusters, draws | Preserved exactly, including all per-stratum vectors and data frames |
| Whole numbers                   | May switch between integer and double (JSON does not distinguish)    |
| Data-frame specifications       | Returned as tibbles; factor columns become character                 |
| Control expressions             | Rebuilt from declarative JSON terms into equivalent quosures         |
| Builder state                   | Not stored; the design is revalidated on read                        |

Files carry a `format_version` so that future versions of `samplyr` can read (and migrate) files written today. `read_design()` reports a clear error when a file was written by a newer version than the installed one.

```{r, include = FALSE}
unlink(path)
```

## Where serialization fits

- `vignette("design-semantics")` explains the frame-independence that makes designs portable in the first place.
- `validate_frame()` (see `vignette("introduction")`) checks any candidate frame against a design's requirements before execution.
- For coordinated samples across waves (see `vignette("sampling-coordination")`), archiving each wave's design file together with the PRN frame gives a complete, replayable history of the sample selection.
