---
title: "Save and Replay a Design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Save and Replay a Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, echo = FALSE}
library(samplyr)
library(dplyr)
```

## What the file contains

A `sampling_design` describes how to sample without embedding the confidential
frame. Saving it creates a versioned JSON recipe that can be inspected,
reviewed, and executed later.

Three functions cover the workflow.

| Function | Role |
|:--|:--|
| `write_design()` | Write a design or execution receipt to JSON |
| `read_design()` | Restore a design from a file or JSON string |
| `design_json()` | Return the same JSON in memory |

The format is native to `samplyr` and remains experimental. It is not a
finalized cross-tool interchange standard.

## Save and restore a design

```{r}
data(bfa_eas)

design <- sampling_design(title = "Burkina Faso EA survey") |>
  stratify_by(region, alloc = "proportional") |>
  draw(
    n = 300,
    method = "systematic",
    control = c(serp(province, commune))
  )

path <- tempfile(fileext = ".json")
write_design(design, path)

restored <- read_design(path)
restored
```

Under the same compatible implementation, frame, and seed, the restored design
executes like the original.

```{r}
original_sample <- execute(design, bfa_eas, seed = 123)
restored_sample <- execute(restored, bfa_eas, seed = 123)

identical(original_sample$ea_id, restored_sample$ea_id)
identical(original_sample$.weight, restored_sample$.weight)
```

Execution equality is the contract. Incidental R storage details can differ.
For example, JSON does not distinguish integers from whole-valued doubles and
restored data-frame specifications are tibbles.

`design_json()` is useful when the document belongs in a database or API.

```{r}
json <- design_json(design)
names(jsonlite::fromJSON(json))
```

## Requirements and frame fingerprints

Every file records the variables needed by each stage and their roles. These
requirements allow a candidate frame to be checked before selection.

```{r}
frame_info <- attr(restored, "frame_info")
bind_rows(frame_info$required_variables)
```

Passing a frame to `write_design()` adds a fingerprint without writing the
frame data.

```{r}
write_design(design, path, frame = bfa_eas)

with_frame <- read_design(path)
fingerprint <- attr(with_frame, "frame_info")[["fingerprint"]]

fingerprint$name
fingerprint$nrow
fingerprint$hash
```

The hash covers column names, values, and row order. It ignores whether the
object is a tibble and ignores column order. It is an integrity check, not a
proof that the register has good coverage, stable identifiers, or the correct
target-population definition.

`validate_frame()` reports structural or probability-relevant drift while
still checking whether the revised frame can execute the design.

```{r}
validate_frame(with_frame, bfa_eas[-(1:10), ])
```

A compatible but changed frame can run a new selection. It cannot reproduce
the old one. `replay_design()` therefore refuses a fingerprint mismatch by
default.

For a design with one register per stage, the file records one fingerprint per
register in supplied order. Replay requires the same number and ordering of
registers.

## Save an execution receipt

Writing a `tbl_sample` adds the arguments that determined its execution. These
include the seed, executed stages, panel or replicate request, selected count,
timestamp, package versions, and RNG configuration.

```{r}
sample <- execute(design, bfa_eas, seed = 2026, panels = 4)
write_design(sample, path, frame = bfa_eas)

saved <- read_design(path)
receipt <- attr(saved, "execution")

receipt$seed
receipt$panels
receipt$n_selected
```

Replay is exact when the recorded frame content, method implementation,
package behavior, and registered custom methods remain compatible.

```{r}
replayed <- replay_design(saved, bfa_eas)

identical(replayed$ea_id, sample$ea_id)
identical(replayed$.weight, sample$.weight)
identical(replayed$.panel, sample$.panel)
```

The execution timestamp is new. The selected rows, their order, design
columns, panels, and replicate assignments are reproduced.

The selected records themselves are not stored in the JSON recipe. Save them
with an ordinary data format and keep the design document beside them as an
audit artifact.

## Replay boundaries

The receipt describes one `execute()` call. A stage continuation or
multi-phase pipeline uses several calls, so its final receipt is marked
`chained` and cannot replay the whole pipeline. Save and replay each stage
batch or phase separately.

A modified sample is also outside the contract. If rows or design columns
changed after execution, the receipt still describes the original selection.

A materialized panel wave is derived from its master and is not independently
replayable. Save the master. A `rotation_program()` is a registry rather than
a sampling design, so archive each cohort execution and the schedule used to
rebuild the registry.

For PRN-coordinated samples, archive each executed sample and the versioned
register containing stable identifiers and PRNs. Those artifacts audit sample
selection. They do not preserve fieldwork outcomes, response histories, or
later register maintenance.

## Format and safety

The JSON stores declarative method descriptors alongside `samplyr` metadata.
Built-in methods use a versioned vocabulary. A custom method is recorded as
tool-specific and must be registered with the same implementation before
replay.

Control sorting is serialized as a restricted data structure for ascending,
descending, and serpentine terms. The reader rebuilds those terms without
parsing arbitrary R code. `write_design()` refuses expressions outside that
grammar.

```{r, error = TRUE}
unsupported_control <- sampling_design() |>
  draw(n = 50, control = c(population / households))

design_json(unsupported_control)
```

Files carry a `format_version`. A reader refuses newer versions it does not
understand and validates the fields required by older versions. Unknown tool
namespaces are retained when a file is read and written again, but preservation
does not imply cross-tool statistical equivalence.

## Archive checklist

Keep these artifacts together:

1. the executed sample data
2. the design JSON and execution receipt
3. the exact frame vintage or an access-controlled reference to it
4. stable identifiers and PRNs when coordination is used
5. custom method definitions and package versions
6. schedules linking separately executed longitudinal cohorts

Use `vignette("design-semantics")` for the detailed design contract and
`vignette("rotating-panels")` for cohort-specific persistence rules.

```{r, include = FALSE}
unlink(path)
```
