---
title: "Save and Replay a Design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Save and Replay a Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, echo = FALSE}
library(samplyr)
library(dplyr)
```

## What the file contains

A `sampling_design` describes how to sample without embedding the confidential
frame. Saving it creates a versioned JSON recipe that can be inspected,
reviewed, and executed later.

That statement applies to a design-only recipe. A file written from an
executed sample can carry its frame digest. The digest does not contain the
ordinary frame columns, but it retains selected-unit identifiers and may contain
per-unit chance metadata. Treat an execution receipt as potentially
confidential.

Three functions cover the workflow.

| Function | Role |
|:--|:--|
| `write_design()` | Write a design or execution receipt to JSON |
| `read_design()` | Restore a design from a file or JSON string |
| `design_json()` | Return the same JSON in memory |

The format is native to `samplyr` and remains experimental. It is not a
finalized cross-tool interchange standard.

Reads and writes use native R checks generated from the same source as the
bundled JSON Schemas. No JavaScript engine is needed. Misspelled executable
fields are errors. Frame compatibility is checked separately by
`validate_frame()` and `execute()`.

Files use method vocabulary version 2, with the probability quantity and
its exact or approximate quality in the common descriptor. Optional named
namespaces in `annotations` and `tools` survive reading and writing. They cannot add selection rules, and
any declared required extension is refused by this implementation.

The schemas and compatibility notes are installed with the package:

```{r, eval=FALSE}
system.file("schema", "design.schema.json", package = "samplyr")
system.file("schema", "README.md", package = "samplyr")
```

## Save and restore a design

```{r}
library(samplyr)
library(dplyr)

data(bfa_eas)

design <- sampling_design(title = "Burkina Faso EA survey") |>
  stratify_by(region, alloc = "proportional") |>
  draw(
    n = 300,
    method = "systematic",
    control = c(serp(province, commune))
  )

path <- tempfile(fileext = ".json")
write_design(design, path)

restored <- read_design(path)
restored
```

Under the same compatible implementation, frame, and seed, the restored design
executes like the original.

```{r}
original_sample <- execute(design, bfa_eas, seed = 123)
restored_sample <- execute(restored, bfa_eas, seed = 123)

identical(original_sample$ea_id, restored_sample$ea_id)
identical(original_sample$.weight, restored_sample$.weight)
```

Execution equality is the contract. Incidental R storage details can differ.
For example, JSON does not distinguish integers from whole-valued doubles and
restored data-frame specifications are tibbles.

`design_json()` is useful when the document belongs in a database or API.

```{r}
json <- design_json(design)
names(jsonlite::fromJSON(json))
```

## Requirements and frame fingerprints

Every file records the variables needed by each stage and their roles. These
requirements allow a candidate frame to be checked before selection.

```{r}
frame_info <- attr(restored, "frame_info")
bind_rows(frame_info$required_variables)
```

Passing a frame to `write_design()` adds a fingerprint without writing the
frame data.

```{r}
write_design(design, path, frame = bfa_eas)

with_frame <- read_design(path)
fingerprint <- attr(with_frame, "frame_info")[["fingerprint"]]

fingerprint$name
fingerprint$nrow
fingerprint$hash
```

The hash covers column names, values, and row order. It ignores whether the
object is a tibble and ignores column order. It is an integrity check, not a
proof that the register has good coverage, stable identifiers, or the correct
target-population definition.

`validate_frame()` reports structural or probability-relevant drift while
still checking whether the revised frame can execute the design.

```{r}
validate_frame(with_frame, bfa_eas[-(1:10), ])
```

A compatible but changed frame can run a new selection. It cannot reproduce
the old one. `replay_design()` therefore refuses a fingerprint mismatch by
default.

For a design with one register per stage, the file records one fingerprint per
register in supplied order. Replay requires the same number and ordering of
registers.

## Save an execution receipt

Writing a `tbl_sample` adds the arguments that determined its execution. These
include the seed, executed stages, panel or replicate request, selected count,
timestamp, package versions, and RNG configuration.

```{r}
sample <- execute(design, bfa_eas, seed = 2026, panels = 4)
write_design(sample, path, frame = bfa_eas)

saved <- read_design(path)
receipt <- attr(saved, "execution")

receipt$seed
receipt$panels
receipt$n_selected
```

Replay is exact when the recorded frame content, method implementation,
package behavior, and registered custom methods remain compatible.

```{r}
replayed <- replay_design(saved, bfa_eas)

identical(replayed$ea_id, sample$ea_id)
identical(replayed$.weight, sample$.weight)
identical(replayed$.panel, sample$.panel)
```

The execution timestamp is new. The selected rows, their order, design
columns, panels, and replicate assignments are reproduced.

The selected records themselves are not stored in the JSON recipe, but their
identifiers occur in the digest's selection trace, as frame-derived cluster
keys or row indices. Save the records with an ordinary
data format and keep the protected design document beside them as an audit
artifact. Use `frame_digest = "none"` at execution when that trace must not be
created, accepting the loss of digest-backed diagnostics.

## Replay boundaries

The receipt describes one `execute()` call. A stage continuation or
multi-phase pipeline uses several calls, so its final receipt is marked
`chained` and cannot replay the whole pipeline. Save and replay each stage
batch or phase separately.

A modified sample is also outside the contract. If rows or design columns
changed after execution, the receipt still describes the original selection.

A materialized panel wave is derived from its master and is not independently
replayable. Save the master. A `rotation_program()` is a registry rather than
a sampling design, so archive each cohort execution and the schedule used to
rebuild the registry.

A sample carrying shared weights from `share_weights()` has a format of its
own, `samplyr/shared-sample`. It records the source selection and the
transformation's arguments, and nothing else. The links and the target
register are supplied again at replay, the way a frame is, so the file
contains no target unit and no link:

```{r}
households <- data.frame(hh = paste0("h", 1:20))
people <- data.frame(
  person = paste0("p", 1:40),
  hh = rep(paste0("h", 1:20), each = 2)
)

shared <- share_weights(
  sampling_design() |> draw(n = 8) |> execute(households, seed = 3),
  targets = people,
  links = people,
  by = c(hh = "hh"),
  to = c(person = "person"),
  within = hh,
  multiplicity = complete_links()
)

shared_path <- tempfile(fileext = ".json")
write_design(shared, shared_path, frame = households)
read_design(shared_path)
```

Replaying it takes the register, the links and the targets. The source
selection is re-executed, then the transformation is re-applied:

```{r}
rebuilt <- replay_design(
  read_design(shared_path),
  frame = households,
  links = people,
  targets = people
)
identical(rebuilt$.weight, shared$.weight)
```

Because the links and the targets are not in the file, they cannot be checked
before use. The file instead records what the source selection and the result
hashed to, and replay is checked against both, so a mismatch says which input
was wrong. A source that does not reproduce means the register is wrong,
whereas a result that does not means the links or the targets are.

`saveRDS()` preserves the object whole, though it records an R object rather
than a design another tool can read, and it does embed the link structure
that the JSON format deliberately leaves out.

A `stack_frames()` collection has a format of its own, `samplyr/frame-stack`,
because it is one selection per component rather than one in total. Give
`frame` as a list named by component, and each component is fingerprinted
against its own register:

```{r}
population <- data.frame(
  person_id = 1:400,
  has_landline = rep(c(TRUE, FALSE), times = c(250, 150)),
  has_cell = rep(c(FALSE, TRUE), times = c(100, 300))
)
landline_frame <- population[population$has_landline, ]
cell_frame <- population[population$has_cell, ]

frames <- stack_frames(
  landline = sampling_design() |>
    draw(n = 50) |>
    execute(landline_frame, seed = 1),
  cell = sampling_design() |>
    draw(n = 60) |>
    execute(cell_frame, seed = 2),
  membership = c(landline = "has_landline", cell = "has_cell"),
  key = person_id
)

path <- tempfile(fileext = ".json")
write_design(frames, path, frame = list(landline = landline_frame,
                                        cell = cell_frame))

restored <- read_design(path)
restored
```

`read_design()` returns the components' designs and receipts rather than the
collection, which needs the registers. `replay_design()` supplies them and
stacks the results:

```{r}
replayed <- replay_design(
  restored,
  frame = list(landline = landline_frame, cell = cell_frame)
)
identical(as.data.frame(replayed), as.data.frame(frames))
```

Two things a collection cannot carry are refused rather than dropped, because
either loss would return a collection that looks complete and estimates
differently: overlaps from `exante_overlaps()`, which resolve to one chance
per selected unit when the collection is formed, and a component carrying
shared weights, which a collection has nowhere to put a link table for.
Overlaps declared with `declared_overlaps()` name columns of the components
and travel with the file.

For PRN-coordinated samples, archive each executed sample and the versioned
register containing stable identifiers and PRNs. Those artifacts audit sample
selection. They do not preserve fieldwork outcomes, response histories, or
later register maintenance.

## Format and safety

The JSON stores declarative method descriptors alongside `samplyr` metadata.
Built-in methods use a versioned vocabulary. A custom method is recorded as
tool-specific and must be registered with the same implementation before
replay.

Control sorting is serialized as a restricted data structure for ascending,
descending, and serpentine terms. The reader rebuilds those terms without
parsing arbitrary R code. `write_design()` refuses expressions outside that
grammar.

```{r, error = TRUE}
unsupported_control <- sampling_design() |>
  draw(n = 50, control = c(population / households))

design_json(unsupported_control)
```

Files carry a `format_version`. A reader refuses newer versions it does not
understand and validates the fields required by older versions. Unknown tool
namespaces are retained when a file is read and written again, but preservation
does not imply cross-tool statistical equivalence.

## Archive checklist

Keep these artifacts together:

1. the executed sample data
2. the design JSON and execution receipt
3. the exact frame vintage or an access-controlled reference to it
4. stable identifiers and PRNs when coordination is used
5. custom method definitions and package versions
6. schedules linking separately executed longitudinal cohorts

Use `vignette("design-semantics")` for the detailed design contract and
`vignette("rotating-panels")` for cohort-specific persistence rules.

```{r, include = FALSE}
unlink(path)
```
