---
title: "Survey Analysis with samplyr"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Survey Analysis with samplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, echo = FALSE}
library(samplyr)
library(dplyr)
library(survey)
```

## Overview

In a typical survey, sampling is one step in a longer process:

1. **Design and execute** the sample using `samplyr` to select units from a sampling frame.
2. **Collect data** in the field on the selected units.
3. **Merge** the collected data back into the sample, preserving the design metadata (strata, clusters, weights, FPC).
4. **Analyze** using the `survey` or `srvyr` package for design-based inference.

`samplyr` handles step 1. For steps 3 and 4, it provides bridge functions that convert a `tbl_sample` into `survey` design objects. The `tbl_sample` carries all the information that `survey::svydesign()` needs (strata, cluster ids, weights, finite population corrections), so the conversion is automatic.

Because `tbl_sample` is a tibble subclass, standard dplyr operations like `left_join()` preserve the sampling metadata. This makes step 3 straightforward: join your collected data by unit identifier and the result is still a valid `tbl_sample`. The condition is that the join is one to one on the sampled units. A join that duplicates rows changes the row set, and the sample is then marked modified; see "Domain (Subpopulation) Analysis" below for what that means for the export.

This vignette covers sample diagnostics (before fieldwork), the merge step, and integration with `survey` and `srvyr` for estimation.

## Sample Diagnostics

After executing a design, `summary()` gives a concise overview: one design line and one realization line per stage, then weight diagnostics. This is what you check before going to the field to verify the sample looks right. Per-pool allocation tables are in `frame_summary(detail = "pool")`.

```{r}
library(samplyr)
library(dplyr)
library(survey)

data(bfa_eas)

sample <- sampling_design() |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 300) |>
  execute(bfa_eas, seed = 1)

summary(sample)
```

The realization line uses compact sampling notation. Its $N_h$, $n_h$, and
$f_h$ labels correspond to the `N`, `n_realized`, and `take_rate` columns in
the pool table.

```{r}
allocation <- frame_summary(sample, detail = "pool")

allocation |>
  select(region, replicate, N, n_target, n_expected, n_realized, take_rate)
```

`n_target` is the nominal allocation requested by the design, `n_expected` is
the sum of the final resolved selection chances, and `n_realized` is what the
execution actually selected. They coincide here because this is a feasible
fixed-size design. For Bernoulli or Poisson sampling, `n_target` records the
requested expectation, which can differ from both the resolved expectation and
the random realized count. `take_rate` is `n_realized / N`; it is not a
unit-level inclusion probability for an unequal-probability design.

Every pool table has a `replicate` column (`1` for an ordinary execution).
With `execute(reps = ...)`, it contains one row per pool and replicate, so the
realized variation remains numeric and can be grouped, plotted, or filtered
without unpacking list columns. The weight diagnostics report the Kish design
effect and effective sample size.

## Merging Collected Data

After fieldwork, you have collected data (survey responses, measurements) for the selected units. Join it into the `tbl_sample` by a shared identifier, one row of collected data per selected unit. The `tbl_sample` class and its design metadata are preserved through a join that does not change the row set.

```{r}
# Simulate household-budget outcomes collected after selecting the EAs.
# These outcomes are not variables on the sampling frame.
set.seed(1)
log_consumption <- rnorm(
  nrow(sample),
  mean = 11 + 0.25 * (sample$urban_rural == "Urban") -
    0.12 * (sample$remoteness == "High"),
  sd = 0.65
)
collected <- tibble(
  ea_id = sample$ea_id,
  consumption_per_capita = round(exp(log_consumption)),
  below_consumption_threshold = as.integer(log_consumption < 11)
)

# Merge into the sample
sample_with_data <- sample |>
  left_join(collected, by = "ea_id")

# Still a tbl_sample with all design metadata intact
class(sample_with_data)
```

The merged object can now be exported to `survey` for design-based analysis of the collected variables.

## Exporting to the survey Package

`as_svydesign()` converts a `tbl_sample` into a `survey::svydesign()` object. It automatically maps strata, clusters, weights, and finite population corrections from the sampling metadata.

### Stratified design

```{r}
svy <- as_svydesign(sample_with_data)
svy

# Estimate collected variables
svymean(~consumption_per_capita, svy)
svymean(~below_consumption_threshold, svy)

# Frame variables are still available for validation
svytotal(~population, svy)
```

### Cluster design with PPS

For fixed-size PPS without-replacement designs (`pps_brewer`,
`pps_systematic`, `pps_cps`, `pps_sampford`, `pps_sps`, and `pps_pareto`),
`as_svydesign()` uses Brewer's variance approximation by default. Here
"Brewer" names the variance approximation used by `survey`, not the selection
algorithm: a Sampford sample, for example, still receives this default unless
you supply its exact joint inclusion probabilities.

```{r}
data(bfa_eas)

cluster_sample <- sampling_design() |>
  cluster_by(ea_id) |>
  draw(n = 50, method = "pps_brewer", mos = households) |>
  execute(bfa_eas, seed = 2)

svy_cluster <- as_svydesign(cluster_sample)
svy_cluster
svymean(~households, svy_cluster)
```

### Multi-stage design

For multi-stage designs, every executed stage maps to one term of the `id` formula, with a matching finite population correction. A final stage without `cluster_by()` gets a synthesized element identifier. Strata are exported per stage, so `survey` applies the full multi-stage linearization [@sarndal1992, ch. 4.3] rather than collapsing the design to its first stage.

The linearization is exact for the design it is given, but each stage still carries whatever variance approximation its selection method uses. The design below has a `pps_brewer` first stage, so its first-stage term uses Brewer's approximation to \(\pi_{kl}\), as described above. Stage-by-stage exactness is a property of the method, not of the export.

```{r}
data(zwe_eas)

zwe_frame <- zwe_eas |>
  mutate(district_hh = sum(households), .by = district)

ms_sample <- sampling_design() |>
  add_stage(label = "Districts") |>
    stratify_by(province) |>
    cluster_by(district) |>
    draw(n = 2, method = "pps_brewer", mos = district_hh) |>
  add_stage(label = "EAs") |>
    draw(n = 3) |>
  execute(zwe_frame, seed = 3)

svy_ms <- as_svydesign(ms_sample)
svy_ms
```

Two things about this frame surface here, and they are separate. First, the capping warning: Bulawayo holds a single district, so a take of two selects it whole. That is reported once per stage, and it is why the province contributes no first-stage variance.

Second, certainty. Harare and Bulawayo each contain a district large enough that its inclusion probability resolves to 1. Those districts are self-representing, so they go into a take-all stratum that contributes no variance. That split leaves Harare with a single probability district, and a stratum holding one probability PSU has no estimable within-stratum variance. `survey` reports this rather than returning a number, so the design needs a rule for it. Centering the lone PSU on the population mean rather than its own stratum mean is the conservative choice, in that it contributes a variance term instead of none.

`survey.lonely.psu` is a global option. Set it for the call rather than for the session, so a later analysis in the same script does not inherit a rule that was chosen for this design.

```{r}
withr::with_options(
  list(survey.lonely.psu = "adjust"),
  svymean(~households, svy_ms)
)
```

## Domain (Subpopulation) Analysis

To estimate for a subpopulation, convert the full sample first and
subset the survey design. Do not filter the `tbl_sample` rows before
converting.

```{r}
urban_svy <- subset(svy, urban_rural == "Urban")
svymean(~consumption_per_capita, urban_svy)
```

The distinction matters for variance. Dropping out-of-domain rows
before conversion produces the same point estimate but understates the
standard error, because the number of sampled units falling in the
domain is itself random and that variation is part of the design
variance [@sarndal1992, ch. 5.8]. `survey::subset()` keeps the full
design structure and applies the correct domain estimator.

For this reason a filtered `tbl_sample` is marked as modified and
`as_svydesign()` refuses it.

```{r, error=TRUE}
urban_only <- sample_with_data |>
  filter(urban_rural == "Urban")

as_svydesign(urban_only)
```

The same rule applies to any operation that removes, adds, or
duplicates rows, or that drops, renames, or overwrites internal design
columns such as `.weight` or `.fpc_1` (a `select()` that keeps only
`.weight` would silently lose the finite population correction).
One-to-one joins (as in the merge above), row reordering, and column
reordering are unaffected. Extracting a single complete replicate with
`filter(.replicate == r)` from a replicated execution remains
supported.

With srvyr, the same domain analysis reads naturally as a pipeline:
convert first, then filter the design object.

```{r, eval=requireNamespace("srvyr", quietly = TRUE)}
library(srvyr)

sample_with_data |>
  as_survey_design() |>
  filter(urban_rural == "Urban") |>
  summarise(mean_consumption = survey_mean(consumption_per_capita))
```

## Replicate-Weight Export

For replicate-based variance estimation (jackknife/BRR/bootstrap), use `as_svrepdesign()`. This converts a `tbl_sample` to `svydesign`, then to `svyrep.design` using `survey::as.svrepdesign()`.

```{r}
rep_svy <- as_svrepdesign(sample, type = "auto")
svymean(~households, rep_svy)
```

You can also export directly to a `srvyr` replicate design, with `srvyr`
already attached from the domain example above.

```{r, eval=requireNamespace("srvyr", quietly = TRUE)}
rep_tbl <- as_survey_rep(sample, type = "auto")
rep_tbl |>
  summarise(mean_hh = survey_mean(households, vartype = "se"))
```

Replicate export supports single-phase designs. For PPS designs, use `type = "subbootstrap"` or `type = "mrbbootstrap"`. Two-phase designs should use `as_svydesign()` (linearization).

LPM2 and SCPS require a different qualification. Their spatial spreading
changes pairwise selection behavior, and samplyr does not currently provide
a design-specific linearization estimator or joint inclusion probabilities
for them. Consequently, `as_svydesign()` refuses these designs.
`as_svrepdesign(type = "subbootstrap")` provides a generic PPS bootstrap
approximation, but it does not reproduce the spatial selection algorithm
inside each replicate and should not be described as exact.

```{r}
spatial_frame <- slice_head(bfa_eas, n = 500)

spatial_sample <- sampling_design() |>
  draw(
    n = 40,
    method = "lpm2",
    spread = c(longitude, latitude)
  ) |>
  execute(spatial_frame, seed = 7)

# Use method = "scps" for spatially correlated Poisson sampling.
spatial_rep <- as_svrepdesign(
  spatial_sample,
  type = "subbootstrap",
  replicates = 20
)
svymean(~households, spatial_rep)
```

## Variance with Joint Expectations

Brewer's approximation is what `as_svydesign()` supplies by default for
fixed-size PPS-WOR stages. When the exact second-order quantities are wanted
instead, `joint_expectation()` reconstructs them directly for supported
methods. For WOR stages, it returns joint inclusion probabilities
$\pi_{kl}$. For WR/PMR stages (`pps_multinomial`, `pps_chromy`), it returns
joint expected hits $E(n_k n_l)$. Depending on the selection method, these
quantities are exact or approximate.

`vignette("validation")` runs a two-stage `pps_brewer` design against a
synthetic population with a known mean and compares the empirical spread of the
estimates with the standard errors this default produces, which is the
approximation being checked rather than assumed.

It replays the design specification against the original frame, computing
quantities per stratum or conditional parent occurrence. Blocks and units
within blocks follow first appearance in the sample. Cross-block entries are
products of marginal chances. Below a WR parent, every parent draw occurrence
defines a separate independent child block, even when several draws hit the
same population cluster.

```{r}
sampford_sample <- sampling_design() |>
  stratify_by(region) |>
  cluster_by(ea_id) |>
  draw(n = 5, method = "pps_sampford", mos = households) |>
  execute(bfa_eas, seed = 2025)

# Sampford joint inclusion probabilities are exact.
jip <- joint_expectation(sampford_sample, bfa_eas, stages = 1)

# Use the exact matrix for single-stage PPS variance estimation.
svy_exact <- as_svydesign(
  sampford_sample,
  pps = ppsmat(jip[[1]])
)
svymean(~households, svy_exact)

# The default export instead uses the generic Brewer approximation.
svy_brewer <- as_svydesign(sampford_sample)
svymean(~households, svy_brewer)
```

Which methods yield exact rather than approximate second-order quantities,
and which yield none at all, is tabulated in `?joint_expectation`. In outline:
`pps_sampford`, `pps_systematic`, `pps_cps` and `pps_poisson` are exact;
`pps_brewer`, `pps_sps`, `pps_pareto` and unconstrained `cube` use the
high-entropy approximation; `lpm2` and `scps` have none, because spatial
spreading is not represented.

For bounded cube designs, joint probabilities are also unavailable because
the count constraints alter pairwise selection behavior. For LPM2, SCPS, and
bounded cube, use the generic bootstrap route described above. For a
multi-stage PPS design, supplying `pps = ppsmat(...)` represents only the
stage covered by that matrix; omit `pps` to retain samplyr's full multi-stage
linearization with the default stage-level approximation.

## End-to-End Workflow

A complete workflow from design to estimates.

```{r}
# 1. Define the design
design <- sampling_design(title = "Burkina Faso EA Survey") |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 300)

# 2. Validate the frame
validate_frame(design, bfa_eas)

# 3. Draw the sample
ea_sample <- execute(design, bfa_eas, seed = 2026)

# 4. Inspect before fieldwork
summary(ea_sample)

# 5. Collect data (simulated here)
set.seed(99)
field_data <- tibble(
  ea_id = ea_sample$ea_id,
  consumption_per_capita = round(
    rlnorm(nrow(ea_sample), meanlog = 11, sdlog = 0.65)
  )
)

# 6. Merge collected data
ea_sample <- ea_sample |>
  left_join(field_data, by = "ea_id")

# 7. Export and analyze
svy <- as_svydesign(ea_sample)
svymean(~consumption_per_capita, svy)
confint(svymean(~consumption_per_capita, svy))
```

## References
