---
title: "Analyze a samplyr Sample"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Analyze a samplyr Sample}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, echo = FALSE}
library(samplyr)
library(dplyr)
library(survey)
```

## From selection to estimation

The usual workflow has four parts. Select the sample, collect outcomes, join
them to the selected units, and export the result to `survey` or `srvyr`.

`samplyr` records the strata, clusters, weights, and finite population
corrections during selection. A safe one-to-one join preserves that metadata.
An operation that adds, removes, or duplicates sampled rows changes the
realization and prevents export.

## Inspect the sample before fieldwork

We begin with a stratified sample of enumeration areas.

```{r}
data(bfa_eas)

sample <- sampling_design(title = "Burkina Faso EA survey") |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 300) |>
  execute(bfa_eas, seed = 1)

summary(sample)
```

`frame_summary()` exposes the same realization at pool level.

```{r}
frame_summary(sample, detail = "pool") |>
  select(region, N, n_target, n_expected, n_realized, take_rate)
```

For this feasible fixed-size design, the target, expectation, and realized
take coincide. Under Bernoulli or Poisson selection the realized count is
random. Under unequal-probability selection, `take_rate` describes the
realized pool count and is not a unit-level inclusion probability.

## Join collected outcomes

The frame need not contain the survey outcomes. Join one collected record to
each sampled unit using a stable identifier.

```{r}
library(dplyr)
set.seed(2)

collected <- tibble(
  ea_id = sample$ea_id,
  consumption_per_capita = round(rlnorm(
    nrow(sample),
    meanlog = 11,
    sdlog = 0.65
  )),
  below_threshold = rbinom(nrow(sample), size = 1, prob = 0.25)
)

sample_with_data <- sample |>
  left_join(collected, by = "ea_id")

class(sample_with_data)
```

The result remains a `tbl_sample` because the join is one to one and leaves
the selected rows intact. Check key uniqueness before joining when collected
data can contain repeated visits, roster members, or duplicate submissions.

## Export to survey

`as_svydesign()` maps the recorded design to a `survey` object.

```{r}
library(survey)
svy <- as_svydesign(sample_with_data)
svy
```

The collected variables can now be analyzed with standard `survey` functions.

```{r}
svymean(~consumption_per_capita, svy)
svymean(~below_threshold, svy)
svytotal(~population, svy)
```

For a multistage sample, every represented stage contributes its sampling-unit
and finite-population terms. This preserves the stage structure. It does not
turn an approximate stage variance treatment into an exact one.

Fixed-size PPS methods use Brewer's approximation by default. In that phrase,
Brewer names the `survey` variance approximation and not necessarily the
selection algorithm. The approximation's accuracy depends on the population
and design. Exact joint inclusion probabilities can be supplied for methods
that provide them.

Equal-probability systematic sampling uses an SRS-style variance approximation.
Its direction depends on frame ordering. `as_svydesign()` warns unless the
caller explicitly accepts or refuses that approximation through
`systematic_variance`.

## Estimate a domain correctly

Convert the complete sample first, then subset the survey design.

```{r}
urban_svy <- subset(svy, urban_rural == "Urban")
svymean(~consumption_per_capita, urban_svy)
```

Filtering the `tbl_sample` first treats the observed domain sample as fixed.
The point estimate can agree, but the variance estimate is generally wrong and
is often too small because domain membership is random under the design
[@sarndal1992, ch. 5.8].

```{r, error = TRUE}
urban_rows <- sample_with_data |>
  filter(urban_rural == "Urban")

as_svydesign(urban_rows)
```

One-to-one joins, row reordering, and new ordinary columns remain supported.
Changing the row set or an internal column such as `.weight` or `.fpc_1` does
not.

With `srvyr`, use the same order. Convert first and then filter the survey
object.

```{r, eval=requireNamespace("srvyr", quietly = TRUE)}
library(srvyr)

sample_with_data |>
  as_survey_design() |>
  filter(urban_rural == "Urban") |>
  summarise(
    mean_consumption = survey_mean(consumption_per_capita)
  )
```

## Replicate weights and PPS variance

For a supported single-phase design, `as_svrepdesign()` creates replicate
weights through `survey::as.svrepdesign()`.

```{r}
replicate_design <- as_svrepdesign(sample, type = "auto")
svymean(~households, replicate_design)
```

Replicate methods are not automatic proofs of exactness. Generic PPS
bootstraps do not recreate spatial selection, hard cube constraints, or a
random Poisson sample size. `as_svydesign()` and `as_svrepdesign()` report or
refuse combinations that cannot be represented under their documented
contracts. See their reference pages before selecting a variance route for an
advanced design.

For supported PPS methods, `joint_expectation()` reconstructs exact or
approximate second-order quantities according to the method.

```{r}
sampford_sample <- sampling_design() |>
  stratify_by(region) |>
  cluster_by(ea_id) |>
  draw(n = 5, method = "pps_sampford", mos = households) |>
  execute(bfa_eas, seed = 2025)

joints <- joint_expectation(sampford_sample, bfa_eas, stages = 1)

exact_pps_design <- as_svydesign(
  sampford_sample,
  pps = ppsmat(joints[[1]])
)

svymean(~households, exact_pps_design)
```

The Sampford matrix above is exact. CPS, systematic PPS, and Poisson selection
also have exact second-order calculations in the supported cases.
Generalized Brewer, SPS, Pareto, and unconstrained cube use the documented
high-entropy approximation. Spatial methods and bounded cube do not provide a
joint matrix. `?joint_expectation` is the authoritative method table.

A user-supplied `pps` matrix describes one stage. For a multistage design,
supplying it exports only the represented PPS stage. Omit it when the intended
analysis is the full multistage linearization with the default stage-level
variance treatment.

## Analysis checklist

1. Inspect the selected counts and weights before fieldwork.
2. Verify that collected-data keys are unique at the sampled-unit level.
3. Join outcomes without changing the selected row set.
4. Convert the complete sample to a survey design.
5. Define domains on the converted design.
6. State every variance approximation used by an advanced selection method.

Use `vignette("serialization")` to archive the selection and
`vignette("design-semantics")` for the detailed export contracts.

## References
