---
title: "Survey Analysis with samplyr"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Survey Analysis with samplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, echo = FALSE}
library(samplyr)
library(dplyr)
library(survey)
```

## Overview

In a typical survey, sampling is one step in a longer process:

1. **Design and execute** the sample using `samplyr` to select units from a sampling frame.
2. **Collect data** in the field on the selected units.
3. **Merge** the collected data back into the sample, preserving the design metadata (strata, clusters, weights, FPC).
4. **Analyze** using the `survey` or `srvyr` package for design-based inference.

`samplyr` handles step 1. For steps 3 and 4, it provides bridge functions that convert a `tbl_sample` into `survey` design objects. The `tbl_sample` carries all the information that `survey::svydesign()` needs (strata, cluster ids, weights, finite population corrections), so the conversion is automatic.

Because `tbl_sample` is a tibble subclass, standard dplyr operations like `left_join()` preserve the sampling metadata. This makes step 3 straightforward: join your collected data by unit identifier and the result is still a valid `tbl_sample`.

This vignette covers sample diagnostics (before fieldwork), the merge step, and integration with `survey` and `srvyr` for estimation.

## Sample Diagnostics

After executing a design, `summary()` gives a concise overview: one design line and one realization line per stage, then weight diagnostics. This is what you check before going to the field to verify the sample looks right. Per-pool allocation tables are in `frame_summary(detail = "pool")`.

```{r}
library(samplyr)
library(dplyr)
library(survey)

data(bfa_eas)

sample <- sampling_design() |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 300) |>
  execute(bfa_eas, seed = 1)

summary(sample)
```

The allocation table shows the population size ($N_h$), sample size ($n_h$), and sampling fraction ($f_h$) per stratum. The weight diagnostics report the Kish design effect and effective sample size.

## Merging Collected Data

After fieldwork, you have collected data (survey responses, measurements) for the selected units. Join it into the `tbl_sample` by a shared identifier. The `tbl_sample` class and its design metadata are preserved through the join.

```{r}
# Simulate household-budget outcomes collected after selecting the EAs.
# These outcomes are not variables on the sampling frame.
set.seed(1)
log_consumption <- rnorm(
  nrow(sample),
  mean = 11 + 0.25 * (sample$urban_rural == "Urban") -
    0.12 * (sample$remoteness == "High"),
  sd = 0.65
)
collected <- tibble(
  ea_id = sample$ea_id,
  consumption_per_capita = round(exp(log_consumption)),
  below_consumption_threshold = as.integer(log_consumption < 11)
)

# Merge into the sample
sample_with_data <- sample |>
  left_join(collected, by = "ea_id")

# Still a tbl_sample with all design metadata intact
class(sample_with_data)
```

The merged object can now be exported to `survey` for design-based analysis of the collected variables.

## Exporting to the survey Package

`as_svydesign()` converts a `tbl_sample` into a `survey::svydesign()` object. It automatically maps strata, clusters, weights, and finite population corrections from the sampling metadata.

### Stratified design

```{r}
svy <- as_svydesign(sample_with_data)
svy

# Estimate collected variables
svymean(~consumption_per_capita, svy)
svymean(~below_consumption_threshold, svy)

# Frame variables are still available for validation
svytotal(~population, svy)
```

### Cluster design with PPS

For fixed-size PPS without-replacement designs (`pps_brewer`,
`pps_systematic`, `pps_cps`, `pps_sampford`, `pps_sps`, and `pps_pareto`),
`as_svydesign()` uses Brewer's variance approximation by default. Here
"Brewer" names the variance approximation used by `survey`, not the selection
algorithm: a Sampford sample, for example, still receives this default unless
you supply its exact joint inclusion probabilities.

```{r}
data(bfa_eas)

cluster_sample <- sampling_design() |>
  cluster_by(ea_id) |>
  draw(n = 50, method = "pps_brewer", mos = households) |>
  execute(bfa_eas, seed = 2)

svy_cluster <- as_svydesign(cluster_sample)
svy_cluster
svymean(~households, svy_cluster)
```

### Multi-stage design

For multi-stage designs, every executed stage maps to one term of the `id` formula, with a matching finite population correction. A final stage without `cluster_by()` gets a synthesized element identifier. Strata are exported per stage, so the result is exact multi-stage linearization [@sarndal1992, ch. 4.3].

```{r}
data(zwe_eas)

zwe_frame <- zwe_eas |>
  mutate(district_hh = sum(households), .by = district)

ms_sample <- sampling_design() |>
  add_stage(label = "Districts") |>
    stratify_by(province) |>
    cluster_by(district) |>
    draw(n = 2, method = "pps_brewer", mos = district_hh) |>
  add_stage(label = "EAs") |>
    draw(n = 3) |>
  execute(zwe_frame, seed = 3)

svy_ms <- as_svydesign(ms_sample)
svy_ms
svymean(~households, svy_ms)
```

## Domain (Subpopulation) Analysis

To estimate for a subpopulation, convert the full sample first and
subset the survey design. Do not filter the `tbl_sample` rows before
converting.

```{r}
urban_svy <- subset(svy, urban_rural == "Urban")
svymean(~consumption_per_capita, urban_svy)
```

The distinction matters for variance. Dropping out-of-domain rows
before conversion produces the same point estimate but understates the
standard error, because the number of sampled units falling in the
domain is itself random and that variation is part of the design
variance [@sarndal1992, ch. 5.8]. `survey::subset()` keeps the full
design structure and applies the correct domain estimator.

For this reason a filtered `tbl_sample` is marked as modified and
`as_svydesign()` refuses it:

```{r, error=TRUE}
urban_only <- sample_with_data |>
  filter(urban_rural == "Urban")

as_svydesign(urban_only)
```

The same rule applies to any operation that removes, adds, or
duplicates rows, or that drops, renames, or overwrites internal design
columns such as `.weight` or `.fpc_1` (a `select()` that keeps only
`.weight` would silently lose the finite population correction).
One-to-one joins (as in the merge above), row reordering, and column
reordering are unaffected. Extracting a single complete replicate with
`filter(.replicate == r)` from a replicated execution remains
supported.

With srvyr, the same domain analysis reads naturally as a pipeline:
convert first, then filter the design object.

```{r, eval=requireNamespace("srvyr", quietly = TRUE)}
library(srvyr)

sample_with_data |>
  as_survey_design() |>
  filter(urban_rural == "Urban") |>
  summarise(mean_consumption = survey_mean(consumption_per_capita))
```

## Replicate-Weight Export

For replicate-based variance estimation (jackknife/BRR/bootstrap), use `as_svrepdesign()`. This converts a `tbl_sample` to `svydesign`, then to `svyrep.design` using `survey::as.svrepdesign()`.

```{r}
rep_svy <- as_svrepdesign(sample, type = "auto")
svymean(~households, rep_svy)
```

You can also export directly to a `srvyr` replicate design:

```{r}
library(srvyr)

rep_tbl <- as_survey_rep(sample, type = "auto")
rep_tbl |>
  summarise(mean_hh = survey_mean(households, vartype = "se"))
```

Replicate export supports single-phase designs. For PPS designs, use `type = "subbootstrap"` or `type = "mrbbootstrap"`. Two-phase designs should use `as_svydesign()` (linearization).

LPM2 and SCPS require a different qualification. Their spatial spreading
changes pairwise selection behaviour, and samplyr does not currently provide
a design-specific linearization estimator or joint inclusion probabilities
for them. Consequently, `as_svydesign()` refuses these designs.
`as_svrepdesign(type = "subbootstrap")` provides a generic PPS bootstrap
approximation, but it does not reproduce the spatial selection algorithm
inside each replicate and should not be described as exact.

```{r}
spatial_frame <- slice_head(bfa_eas, n = 500)

spatial_sample <- sampling_design() |>
  draw(
    n = 40,
    method = "lpm2",
    spread = c(longitude, latitude)
  ) |>
  execute(spatial_frame, seed = 7)

# Use method = "scps" for spatially correlated Poisson sampling.
spatial_rep <- as_svrepdesign(
  spatial_sample,
  type = "subbootstrap",
  replicates = 20
)
svymean(~households, spatial_rep)
```

## Variance with Joint Expectations

The Brewer approximation works well in practice, but
`joint_expectation()` can reconstruct second-order quantities directly for
supported methods. For WOR stages, it returns joint inclusion probabilities
$\pi_{kl}$. For WR/PMR stages (`pps_multinomial`, `pps_chromy`), it returns
joint expected hits $E(n_k n_l)$. Depending on the selection method, these
quantities are exact or approximate.

It replays the design specification against the original frame, computing quantities per stratum and assembling a block-diagonal matrix.

```{r}
sampford_sample <- sampling_design() |>
  stratify_by(region) |>
  cluster_by(ea_id) |>
  draw(n = 5, method = "pps_sampford", mos = households) |>
  execute(bfa_eas, seed = 2025)

# Sampford joint inclusion probabilities are exact.
jip <- joint_expectation(sampford_sample, bfa_eas, stage = 1)

# Use the exact matrix for single-stage PPS variance estimation.
svy_exact <- as_svydesign(
  sampford_sample,
  pps = ppsmat(jip[[1]])
)
svymean(~households, svy_exact)

# The default export instead uses the generic Brewer approximation.
svy_brewer <- as_svydesign(sampford_sample)
svymean(~households, svy_brewer)
```

The method-specific availability is:

| samplyr method    | `joint_expectation()` result | Precision or status |
|:------------------|:-----------------------------|:--------------------|
| `pps_brewer`      | $\pi_{kl}$                   | Approximate          |
| `pps_sampford`    | $\pi_{kl}$                   | Exact                |
| `pps_systematic`  | $\pi_{kl}$                   | Exact                |
| `pps_cps`         | $\pi_{kl}$                   | Exact                |
| `pps_poisson`     | $\pi_{kl}$                   | Exact                |
| `pps_sps`         | $\pi_{kl}$                   | Approximate          |
| `pps_pareto`      | $\pi_{kl}$                   | Approximate          |
| `pps_multinomial` | $E(n_k n_l)$                 | Exact                |
| `pps_chromy`      | $E(n_k n_l)$                 | Approximate          |
| `cube`            | $\pi_{kl}$                   | Approximate when unconstrained |
| `lpm2`            | Unavailable                  | Spatial bootstrap only |
| `scps`            | Unavailable                  | Spatial bootstrap only |

For bounded cube designs, joint probabilities are also unavailable because
the count constraints alter pairwise selection behaviour. For LPM2, SCPS, and
bounded cube, use the generic bootstrap route described above. For a
multi-stage PPS design, supplying `pps = ppsmat(...)` represents only the
stage covered by that matrix; omit `pps` to retain samplyr's full multi-stage
linearization with the default stage-level approximation.

## End-to-End Workflow

A complete workflow from design to estimates:

```{r}
# 1. Define the design
design <- sampling_design(title = "Burkina Faso EA Survey") |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 300)

# 2. Validate the frame
validate_frame(design, bfa_eas)

# 3. Draw the sample
ea_sample <- execute(design, bfa_eas, seed = 2026)

# 4. Inspect before fieldwork
summary(ea_sample)

# 5. Collect data (simulated here)
set.seed(99)
field_data <- tibble(
  ea_id = ea_sample$ea_id,
  consumption_per_capita = round(
    rlnorm(nrow(ea_sample), meanlog = 11, sdlog = 0.65)
  )
)

# 6. Merge collected data
ea_sample <- ea_sample |>
  left_join(field_data, by = "ea_id")

# 7. Export and analyze
svy <- as_svydesign(ea_sample)
svymean(~consumption_per_capita, svy)
confint(svymean(~consumption_per_capita, svy))
```

## Reference
