---
title: "Survey Planning with svyplan"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Survey Planning with svyplan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, echo = FALSE}
library(samplyr)
library(svyplan)
library(dplyr)
```

## Overview

Before drawing a sample, you need to answer two questions: how many units to select, and how to divide them across strata. The `svyplan` package provides tools for both. `samplyr` integrates with `svyplan` so that planning results flow directly into `draw()` and `execute()`.

The three packages form a pipeline:

- **svyplan** determines sample sizes, evaluates precision, and computes strata boundaries (planning).
- **samplyr** specifies and executes the sampling design (selection).
- **sondage** provides the low-level sampling algorithms (engine).

This vignette uses the `ken_enterprises` dataset, a synthetic frame of 17,004 Kenyan establishments with revenue, employee counts, and export status.

```{r}
data(ken_enterprises)
glimpse(ken_enterprises)
```

## Sample Size for a Proportion

Suppose we want to estimate the proportion of exporting firms. For this
demonstration, we use the proportion in the synthetic frame as the planning
value. We want a margin of error of 3 percentage points at 95% confidence.

```{r}
p_export <- mean(ken_enterprises$exporter)
round(p_export, 3)

n_export <- n_prop(p = p_export, moe = 0.03)
n_export
```

The result is a `svyplan_n` object. You can pass it directly to `draw()`:

```{r}
samp <- sampling_design() |>
  draw(n = n_export) |>
  execute(ken_enterprises, seed = 1)
nrow(samp)
```

## Sample Size for a Mean

We also want to estimate mean revenue with a margin of error of 30 million KES. We compute the population variance from the frame:

```{r}
rev_var <- var(ken_enterprises$revenue_millions)
n_rev <- n_mean(var = rev_var, moe = 30)
n_rev
```

## Multiple Indicators

In practice, a survey measures several things at once. `n_multi()` takes the binding constraint across indicators:

```{r}
indicators <- data.frame(
  name = c("exporter_rate", "mean_revenue"),
  p = c(p_export, NA),
  var = c(NA, rev_var),
  moe = c(0.03, 30)
)
n_survey <- n_multi(indicators)
n_survey
```

The sample size is driven by whichever indicator requires the most units. Pass it to `draw()` like any other sample size:

```{r}
samp <- sampling_design() |>
  draw(n = n_survey) |>
  execute(ken_enterprises, seed = 2)
nrow(samp)
```

### Per-domain requirements

With `domains`, `n_multi()` sizes each reporting domain separately. The
result hands `draw()` a per-domain table, so the design needs a matching
`stratify_by()`:

```{r}
indicators_dom <- data.frame(
  name = "exporter_rate",
  size_class = c("Small", "Medium", "Large"),
  p = c(0.10, 0.22, 0.45),
  cv = 0.10
)
n_dom <- n_multi(indicators_dom, domains = "size_class")

samp_dom <- sampling_design() |>
  stratify_by(size_class) |>
  draw(n = n_dom) |>
  execute(ken_enterprises, seed = 2)
table(samp_dom$size_class)
```

## Power Analysis

If the goal is to detect a difference (for example, comparing exporter rates between two regions), `power_prop()` gives the required per-group sample size:

```{r}
n_power <- power_prop(p1 = p_export, p2 = p_export + 0.08, power = 0.8)
n_power
```

This also works directly with `draw()`:

```{r}
samp <- sampling_design() |>
  draw(n = n_power) |>
  execute(ken_enterprises, seed = 3)
nrow(samp)
```

## Response Rate Adjustment

All svyplan functions accept a `resp_rate` parameter. The sample size is inflated by `1 / resp_rate` to account for expected non-response:

```{r}
n_prop(p = p_export, moe = 0.03, resp_rate = 0.85)
```

The net sample size (after non-response) matches the original target. This works in `n_mean()`, `n_multi()`, `n_cluster()`, and `power_*()` functions as well.

## Precision Analysis

Given a fixed sample size, how precise will the estimates be? The `prec_*()` functions are the inverse of `n_*()`:

```{r}
prec_prop(p = p_export, n = 600)
```

```{r}
prec_mean(var = rev_var, n = 300, mu = mean(ken_enterprises$revenue_millions))
```

### Round-trip between size and precision

All `n_*()` and `prec_*()` functions are S3 generics. Pass a precision result to `n_*()` to recover the sample size, or a sample size result to `prec_*()` to compute the achieved precision:

```{r}
s <- n_prop(p = p_export, moe = 0.03)
p <- prec_prop(s)
p

# Recover the original n
n_prop(p)
```

You can override a parameter on the way back:

```{r}
n_prop(p, cv = 0.10)
```

### Confidence intervals

`confint()` extracts the expected confidence interval from any sample size or precision object:

```{r}
confint(s)
confint(p, level = 0.99)
```

## Sensitivity Analysis

How sensitive is a sample size to assumptions about the design effect or response rate? The `predict()` method evaluates any svyplan result at new parameter combinations:

```{r}
x <- n_prop(p = p_export, moe = 0.03, deff = 1.5)
predict(x, expand.grid(
  deff = c(1.0, 1.5, 2.0, 2.5),
  resp_rate = c(0.8, 0.9, 1.0)
))
```

This works for all svyplan object types: sample size, precision, cluster, and power.

## Multistage Cluster Planning

For cluster designs, `n_cluster()` finds the optimal allocation across stages given costs and the degree of clustering.

### Variance components

If a frame with cluster identifiers is available, `varcomp()` estimates the between-cluster and within-cluster variance components:

```{r}
vc <- varcomp(
  revenue_millions ~ county,
  data = ken_enterprises
)
vc
as.data.frame(vc)
```

The `icc` value (measure of homogeneity) feeds directly into `n_cluster()` and `design_effect()`.

### Optimal allocation

```{r}
# Minimize cost to achieve CV = 0.10
n_cluster(stage_cost = c(500, 50), icc = vc, cv = 0.10)

# Maximize precision within a budget
n_cluster(stage_cost = c(500, 50), icc = vc, budget = 30000)
```

The precision-target result is diagnostic: its required number of county PSUs
far exceeds the 47 available in this frame, so a 10% CV is infeasible under
these assumptions. The budget-constrained result is feasible and is the plan
used below.

A cluster plan feeds a two-stage design directly. In a design with
`cluster_by()`, `draw(n = cl)` hands stage 1 the number of PSUs and
stage 2 the per-cluster take, using the plan's integerized field
design (`as.integer(cl)`, which svyplan re-optimizes to whole units
rather than rounding each stage up):

```{r}
cl <- n_cluster(stage_cost = c(500, 50), icc = vc, budget = 30000)
as.integer(cl)

samp <- sampling_design() |>
  cluster_by(county) |>
  draw(n = cl) |>
  add_stage() |>
  draw(n = cl) |>
  execute(ken_enterprises, seed = 5)
length(unique(samp$county))
nrow(samp)
```

In a flat, unclustered single-stage design the same object contributes
its operational element total (`prod(as.integer(cl))`) instead, for
sizing an element sample.

## Stratification Boundaries

Enterprise surveys typically stratify by size. Rather than using predefined size classes, `strata_bound()` finds optimal boundaries on a continuous variable. The cumulative square root of frequency method minimizes the coefficient of variation for a given number of strata and total sample size:

```{r}
bounds <- strata_bound(
  ken_enterprises$revenue_millions,
  n_strata = 4,
  method = "cumrootf",
  cv = 0.05
)
bounds
```

The result includes jointly optimized boundaries and stratum sample sizes. The `predict()` method assigns each frame unit to a stratum:

```{r}
labels <- paste0("S", 1:4)
ken_enterprises$rev_stratum <- predict(
  bounds,
  ken_enterprises$revenue_millions,
  labels = labels
)
table(ken_enterprises$rev_stratum)
```

Use the stratum allocations from `strata_bound()` directly as a named vector for `draw()`. The `as.data.frame()` method exposes the boundary table. This preserves the joint optimization between boundaries and allocation:

```{r}
n_alloc <- setNames(ceiling(as.data.frame(bounds)$n), labels)
n_alloc
```

```{r}
samp <- sampling_design() |>
  stratify_by(rev_stratum) |>
  draw(n = n_alloc) |>
  execute(ken_enterprises, seed = 4)
samp |>
  count(rev_stratum, name = "n_sampled")
```

Note that the boundaries and allocation are coupled. The `cumrootf` method optimizes them together to achieve the target CV. Using `strata_bound()` for boundaries but a different allocation in `samplyr` (for example, proportional instead of Neyman) might break this optimality. If you need a different allocation method, compute boundaries and allocation separately.

## Stratified Allocation

When strata are defined by administrative or domain boundaries rather than
optimal boundaries from `strata_bound()`, use `n_alloc()` to compute the
allocation. It supports Neyman, proportional, optimal (cost-weighted), and
power allocation with optional constraints on minimum stratum size.

First, summarize the frame to get stratum population sizes and variabilities:

```{r}
alloc_frame <- ken_enterprises |>
  summarize(
    N = n(),
    sd = sd(revenue_millions),
    mean = mean(revenue_millions),
    .by = sector
  ) |>
  rename(stratum = sector)
alloc_frame
```

Then compute the allocation. Here we use Neyman allocation targeting 600 units
with a minimum of 20 per stratum:

```{r}
alloc <- n_alloc(alloc_frame, n = 600, alloc = "neyman", min_n_stratum = 20)
alloc
as.data.frame(alloc)[, c("stratum", "N", "n_int", "weight")]
```

The result passes directly to `draw()`. The per-stratum allocations are
matched by stratum name to the values of the stratification variable:

```{r}
samp <- sampling_design() |>
  stratify_by(sector) |>
  draw(n = alloc) |>
  execute(ken_enterprises, seed = 10)
samp |>
  count(sector, name = "n_sampled")
```

Because `n_alloc()` already provides per-stratum sizes, do not combine it
with `alloc` in `stratify_by()`. The two allocation mechanisms are
mutually exclusive: either let `svyplan` allocate (via `n_alloc()`) or let
`samplyr` allocate (via `alloc` in `stratify_by()`), but not both.

### Budget-constrained allocation

When field costs vary across strata, `n_alloc()` can optimize under a budget
constraint using cost-weighted (optimal) allocation:

```{r}
alloc_frame_cost <- alloc_frame |>
  mutate(unit_cost = c(120, 200, 150, 130, 80, 100, 90))

budget_alloc <- n_alloc(
  alloc_frame_cost,
  budget = 50000,
  alloc = "optimal",
  min_n_stratum = 20
)
budget_alloc
as.data.frame(budget_alloc)[, c("stratum", "N", "unit_cost", "n_int", "weight")]
```

```{r}
samp_budget <- sampling_design() |>
  stratify_by(sector) |>
  draw(n = budget_alloc) |>
  execute(ken_enterprises, seed = 11)
samp_budget |>
  count(sector, name = "n_sampled")
```

## Stratified Two-Stage Designs

The workhorse of household surveys stratifies by region, samples
enumeration areas within each region, then households within each EA.
`n_alloc()` plans this design when the frame carries a `icc_psu`
column. The whole plan then drives both stages of the samplyr design.

Build the planning frame from the EA frame's own aggregates. The
homogeneity `icc_psu` typically comes from a previous survey round
(via `varcomp()` with its `strata` argument). Here we use a flat 0.05:

```{r}
frame_bfa <- bfa_eas |>
  summarise(
    N = sum(households),
    cost_psu = round(mean(fieldwork_cost)),
    .by = region
  ) |>
  left_join(bfa_eas_variance, by = "region") |>
  mutate(
    stratum = region,
    sd = sqrt(var),
    icc_psu = 0.05,
    cost_ssu = 15
  )

plan_2stage <- n_alloc(
  frame_bfa[, c("stratum", "N", "sd", "icc_psu", "cost_psu", "cost_ssu")],
  n = 6000,
  alloc = "neyman"
)
head(as.data.frame(plan_2stage)[, c("stratum", "n_int", "n_psu_int", "n_per_psu_int")])
```

Each region gets a whole-unit cluster take (`n_per_psu_int`) and a
number of EAs (`n_psu_int`), integerized together so their product
stays on target. Execute stage 1 against the EA frame, then stage 2
within each EA. Note the `cluster_by(ea_id)` even though rows already
are EAs: it declares the cluster structure the plan assumes.

```{r}
design_2stage <- sampling_design() |>
  stratify_by(region) |>
  cluster_by(ea_id) |>
  draw(n = plan_2stage) |>
  add_stage() |>
  stratify_by(region) |>
  draw(n = plan_2stage)

eas_selected <- execute(design_2stage, bfa_eas, stages = 1, seed = 7)
nrow(eas_selected)
```

Stage 2 runs later as a continuation, once households have been listed
in the selected EAs (the listing must carry the `region` column):

```{r, eval = FALSE}
eas_selected |>
  execute(household_listing, seed = 8)
```

Two conventions to keep in mind. First, the meaning of `n` shifts
across stages: at stage 1 a value is the stratum *total* of PSUs, while
at stage 2 it is the take *per PSU*. Second, per-stratum values at
stage 2 need the explicit `stratify_by(region)` shown above. Strata do
not carry over from stage 1, and the second `stratify_by()` only routes
the takes (every EA lies in one region), so weights are unaffected.

## Two-Phase Designs

When the variable you care about is expensive to measure but a cheap
screener can classify units into useful strata, a two-phase design pays
for the screener out of the savings. `n_twophase()` plans one: a large
phase-1 sample that classifies, and a stratified subsample of it that
carries the expensive measurement.

Here a short screening call establishes each enterprise's sector, and the
full financial interview is taken from a subsample. The phase-2 frame
describes what that interview costs and how variable revenue is inside
each sector:

```{r}
frame_twophase <- ken_enterprises |>
  summarize(
    N = n(),
    sd = sd(revenue_millions),
    mean = mean(revenue_millions),
    .by = sector
  ) |>
  rename(stratum = sector) |>
  mutate(unit_cost = c(120, 200, 150, 130, 80, 100, 90))

plan2 <- n_twophase(
  frame_twophase,
  phase1_cost = 5,
  budget = 60000,
  N = nrow(ken_enterprises)
)
plan2
```

The allocation is a subsampling fraction per stratum. Sectors where
revenue varies a lot, or where the interview is cheap, are carried into
phase 2 at a higher rate; `nu = 1` means the screener found so few units
that all of them are worth interviewing:

```{r}
as.data.frame(plan2$detail)[, c("stratum", "share", "nu", "n_issued")]
```

Phase 1 is an ordinary design executed against the frame. Give it a unit
identifier with `cluster_by()`: the two phases have to share one, or the
export in the next step cannot link them.

```{r}
phase1 <- sampling_design() |>
  cluster_by(enterprise_id) |>
  draw(n = plan2$operational$n[["n_phase1"]]) |>
  execute(ken_enterprises, seed = 2026)

phase1 |> count(sector, name = "screened")
```

Phase 2 is a *new* design executed with the phase-1 sample as its frame.
Pass the subsampling fractions with `frac`, not fixed counts: `nu` is a
share of whatever phase 1 actually found in each stratum, and what it
found is random. A take-all stratum needs the entire realized pool, so a
fixed count would ask for more units than the screener produced roughly
half the time.

```{r}
nu <- setNames(plan2$detail$nu, plan2$detail$stratum)

phase2 <- sampling_design() |>
  stratify_by(sector) |>
  cluster_by(enterprise_id) |>
  draw(frac = nu) |>
  execute(phase1, seed = 7)

phase2 |> count(sector, name = "interviewed")
```

Weights compound across the phases, so they still sum to the frame:

```{r}
c(sum_of_weights = sum(phase2$.weight), frame_size = nrow(ken_enterprises))
```

`as_svydesign()` recognizes the previous-phase link and exports through
`survey::twophase()`, so the two-phase variance is computed rather than
approximated by treating phase 2 as a simple sample:

```{r}
svy2 <- as_svydesign(phase2)
class(svy2)
survey::svymean(~revenue_millions, svy2)
```

The realized standard error should land near the coefficient of variation
the plan was built for, but only on average. The planned figure is a
property of the design; the realized one is an estimate from the single
sample you drew, and on a skewed outcome it moves a good deal from draw to
draw. Repeating this design over many seeds gives a spread of realized
coefficients of variation centered close to the planned one.

```{r}
est <- survey::svymean(~revenue_millions, svy2)
c(
  planned_cv = plan2$cv,
  realized_cv = as.numeric(survey::SE(est) / coef(est))
)
```

## Design Effects After Sampling

After executing a design, `design_effect()` and `effective_n()` report what
the weights cost you. These are re-exported from `svyplan` and work directly
on `tbl_sample` objects.

### Weighting Loss

The `tbl_sample` methods report the weighting loss: how much precision the
realized weights cost relative to a self-weighting sample of the same size.
This is Kish's design effect, computed from `.weight`, so it equals 1 for a
self-weighting design and rises with weight variability:

```{r}
design_effect(samp)
effective_n(samp)
```

The effective sample size is the number of observations an SRS would need to
achieve the same precision.

This is one component of a design effect, not a substitute for one.
Clustering and stratification also move precision, and neither is visible in
the weights, so neither can be read off the sample without an outcome. For a
design effect that reflects them, hand the design to survey and let the
estimator do it:

```{r, eval = FALSE}
svy <- as_svydesign(samp)
survey::svymean(~revenue_millions, svy, deff = TRUE)
```

### Cluster Planning

Before collecting data, you can anticipate the design effect for a cluster design using the survey-planning homogeneity measure `icc` and cluster size. This `icc` is not a generic mixed-model ICC. The calculation uses `svyplan::design_effect()` directly, not a `tbl_sample` method:

```{r}
design_effect(icc = 0.05, n_per_psu = 10)
design_effect(icc = 0.15, n_per_psu = 10)
```

With an `icc` of 0.05 and 10 units per cluster, the design effect is 1.45. An `icc` of 0.15 raises it to 2.35. This helps decide the number of clusters and units per cluster during the planning stage.

## Planning the Next Wave

An executed wave records a frame digest: the population counts and the realized allocation the selection resolved. `frame_summary()` returns them as tibbles, and its pool detail is a valid `frame` input for `n_alloc()`. The next wave can therefore be planned from the previous one, without going back to the sampling frame.

```{r}
wave1 <- sampling_design(title = "Wave 1") |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 300) |>
  execute(ken_enterprises, seed = 71)

pools <- frame_summary(wave1, detail = "pool")
pools
```

The digest deliberately records no variances, costs, or homogeneity measures. Those are planning assumptions, not sampling facts, so they stay explicit inputs. With a constant `sd` the allocation is proportional to size. Replace it with prior-wave estimates when you have them.

```{r}
plan_frame <- pools |>
  transmute(stratum = as.character(region), N, sd = 1)

wave2_alloc <- n_alloc(plan_frame, n = 450)

wave2 <- sampling_design(title = "Wave 2") |>
  stratify_by(region) |>
  draw(n = wave2_alloc) |>
  execute(ken_enterprises, seed = 72)
wave2
```

A design file written from an executed sample carries the digest inside its execution receipt, so the same loop works from the file alone. The confidential frame never has to travel with it:

```{r}
path <- tempfile(fileext = ".json")
write_design(wave1, path, frame = ken_enterprises)

restored <- read_design(path)
frame_summary(restored)
```

```{r, include = FALSE}
unlink(path)
```

## Summary

The typical planning workflow can be:

1. Determine sample size with `n_prop()`, `n_mean()`, `n_multi()`, or
   `power_prop()` / `power_mean()`.
2. Evaluate precision at the chosen *n* with `prec_prop()`, `prec_mean()`.
3. Run sensitivity analysis with `predict()` to check robustness.
4. Optionally create stratification boundaries with `strata_bound()` and
   assign strata with `predict()`.
5. Allocate across strata with `n_alloc()` (Neyman, proportional, optimal,
   or power allocation with constraints).
6. For cluster designs, estimate variance components with `varcomp()` and
   optimize the allocation with `n_cluster()`.
7. When a cheap screener can classify units into strata that an expensive
   measurement then subsamples, plan both phases with `n_twophase()`.
8. Pass the sample size or allocation to `draw()` and execute the design.
9. Read the realized weighting loss with `design_effect()` and
   `effective_n()`, and for a design effect that also reflects clustering and
   stratification, export with `as_svydesign()` and use
   `survey::svymean(deff = TRUE)`.

All `svyplan` sample size objects work directly with `draw()`, and the
handoff is stage-aware. `n_alloc()` results become per-stratum
allocations. Their stratified two-stage variant feeds PSU counts to a
clustered stage 1 and per-cluster takes to stage 2. `n_cluster()`
results contribute the matching stage value in a clustered design (and
their operational total in a flat single-stage one). `n_multi()` domain
plans become per-domain tables for a design stratified on the domain
variables. Scalar objects (`n_prop()`, `n_mean()`, `power_*()`)
contribute their operational total. The stable tabular form of any plan
is available via `as.data.frame()` when you need the numbers yourself.

`n_twophase()` is the exception to that handoff, because its allocation is
a set of subsampling fractions rather than a count. Execute phase 1 as an
ordinary design, then execute a second design against the phase-1 sample
with `draw(frac = )` carrying the plan's `nu`. Both phases need a shared
unit identifier for `as_svydesign()` to link them into a
`survey::twophase()` object.
