---
title: "Analyze a samplyr sample"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Analyze a samplyr sample}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
has_suggests <- requireNamespace("survey", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_suggests
)
```

```{r setup, message = FALSE, echo = FALSE}
library(samplyr)
library(dplyr)
library(survey)
```

## From selection to estimation

The usual workflow has four parts. Select the sample, collect outcomes, join
them to the selected units, and export the result to `survey` or `srvyr`.

`samplyr` records the strata, clusters, weights, and finite population
corrections during selection. A safe one-to-one join preserves that metadata.
An operation that adds, removes, or duplicates sampled rows changes the
realization and prevents export.

The main path reaches this article after `vignette("introduction")`, which
includes a three-stage household sample. The sections on joins, export and domains
apply to an ordinary sample. Read the later two-phase or overlapping-frame
examples when your design uses them.

We use a small stratified EA sample so the analysis examples run independently.
Sample inspection is covered in the "Inspecting the result" section of
`vignette("introduction")`.

```{r analysis-sample}
sample <- sampling_design(title = "Burkina Faso EA survey") |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 300) |>
  execute(bfa_eas, seed = 1)
```

`variance_estimators()` reports, from the design and the frame and before
anything is drawn, which variance estimators the design supports, which only
approximately, and which the export would refuse, with the condition each
would raise. A stratified first stage rules out survey's JK1, for instance.

```{r}
estimators <- variance_estimators(get_design(sample), bfa_eas)
estimators[c("estimator", "status", "class")]
```

## Join collected outcomes

The frame need not contain the survey outcomes. Join one collected record to
each sampled unit using a stable identifier.

```{r, message = FALSE}
library(dplyr)
set.seed(2)

collected <- tibble(
  ea_id = sample$ea_id,
  consumption_per_capita = round(rlnorm(
    nrow(sample),
    meanlog = 11,
    sdlog = 0.65
  )),
  below_threshold = rbinom(nrow(sample), size = 1, prob = 0.25)
)

sample_with_data <- sample |>
  left_join(collected, by = "ea_id")

class(sample_with_data)
```

The result remains a `tbl_sample` because the join is one to one and leaves
the selected rows intact. Check key uniqueness before joining when collected
data can contain repeated visits, roster members, or duplicate submissions.

### Nonresponse and calibration

Keep every issued row and record each unit's disposition next to its
outcomes. Missing outcomes do not change the selection weights, and neither
dropping nonrespondents before export nor `na.rm = TRUE` is a nonresponse
adjustment. The adjustment and any calibration belong to the analysis. Applied
to the exported design, they leave `.weight` holding the base weights while
the adjusted weights live in the design object.

```{r nonresponse, eval = has_suggests && requireNamespace("svrep", quietly = TRUE)}
set.seed(3)
issued <- sample_with_data |>
  mutate(disposition = if_else(runif(n()) < 0.85, "respondent",
                               "nonrespondent"))
issued |> as.data.frame() |> count(disposition)

set.seed(4)
issued_replicates <- as_svrepdesign(issued, type = "rwyb", replicates = 500)
adjusted <- svrep::redistribute_weights(
  issued_replicates,
  reduce_if = disposition == "nonrespondent",
  increase_if = disposition == "respondent",
  by = "region"
)
respondents <- subset(adjusted, disposition == "respondent")
```

The adjustment moves each nonrespondent's weight to the respondents of its
region, in every replicate as well as in the full-sample weights, so the
replicate variance carries it. It assumes that response is unrelated to the
outcome within a region, which the analyst has to justify. `survey` warns
that the respondent subset has fewer rows than the design's degrees of
freedom. The replicate variance is computed as for any domain and is not
affected. Calibration then matches the respondents to known population
totals, here the number of urban EAs in the frame.

```{r calibration, eval = has_suggests && requireNamespace("svrep", quietly = TRUE)}
ea_totals <- c(`(Intercept)` = nrow(bfa_eas),
               urban_ruralUrban = sum(bfa_eas$urban_rural == "Urban"))
calibrated <- calibrate(respondents, ~urban_rural, population = ea_totals)
svymean(~consumption_per_capita, calibrated)
```

`weights(calibrated)` returns the adjusted and calibrated weights, while
`issued$.weight` keeps the base weights, so the two stay apart for reporting.
The example uses RWYB replicates because svrep's weight adjustments act on a
replicate design. `?variance-estimation` gives what RWYB assumes for each
selection method.

How each selection method's variance is estimated is in `?variance-estimation`,
and `?as_svydesign` and `?as_svrepdesign` give the export restrictions.

## Export to survey

`as_svydesign()` maps the recorded design to a `survey` object.

```{r}
library(survey)
svy <- as_svydesign(sample_with_data)
svy
```

The collected variables can now be analyzed with standard `survey` functions.

```{r}
svymean(~consumption_per_capita, svy)
svymean(~below_threshold, svy)
svytotal(~population, svy)
```

For a multi-stage sample, every represented stage contributes its sampling-unit
and finite-population terms. This preserves the stage structure. It does not
turn an approximate stage variance treatment into an exact one.

Fixed-size PPS methods without replacement use Brewer's approximation by
default. In that phrase,
Brewer names the `survey` variance approximation and not necessarily the
selection algorithm. The approximation's accuracy depends on the population
and design. Exact joint inclusion probabilities can be supplied for methods
that provide them.

Systematic sampling uses an SRS-style variance approximation, and systematic
PPS uses Brewer's. Neither sees frame order, so the direction of the error
depends on it. `as_svydesign()` warns unless the caller explicitly accepts or
refuses the approximation through `systematic_variance`.

### Strata with one sampled unit

A stratum that ends up with one sampled unit outside certainty, because its
allocation was one or because certainty units took the rest, has no
within-stratum variance estimate. `execute()` says so when it selects such a
stratum, and `as_svydesign()` warns again at export
(`samplyr_warning_lonely_psu`) while `survey` is left at its default
`survey.lonely.psu = "fail"`, under which estimation would stop. Setting the
option, as below, also silences that export warning.

```{r lonely}
one_per_region <- sampling_design() |>
  stratify_by(region) |>
  draw(n = 1) |>
  execute(bfa_eas, seed = 5)

old <- options(survey.lonely.psu = "adjust")
svytotal(~households, as_svydesign(one_per_region))
options(old)
```

`"adjust"` centers a lonely unit at the population mean, which is
conservative. Collapsing each such stratum with a similar one before export
is the other usual choice, and it is a design decision the analyst has to
make. Allocating at least two units outside certainty to every stratum avoids
the problem at the source. RWYB replicates refuse a singleton whose variance
is needed, except under Poisson sampling and, with `lonely.psu =
"certainty"`, at the final stage. `?variance-estimation` gives the certainty case.

## Estimate a domain correctly

Convert the complete sample first, then subset the survey design.

```{r}
urban_svy <- subset(svy, urban_rural == "Urban")
svymean(~consumption_per_capita, urban_svy)
```

Filtering the `tbl_sample` first treats the observed domain sample as fixed.
The point estimate can agree, but the variance estimate is generally wrong and
is often too small because domain membership is random under the design
[@sarndal1992, ch. 5.8].

```{r, error = TRUE}
urban_rows <- sample_with_data |>
  filter(urban_rural == "Urban")

as_svydesign(urban_rows)
```

One-to-one joins, row reordering, and new ordinary columns remain supported.
Changing the row set or an internal column such as `.weight` or `.fpc_1` does
not.

With `srvyr`, use the same order. Convert first and then filter the survey
object.

```{r, eval = has_suggests && requireNamespace("srvyr", quietly = TRUE)}
library(srvyr)

sample_with_data |>
  as_survey_design() |>
  filter(urban_rural == "Urban") |>
  summarise(
    mean_consumption = survey_mean(consumption_per_capita)
)
```

For a strictly positive domain total, a log-scale delta interval is an
available finite-sample sensitivity analysis. It keeps the endpoints positive
and allows asymmetric uncertainty.

```{r}
urban_total <- svytotal(~population, urban_svy)
urban_estimate <- unname(coef(urban_total)[[1]])
urban_se <- unname(SE(urban_total)[[1]])
z <- qnorm(0.975)

exp(log(urban_estimate) + c(-1, 1) * z * urban_se / urban_estimate)
```

This transformation does not repair a poor variance estimator, and it
applies only when the estimate is positive. It helps where the sampling
distribution of a positive total is skewed, as for a small domain, and
matters little for a full-population total. For a skewed ratio, neither a
log-ratio nor a Fieller interval is uniformly reliable, and the ratio
estimator itself can carry material finite-sample bias.

## Replicate weights and PPS variance

For a supported single-phase design, `as_svrepdesign()` creates replicate
weights through `survey::as.svrepdesign()`, through the optional **svrep**
package when `type = "rwyb"` is requested, or from the replicates of
`execute(reps = R)` when `type = "random_groups"` is requested.

```{r}
replicate_design <- as_svrepdesign(sample, type = "auto")
svymean(~households, replicate_design)
```

Replicate methods are not automatic proofs of exactness. Generic PPS
bootstraps do not recreate spatial selection, hard cube constraints, or a
random Poisson sample size. `as_svydesign()` and `as_svrepdesign()` report or
refuse combinations that cannot be represented under their documented
contracts. `?variance-estimation` gives how each selection method's variance
is estimated and how exact it is.

For independent Poisson selection, including clustered and multi-stage
designs, the generic replicate types, `"auto"` included, refuse the sample,
so request Rao-Wu-Yue-Beaumont replicates explicitly:

```{r, eval = has_suggests && requireNamespace("svrep", quietly = TRUE)}
poisson_sample <- sampling_design() |>
  draw(frac = 0.4, method = "bernoulli") |>
  execute(data.frame(id = 1:100, y = 1:100), seed = 12)
set.seed(42)
poisson_replicates <- as_svrepdesign(
  poisson_sample, type = "rwyb", replicates = 1000
)
svytotal(~y, poisson_replicates)
```

RWYB retains the variation in Poisson sample size. It also supports SRS,
independent draws with replacement, and Brewer, CPS, Sampford and systematic
PPS stages, alone or mixed across stages. PPS WOR variance is an
approximation and is reported as such. Increase the replicate count to reduce
simulation error. svrep's `estimate_boot_sim_cv()` helps assess it for chosen
estimates. No new sampling verbs are needed.

Two-phase replicate export is not yet supported. Multi-stage RWYB export
requires every selected parent to remain represented in the final sample.
Later Poisson stages require a complete frame digest to check this. A
noncertainty singleton is refused when its variance is not estimable, while
a singleton Poisson sample can contribute a valid variance estimate.

When exporting a materialized wave or a second-phase sample with
`as_svydesign()`, current analysis columns take precedence over same-named
columns in the retained first-phase sample. Missing current measurements
remain missing. Keep earlier measurements under separate names if you need
both. Identifiers and sampling metadata retain their design meanings.

## Two phases: screening followed by a detailed interview

A second phase samples units already selected at phase 1, using information
collected in that phase. It is a new selection law on the realized sample.
A later stage in a household design instead samples children within selected
parents. Continuing that design with a household listing is not a new phase.

The bundled `ken_enterprises` is entirely synthetic. Here a cheap screening
interview determines phase-two strata, and only the subsample receives a
detailed interview. The stable `enterprise_id` links the phases. At phase 1,
`cluster_by(enterprise_id)` declares that identity. Each such unit has one row.

```{r screening}
screening_design <- sampling_design("Establishment screening") |>
  cluster_by(enterprise_id) |>
  draw(n = 500)

phase1 <- execute(screening_design, ken_enterprises, seed = 71)
set.seed(72)
phase1 <- phase1 |>
  mutate(
    revenue_baseline = revenue_millions,
    screen_score = log1p(employees) + rnorm(n(), sd = 0.5),
    priority = if_else(screen_score >= 4, "high", "ordinary")
  )
phase1 |> as.data.frame() |> count(priority)
```

The synthetic screening variables are observed for every phase-one unit,
including those that will not enter phase 2. They remain in the retained
phase-one sample so the two-phase export can reconstruct the selection pools.

```{r detailed-interview}
interview_design <- sampling_design("Detailed establishment interview") |>
  stratify_by(priority) |>
  draw(n = c(high = 60, ordinary = 90))

phase2 <- execute(interview_design, phase1, seed = 73)

set.seed(74)
phase2 <- phase2 |>
  mutate(
    revenue_millions = revenue_baseline * exp(rnorm(n(), 0, 0.15)),
    audit_hours = rgamma(n(), shape = 3, rate = 0.5)
  )
```

`audit_hours` is a new phase-two measurement. `revenue_millions` is an updated
measurement with the same name as its baseline counterpart. Export uses the
current phase-two values. `revenue_baseline` keeps the earlier values available
explicitly. All detailed interviews are complete in this example.

```{r two-phase-estimation}
two_phase_svy <- as_svydesign(phase2)
svytotal(~revenue_millions + audit_hours, two_phase_svy)

# The export uses the current observations and the combined phase weights.
stopifnot(isTRUE(all.equal(
  unname(coef(svytotal(~revenue_millions, two_phase_svy))),
  sum(phase2$.weight * phase2$revenue_millions)
)))
```

These estimates illustrate inference for the synthetic establishment
population, not substantive estimates for Kenya. The example uses SRS at
phase 1 and stratified SRS at phase 2, a supported two-phase path. A phase 2
drawn by Sampford, CPS, Brewer, SPS or Pareto also exports, with joint
probabilities computed on the phase-1 sample. A phase 1 drawn with unequal
probabilities without replacement is refused, and so is Poisson selection at
either phase. Phase chains longer
than two and two-phase replicate export are currently unsupported. See
`?as_svydesign` before changing the selection methods.

## Joint information for PPS variance

For supported PPS methods, `joint_expectation()` reconstructs exact or
approximate second-order quantities according to the method.

```{r}
sampford_sample <- sampling_design() |>
  stratify_by(region) |>
  cluster_by(ea_id) |>
  draw(n = 5, method = "pps_sampford", mos = households) |>
  execute(bfa_eas, seed = 2025)

joints <- joint_expectation(sampford_sample, bfa_eas, stages = 1)

exact_pps_design <- as_svydesign(
  sampford_sample,
  pps = ppsmat(joints[[1]])
)

svymean(~households, exact_pps_design)
```

The Sampford matrix above is exact. CPS, systematic PPS, and Poisson selection
also have exact second-order calculations in the supported cases.
Generalized Brewer, SPS, Pareto, and unconstrained cube use the documented
high-entropy approximation. Spatial methods and bounded cube do not provide a
joint matrix. `?joint_expectation` is the authoritative method table.

An exact systematic PPS matrix can contain zero pair probabilities. Exact
design quantities do not then guarantee an estimable or stable sample variance
through `survey::ppsmat()`. A sampled matrix cannot show whether unseen
population pairs have zero probability, so `as_svydesign()` warns when a
`ppsmat` object is supplied for systematic PPS. It does not refuse the route,
because full pair positivity can hold at high sampling fractions.

A user-supplied `pps` matrix describes stage 1. For a multi-stage sample it is
applied to stage 1 alone, with a warning, and only when the sample has one row
per stage-1 unit. Otherwise it is refused (`samplyr_error_pps_rows_per_psu`).
Omit it when the intended analysis is the full multi-stage linearization with
the default stage-level variance treatment.

## Frames that overlap on one population

Two registers of the same population, each incomplete: a landline frame and a
cell frame, an area frame and a list frame. The samples are selected
independently, one design and one seed each, and everything the feature
contains happens after that.

`stack_frames()` records which frames each sampled unit belongs to. Every
component has to carry every frame's membership column, not only its own,
because a unit selected from frame A has to say whether it was also listed in
frame B, and that is the information the compositing needs.

```{r frames}
population <- data.frame(
  person_id = 1:600,
  spend = round(stats::rlnorm(600, log(120), 0.4), 2),
  in_landline = rep(c(TRUE, FALSE), times = c(400, 200)),
  in_cell = rep(c(FALSE, TRUE), times = c(150, 450))
)

frames <- stack_frames(
  landline = sampling_design() |>
    draw(n = 80) |>
    execute(population[population$in_landline, ], seed = 1),
  cell = sampling_design() |>
    draw(n = 120) |>
    execute(population[population$in_cell, ], seed = 2),
  membership = c(landline = "in_landline", cell = "in_cell"),
  key = person_id
)

summary(frames)
```

The domain counts are rows, not distinct people. Someone listed in both frames
and selected from both appears twice, once per frame that selected them, and
that is exactly the double count a composite weight has to resolve.

### Two export routes

`as_svydesign()` hands the stack to `survey::multiframe()`. Each component is
exported on its own, so it keeps its strata, clusters and variance treatment,
and the compositing sits on top.

```{r frames-linearized}
svytotal(~spend, as_svydesign(frames))
```

`theta = NULL`, the default, is the multiplicity estimator: a unit reached by
both frames contributes half its weight through each. An explicit `theta` is
Hartley's constant factor [@hartley1962], and it belongs to the **first** frame
of the stack, so reversing the components and wanting the same estimator means
asking for `1 - theta`.

```{r frames-hartley}
svytotal(~spend, as_svydesign(frames, theta = 0.74))
```

The expected multiframe estimator instead needs each unit's selection
probability in every frame containing it. Supply the original registers through
`overlaps = exante_overlaps(list(landline = landline_register,
cell = cell_register), by = c(person_id = "person_id"))` when building the stack, using the stack's
component names and key, then export with `estimator = "expected"`.
The resolver requires exact inclusion probabilities by default. SPS and Pareto
targets require `allow_approximate = TRUE` in `exante_overlaps()`. This accepts
an approximation that can affect design unbiasedness. The stack records quality
by frame, and exported designs retain it in their
`samplyr_overlap_probability_quality` attribute. For inspecting one register,
`exante_probabilities()` returns the same quality beside each probability.

`samplyr` never forwards a missing `theta` to `survey`. Left unset there, it
defaults to the ratio of the frames' mean sampling weights, which is a
data-dependent stand-in for frame size over sample size rather than a neutral
choice.

The replicate route builds one system per frame and combines them in blocks: in
a replicate column belonging to one frame, only that frame varies and every
other stays at its full-sample weight. The combined variance is then the sum of
the frames' own contributions, which is what independent selection from each
frame gives.

```{r frames-replicate}
svytotal(~spend, as_svrepdesign(frames, type = "bootstrap", replicates = 100))
```

The two routes are not interchangeable, and the differences are the reason to
know which one you are on.

| | `as_svydesign()` | `as_svrepdesign()` |
|---|---|---|
| Number of frames | Two. `survey::multiframe()` composites two | Any number |
| Compositing | Multiplicity, or Hartley with a stated `theta` | The same, with `theta` above two frames refused |
| A component with shared weights | Refused | Supported |
| Replicate methods | Not applicable | May differ between frames. `type = "auto"` picks one per component |

The two-frame limit belongs to the export and is stated there, not at the
design layer: `stack_frames()` records any number, exactly as `execute()`
chains any number of phases while the two-phase export stops at two.

### Where the theory comes from

The multiplicity estimator for overlapping frames is the generalized weight
share method with the standardized link matrix, because the number of links to
a unit is the number of frames containing it [@lohr2021, section 3.2;
@mecatti2007]. That is why a stack composed of frames and a sample whose
weights were shared through a link table are handled by one mechanism rather
than two, and why `vignette("design-semantics")` documents them together.

## Analysis checklist

1. Inspect the selected counts and weights before fieldwork.
2. Verify that collected-data keys are unique at the sampled-unit level.
3. Join outcomes without changing the selected row set.
4. Convert the complete sample to a survey design.
5. Define domains on the converted design.
6. State every variance approximation used by an advanced selection method.
7. Where the estimates combine frames, say which compositing factor was used
   and which frame an explicit `theta` was applied to.

Use `vignette("serialization")` to archive the selection and
`vignette("design-semantics")` for the detailed export contracts, the weight
share method, and the contract separating design weights from estimation
weights.

## References
