---
title: "Analyze a samplyr Sample"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Analyze a samplyr Sample}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, echo = FALSE}
library(samplyr)
library(dplyr)
library(survey)
```

## From selection to estimation

The usual workflow has four parts. Select the sample, collect outcomes, join
them to the selected units, and export the result to `survey` or `srvyr`.

`samplyr` records the strata, clusters, weights, and finite population
corrections during selection. A safe one-to-one join preserves that metadata.
An operation that adds, removes, or duplicates sampled rows changes the
realization and prevents export.

## Inspect the sample before fieldwork

We begin with a stratified sample of enumeration areas.

```{r}
data(bfa_eas)

sample <- sampling_design(title = "Burkina Faso EA survey") |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 300) |>
  execute(bfa_eas, seed = 1)

summary(sample)
```

`frame_summary()` exposes the same realization at pool level.

```{r}
frame_summary(sample, detail = "pool") |>
  select(region, N, n_target, n_expected, n_realized, take_rate)
```

For this feasible fixed-size design, the target, expectation, and realized
take coincide. Under Bernoulli or Poisson selection the realized count is
random. Under unequal-probability selection, `take_rate` describes the
realized pool count and is not a unit-level inclusion probability.

## Join collected outcomes

The frame need not contain the survey outcomes. Join one collected record to
each sampled unit using a stable identifier.

```{r}
library(dplyr)
set.seed(2)

collected <- tibble(
  ea_id = sample$ea_id,
  consumption_per_capita = round(rlnorm(
    nrow(sample),
    meanlog = 11,
    sdlog = 0.65
  )),
  below_threshold = rbinom(nrow(sample), size = 1, prob = 0.25)
)

sample_with_data <- sample |>
  left_join(collected, by = "ea_id")

class(sample_with_data)
```

The result remains a `tbl_sample` because the join is one to one and leaves
the selected rows intact. Check key uniqueness before joining when collected
data can contain repeated visits, roster members, or duplicate submissions.

## Export to survey

`as_svydesign()` maps the recorded design to a `survey` object.

```{r}
library(survey)
svy <- as_svydesign(sample_with_data)
svy
```

The collected variables can now be analyzed with standard `survey` functions.

```{r}
svymean(~consumption_per_capita, svy)
svymean(~below_threshold, svy)
svytotal(~population, svy)
```

For a multistage sample, every represented stage contributes its sampling-unit
and finite-population terms. This preserves the stage structure. It does not
turn an approximate stage variance treatment into an exact one.

Fixed-size PPS methods use Brewer's approximation by default. In that phrase,
Brewer names the `survey` variance approximation and not necessarily the
selection algorithm. The approximation's accuracy depends on the population
and design. Exact joint inclusion probabilities can be supplied for methods
that provide them.

Equal-probability systematic sampling uses an SRS-style variance approximation.
Its direction depends on frame ordering. `as_svydesign()` warns unless the
caller explicitly accepts or refuses that approximation through
`systematic_variance`.

## Estimate a domain correctly

Convert the complete sample first, then subset the survey design.

```{r}
urban_svy <- subset(svy, urban_rural == "Urban")
svymean(~consumption_per_capita, urban_svy)
```

Filtering the `tbl_sample` first treats the observed domain sample as fixed.
The point estimate can agree, but the variance estimate is generally wrong and
is often too small because domain membership is random under the design
[@sarndal1992, ch. 5.8].

```{r, error = TRUE}
urban_rows <- sample_with_data |>
  filter(urban_rural == "Urban")

as_svydesign(urban_rows)
```

One-to-one joins, row reordering, and new ordinary columns remain supported.
Changing the row set or an internal column such as `.weight` or `.fpc_1` does
not.

With `srvyr`, use the same order. Convert first and then filter the survey
object.

```{r, eval=requireNamespace("srvyr", quietly = TRUE)}
library(srvyr)

sample_with_data |>
  as_survey_design() |>
  filter(urban_rural == "Urban") |>
  summarise(
    mean_consumption = survey_mean(consumption_per_capita)
  )
```

## Replicate weights and PPS variance

For a supported single-phase design, `as_svrepdesign()` creates replicate
weights through `survey::as.svrepdesign()`.

```{r}
replicate_design <- as_svrepdesign(sample, type = "auto")
svymean(~households, replicate_design)
```

Replicate methods are not automatic proofs of exactness. Generic PPS
bootstraps do not recreate spatial selection, hard cube constraints, or a
random Poisson sample size. `as_svydesign()` and `as_svrepdesign()` report or
refuse combinations that cannot be represented under their documented
contracts. See their reference pages before selecting a variance route for an
advanced design.

For supported PPS methods, `joint_expectation()` reconstructs exact or
approximate second-order quantities according to the method.

```{r}
sampford_sample <- sampling_design() |>
  stratify_by(region) |>
  cluster_by(ea_id) |>
  draw(n = 5, method = "pps_sampford", mos = households) |>
  execute(bfa_eas, seed = 2025)

joints <- joint_expectation(sampford_sample, bfa_eas, stages = 1)

exact_pps_design <- as_svydesign(
  sampford_sample,
  pps = ppsmat(joints[[1]])
)

svymean(~households, exact_pps_design)
```

The Sampford matrix above is exact. CPS, systematic PPS, and Poisson selection
also have exact second-order calculations in the supported cases.
Generalized Brewer, SPS, Pareto, and unconstrained cube use the documented
high-entropy approximation. Spatial methods and bounded cube do not provide a
joint matrix. `?joint_expectation` is the authoritative method table.

A user-supplied `pps` matrix describes one stage. For a multistage design,
supplying it exports only the represented PPS stage. Omit it when the intended
analysis is the full multistage linearization with the default stage-level
variance treatment.

## Frames that overlap on one population

Two registers of the same population, each incomplete: a landline frame and a
cell frame, an area frame and a list frame. The samples are selected
independently, one design and one seed each, and everything the feature
contains happens after that.

`stack_frames()` records which frames each sampled unit belongs to. Every
component has to carry every frame's membership column, not only its own,
because a unit selected from frame A has to say whether it was also listed in
frame B, and that is the information the compositing needs.

```{r frames}
population <- data.frame(
  person_id = 1:600,
  spend = round(stats::rlnorm(600, log(120), 0.4), 2),
  in_landline = rep(c(TRUE, FALSE), times = c(400, 200)),
  in_cell = rep(c(FALSE, TRUE), times = c(150, 450))
)

frames <- stack_frames(
  landline = sampling_design() |>
    draw(n = 80) |>
    execute(population[population$in_landline, ], seed = 1),
  cell = sampling_design() |>
    draw(n = 120) |>
    execute(population[population$in_cell, ], seed = 2),
  membership = c(landline = "in_landline", cell = "in_cell"),
  key = person_id
)

summary(frames)
```

The domain counts are rows, not distinct people. Someone listed in both frames
and selected from both appears twice, once per frame that selected them, and
that is exactly the double count a composite weight has to resolve.

### Two export routes

`as_svydesign()` hands the stack to `survey::multiframe()`. Each component is
exported on its own, so it keeps its strata, clusters and variance treatment,
and the compositing sits on top.

```{r frames-linearized}
svytotal(~spend, as_svydesign(frames))
```

`theta = NULL`, the default, is the multiplicity estimator: a unit reached by
both frames contributes half its weight through each. An explicit `theta` is
Hartley's constant factor [@hartley1962], and it belongs to the **first** frame
of the stack, so reversing the components and wanting the same estimator means
asking for `1 - theta`.

```{r frames-hartley}
svytotal(~spend, as_svydesign(frames, theta = 0.74))
```

`samplyr` never forwards a missing `theta` to `survey`. Left unset there, it
defaults to the ratio of the frames' mean sampling weights, which is a
data-dependent stand-in for frame size over sample size rather than a neutral
choice.

The replicate route builds one system per frame and combines them in blocks: in
a replicate column belonging to one frame, only that frame varies and every
other stays at its full-sample weight. The combined variance is then the sum of
the frames' own contributions, which is what independent selection from each
frame gives.

```{r frames-replicate}
svytotal(~spend, as_svrepdesign(frames, type = "bootstrap", replicates = 100))
```

The two routes are not interchangeable, and the differences are the reason to
know which one you are on.

| | `as_svydesign()` | `as_svrepdesign()` |
|---|---|---|
| Number of frames | Two. `survey::multiframe()` composites two | Any number |
| Compositing | Multiplicity, or Hartley with a stated `theta` | The same, with `theta` above two frames refused |
| A component with shared weights | Refused | Supported |
| Replicate methods | Not applicable | May differ between frames; `type = "auto"` picks one per component |

The two-frame limit belongs to the export and is stated there, not at the
design layer: `stack_frames()` records any number, exactly as `execute()`
chains any number of phases while the two-phase export stops at two.

### Where the theory comes from

The multiplicity estimator for overlapping frames is the generalized weight
share method with the standardized link matrix, because the number of links to
a unit is the number of frames containing it [@lohr2021, section 3.2;
@mecatti2007]. That is why a stack composed of frames and a sample whose
weights were shared through a link table are handled by one mechanism rather
than two, and why `vignette("design-semantics")` documents them together.

## Analysis checklist

1. Inspect the selected counts and weights before fieldwork.
2. Verify that collected-data keys are unique at the sampled-unit level.
3. Join outcomes without changing the selected row set.
4. Convert the complete sample to a survey design.
5. Define domains on the converted design.
6. State every variance approximation used by an advanced selection method.
7. Where the estimates combine frames, say which compositing factor was used
   and which frame an explicit `theta` was applied to.

Use `vignette("serialization")` to archive the selection and
`vignette("design-semantics")` for the detailed export contracts, the weight
share method, and the contract separating design weights from estimation
weights.

## References
