---
title: "Plan a Sample with svyplan"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Plan a Sample with svyplan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, echo = FALSE}
library(samplyr)
library(svyplan)
library(dplyr)
```

## From a precision target to a draw

Planning answers how many units to issue and how to distribute them. Selection
then applies that plan to a frame. `svyplan` owns the first calculation and
`samplyr` owns the second.

We use the synthetic `ken_enterprises` frame. Suppose the main estimate is the
proportion of exporting enterprises, with a margin of error of three percentage
points at 95 percent confidence.

```{r}
data(ken_enterprises)

p_export <- mean(ken_enterprises$exporter)
target <- n_prop(p = p_export, moe = 0.03)
target
```

The result is a `svyplan_n` object. `draw()` reads its planned sample size
directly.

```{r}
sample <- sampling_design(title = "Enterprise survey") |>
  draw(n = target) |>
  execute(ken_enterprises, seed = 1)

nrow(sample)
```

Use a plausible planning value from earlier rounds or a pilot when the true
population value is unknown. For a proportion, `p = 0.5` is the conservative
choice under the ordinary Wald variance because it maximizes `p * (1 - p)`.

## Means, response, and design effects

For a mean, supply the population variance or a prior estimate of it.

```{r}
revenue_target <- n_mean(
  var = var(ken_enterprises$revenue_millions),
  moe = 30
)
revenue_target
```

Expected nonresponse increases the number issued. Here the target remains the
same, but only 85 percent of issued units are expected to respond.

```{r}
response_target <- n_prop(
  p = p_export,
  moe = 0.03,
  resp_rate = 0.85
)

c(
  full_response = as.integer(target),
  issued_at_85_percent = as.integer(response_target)
)
```

`deff` expresses the variance relative to simple random sampling under the
planning model. It can reflect clustering, unequal weights, or another design
feature, but the value should match the estimand being planned.

```{r}
n_prop(p = p_export, moe = 0.03, deff = 1.5, resp_rate = 0.85)
```

A cross-sectional design effect does not automatically describe a change or a
pooled longitudinal estimate. Correlation across occasions can help one
estimand and hurt another.

## Read the calculation in reverse

The `prec_*()` functions evaluate a sample size rather than solving for it.

```{r}
prec_prop(p = p_export, n = 600)
prec_mean(
  var = var(ken_enterprises$revenue_millions),
  n = 300,
  mu = mean(ken_enterprises$revenue_millions)
)
```

This is useful when budget fixes the take. `confint()` gives the corresponding
planning interval, while `predict()` can compare supported assumptions such as
response rates and design effects.

```{r}
plan <- n_prop(p = p_export, moe = 0.03, deff = 1.5)

predict(
  plan,
  expand.grid(
    deff = c(1, 1.5, 2),
    resp_rate = c(0.8, 0.9, 1)
  )
)
```

## Several indicators or domains

`n_multi()` finds the binding requirement across several means and
proportions.

```{r}
indicators <- data.frame(
  name = c("exporter_rate", "mean_revenue"),
  p = c(p_export, NA),
  var = c(NA, var(ken_enterprises$revenue_millions)),
  moe = c(0.03, 30)
)

multi_target <- n_multi(indicators)
multi_target
```

A domain-specific plan needs enough observations within every reporting
domain. When `n_multi()` returns a domain table, the sampling design needs a
matching `stratify_by()` so those counts have an operational meaning.

```{r}
domain_indicators <- data.frame(
  name = "exporter_rate",
  size_class = c("Small", "Medium", "Large"),
  p = c(0.10, 0.22, 0.45),
  cv = 0.10
)

domain_target <- n_multi(domain_indicators, domains = "size_class")

domain_sample <- sampling_design() |>
  stratify_by(size_class) |>
  draw(n = domain_target) |>
  execute(ken_enterprises, seed = 2)

count(domain_sample, size_class, name = "n_sampled")
```

## Allocate a total across strata

When strata are already defined, `n_alloc()` distributes a total using the
chosen planning criterion. This example uses Neyman allocation for mean
revenue, with at least 20 enterprises in every sector.

```{r}
allocation_frame <- ken_enterprises |>
  summarise(
    N = n(),
    sd = sd(revenue_millions),
    .by = sector
  ) |>
  transmute(stratum = sector, N, sd)

allocation <- n_alloc(
  allocation_frame,
  n = 600,
  alloc = "neyman",
  min_n_stratum = 20
)

as.data.frame(allocation)[, c("stratum", "N", "n_int", "weight")]
```

The result maps its stratum names to the values used by `stratify_by()`.

```{r}
allocated_sample <- sampling_design(title = "Allocated enterprise survey") |>
  stratify_by(sector) |>
  draw(n = allocation) |>
  execute(ken_enterprises, seed = 3)

count(allocated_sample, sector, name = "n_sampled")
```

Do not also request an allocation in `stratify_by()`. Either `svyplan` supplies
the stratum takes or `samplyr` computes them, not both.

Neyman allocation minimizes the planned variance for the variable represented
by `sd`, subject to the stated constraints. A cost-weighted allocation adds the
declared costs. Neither choice is automatically optimal for other variables or
domains.

## Allocate for several indicators at once

Given per-stratum moments for several indicators and a precision target for
each, `n_alloc()` finds the cheapest allocation that satisfies all of them
together. `measures` carries the moments, one row per stratum and indicator,
and `targets` carries one requirement per indicator.

```{r}
sector_moments <- ken_enterprises |>
  summarise(
    N = n(),
    mean = mean(revenue_millions),
    sd = sd(revenue_millions),
    p = mean(exporter),
    .by = sector
  )

joint_measures <- bind_rows(
  mutate(sector_moments, stratum = sector, name = "revenue", mean, sd, .keep = "none"),
  mutate(sector_moments, stratum = sector, name = "exporter", p, .keep = "none")
)

joint_plan <- n_alloc(
  transmute(sector_moments, stratum = sector, N, .keep = "none"),
  measures = joint_measures,
  targets = data.frame(
    name = c("revenue", "exporter"),
    cv = c(0.10, 0.05)
  )
)

joint_plan
```

One requirement determines the sample size and the others have slack. The
result reports which.

```{r}
joint_plan$constraints[, c("name", ".target", ".achieved", ".binding")]
```

The exporter proportion binds. Mean revenue was asked for a CV of 0.10 and the
design delivers about 0.089, because the sample the exporter target requires is
larger than the one revenue needs. A total sample size on its own would not
show that.

## Check a plan against the design that runs it

A plan states the precision it expects. Executing the design repeatedly against
a known population shows the precision it delivers. Comparing the two is the
most direct check that the variance model, the allocation, and the executed
design agree with each other.

```{r}
joint_design <- sampling_design(title = "Enterprise indicators") |>
  stratify_by(sector) |>
  draw(n = joint_plan)

estimate_once <- function(seed) {
  smp <- execute(joint_design, ken_enterprises, seed = seed)
  c(
    revenue = sum(smp$.weight * smp$revenue_millions) / sum(smp$.weight),
    exporter = sum(smp$.weight * smp$exporter) / sum(smp$.weight)
  )
}

replicates <- 200
estimates <- vapply(seq_len(replicates), estimate_once, numeric(2))
```

Fix the acceptance band before reading the result. The Monte Carlo error on an
estimated CV is about `1 / sqrt(2 * (replicates - 1))` in relative terms, so
three of those is a defensible band for a check of this size.

```{r}
truth <- c(
  revenue = mean(ken_enterprises$revenue_millions),
  exporter = mean(ken_enterprises$exporter)
)
tolerance <- 3 / sqrt(2 * (replicates - 1))

check <- data.frame(
  indicator = joint_plan$constraints$name,
  planned_cv = joint_plan$constraints$.achieved,
  empirical_cv = apply(estimates, 1, sd) / abs(rowMeans(estimates)),
  rel_bias = rowMeans(estimates) / truth[joint_plan$constraints$name] - 1,
  row.names = NULL
)
check$ratio <- check$empirical_cv / check$planned_cv
check$within_tolerance <- abs(check$ratio - 1) < tolerance

check
```

Compare each indicator against the precision the plan plans for it, which is
`.achieved`, and not against the target it was asked for. Revenue is checked
against 0.089 rather than 0.10 for the reason the previous section gives.

This validates the planning model, not the survey. The frame is used as its own
population, so response, coverage, and measurement play no part, and the
moments the plan was given are true by construction. It cannot tell you whether
those moments describe the population that will actually be surveyed.

## Plan a clustered design

Clustering usually increases variance because units in the same cluster are
similar. For a one-stage approximation with equal cluster size, the familiar
design effect is determined by the homogeneity measure `icc` and the take per
cluster.

```{r}
design_effect(icc = 0.05, n_per_psu = 10)
```

`n_cluster()` can trade the number of clusters against the number of units per
cluster under declared costs and homogeneity. Its result is optimal only under
that planning model.

```{r}
cluster_plan <- n_cluster(
  stage_cost = c(500, 50),
  icc = 0.05,
  budget = 30000
)

cluster_plan
as.integer(cluster_plan)
```

In a two-stage `samplyr` design, the cluster plan supplies the stage-1 cluster
count and the within-cluster take. See `?n_cluster` for the feasibility checks
and `vignette("introduction")` for multistage execution.

## Field a certainty-aware allocation

A few large PSUs often dominate a stratum. Under proportional-to-size
selection their inclusion probability reaches one, so a plan that pretends
they are sampled understates what the field work fixes in advance. Given a
register with one row per PSU, `n_alloc()` separates the certainty part from
the sampled remainder and sizes both together. Here the ultimate units are
households and the register carries each enumeration area's household count.

```{r}
ea_register <- data.frame(
  psu_id = c(sprintf("U%02d", 1:8), sprintf("R%02d", 1:8)),
  stratum = rep(c("urban", "rural"), each = 8),
  N = c(600, rep(200, 7), 450, 220, 165, rep(153, 5))
)

certainty_plan <- n_alloc(
  data.frame(stratum = c("urban", "rural"), N = c(2000, 1600), n_per_psu = 10),
  measures = data.frame(
    stratum = c("urban", "rural"), name = "literacy",
    p = c(0.5, 0.4), icc_psu = 0.05
  ),
  targets = data.frame(name = "literacy", cv = 0.10),
  psu = ea_register
)

certainty_plan$detail[, c("stratum", "n_int", "n_psu_certain", "n_psu_draw", "threshold")]
```

The plan's `$psu` table classifies every PSU and states the take the
operational design fields in it. A certainty PSU carries its own whole take.
The remainder carries `n_per_psu`.

```{r}
filter(certainty_plan$psu, certainty)
```

Passing the plan to `draw()` fields that classification as stored. Stage 1
must be stratified and clustered, with an exact-pik fixed-size PPS method and
the register's own sizes as `mos`: every certainty PSU enters with
probability one, and exactly `n_psu_draw` more are drawn per stratum. The
same plan at stage 2 applies each PSU's take. The frame here is the household
listing the register describes.

```{r}
households <- ea_register[rep(seq_len(nrow(ea_register)), ea_register$N), ] |>
  mutate(household = row_number(), .by = psu_id)

certainty_sample <- sampling_design(title = "Certainty-aware household survey") |>
  stratify_by(stratum) |>
  cluster_by(psu_id) |>
  draw(n = certainty_plan, method = "pps_systematic", mos = N) |>
  add_stage() |>
  draw(n = certainty_plan) |>
  execute(households, seed = 4)

count(certainty_sample, stratum, name = "n_sampled")
```

The totals equal the plan's `n_int` exactly, because both stages field the
plan's own whole-unit numbers. Before anything is drawn, `execute()`
reconciles the frame against the register on PSU identity, stratum, and size,
and refuses a frame the plan was not solved for. It also refuses a plan whose
remainder draw would push another PSU to probability one, naming the PSU to
flag in the register's `certainty` column before refitting. The plan's
classification is never silently altered.

Stratifying within PSUs at stage 2 is allowed when an allocation method
states how each take splits. A bare `stratify_by()` is refused there, because
a scalar size at a stratified stage means that size in every stratum.

```{r}
households$sex <- c("f", "m")[households$household %% 2 + 1]

stratified_takes <- sampling_design() |>
  stratify_by(stratum) |>
  cluster_by(psu_id) |>
  draw(n = certainty_plan, method = "pps_systematic", mos = N) |>
  add_stage() |>
  stratify_by(sex, alloc = "proportional") |>
  draw(n = certainty_plan) |>
  execute(households, seed = 4)

count(stratified_takes, stratum, name = "n_sampled")
```

Designs holding a certainty plan serialize with `write_design()` like any
other and replay identically. The file carries the register, so an edited
copy is refused at execution rather than fielded.

## Plan a repeated survey

A repeated survey often needs precision for a change. That calculation depends
on how much of the responding sample appears at both occasions and how strongly
the outcome is associated within those shared units.

`design_overlap()` computes issued-sample overlap from a rotation schedule. A
cohort interviewed on four consecutive occasions produces 75 percent overlap
between adjacent steady-state samples.

```{r}
membership <- design_overlap("4")
membership[1]
```

At complete response, this issued overlap can be passed directly to
`n_change()`. Below complete response, convert it to the expected overlap among
respondents under a stated response-persistence assumption. For illustration,
assume response is independent between occasions at rate 0.80. A shared issued
unit then responds again with probability 0.80, so respondent overlap is issued
overlap multiplied by that rate.

```{r}
change_response_rate <- 0.80
respondent_overlap <- membership[1] * change_response_rate

change_target <- n_change(
  p = c(p_export, p_export + 0.05),
  moe = 0.02,
  overlap = respondent_overlap,
  overlap_cor = 0.6
)

change_target
```

`overlap_cor` is an outcome assumption, not another name for membership
overlap. Use prior rounds or a sensitivity range rather than deriving it from
the rotation pattern.

## Recruit a rotating panel

`n_panel()` translates a cross-sectional responding target into recruitment
under initial response and conditional retention.

```{r}
panel_plan <- n_panel(
  n_prop(p = p_export, moe = 0.03),
  retention = c(0.90, 0.95, 0.95),
  resp_rate = 0.80,
  design = "rotating",
  start = "immediate"
)

panel_plan
```

A rotating `svyplan_panel` contains several counts because start-up and later
intakes are different operations.

```{r}
intake_n <- ceiling(panel_plan$n_entrants)

fieldwork_counts <- c(
  startup = panel_plan$n_cohorts * intake_n,
  intake_each_occasion = intake_n,
  expected_respondents = panel_plan$n_resp
)

fieldwork_counts
```

Use `startup` for the immediate start-up master and partition it into four
cohort ages. Use `intake_each_occasion` for every later refreshment draw.
`draw(n = panel_plan)` is intentionally unsupported because it cannot know
which operation the caller means.

`n_panel()` treats retention as an unbroken chain. It does not yet turn a
gapped life such as 4-8-4 into a recruitment and reserve-cohort plan. It also
does not estimate attrition, adjust nonresponse weights, or combine cohort
weights. Those are separate design and analysis decisions.

The complete execution handoff is in `vignette("rotating-panels")`.

## What planning does not decide

Planning results depend on their inputs. Before treating a number as a field
requirement, record:

- the estimand and reporting domains
- the source of variances, proportions, and homogeneity assumptions
- whether the count means issued units or expected respondents
- the response and retention model
- the design effect and the estimand it describes
- the rounding rule and any finite-frame limit

`svyplan` does not choose a sampling algorithm or validate a register.
`samplyr` does not decide the precision target. Keeping that boundary explicit
makes it possible to revisit assumptions without silently changing the
selection design.

## Next steps

- Use `vignette("introduction")` to execute an ordinary or multistage design.
- Use `vignette("rotating-panels")` for the longitudinal fieldwork handoff.
- Use `vignette("survey-analysis")` after data collection.
- Consult the svyplan reference pages for power, two-phase designs,
  stratification boundaries, pooled estimates, and detailed sensitivity
  analysis.

## References
