---
title: "Analyze a samplyr Sample"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Analyze a samplyr Sample}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, echo = FALSE}
library(samplyr)
library(dplyr)
library(survey)
```

## From selection to estimation

The usual workflow has four parts. Select the sample, collect outcomes, join
them to the selected units, and export the result to `survey` or `srvyr`.

`samplyr` records the strata, clusters, weights, and finite population
corrections during selection. A safe one-to-one join preserves that metadata.
An operation that adds, removes, or duplicates sampled rows changes the
realization and prevents export.

The main path reaches this article after `vignette("introduction")` and
`vignette("three-stage-sampling")`. The sections on joins, export and domains
apply to an ordinary sample. Read the later two-phase or overlapping-frame
examples when your design uses them.

We use a small stratified EA sample so the analysis examples run independently.
Sample inspection is covered in the selection tutorials.

```{r analysis-sample}
sample <- sampling_design(title = "Burkina Faso EA survey") |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 300) |>
  execute(bfa_eas, seed = 1)
```

## Join collected outcomes

The frame need not contain the survey outcomes. Join one collected record to
each sampled unit using a stable identifier.

```{r, message = FALSE}
library(dplyr)
set.seed(2)

collected <- tibble(
  ea_id = sample$ea_id,
  consumption_per_capita = round(rlnorm(
    nrow(sample),
    meanlog = 11,
    sdlog = 0.65
  )),
  below_threshold = rbinom(nrow(sample), size = 1, prob = 0.25)
)

sample_with_data <- sample |>
  left_join(collected, by = "ea_id")

class(sample_with_data)
```

The result remains a `tbl_sample` because the join is one to one and leaves
the selected rows intact. Check key uniqueness before joining when collected
data can contain repeated visits, roster members, or duplicate submissions.

Keep the issued rows when some units do not respond: use a left join and
record a response indicator. Missing outcomes do not change the selection
weights. Neither dropping those rows before export nor using `na.rm = TRUE`
is, by itself, a nonresponse adjustment. Choosing response adjustments or
calibration belongs to the downstream analysis and its assumptions.

The compact selection and inference table is in `?selection-methods`.
`?as_svydesign` and `?as_svrepdesign` give the detailed export restrictions.

## Export to survey

`as_svydesign()` maps the recorded design to a `survey` object.

```{r}
library(survey)
svy <- as_svydesign(sample_with_data)
svy
```

The collected variables can now be analyzed with standard `survey` functions.

```{r}
svymean(~consumption_per_capita, svy)
svymean(~below_threshold, svy)
svytotal(~population, svy)
```

For a multistage sample, every represented stage contributes its sampling-unit
and finite-population terms. This preserves the stage structure. It does not
turn an approximate stage variance treatment into an exact one.

Fixed-size PPS methods use Brewer's approximation by default. In that phrase,
Brewer names the `survey` variance approximation and not necessarily the
selection algorithm. The approximation's accuracy depends on the population
and design. Exact joint inclusion probabilities can be supplied for methods
that provide them.

Equal-probability systematic sampling uses an SRS-style variance approximation.
Its direction depends on frame ordering. `as_svydesign()` warns unless the
caller explicitly accepts or refuses that approximation through
`systematic_variance`.

## Estimate a domain correctly

Convert the complete sample first, then subset the survey design.

```{r}
urban_svy <- subset(svy, urban_rural == "Urban")
svymean(~consumption_per_capita, urban_svy)
```

Filtering the `tbl_sample` first treats the observed domain sample as fixed.
The point estimate can agree, but the variance estimate is generally wrong and
is often too small because domain membership is random under the design
[@sarndal1992, ch. 5.8].

```{r, error = TRUE}
urban_rows <- sample_with_data |>
  filter(urban_rural == "Urban")

as_svydesign(urban_rows)
```

One-to-one joins, row reordering, and new ordinary columns remain supported.
Changing the row set or an internal column such as `.weight` or `.fpc_1` does
not.

With `srvyr`, use the same order. Convert first and then filter the survey
object.

```{r, eval=requireNamespace("srvyr", quietly = TRUE)}
library(srvyr)

sample_with_data |>
  as_survey_design() |>
  filter(urban_rural == "Urban") |>
  summarise(
    mean_consumption = survey_mean(consumption_per_capita)
)
```

For a strictly positive domain total, a log-scale delta interval is an
available finite-sample sensitivity analysis. It keeps the endpoints positive
and allows asymmetric uncertainty.

```{r}
urban_total <- svytotal(~population, urban_svy)
urban_estimate <- unname(coef(urban_total)[[1]])
urban_se <- unname(SE(urban_total)[[1]])
z <- qnorm(0.975)

exp(log(urban_estimate) + c(-1, 1) * z * urban_se / urban_estimate)
```

This transformation does not repair a poor variance estimator. It applies
only when the estimate is positive. Extended PPS qualification found a clear
coverage improvement for positive domain totals, but not for full-population
totals. Log-ratio and Fieller intervals did not provide a uniformly reliable
correction for skewed ratios, whose point estimators can also have material
finite-sample bias.

## Replicate weights and PPS variance

For a supported single-phase design, `as_svrepdesign()` creates replicate
weights through `survey::as.svrepdesign()`, or through the optional **svrep**
package when `type = "rwyb"` is requested.

```{r}
replicate_design <- as_svrepdesign(sample, type = "auto")
svymean(~households, replicate_design)
```

Replicate methods are not automatic proofs of exactness. Generic PPS
bootstraps do not recreate spatial selection, hard cube constraints, or a
random Poisson sample size. `as_svydesign()` and `as_svrepdesign()` report or
refuse combinations that cannot be represented under their documented
contracts. See their reference pages before selecting a variance route for an
advanced design.

For independent Poisson selection, including clustered and multistage
designs, request Rao-Wu-Yue-Beaumont replicates explicitly:

```{r, eval=requireNamespace("svrep", quietly = TRUE)}
poisson_sample <- sampling_design() |>
  draw(frac = 0.4, method = "bernoulli") |>
  execute(data.frame(id = 1:100, y = 1:100), seed = 12)
set.seed(42)
poisson_replicates <- as_svrepdesign(
  poisson_sample, type = "rwyb", replicates = 1000
)
svytotal(~y, poisson_replicates)
```

RWYB retains the variation in Poisson sample size that generic fixed-size
bootstraps can lose. It also supports mixed SRS, independent draws with
replacement, and fixed-size PPS WOR stages. PPS WOR variance is an
approximation and is reported as such. Increase the replicate count to reduce
simulation error. svrep's `estimate_boot_sim_cv()` helps assess it for chosen
estimates. No new sampling verbs are needed.

Two-phase replicate export is not yet supported. Multistage RWYB export
requires every selected parent to remain represented in the final sample.
Later Poisson stages require a complete frame digest to check this. A
noncertainty singleton is refused when its variance is not estimable, while
a singleton Poisson sample can contribute a valid variance estimate.

When exporting a materialized wave or a second-phase sample with
`as_svydesign()`, current analysis columns take precedence over same-named
columns in the retained first-phase sample. Missing current measurements
remain missing. Keep earlier measurements under separate names if you need
both. Identifiers and sampling metadata retain their design meanings.

## Two phases: screening followed by a detailed interview

A second phase samples units already selected at phase 1, using information
collected in that phase. It is a new selection law on the realized sample.
A later stage in a household design instead samples children within selected
parents. Continuing that design with a household listing is not a new phase.

The bundled `ken_enterprises` is entirely synthetic. Here a cheap screening
interview determines phase-two strata, and only the subsample receives a
detailed interview. The stable `enterprise_id` links the phases. At phase 1,
`cluster_by(enterprise_id)` declares that identity. Each such unit has one row.

```{r screening}
screening_design <- sampling_design("Establishment screening") |>
  cluster_by(enterprise_id) |>
  draw(n = 500)

phase1 <- execute(screening_design, ken_enterprises, seed = 71)
set.seed(72)
phase1 <- phase1 |>
  mutate(
    revenue_baseline = revenue_millions,
    screen_score = log1p(employees) + rnorm(n(), sd = 0.5),
    priority = if_else(screen_score >= 4, "high", "ordinary")
  )
phase1 |> as.data.frame() |> count(priority)
```

The synthetic screening variables are observed for every phase-one unit,
including those that will not enter phase 2. They remain in the retained
phase-one sample so the two-phase export can reconstruct the selection pools.

```{r detailed-interview}
interview_design <- sampling_design("Detailed establishment interview") |>
  stratify_by(priority) |>
  draw(n = c(high = 90, ordinary = 60))

phase2 <- execute(interview_design, phase1, seed = 73)

set.seed(74)
phase2 <- phase2 |>
  mutate(
    revenue_millions = revenue_baseline * exp(rnorm(n(), 0, 0.15)),
    audit_hours = rgamma(n(), shape = 3, rate = 0.5)
  )
```

`audit_hours` is a new phase-two measurement. `revenue_millions` is an updated
measurement with the same name as its baseline counterpart. Export uses the
current phase-two values. `revenue_baseline` keeps the earlier values available
explicitly. All detailed interviews are complete in this example.

```{r two-phase-estimation}
two_phase_svy <- as_svydesign(phase2)
svytotal(~revenue_millions + audit_hours, two_phase_svy)

# The export uses the current observations and the combined phase weights.
stopifnot(isTRUE(all.equal(
  unname(coef(svytotal(~revenue_millions, two_phase_svy))),
  sum(phase2$.weight * phase2$revenue_millions)
)))
```

These estimates illustrate inference for the synthetic establishment
population, not substantive estimates for Kenya. The example uses SRS at
phase 1 and stratified SRS at phase 2, a supported two-phase path. PPS, WR and
Poisson phase designs have additional export restrictions. Phase chains longer
than two and two-phase replicate export are currently unsupported. See
`?as_svydesign` before changing the selection methods.

## Joint information for PPS variance

For supported PPS methods, `joint_expectation()` reconstructs exact or
approximate second-order quantities according to the method.

```{r}
sampford_sample <- sampling_design() |>
  stratify_by(region) |>
  cluster_by(ea_id) |>
  draw(n = 5, method = "pps_sampford", mos = households) |>
  execute(bfa_eas, seed = 2025)

joints <- joint_expectation(sampford_sample, bfa_eas, stages = 1)

exact_pps_design <- as_svydesign(
  sampford_sample,
  pps = ppsmat(joints[[1]])
)

svymean(~households, exact_pps_design)
```

The Sampford matrix above is exact. CPS, systematic PPS, and Poisson selection
also have exact second-order calculations in the supported cases.
Generalized Brewer, SPS, Pareto, and unconstrained cube use the documented
high-entropy approximation. Spatial methods and bounded cube do not provide a
joint matrix. `?joint_expectation` is the authoritative method table.

An exact systematic PPS matrix can contain zero pair probabilities. Exact
design quantities do not then guarantee an estimable or stable sample variance
through `survey::ppsmat()`. A sampled matrix cannot show whether unseen
population pairs have zero probability, so `as_svydesign()` warns when a
`ppsmat` object is supplied for systematic PPS. It does not refuse the route,
because full pair positivity can hold at high sampling fractions.

A user-supplied `pps` matrix describes one stage. For a multistage design,
supplying it exports only the represented PPS stage. Omit it when the intended
analysis is the full multistage linearization with the default stage-level
variance treatment.

## Frames that overlap on one population

Two registers of the same population, each incomplete: a landline frame and a
cell frame, an area frame and a list frame. The samples are selected
independently, one design and one seed each, and everything the feature
contains happens after that.

`stack_frames()` records which frames each sampled unit belongs to. Every
component has to carry every frame's membership column, not only its own,
because a unit selected from frame A has to say whether it was also listed in
frame B, and that is the information the compositing needs.

```{r frames}
population <- data.frame(
  person_id = 1:600,
  spend = round(stats::rlnorm(600, log(120), 0.4), 2),
  in_landline = rep(c(TRUE, FALSE), times = c(400, 200)),
  in_cell = rep(c(FALSE, TRUE), times = c(150, 450))
)

frames <- stack_frames(
  landline = sampling_design() |>
    draw(n = 80) |>
    execute(population[population$in_landline, ], seed = 1),
  cell = sampling_design() |>
    draw(n = 120) |>
    execute(population[population$in_cell, ], seed = 2),
  membership = c(landline = "in_landline", cell = "in_cell"),
  key = person_id
)

summary(frames)
```

The domain counts are rows, not distinct people. Someone listed in both frames
and selected from both appears twice, once per frame that selected them, and
that is exactly the double count a composite weight has to resolve.

### Two export routes

`as_svydesign()` hands the stack to `survey::multiframe()`. Each component is
exported on its own, so it keeps its strata, clusters and variance treatment,
and the compositing sits on top.

```{r frames-linearized}
svytotal(~spend, as_svydesign(frames))
```

`theta = NULL`, the default, is the multiplicity estimator: a unit reached by
both frames contributes half its weight through each. An explicit `theta` is
Hartley's constant factor [@hartley1962], and it belongs to the **first** frame
of the stack, so reversing the components and wanting the same estimator means
asking for `1 - theta`.

```{r frames-hartley}
svytotal(~spend, as_svydesign(frames, theta = 0.74))
```

The expected multiframe estimator instead needs each unit's selection
probability in every frame containing it. Supply the original registers through
`overlaps = exante_overlaps(list(area = area_register, list = list_register),
by = c(person_id = "person_id"))` when building the stack, using the stack's
component names and key, then export with `estimator = "expected"`.
The resolver requires exact inclusion probabilities by default. SPS and Pareto
targets require `allow_approximate = TRUE` in `exante_overlaps()`. This accepts
an approximation that can affect design unbiasedness. The stack records quality
by frame, and exported designs retain it in their
`samplyr_overlap_probability_quality` attribute. For inspecting one register,
`exante_probabilities()` returns the same quality beside each probability.

`samplyr` never forwards a missing `theta` to `survey`. Left unset there, it
defaults to the ratio of the frames' mean sampling weights, which is a
data-dependent stand-in for frame size over sample size rather than a neutral
choice.

The replicate route builds one system per frame and combines them in blocks: in
a replicate column belonging to one frame, only that frame varies and every
other stays at its full-sample weight. The combined variance is then the sum of
the frames' own contributions, which is what independent selection from each
frame gives.

```{r frames-replicate}
svytotal(~spend, as_svrepdesign(frames, type = "bootstrap", replicates = 100))
```

The two routes are not interchangeable, and the differences are the reason to
know which one you are on.

| | `as_svydesign()` | `as_svrepdesign()` |
|---|---|---|
| Number of frames | Two. `survey::multiframe()` composites two | Any number |
| Compositing | Multiplicity, or Hartley with a stated `theta` | The same, with `theta` above two frames refused |
| A component with shared weights | Refused | Supported |
| Replicate methods | Not applicable | May differ between frames. `type = "auto"` picks one per component |

The two-frame limit belongs to the export and is stated there, not at the
design layer: `stack_frames()` records any number, exactly as `execute()`
chains any number of phases while the two-phase export stops at two.

### Where the theory comes from

The multiplicity estimator for overlapping frames is the generalized weight
share method with the standardized link matrix, because the number of links to
a unit is the number of frames containing it [@lohr2021, section 3.2;
@mecatti2007]. That is why a stack composed of frames and a sample whose
weights were shared through a link table are handled by one mechanism rather
than two, and why `vignette("design-semantics")` documents them together.

## Analysis checklist

1. Inspect the selected counts and weights before fieldwork.
2. Verify that collected-data keys are unique at the sampled-unit level.
3. Join outcomes without changing the selected row set.
4. Convert the complete sample to a survey design.
5. Define domains on the converted design.
6. State every variance approximation used by an advanced selection method.
7. Where the estimates combine frames, say which compositing factor was used
   and which frame an explicit `theta` was applied to.

Use `vignette("serialization")` to archive the selection and
`vignette("design-semantics")` for the detailed export contracts, the weight
share method, and the contract separating design weights from estimation
weights.

## References
