---
title: "Analyze a samplyr sample"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Analyze a samplyr sample}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
has_suggests <- requireNamespace("survey", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_suggests
)
```

```{r setup, message = FALSE, echo = FALSE}
library(samplyr)
library(dplyr)
library(survey)
```

## From selection to estimation

The usual workflow has four parts. Select the sample, collect outcomes, join
them to the selected units, and export the result to `survey` or `srvyr`.

`samplyr` records the strata, clusters, weights, and finite population
corrections during selection. A safe one-to-one join preserves that metadata.
An operation that adds, removes, or duplicates sampled rows changes the
realization and prevents export.

The main path reaches this article after `vignette("introduction")`, which
includes a three-stage household sample. The sections on joins, export and
domains apply to any supported sample.

We use a small stratified EA sample so the analysis examples run independently.
Sample inspection is covered in the "Inspecting the result" section of
`vignette("introduction")`.

```{r analysis-sample}
sample <- sampling_design(title = "Burkina Faso EA survey") |>
  stratify_by(region, alloc = "proportional") |>
  draw(n = 300) |>
  execute(bfa_eas, seed = 1)
```

`variance_estimators()` reports, from the design and the frame and before
anything is drawn, which variance estimators the design supports, which only
approximately, and which the export would refuse, with the condition each
would raise. A stratified first stage rules out survey's JK1, for instance.

```{r}
estimators <- variance_estimators(get_design(sample), bfa_eas)
estimators[c("estimator", "status", "class")]
```

## Join collected outcomes

The frame need not contain the survey outcomes. Join one collected record to
each sampled unit using a stable identifier.

```{r, message = FALSE}
library(dplyr)
set.seed(2)

collected <- tibble(
  ea_id = sample$ea_id,
  consumption_per_capita = round(rlnorm(
    nrow(sample),
    meanlog = 11,
    sdlog = 0.65
  )),
  below_threshold = rbinom(nrow(sample), size = 1, prob = 0.25)
)

sample_with_data <- sample |>
  left_join(collected, by = "ea_id")

class(sample_with_data)
```

The result remains a `tbl_sample` because the join is one to one and leaves
the selected rows intact. Check key uniqueness before joining when collected
data can contain repeated visits, roster members, or duplicate submissions.

### Nonresponse and calibration

Keep every issued row and record each unit's disposition next to its
outcomes. Missing outcomes do not change the selection weights, and neither
dropping nonrespondents before export nor `na.rm = TRUE` is a nonresponse
adjustment. The adjustment and any calibration belong to the analysis. Applied
to the exported design, they leave `.weight` holding the base weights while
the adjusted weights live in the design object.

```{r nonresponse, eval = has_suggests && requireNamespace("svrep", quietly = TRUE)}
set.seed(3)
issued <- sample_with_data |>
  mutate(disposition = if_else(runif(n()) < 0.85, "respondent",
                               "nonrespondent"))
issued |> as.data.frame() |> count(disposition)

set.seed(4)
issued_replicates <- as_svrepdesign(issued, type = "rwyb", replicates = 500)
adjusted <- svrep::redistribute_weights(
  issued_replicates,
  reduce_if = disposition == "nonrespondent",
  increase_if = disposition == "respondent",
  by = "region"
)
respondents <- subset(adjusted, disposition == "respondent")
```

The adjustment moves each nonrespondent's weight to the respondents of its
region, in every replicate as well as in the full-sample weights, so the
replicate variance carries it. It assumes that response is unrelated to the
outcome within a region, which the analyst has to justify. `survey` warns
that the respondent subset has fewer rows than the design's degrees of
freedom. The replicate variance is computed as for any domain and is not
affected. Calibration then matches the respondents to known population
totals, here the number of urban EAs in the frame.

```{r calibration, eval = has_suggests && requireNamespace("svrep", quietly = TRUE)}
ea_totals <- c(`(Intercept)` = nrow(bfa_eas),
               urban_ruralUrban = sum(bfa_eas$urban_rural == "Urban"))
calibrated <- calibrate(respondents, ~urban_rural, population = ea_totals)
svymean(~consumption_per_capita, calibrated)
```

`weights(calibrated)` returns the adjusted and calibrated weights, while
`issued$.weight` keeps the base weights, so the two stay apart for reporting.
The example uses RWYB replicates because svrep's weight adjustments act on a
replicate design. `?variance-estimation` gives what RWYB assumes for each
selection method.

How each selection method's variance is estimated is in `?variance-estimation`,
and `?as_svydesign` and `?as_svrepdesign` give the export restrictions.

## Export to survey

`as_svydesign()` maps the recorded design to a `survey` object.

```{r}
library(survey)
svy <- as_svydesign(sample_with_data)
svy
```

The collected variables can now be analyzed with standard `survey` functions.

```{r}
svymean(~consumption_per_capita, svy)
svymean(~below_threshold, svy)
svytotal(~population, svy)
```

For a multi-stage sample, every represented stage contributes its sampling-unit
and finite-population terms. This preserves the stage structure. It does not
turn an approximate stage variance treatment into an exact one.

Fixed-size PPS methods without replacement use Brewer's approximation by
default. In that phrase,
Brewer names the `survey` variance approximation and not necessarily the
selection algorithm. The approximation's accuracy depends on the population
and design. Exact joint inclusion probabilities can be supplied for methods
that provide them.

Systematic sampling uses an SRS-style variance approximation, and systematic
PPS uses Brewer's. Neither sees frame order, so the direction of the error
depends on it. `as_svydesign()` warns unless the caller explicitly accepts or
refuses the approximation through `systematic_variance`.

### Strata with one sampled unit

A stratum that ends up with one sampled unit outside certainty, because its
allocation was one or because certainty units took the rest, has no
within-stratum variance estimate. `execute()` says so when it selects such a
stratum, and `as_svydesign()` warns again at export
(`samplyr_warning_lonely_psu`) while `survey` is left at its default
`survey.lonely.psu = "fail"`, under which estimation would stop. Setting the
option, as below, also silences that export warning.

```{r lonely}
one_per_region <- sampling_design() |>
  stratify_by(region) |>
  draw(n = 1) |>
  execute(bfa_eas, seed = 5)

old <- options(survey.lonely.psu = "adjust")
svytotal(~households, as_svydesign(one_per_region))
options(old)
```

`"adjust"` centers a lonely unit at the population mean, which is
conservative. Collapsing each such stratum with a similar one before export
is the other usual choice, and it is a design decision the analyst has to
make. Allocating at least two units outside certainty to every stratum avoids
the problem at the source. RWYB replicates refuse a singleton whose variance
is needed, except under Poisson sampling and, with `lonely.psu =
"certainty"`, at the final stage. `?variance-estimation` gives the certainty case.

## Estimate a domain correctly

Convert the complete sample first, then subset the survey design.

```{r}
urban_svy <- subset(svy, urban_rural == "Urban")
svymean(~consumption_per_capita, urban_svy)
```

Filtering the `tbl_sample` first treats the observed domain sample as fixed.
The point estimate can agree, but the variance estimate is generally wrong and
is often too small because domain membership is random under the design
[@sarndal1992, ch. 5.8].

```{r, error = TRUE}
urban_rows <- sample_with_data |>
  filter(urban_rural == "Urban")

as_svydesign(urban_rows)
```

One-to-one joins, row reordering, and new ordinary columns remain supported.
Changing the row set or an internal column such as `.weight` or `.fpc_1` does
not.

With `srvyr`, use the same order. Convert first and then filter the survey
object.

```{r, eval = has_suggests && requireNamespace("srvyr", quietly = TRUE)}
library(srvyr)

sample_with_data |>
  as_survey_design() |>
  filter(urban_rural == "Urban") |>
  summarise(
    mean_consumption = survey_mean(consumption_per_capita)
)
```

For a strictly positive domain total, a log-scale delta interval is an
available finite-sample sensitivity analysis. It keeps the endpoints positive
and allows asymmetric uncertainty.

```{r}
urban_total <- svytotal(~population, urban_svy)
urban_estimate <- unname(coef(urban_total)[[1]])
urban_se <- unname(SE(urban_total)[[1]])
z <- qnorm(0.975)

exp(log(urban_estimate) + c(-1, 1) * z * urban_se / urban_estimate)
```

This transformation does not repair a poor variance estimator, and it
applies only when the estimate is positive. It helps where the sampling
distribution of a positive total is skewed, as for a small domain, and
matters little for a full-population total. For a skewed ratio, neither a
log-ratio nor a Fieller interval is uniformly reliable, and the ratio
estimator itself can carry material finite-sample bias.

## Replicate weights and PPS variance

For a supported single-phase design, `as_svrepdesign()` creates replicate
weights through `survey::as.svrepdesign()`, through the optional **svrep**
package when `type = "rwyb"` is requested, or from the replicates of
`execute(reps = R)` when `type = "random_groups"` is requested.

```{r}
replicate_design <- as_svrepdesign(sample, type = "auto")
svymean(~households, replicate_design)
```

Replicate methods are not automatic proofs of exactness. Generic PPS
bootstraps do not recreate spatial selection, hard cube constraints, or a
random Poisson sample size. `as_svydesign()` and `as_svrepdesign()` report or
refuse combinations that cannot be represented under their documented
contracts. `?variance-estimation` gives how each selection method's variance
is estimated and how exact it is.

For independent Poisson selection, including clustered and multi-stage
designs, the generic replicate types, `"auto"` included, refuse the sample,
so request Rao-Wu-Yue-Beaumont replicates explicitly:

```{r, eval = has_suggests && requireNamespace("svrep", quietly = TRUE)}
poisson_sample <- sampling_design() |>
  draw(frac = 0.4, method = "bernoulli") |>
  execute(data.frame(id = 1:100, y = 1:100), seed = 12)
set.seed(42)
poisson_replicates <- as_svrepdesign(
  poisson_sample, type = "rwyb", replicates = 1000
)
svytotal(~y, poisson_replicates)
```

RWYB retains the variation in Poisson sample size. It also supports SRS,
independent draws with replacement, and Brewer, CPS, Sampford and systematic
PPS stages, alone or mixed across stages. PPS WOR variance is an
approximation and is reported as such. Increase the replicate count to reduce
simulation error. svrep's `estimate_boot_sim_cv()` helps assess it for chosen
estimates. No new sampling verbs are needed.

Two-phase replicate export is not yet supported. Multi-stage RWYB export
requires every selected parent to remain represented in the final sample.
Later Poisson stages require a complete frame digest to check this. A
noncertainty singleton is refused when its variance is not estimable, while
a singleton Poisson sample can contribute a valid variance estimate.

When exporting a materialized wave or a second-phase sample with
`as_svydesign()`, current analysis columns take precedence over same-named
columns in the retained first-phase sample. Missing current measurements
remain missing. Keep earlier measurements under separate names if you need
both. Identifiers and sampling metadata retain their design meanings.

## Other designs

The same export functions handle designs this article does not work through,
and each help page carries a runnable example.

| Design | Where it is documented |
|:--|:--|
| A second phase selected from the first-phase sample | `?execute` for the selection, `?as_svydesign` for the `survey::twophase()` export |
| Exact or approximate joint inclusion probabilities for PPS | `?joint_expectation`, passed to `as_svydesign(pps = )` |
| Independent samples from frames that overlap on one population | `?stack_frames`, `?as_svydesign.frame_stack` and `?as_svrepdesign.frame_stack` |
| Weights shared to a linked population | `?share_weights` |

## Analysis checklist

1. Inspect the selected counts and weights before fieldwork.
2. Verify that collected-data keys are unique at the sampled-unit level.
3. Join outcomes without changing the selected row set.
4. Convert the complete sample to a survey design.
5. Define domains on the converted design.
6. State every variance approximation used by an advanced selection method.
7. Where the estimates combine frames, say which compositing factor was used
   and which frame an explicit `theta` was applied to.

Use `vignette("serialization")` to archive the selection and
`vignette("design-semantics")` for the detailed export contracts and the
contract separating design weights from estimation weights.

## References
