---
title: "Select a Three-Stage Household Sample"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Select a Three-Stage Household Sample}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(samplyr)
library(dplyr)
```

## The selection task

Select communes within regions, enumeration areas (EAs) within those communes,
and households within those EAs. The first two stages use PPS without
replacement and the last uses simple random sampling without replacement.
Selection pauses while the household listing is prepared, then continues
under the same design.

We use `bfa_eas`, the package's derived Burkina Faso frame. Its modeled EAs,
household counts and historical administrative labels are teaching inputs.
They are not a current official household register. Household listings and
interview outcomes below are explicitly synthetic.

## Give each register its own sampling unit and MOS

```{r registers}
library(dplyr)
ea_register <- bfa_eas |>
  mutate(
    region,
    commune_id = paste(region, province, commune, sep = "/"),
    ea_id,
    ea_households = households,
    .keep = "none"
  )

commune_register <- ea_register |>
  summarise(commune_households = sum(ea_households),
            .by = c(region, commune_id))

head(commune_register, 3)
head(ea_register, 3)
```

| Register                  | One row represents           | Key                                      | Measure of size                            |
|---------------------------|------------------------------|------------------------------------------|--------------------------------------------|
| `commune_register`        | A commune                    | `commune_id`, qualified by its geography | Sum of EA household counts in that commune |
| `ea_register`             | An EA                        | `ea_id`, with its parent `commune_id`    | That EA's modeled household count          |
| `household_listing` below | A household in a selected EA | `ea_id` and local `hh_id`                | None: equal-probability sampling           |

MOS belongs to the unit being selected. A commune's total must not vary across
its EA rows if using one shared hierarchy. Separate registers make the two
levels explicit. Their order in `execute()` determines their stage mapping.
Names on a list are labels, not a join specification.

## Declare all three stages

```{r design}
library(samplyr)
household_design <- sampling_design("Three-stage household survey") |>
  add_stage("Communes") |>
    stratify_by(region) |>
    cluster_by(commune_id) |>
    draw(n = 4, method = "pps_brewer", mos = commune_households) |>
  add_stage("EAs") |>
    cluster_by(ea_id) |>
    draw(n = 3, method = "pps_brewer", mos = ea_households) |>
  add_stage("Households") |>
    draw(n = 12)

household_design
```

The scopes are four communes **per region**, three EAs **per selected commune**,
and twelve households **per selected EA**. There is no allocation rule at the
first stage, so `n = 4` is per stratum. A pool smaller than its requested take
is a census, which execution reports. Region stratification does not implicitly
become a new stratum specification at later stages.

## Select areas, then pause for listing

```{r select-areas}
selected_eas <- execute(
  household_design,
  list(commune_register, ea_register),
  stages = 1:2,
  seed = 91
)

selected_eas |>
  as.data.frame() |>
  count(region, name = "selected_eas")
```

The EA register covers all communes and execution restricts it to the selected
parents. The retained object contains the two completed stages and their
weights. It is the starting point for continuation.

In fieldwork, a listing team enumerates households in the selected EAs. Here
we create artificial listings whose lengths equal the modeled counts. This
creates identifiers, not observed household records, and expands only the
selected EAs.

```{r listing}
library(tidyr)
household_listing <- selected_eas |>
  as.data.frame() |>
  select(commune_id, ea_id, ea_households) |>
  uncount(ea_households, .id = "hh_id")

stopifnot(!anyDuplicated(household_listing[c("ea_id", "hh_id")]))
head(household_listing, 3)
```

Converting to a plain data frame is intentional here: a listing enumerates
candidates for the next stage. It is not an expanded final sample for analysis.
The parent EA disambiguates local household IDs that repeat across EAs.

```{r select-households}
issued <- execute(selected_eas, household_listing, seed = 92)

issued |>
  as.data.frame() |>
  select(commune_id, ea_id, hh_id, .weight_1, .weight_2, .weight_3, .weight) |>
  head(4)

stopifnot(isTRUE(all.equal(
  issued$.weight,
  issued$.weight_1 * issued$.weight_2 * issued$.weight_3
)))
frame_summary(issued)
```

The final weight is the reciprocal of the product of the three conditional
selection probabilities. Continuation retains the earlier factors. Starting
a new design on `selected_eas` would instead declare a new sampling phase.
The two seeds above specify separate reproducible randomization steps. Matching
a one-call execution exactly would require the same uninterrupted RNG stream.

## Join observations and hand off to analysis

All issued households respond in this example. The artificial consumption
values illustrate a collected outcome that was absent from the sampling frames.

```{r observations}
set.seed(93)
interviews <- issued |>
  as.data.frame() |>
  mutate(ea_id, hh_id,
         consumption = round(rlnorm(n(), log(100000), 0.5)),
         .keep = "none")

observed <- issued |>
  left_join(interviews, by = c("ea_id", "hh_id"), relationship = "one-to-one")
stopifnot(nrow(observed) == nrow(issued))
```

With real missing responses, retain the issued rows and join response
indicators and measurements to them. Sampling weights describe selection.
Removing nonrespondents does not perform response adjustment. The analyst
chooses and justifies that adjustment downstream in `survey` or another
analysis package.

```{r analysis}
linearized <- as_svydesign(observed)
survey::svymean(~consumption, linearized)
```

This export uses Brewer's variance approximation for the PPS stages. Exact
first-order probabilities do not make that variance approximation exact.
An optional alternative uses svrep's RWYB replicates, also approximate for
PPS without replacement:

```{r replication, eval=requireNamespace("svrep", quietly=TRUE)}
set.seed(94)
replicated <- as_svrepdesign(observed, type = "rwyb", replicates = 500)
survey::svymean(~consumption, replicated)
```

The warning describes the PPS approximation. Increasing the number of
replicates reduces simulation error, not error in the variance approximation.
Continue with `vignette("survey-analysis")` for outcome joins, domains and
other analysis designs. Use `?selection-methods` for a compact comparison of
selection and inference support.
