---
title: "Validation on Synthetic Populations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validation on Synthetic Populations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Why this vignette

This vignette runs a set of smoke tests for `samplyr` on synthetic finite populations with known truths. It uses two complementary layers:

1. Deterministic checks on design metadata and invariants.
2. Monte Carlo checks for estimator bias, standard-error calibration, and 95% CI coverage.

The goal is not to benchmark speed, and it is not a proof of correctness. A run at `R = 199` can detect a design that is plainly wrong. It cannot establish that one is right. A pass says that the package behaved consistently with correctness on these three designs at this replication count. That is the level of assurance a smoke test provides. The scale of the Monte Carlo uncertainty is reported below so the bands can be read for what they are.

## Synthetic population with known properties

```{r, message = FALSE}
library(dplyr)
set.seed(20260302)

make_population <- function(
    n_strata = 6L,
    psu_per_stratum = 50L,
    units_per_psu = 30L) {
  n_psu <- n_strata * psu_per_stratum
  stratum <- rep(LETTERS[seq_len(n_strata)], each = psu_per_stratum * units_per_psu)
  psu <- rep(sprintf("PSU%03d", seq_len(n_psu)), each = units_per_psu)

  psu_mos <- round(runif(n_psu, min = 80, max = 420))
  psu_re <- rnorm(n_psu, sd = 18)
  stratum_re <- rep(
    seq(-12, 16, length.out = n_strata),
    each = psu_per_stratum * units_per_psu
  )

  x <- rep(psu_mos, each = units_per_psu) + rnorm(n_psu * units_per_psu, sd = 10)
  y <- 200 + stratum_re + rep(psu_re, each = units_per_psu) + 0.08 * x + rnorm(n_psu * units_per_psu, sd = 20)

  tibble(
    id = seq_len(n_psu * units_per_psu),
    stratum = factor(stratum),
    psu = factor(psu),
    psu_mos = rep(psu_mos, each = units_per_psu),
    x = x,
    y = y
  )
}

population <- make_population()

truth <- population |>
  summarise(
    N = dplyr::n(),
    Y_total = sum(y),
    Y_mean = mean(y)
  )

truth
```

## Layer 1: deterministic checks

```{r, message = FALSE}
library(samplyr)
library(survey)

srs_sample <- sampling_design() |>
  draw(n = 400) |>
  execute(population, seed = 1001)

strat_sample <- sampling_design() |>
  stratify_by(stratum, alloc = "proportional") |>
  draw(n = 400) |>
  execute(population, seed = 1002)

pps_two_stage <- sampling_design() |>
  add_stage("PSU") |>
    stratify_by(stratum) |>
    cluster_by(psu) |>
    draw(
      n = 8,
      method = "pps_brewer",
      mos = psu_mos,
      certainty_size = 390
    ) |>
  add_stage("SSU") |>
    draw(n = 6) |>
  execute(population, seed = 1003)

stratum_sizes <- population |>
  count(stratum, name = "N_h")

strat_weight_by_stratum <- strat_sample |>
  summarise(weight_sum = sum(.weight), .by = stratum) |>
  left_join(stratum_sizes, by = "stratum")

# all(logical(0)) is TRUE, so the certainty check has to establish that the
# set it quantifies over is non-empty before it means anything.
n_certainty <- sum(pps_two_stage$.certainty_1)

checks <- tibble(
  check = c(
    "SRS: sum(weights) = N",
    "Stratified SRS: sum_h weights_h = N_h",
    "Two-stage PPS: .weight = .weight_1 * .weight_2",
    "Two-stage PPS: the certainty_size threshold selected some PSUs",
    "Two-stage PPS certainty: certainty PSUs have .weight_1 = 1"
  ),
  pass = c(
    isTRUE(all.equal(sum(srs_sample$.weight), truth$N)),
    isTRUE(all(abs(strat_weight_by_stratum$weight_sum - strat_weight_by_stratum$N_h) < 1e-8)),
    isTRUE(all.equal(pps_two_stage$.weight, pps_two_stage$.weight_1 * pps_two_stage$.weight_2, tolerance = 1e-10)),
    n_certainty > 0,
    n_certainty > 0 && all(pps_two_stage$.weight_1[pps_two_stage$.certainty_1] == 1)
  )
)

checks
```

These are structural, non-random checks. If they fail, the design object is wrong before analysis begins.

The fourth row is not decoration. `all()` over an empty set returns `TRUE`, so a certainty check written without it would pass on a design in which `certainty_size = 390` happened to select nothing at all, and the pass would say nothing about certainty handling.

## Layer 2: Monte Carlo validation

Below we repeatedly sample from the same finite population and analyze with `survey`.
For each design we report:

- `rel_bias_pct`: relative bias of \(\hat{\bar{Y}}\),
- `bias_z`: the same bias divided by its own Monte Carlo standard error,
- `se_ratio`: empirical SD of estimates divided by mean estimated SE,
- `coverage_95`: fraction of 95% CIs covering the true finite-population mean.

`rel_bias_pct` and `bias_z` measure the same thing on different scales, and only the second is comparable across designs. The true mean here is near 200, so a 2% relative bias is about 4 units, which is large next to the Monte Carlo noise: relative bias against a large mean is a generous criterion. `bias_z` divides by the Monte Carlo standard error of the estimated bias, so a value near zero says the bias is not distinguishable from simulation noise, and a value past 2 or 3 says it is, whatever the units of `y` happen to be.

This vignette fixes Monte Carlo replication at `R = 199`.

```{r}
evaluate_design <- function(design_fn, population, true_mean, R = 199L, seed_init = 3000L) {
  est <- numeric(R)
  se <- numeric(R)
  cover <- logical(R)

  for (i in seq_len(R)) {
    sample_i <- design_fn(population, seed_init + i)
    svy_i <- as_svydesign(sample_i)
    fit <- svymean(~y, svy_i)

    est[i] <- unname(coef(fit)[1])
    se[i] <- unname(SE(fit)[1])

    ci <- suppressWarnings(confint(fit, level = 0.95))
    cover[i] <- ci[1, 1] <= true_mean && ci[1, 2] >= true_mean
  }

  # Delete one independent simulation at a time. This accounts for both
  # parts of the SD/mean-SE ratio and their dependence within each draw.
  leave_one_ratio <- vapply(seq_len(R), function(i) {
    sd(est[-i]) / mean(se[-i])
  }, numeric(1))
  ratio_mcse <- sqrt((R - 1) * mean(
    (leave_one_ratio - mean(leave_one_ratio))^2
  ))
  coverage_ci <- binom.test(sum(cover), R)$conf.int

  tibble(R = R,
         bias = mean(est - true_mean),
         rel_bias_pct = 100 * mean(est - true_mean) / true_mean,
         bias_z = mean(est - true_mean) / (sd(est) / sqrt(R)),
         emp_sd = sd(est),
         mean_se = mean(se),
         se_ratio = sd(est) / mean(se),
         coverage_95 = mean(cover),
         mcse_rel_bias_pct = 100 * sd(est) / sqrt(R) / abs(true_mean),
         mcse_se_ratio = ratio_mcse,
         coverage_lower = coverage_ci[1],
         coverage_upper = coverage_ci[2])
}

mcse_coverage <- function(R, p = 0.95) {
  sqrt(p * (1 - p) / R)
}

design_srs <- function(pop, seed) {
  sampling_design() |>
    draw(n = 400, method = "srswor") |>
    execute(pop, seed = seed)
}

design_strat <- function(pop, seed) {
  sampling_design() |>
    stratify_by(stratum, alloc = "proportional") |>
    draw(n = 400, method = "srswor") |>
    execute(pop, seed = seed)
}

design_two_stage <- function(pop, seed) {
  sampling_design() |>
    add_stage("PSU") |>
      stratify_by(stratum) |>
      cluster_by(psu) |>
      draw(n = 8, method = "pps_brewer", mos = psu_mos) |>
    add_stage("SSU") |>
      draw(n = 6, method = "srswor") |>
    execute(pop, seed = seed)
}

R_mc <- 199L
R_mc

mc_results <- bind_rows(
  evaluate_design(design_srs, population, truth$Y_mean, R = R_mc) |>
    mutate(design = "SRSWOR"),
  evaluate_design(design_strat, population, truth$Y_mean, R = R_mc) |>
    mutate(design = "Stratified SRS"),
  evaluate_design(design_two_stage, population, truth$Y_mean, R = R_mc) |>
    mutate(design = "Two-stage PPS->SRS")
) |>
  select(design, everything())

mc_results
```

The third design's first stage is `pps_brewer`, and `as_svydesign()` exports it with Brewer's approximation to the joint inclusion probabilities. Its `mean_se` is therefore an approximate quantity, and `se_ratio` for that row compares the empirical spread against an approximation rather than against an exact variance. The first two rows have no such caveat.

## Separate tolerances from Monte Carlo uncertainty

A practical tolerance states how much error is acceptable for an application.
Monte Carlo uncertainty states how precisely this simulation measured the
error. They answer different questions. For illustration, choose tolerances
of two percentage points of relative bias, 15% around an SD/mean-SE ratio of
one, and three percentage points around nominal 95% coverage. These are
teaching choices, not universal standards for qualifying a survey design.

The table below compares those fixed tolerances with 95% Monte Carlo intervals.
Bias uses the empirical SD divided by the square root of `R`. The ratio uses
a jackknife over independent simulation runs, accounting for uncertainty in
both its numerator and denominator. These two intervals are normal
approximations. Coverage uses an exact binomial interval. The intervals
measure simulation uncertainty, not uncertainty in a particular survey estimate.

```{r mc-precision}
precision <- bind_rows(
  mc_results |>
    transmute(design, metric = "Relative bias (%)", estimate = rel_bias_pct,
              lower = estimate - 1.96 * mcse_rel_bias_pct,
              upper = estimate + 1.96 * mcse_rel_bias_pct,
              acceptable_lower = -2, acceptable_upper = 2),
  mc_results |>
    transmute(design, metric = "SD / mean SE", estimate = se_ratio,
              lower = pmax(0, estimate - 1.96 * mcse_se_ratio),
              upper = estimate + 1.96 * mcse_se_ratio,
              acceptable_lower = 0.85, acceptable_upper = 1.15),
  mc_results |>
    transmute(design, metric = "95% coverage", estimate = coverage_95,
              lower = coverage_lower, upper = coverage_upper,
              acceptable_lower = 0.92, acceptable_upper = 0.98)
) |>
  mutate(assessment = case_when(
    lower >= acceptable_lower & upper <= acceptable_upper ~ "Within tolerance",
    upper < acceptable_lower | lower > acceptable_upper ~ "Outside tolerance",
    TRUE ~ "Inconclusive at this replication count"
  ))

precision |>
  select(design, metric, estimate, lower, upper, assessment) |>
  knitr::kable(digits = 3,
               col.names = c("Design", "Metric", "Estimate", "MC lower",
                             "MC upper", "Assessment"))
```

An estimate can fall inside a tolerance while its Monte Carlo interval
extends outside. That result is inconclusive, not a pass. Conversely, a
small bias can be statistically distinguishable from zero (`bias_z`) while
remaining practically tolerable. Report both size and uncertainty.

```{r}
tibble(
  R = R_mc,
  mcse_coverage_95 = mcse_coverage(R_mc, p = 0.95)
)
```

At `R = 199`, nominal 95% coverage has Monte Carlo SE of about 1.5 percentage
points. Resolving a narrow coverage tolerance may therefore need substantially
more draws. Increasing `R` typically narrows the Monte Carlo intervals at
roughly the inverse-square-root rate. It does **not** shrink the chosen
practical tolerances. The interval calculations above are recomputed from the
new simulations. The tolerance constants remain unchanged unless the
application's requirements change.

These are pointwise diagnostics for three particular populations/designs,
not simultaneous certification of the package. Deterministic checks establish
the stated identities for the realized examples. The simulations assess mean
bias, the SD/mean-SE ratio and interval coverage in these cases. They do not
establish unbiased variance estimation for all totals or all populations.
Random-size sampling, extreme MOS, small pools and other compositions need
separate checks. See `?selection-methods` for the inference overview and `?as_svydesign`
and `?as_svrepdesign` for the detailed export restrictions.
