---
title: "Validation on Synthetic Populations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validation on Synthetic Populations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Why this vignette

This vignette validates `samplyr` on synthetic finite populations with known truths. It uses two complementary layers:

1. Deterministic checks on design metadata and invariants.
2. Monte Carlo checks for estimator bias, standard-error calibration, and 95% CI coverage.

The goal is not to benchmark speed. The goal is to show statistical correctness.

## Synthetic population with known properties

```{r}
library(dplyr)
set.seed(20260302)

make_population <- function(
    n_strata = 6L,
    psu_per_stratum = 50L,
    units_per_psu = 30L) {
  n_psu <- n_strata * psu_per_stratum
  stratum <- rep(LETTERS[seq_len(n_strata)], each = psu_per_stratum * units_per_psu)
  psu <- rep(sprintf("PSU%03d", seq_len(n_psu)), each = units_per_psu)

  psu_mos <- round(runif(n_psu, min = 80, max = 420))
  psu_re <- rnorm(n_psu, sd = 18)
  stratum_re <- rep(
    seq(-12, 16, length.out = n_strata),
    each = psu_per_stratum * units_per_psu
  )

  x <- rep(psu_mos, each = units_per_psu) + rnorm(n_psu * units_per_psu, sd = 10)
  y <- 200 + stratum_re + rep(psu_re, each = units_per_psu) + 0.08 * x + rnorm(n_psu * units_per_psu, sd = 20)

  tibble(
    id = seq_len(n_psu * units_per_psu),
    stratum = factor(stratum),
    psu = factor(psu),
    psu_mos = rep(psu_mos, each = units_per_psu),
    x = x,
    y = y
  )
}

population <- make_population()

truth <- population |>
  summarise(
    N = dplyr::n(),
    Y_total = sum(y),
    Y_mean = mean(y)
  )

truth
```

## Layer 1: deterministic checks

```{r}
library(samplyr)
library(survey)

srs_sample <- sampling_design() |>
  draw(n = 400) |>
  execute(population, seed = 1001)

strat_sample <- sampling_design() |>
  stratify_by(stratum, alloc = "proportional") |>
  draw(n = 400) |>
  execute(population, seed = 1002)

pps_two_stage <- sampling_design() |>
  add_stage("PSU") |>
    stratify_by(stratum) |>
    cluster_by(psu) |>
    draw(
      n = 8,
      method = "pps_brewer",
      mos = psu_mos,
      certainty_size = 390
    ) |>
  add_stage("SSU") |>
    draw(n = 6) |>
  execute(population, seed = 1003)

stratum_sizes <- population |>
  count(stratum, name = "N_h")

strat_weight_by_stratum <- strat_sample |>
  summarise(weight_sum = sum(.weight), .by = stratum) |>
  left_join(stratum_sizes, by = "stratum")

checks <- tibble(
  check = c(
    "SRS: sum(weights) = N",
    "Stratified SRS: sum_h weights_h = N_h",
    "Two-stage PPS: .weight = .weight_1 * .weight_2",
    "Two-stage PPS certainty: certainty PSUs have .weight_1 = 1"
  ),
  pass = c(
    isTRUE(all.equal(sum(srs_sample$.weight), truth$N)),
    isTRUE(all(abs(strat_weight_by_stratum$weight_sum - strat_weight_by_stratum$N_h) < 1e-8)),
    isTRUE(all.equal(pps_two_stage$.weight, pps_two_stage$.weight_1 * pps_two_stage$.weight_2, tolerance = 1e-10)),
    all(pps_two_stage$.weight_1[pps_two_stage$.certainty_1] == 1)
  )
)

checks
```

These are structural, non-random checks. If they fail, the design object is wrong before analysis begins.

## Layer 2: Monte Carlo validation

Below we repeatedly sample from the same finite population and analyze with `survey`.
For each design we report:

- `rel_bias_pct`: relative bias of \(\hat{\bar{Y}}\),
- `se_ratio`: empirical SD of estimates divided by mean estimated SE,
- `coverage_95`: fraction of 95% CIs covering the true finite-population mean.

This vignette fixes Monte Carlo replication at `R = 199`.

```{r}
evaluate_design <- function(design_fn, population, true_mean, R = 199L, seed_init = 3000L) {
  est <- numeric(R)
  se <- numeric(R)
  cover <- logical(R)

  for (i in seq_len(R)) {
    sample_i <- design_fn(population, seed_init + i)
    svy_i <- as_svydesign(sample_i)
    fit <- svymean(~y, svy_i)

    est[i] <- unname(coef(fit)[1])
    se[i] <- unname(SE(fit)[1])

    ci <- suppressWarnings(confint(fit, level = 0.95))
    cover[i] <- ci[1, 1] <= true_mean && ci[1, 2] >= true_mean
  }

  tibble(R = R,
         bias = mean(est - true_mean),
         rel_bias_pct = 100 * mean(est - true_mean) / true_mean,
         emp_sd = sd(est),
         mean_se = mean(se),
         se_ratio = sd(est) / mean(se),
         coverage_95 = mean(cover))
}

mcse_coverage <- function(R, p = 0.95) {
  sqrt(p * (1 - p) / R)
}

design_srs <- function(pop, seed) {
  sampling_design() |>
    draw(n = 400, method = "srswor") |>
    execute(pop, seed = seed)
}

design_strat <- function(pop, seed) {
  sampling_design() |>
    stratify_by(stratum, alloc = "proportional") |>
    draw(n = 400, method = "srswor") |>
    execute(pop, seed = seed)
}

design_two_stage <- function(pop, seed) {
  sampling_design() |>
    add_stage("PSU") |>
      stratify_by(stratum) |>
      cluster_by(psu) |>
      draw(n = 8, method = "pps_brewer", mos = psu_mos) |>
    add_stage("SSU") |>
      draw(n = 6, method = "srswor") |>
    execute(pop, seed = seed)
}

R_mc <- 199L
R_mc

mc_results <- bind_rows(
  evaluate_design(design_srs, population, truth$Y_mean, R = R_mc) |>
    mutate(design = "SRSWOR"),
  evaluate_design(design_strat, population, truth$Y_mean, R = R_mc) |>
    mutate(design = "Stratified SRS"),
  evaluate_design(design_two_stage, population, truth$Y_mean, R = R_mc) |>
    mutate(design = "Two-stage PPS->SRS")
) |>
  select(design, everything())

mc_results
```

## Interpreting the results

For this short validation, practical acceptance bands are:

- `abs(rel_bias_pct) <= 2`,
- `0.85 <= se_ratio <= 1.15`,
- `0.90 <= coverage_95 <= 0.98`.

Coverage has Monte Carlo uncertainty. At nominal 95% coverage, the Monte Carlo standard error is approximately \(\sqrt{0.95 \times 0.05 / R}\). With `R = 199`, this is about 1.5 percentage points.

```{r}
acceptance <- mc_results |>
  mutate(
    pass_bias = abs(rel_bias_pct) <= 2,
    pass_se = se_ratio >= 0.85 & se_ratio <= 1.15,
    pass_cov = coverage_95 >= 0.90 & coverage_95 <= 0.98,
    pass_all = pass_bias & pass_se & pass_cov
  ) |>
  select(design, pass_bias, pass_se, pass_cov, pass_all)

acceptance
```

```{r}
tibble(
  R = R_mc,
  mcse_coverage_95 = mcse_coverage(R_mc, p = 0.95)
)
```

If these checks hold across core designs, we have strong evidence that:

1. `samplyr` design metadata is internally consistent,
2. exported `survey` designs reproduce correct design-based inference on known finite populations.

In this vignette, `R` is fixed at 199 as a practical default. If you need higher confidence when interpreting package accuracy, rerun the same validation with a larger `R`.
